---
title: "Freeze 2 vs Freeze 3 — Gene-based ExWAS meta-analysis"
author: "Brad Wubbenhorst"
date: "`r Sys.Date()`"
knit: !expr function(inputFile, encoding) rmarkdown::render(inputFile, encoding = encoding, output_file = file.path(dirname(inputFile), paste0(tools::file_path_sans_ext(basename(inputFile)), "_", Sys.Date(), ".html")))
output:
  html_document:
    theme: paper
    toc: true
    toc_float: true
    toc_depth: 2
    code_folding: hide
params:
  out_dir: "meta"
  regenie_dir_f2: "FREEZE2/data/final"
  regenie_dir_f3: "FREEZE3/data/final"
  project_f2: "PMBB-F2-TGCT-ExWAS-17a"
  project_f3: "PMBB-F3-TGCT-ExWAS-17a"
  tests_keep: ""
  masks_keep: ""
  thresh_keep: ""
  p_strong: 0.001
  p_support: 0.05
  p_weak: 0.1
  abs_beta_max_reasonable: 5
  se_max_reasonable: 2
  abs_beta_diff_max: 1.5
  genes_sanity: "CHEK2"
  n_label_volcano: 5
---

```{=html}
<style type="text/css">
/* body */
body { font-size: 1.24rem; line-height: 1.5; }
.main-container { max-width: 52rem; }
/* title block */
h1.title { font-size: 1.65rem; margin-top: 0.5rem; }
h4.author, h4.date { font-size: 0.95rem; }
/* headings: modest vs body */
h1, h2, h3 {
  margin-top: 0.6rem;
}
h1 { font-size: 1.3rem; margin-bottom: 0.45rem; font-weight: 600; }
h2 { font-size: 1.08rem; margin-bottom: 0.3rem; font-weight: 600; }
h3 { font-size: 0.98rem; margin-bottom: 0.22rem; font-weight: 600; }
h1:first-of-type,
h2:first-of-type {
  margin-top: 0.3rem;
}
#TOC, #toc { font-size: 0.88rem; line-height: 1.35; }
#TOC .tocify-header { margin-top: 0.2rem; }
.table, table { font-size: 0.95rem; }
pre.sourceCode, pre { padding-top: 0.3rem !important; padding-bottom: 0.3rem !important; margin-top: 0.35rem; margin-bottom: 0.35rem; }
.code-folding-btn { margin-bottom: 0.25rem; }
</style>
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 10,
  fig.height = 6
)
library(tidyverse)
library(readr)
library(knitr)
library(kableExtra)
library(ggrepel)

theme_set(theme_minimal(base_size = 13))

dir.create(params$out_dir, recursive = TRUE, showWarnings = FALSE)
```

# Goal

Compare Freeze 2 vs Freeze 3 gene-based results.

This report is for exploratory comparison only. Differences between freezes are treated as signals to investigate (e.g. missingness, capture differences, or power), not as conclusions.

This report is not intended to determine whether signals are true or false, only to identify where cross-freeze discrepancies exist.

Methods in brief: **ADD** uses inverse-variance meta (same-direction rows only); **SKAT / SKATO** use Fisher combination of *p*-values. Full detail is in CSVs; this document supports comparison and diagnostic follow-up.

### How to use this report

- Use **Table B** (consistent same-direction ADD rows) for cross-freeze consistency.  
- Use **Table C** (Freeze 3 stronger than Freeze 2) for Freeze-3-driven signals.  
- Use **Table D** and **CHEK2 discrepancy** for opposite-sign or pattern mismatches and discrepancy follow-up (gene-level only).

---

# 1. Load

```{r load-files, echo=FALSE}
#' Read chr1–22 step2 gene-based .regenie for one freeze; same parsing as report_regenie.Rmd.
load_regenie_gene_based <- function(project_name, regenie_dir) {
  lst <- list()
  for (i in 1:22) {
    fp <- file.path(regenie_dir, paste0(project_name, ".chr", i, ".step2_gene_based_STATUS.regenie"))
    if (file.exists(fp)) {
      lst[[length(lst) + 1]] <- readr::read_delim(
        fp, delim = " ", col_names = TRUE, trim_ws = TRUE, comment = "#", show_col_types = FALSE
      )
    }
  }
  if (!length(lst)) {
    stop("No .step2_gene_based_STATUS.regenie files for project ", project_name, " under ", regenie_dir)
  }
  tb <- bind_rows(lst)
  for (col in c("A1FREQ", "A1FREQ_CASES", "A1FREQ_CONTROLS")) {
    if (!col %in% names(tb)) tb[[col]] <- NA_real_
  }
  tb %>%
    mutate(
      across(c(CHROM, GENPOS, N, N_CASES, N_CONTROLS), as.integer),
      across(c(A1FREQ, A1FREQ_CASES, A1FREQ_CONTROLS, BETA, SE, CHISQ, LOG10P), as.numeric),
      PVAL = 10^(-LOG10P),
      GENE = sub("\\..*", "", ID),
      MASK = sub(".*\\.(M[0-9]+).*", "\\1", ID),
      THRESH_RAW = sub(".*\\.(.*)$", "\\1", ID),
      # Keep THRESH as character only (e.g. "0.01", "All variants"); do not coerce to numeric downstream.
      THRESH = as.character(ifelse(
        grepl("^\\d+\\.\\d+$", THRESH_RAW), THRESH_RAW,
        ifelse(THRESH_RAW == "all", "All variants", THRESH_RAW)
      )),
      TEST = as.character(TEST),
      OR_ADD = if_else(TEST == "ADD" & is.finite(BETA), exp(BETA), NA_real_)
    ) %>%
    transmute(
      GENE, MASK, THRESH, TEST, BETA, SE, PVAL, N_CASES, N_CONTROLS,
      OR_ADD, A1FREQ_CASES, A1FREQ_CONTROLS
    )
}

f2_long <- load_regenie_gene_based(params$project_f2, params$regenie_dir_f2)
f3_long <- load_regenie_gene_based(params$project_f3, params$regenie_dir_f3)

message(
  "REGENIE gene-based dirs — F2: ", params$regenie_dir_f2,
  " (project ", params$project_f2, ") | F3: ", params$regenie_dir_f3,
  " (project ", params$project_f3, ")"
)

readr::write_csv(
  dplyr::bind_rows(
    f2_long %>% mutate(FREEZE = "F2"),
    f3_long %>% mutate(FREEZE = "F3")
  ),
  file.path(params$out_dir, "gene_regenie_standardized_F2_F3_long.csv")
)
message("Wrote: ", file.path(params$out_dir, "gene_regenie_standardized_F2_F3_long.csv"))

f2 <- f2_long %>%
  rename(
    BETA_F2 = BETA, SE_F2 = SE, PVAL_F2 = PVAL, OR_F2 = OR_ADD,
    N_CASES_F2 = N_CASES, N_CONTROLS_F2 = N_CONTROLS,
    A1FREQ_CASES_F2 = A1FREQ_CASES, A1FREQ_CONTROLS_F2 = A1FREQ_CONTROLS
  )

f3 <- f3_long %>%
  rename(
    BETA_F3 = BETA, SE_F3 = SE, PVAL_F3 = PVAL, OR_F3 = OR_ADD,
    N_CASES_F3 = N_CASES, N_CONTROLS_F3 = N_CONTROLS,
    A1FREQ_CASES_F3 = A1FREQ_CASES, A1FREQ_CONTROLS_F3 = A1FREQ_CONTROLS
  )

message("Freeze 2 rows: ", nrow(f2), " | Freeze 3 rows: ", nrow(f3))
```

---

# 2. Filter

Optional YAML: `tests_keep`, `masks_keep`, `thresh_keep` (REGENIE `TEST` strings, e.g. `ADD`, `ADD-SKAT`, `ADD-SKATO`).

```{r filter-tests-masks, echo=FALSE}
parse_csv_param <- function(x) {
  x <- trimws(x)
  if (is.null(x) || length(x) != 1 || x == "") return(character(0))
  strsplit(x, ",", fixed = TRUE)[[1]] %>% trimws() %>% .[. != ""]
}

tests_keep <- parse_csv_param(params$tests_keep)
masks_keep <- parse_csv_param(params$masks_keep)
thresh_keep <- parse_csv_param(params$thresh_keep)

filter_slice <- function(df) {
  out <- df
  if (length(tests_keep)) out <- out %>% filter(TEST %in% tests_keep)
  if (length(masks_keep)) out <- out %>% filter(MASK %in% masks_keep)
  if (length(thresh_keep)) out <- out %>% filter(THRESH %in% thresh_keep)
  out
}

f2_f <- filter_slice(f2)
f3_f <- filter_slice(f3)

message(
  "After filters — F2: ", nrow(f2_f), " rows; F3: ", nrow(f3_f), " rows. ",
  "TEST filter: ", if (length(tests_keep)) paste(tests_keep, collapse = ", ") else "(none)",
  " | MASK filter: ", if (length(masks_keep)) paste(masks_keep, collapse = ", ") else "(none)"
)
```

---

# 3. Join

```{r merge-check, echo=FALSE}
key <- c("GENE", "MASK", "THRESH", "TEST")

dup_f2 <- f2_f %>% count(!!!syms(key), name = "n") %>% filter(n > 1)
dup_f3 <- f3_f %>% count(!!!syms(key), name = "n") %>% filter(n > 1)

merged <- inner_join(f2_f, f3_f, by = key)

only_f2 <- anti_join(f2_f, f3_f, by = key)
only_f3 <- anti_join(f3_f, f2_f, by = key)

message(
  "Matched rows (inner join): ", nrow(merged), "\n",
  "Only in Freeze 2: ", nrow(only_f2), "\n",
  "Only in Freeze 3: ", nrow(only_f3), "\n",
  "Duplicate keys in F2 (after filter): ", nrow(dup_f2), "\n",
  "Duplicate keys in F3 (after filter): ", nrow(dup_f3)
)

if (nrow(dup_f2)) knitr::kable(head(dup_f2, 20), caption = "Unexpected duplicate keys — Freeze 2 (first 20)")
if (nrow(dup_f3)) knitr::kable(head(dup_f3, 20), caption = "Unexpected duplicate keys — Freeze 3 (first 20)")

# Freeze-specific keys (capture / join debugging; do not discard)
readr::write_csv(only_f2, file.path(params$out_dir, "only_F2.csv"))
readr::write_csv(only_f3, file.path(params$out_dir, "only_F3.csv"))
readr::write_csv(
  dplyr::filter(only_f2, TEST == "ADD-SKATO"),
  file.path(params$out_dir, "only_F2_SKATO.csv")
)
readr::write_csv(
  dplyr::filter(only_f3, TEST == "ADD-SKATO"),
  file.path(params$out_dir, "only_F3_SKATO.csv")
)
message("Wrote unmatched keys: ", file.path(params$out_dir, "only_F2.csv"), " | ", file.path(params$out_dir, "only_F3.csv"))
message("Wrote ADD-SKATO-only unmatched: ", file.path(params$out_dir, "only_F2_SKATO.csv"), " | ", file.path(params$out_dir, "only_F3_SKATO.csv"))
```

# 3b. SKAT/SKATO

```{r meta-fisher-vc, echo=FALSE}
fisher_two_p <- function(p1, p2) {
  eps <- .Machine$double.eps
  p_floor <- 1e-300
  ok <- is.finite(p1) & is.finite(p2) & p1 > 0 & p1 <= 1 & p2 > 0 & p2 <= 1
  p1c <- pmax(pmin(p1, 1 - eps), p_floor)
  p2c <- pmax(pmin(p2, 1 - eps), p_floor)
  chi <- -2 * (log(p1c) + log(p2c))
  out <- rep(NA_real_, length(p1))
  out[ok] <- pchisq(chi[ok], df = 4, lower.tail = FALSE)
  out
}

psup <- as.numeric(params$p_support)
pwk <- as.numeric(params$p_weak)

meta_skat <- merged %>%
  dplyr::filter(TEST == "ADD-SKAT") %>%
  mutate(
    FISHER_P = fisher_two_p(PVAL_F2, PVAL_F3),
    BOTH_FREEZE_NOMINAL = PVAL_F2 <= psup & PVAL_F3 <= psup,
    BOTH_FREEZE_WEAK = PVAL_F2 <= pwk & PVAL_F3 <= pwk
  ) %>%
  mutate(FISHER_P_ADJ = p.adjust(FISHER_P, method = "BH")) %>%
  arrange(FISHER_P_ADJ, FISHER_P)

meta_skato <- merged %>%
  dplyr::filter(TEST == "ADD-SKATO") %>%
  mutate(
    FISHER_P = fisher_two_p(PVAL_F2, PVAL_F3),
    BOTH_FREEZE_NOMINAL = PVAL_F2 <= psup & PVAL_F3 <= psup,
    BOTH_FREEZE_WEAK = PVAL_F2 <= pwk & PVAL_F3 <= pwk,
    DOMINANT_FREEZE = case_when(
      !is.finite(PVAL_F2) & !is.finite(PVAL_F3) ~ NA_character_,
      !is.finite(PVAL_F2) ~ "F3",
      !is.finite(PVAL_F3) ~ "F2",
      PVAL_F3 < PVAL_F2 ~ "F3",
      PVAL_F2 < PVAL_F3 ~ "F2",
      TRUE ~ "both"
    )
  ) %>%
  mutate(FISHER_P_ADJ = p.adjust(FISHER_P, method = "BH"))

meta_skato <- meta_skato %>%
  mutate(
    priority = case_when(
      FISHER_P_ADJ <= 0.05 & BOTH_FREEZE_NOMINAL ~ "High",
      FISHER_P_ADJ <= 0.1 & BOTH_FREEZE_WEAK ~ "Medium",
      TRUE ~ "Review"
    )
  ) %>%
  arrange(FISHER_P_ADJ, FISHER_P)

message("ADD-SKAT rows (merged): ", nrow(meta_skat), " | ADD-SKATO rows (merged): ", nrow(meta_skato))
```

---

# 4. ADD

```{r meta-ivw, echo=FALSE}
# w_i = 1/SE^2; applies only to TEST == ADD (directional log-OR scale)

meta_ivw2 <- function(b1, se1, b2, se2) {
  ok <- is.finite(b1) & is.finite(b2) & is.finite(se1) & is.finite(se2) & se1 > 0 & se2 > 0
  w1 <- ifelse(ok, 1 / se1^2, NA_real_)
  w2 <- ifelse(ok, 1 / se2^2, NA_real_)
  sw <- w1 + w2
  beta_m <- ifelse(ok, (w1 * b1 + w2 * b2) / sw, NA_real_)
  se_m <- ifelse(ok, sqrt(1 / sw), NA_real_)
  z <- beta_m / se_m
  p <- 2 * pnorm(-abs(z))
  tibble(META_BETA = beta_m, META_SE = se_m, META_Z = z, META_P = p)
}

merged_add_all <- merged %>% dplyr::filter(TEST == "ADD")
merged_add <- merged_add_all %>%
  dplyr::filter(is.finite(BETA_F2), is.finite(BETA_F3), sign(BETA_F2) == sign(BETA_F3))

message(
  "ADD rows (inner merge): ", nrow(merged_add_all),
  " | IVW meta after sign(BETA_F2)==sign(BETA_F3): ", nrow(merged_add),
  " (excluded ", nrow(merged_add_all) - nrow(merged_add), " discordant or non-finite beta rows)"
)

meta_add_df <- merged_add %>%
  bind_cols(meta_ivw2(.$BETA_F2, .$SE_F2, .$BETA_F3, .$SE_F3)) %>%
  mutate(
    META_OR = exp(META_BETA),
    META_P_ADJ = p.adjust(META_P, method = "BH")
  )

message("ADD rows for IVW meta: ", nrow(meta_add_df))
```

---

# 5. Flags

```{r flags, echo=FALSE}
signz <- function(x) {
  x <- as.numeric(x)
  ifelse(!is.finite(x), NA_integer_, ifelse(x > 0, 1L, ifelse(x < 0, -1L, 0L)))
}

p_strong <- as.numeric(params$p_strong)
p_support <- as.numeric(params$p_support)
p_weak <- as.numeric(params$p_weak)
ab_max <- as.numeric(params$abs_beta_max_reasonable)
se_max <- as.numeric(params$se_max_reasonable)
abd_max <- as.numeric(params$abs_beta_diff_max)

meta_add_df <- meta_add_df %>%
  mutate(
    WEIGHT_F2 = ifelse(is.finite(SE_F2) & SE_F2 > 0, 1 / SE_F2^2, NA_real_),
    WEIGHT_F3 = ifelse(is.finite(SE_F3) & SE_F3 > 0, 1 / SE_F3^2, NA_real_),
    WEIGHT_RATIO = WEIGHT_F3 / (WEIGHT_F2 + WEIGHT_F3),
    DOMINATED_BY_F3 = is.finite(WEIGHT_RATIO) & WEIGHT_RATIO > 0.8,
    Q = ifelse(
      is.finite(META_BETA) & is.finite(SE_F2) & is.finite(SE_F3) & SE_F2 > 0 & SE_F3 > 0,
      (BETA_F2 - META_BETA)^2 / SE_F2^2 + (BETA_F3 - META_BETA)^2 / SE_F3^2,
      NA_real_
    ),
    Q_P = ifelse(is.finite(Q), pchisq(Q, df = 1, lower.tail = FALSE), NA_real_),
    P_HET = Q_P,
    HETEROGENEITY = is.finite(Q_P) & (Q_P < 0.05),
    HET_FLAG = is.finite(Q_P) & (Q_P < 0.05),
    SIGN_F2 = signz(BETA_F2),
    SIGN_F3 = signz(BETA_F3),
    CONSISTENT_DIRECTION = (SIGN_F2 == SIGN_F3) & (SIGN_F2 != 0) & (SIGN_F3 != 0),
    PARTIAL_DIRECTION = (SIGN_F2 == SIGN_F3) | (SIGN_F2 == 0 | SIGN_F3 == 0),
    ABS_BETA_DIFF = abs(BETA_F2 - BETA_F3),
    BEST_FREEZE = case_when(
      !is.finite(PVAL_F2) & !is.finite(PVAL_F3) ~ NA_character_,
      !is.finite(PVAL_F2) ~ "F3",
      !is.finite(PVAL_F3) ~ "F2",
      PVAL_F3 < PVAL_F2 ~ "F3",
      PVAL_F2 < PVAL_F3 ~ "F2",
      TRUE ~ "tie"
    ),
    OPPOSITE_DIRECTION = (SIGN_F2 != SIGN_F3) & (SIGN_F2 != 0) & (SIGN_F3 != 0),
    strong_F2 = is.finite(PVAL_F2) & (PVAL_F2 <= p_strong),
    strong_F3 = is.finite(PVAL_F3) & (PVAL_F3 <= p_strong),
    support_F2 = is.finite(PVAL_F2) & (PVAL_F2 <= p_support),
    support_F3 = is.finite(PVAL_F3) & (PVAL_F3 <= p_support),
    weak_F2 = is.finite(PVAL_F2) & (PVAL_F2 <= p_weak),
    weak_F3 = is.finite(PVAL_F3) & (PVAL_F3 <= p_weak),
    BOTH_WEAK_SUPPORT = weak_F2 & weak_F3,
    BOTH_SUPPORT = (strong_F2 & support_F3 & PARTIAL_DIRECTION) |
      (strong_F3 & support_F2 & PARTIAL_DIRECTION),
    ALT_CT_CASES_F2_EST = ifelse(
      is.finite(N_CASES_F2) & is.finite(A1FREQ_CASES_F2),
      2 * N_CASES_F2 * A1FREQ_CASES_F2,
      NA_real_
    ),
    ALT_CT_CONTROLS_F2_EST = ifelse(
      is.finite(N_CONTROLS_F2) & is.finite(A1FREQ_CONTROLS_F2),
      2 * N_CONTROLS_F2 * A1FREQ_CONTROLS_F2,
      NA_real_
    ),
    ALT_CT_CASES_F3_EST = ifelse(
      is.finite(N_CASES_F3) & is.finite(A1FREQ_CASES_F3),
      2 * N_CASES_F3 * A1FREQ_CASES_F3,
      NA_real_
    ),
    ALT_CT_CONTROLS_F3_EST = ifelse(
      is.finite(N_CONTROLS_F3) & is.finite(A1FREQ_CONTROLS_F3),
      2 * N_CONTROLS_F3 * A1FREQ_CONTROLS_F3,
      NA_real_
    )
  )

meta_add_df <- meta_add_df %>%
  mutate(
    reasonable_se = is.finite(SE_F2) & is.finite(SE_F3) & SE_F2 <= se_max & SE_F3 <= se_max,
    reasonable_beta = abs(BETA_F2) <= ab_max & abs(BETA_F3) <= ab_max,
    not_wild_diff = ABS_BETA_DIFF <= abd_max,
    min_support_ok = support_F2 | support_F3
  )

meta_add_df <- meta_add_df %>%
  mutate(
    priority = case_when(
      OPPOSITE_DIRECTION ~ "Discordant",
      is.finite(META_P) & META_P <= p_strong & CONSISTENT_DIRECTION & reasonable_se &
        reasonable_beta & !HETEROGENEITY & min_support_ok ~ "High",
      ((strong_F2 & weak_F3) | (strong_F3 & weak_F2)) & !OPPOSITE_DIRECTION ~ "Medium",
      TRUE ~ "Review"
    )
  )
```

`P_HET` = Cochran-style *Q* p-value (same as `Q_P`); `HET_FLAG` marks *P*~het~ \< 0.05 (review, do not auto-drop). Longer definitions → **Notes**.

---

# 5b. SKAT/SKATO

**Columns in the tables below:** **`PVAL_F2`**, **`PVAL_F3`** — REGENIE gene-based *p*-values per freeze. **`FISHER_P`** — Fisher’s method combining those two *p*-values (χ² on 4 df; non-directional, no joint effect size). **`FISHER_P_ADJ`** — Benjamini–Hochberg adjustment of **`FISHER_P`** within this SKAT or SKATO block. **`DOMINANT_FREEZE`**, **`BOTH_FREEZE_NOMINAL`**, **`BOTH_FREEZE_WEAK`**, **`priority`** — exploratory triage labels (see **Notes**).

```{r tables-vc, echo=FALSE}
vc_show <- function(df, title) {
  cat("### ", title, "\n\n", sep = "")
  if (!nrow(df)) {
    cat("_No rows after merge._\n\n")
    return(invisible(NULL))
  }
  show_tbl <- function(x, n = Inf) {
    if (is.finite(n)) x <- head(x, n)
    x %>%
      kable(digits = 4, format.args = list(scientific = FALSE)) %>%
      kable_styling(bootstrap_options = c("striped", "hover"), font_size = 13, full_width = FALSE) %>%
      scroll_box(width = "100%", height = "520px")
  }
  show_tbl(
    df %>%
      select(any_of(c(
        "GENE", "MASK", "THRESH", "PVAL_F2", "PVAL_F3", "FISHER_P", "FISHER_P_ADJ",
        "DOMINANT_FREEZE", "BOTH_FREEZE_NOMINAL", "BOTH_FREEZE_WEAK", "priority"
      )))
  )
  cat("\n")
}
vc_show(meta_skato, "SKATO")
vc_show(meta_skat, "SKAT")
```

---

# 6. Results

ADD meta uses **same-direction** rows only; **Discordant** is the full-overlap opposite-sign set (no IVW *p*).

Only **same-direction ADD** overlap rows are inverse-variance meta-analyzed (IVW *p* and FDR); opposite-sign rows appear in Table D without a combined effect.

**Columns in Tables A–C and the comparison shortlist:** **`PVAL_F2`**, **`PVAL_F3`** — REGENIE ADD *p*-values per freeze. **`META_BETA`**, **`META_SE`**, **`META_Z`**, **`META_P`** — inverse-variance meta across freezes (`1/SE²` weights) when `sign(BETA_F2)==sign(BETA_F3)`; **`META_P`** is two-sided from **`META_Z`** (computed in this report, not a REGENIE column). **`META_P_ADJ`** — BH FDR of **`META_P`** across all same-direction ADD rows (`meta_add_df`). **`HET_FLAG`** — TRUE if heterogeneity **`Q_P`** \< 0.05 (Cochran *Q* vs **`META_BETA`**; same as **`P_HET`** in CSVs). Full exports also include **`WEIGHT_F2`**, **`WEIGHT_F3`**, **`WEIGHT_RATIO`**, betas/SEs, **`META_OR`**.

```{r tables-abcd}
show_tbl <- function(df, n = 50) {
  if (is.finite(n)) df <- head(df, n)
  df %>%
    kable(digits = 4, format.args = list(scientific = FALSE)) %>%
    kable_styling(bootstrap_options = c("striped", "hover"), font_size = 13, full_width = FALSE) %>%
    scroll_box(width = "100%", height = "560px")
}

## A. All same-direction ADD (IVW)
table_a <- meta_add_df %>% arrange(META_P, PVAL_F3, PVAL_F2)

## B. Consistent (Table B filters)
table_b <- meta_add_df %>%
  filter(CONSISTENT_DIRECTION, reasonable_se, reasonable_beta, not_wild_diff) %>%
  arrange(META_P)

## C. Freeze 3 stronger (Table C filters)
table_c <- meta_add_df %>%
  filter(
    strong_F3,
    is.finite(PVAL_F2) & PVAL_F2 > p_support,
    PARTIAL_DIRECTION
  ) %>%
  arrange(PVAL_F3, META_P)

## D. Discordant ADD overlap
table_d <- merged_add_all %>%
  dplyr::filter(
    is.finite(BETA_F2),
    is.finite(BETA_F3),
    sign(BETA_F2) != sign(BETA_F3)
  ) %>%
  arrange(pmin(PVAL_F2, PVAL_F3, na.rm = TRUE))
```

## A. All same-direction ADD rows

All same-direction ADD rows used for IVW comparison (`meta_add_df`), sorted by `META_P`.
**Display cap: first 100 rows shown in HTML (full table is still written to CSV outputs).**

```{r table-a, echo=FALSE}
show_tbl(
  table_a %>%
    select(GENE, MASK, THRESH, META_P, META_P_ADJ, PVAL_F2, PVAL_F3, HET_FLAG),
  n = 100
)
```

## B. Consistent same-direction ADD rows

Subset of A with same direction and QC-style sanity filters (`reasonable_se`, `reasonable_beta`, `not_wild_diff`).
**Display cap: first 100 rows shown in HTML (full table is still written to CSV outputs).**

```{r table-b, echo=FALSE}
show_tbl(
  table_b %>%
    select(GENE, MASK, THRESH, META_P, META_P_ADJ, PVAL_F2, PVAL_F3, HET_FLAG),
  n = 100
)
```

## C. Freeze 3 stronger than Freeze 2

Rows with stronger evidence in Freeze 3 and weaker/non-nominal evidence in Freeze 2 (exploratory “F3-driven” view).
**Display cap: first 100 rows shown in HTML (full table is still written to CSV outputs).**

```{r table-c, echo=FALSE}
show_tbl(
  table_c %>%
    select(GENE, MASK, THRESH, META_P, META_P_ADJ, PVAL_F2, PVAL_F3, HET_FLAG),
  n = 100
)
```

## D. Discordant ADD rows

**Columns:** **`PVAL_F2`**, **`PVAL_F3`** only — opposite-sign overlap rows are excluded from IVW meta, so there is no **`META_P`** / **`META_P_ADJ`** for these genes in this view.

ADD overlap rows with opposite beta signs between freezes (not used in same-direction IVW meta rows).
**Display cap: first 100 rows shown in HTML (full table is still written to CSV outputs).**

```{r table-d, echo=FALSE}
show_tbl(
  table_d %>%
    select(GENE, MASK, THRESH, PVAL_F2, PVAL_F3),
  n = 100
)
```

## Cross-freeze comparison shortlist (ADD, same-direction)

Sorted by `META_P`. For exploratory triage; full columns in CSV.
**Display cap: first 100 rows shown in HTML (full table is still written to CSV outputs).**

```{r comparison-table, echo=FALSE}
comparison_tbl <- meta_add_df %>%
  transmute(
    GENE,
    MASK,
    THRESH,
    P_F2 = PVAL_F2,
    P_F3 = PVAL_F3,
    P_META = META_P,
    HET_FLAG,
    META_P_ADJ
  ) %>%
  arrange(P_META, P_F3, P_F2)
```

```{r comparison-table-show, echo=FALSE}
if (nrow(comparison_tbl) > 0) {
  show_tbl(
    comparison_tbl %>%
      select(GENE, MASK, THRESH, P_META, META_P_ADJ, P_F2, P_F3, HET_FLAG),
    n = 100
  )
} else {
  cat("_No rows in comparison shortlist._\n\n")
}
```

---

# 7. CHEK2 discrepancy

CHEK2 is used here as a reference gene to understand differences between freezes (gene-level only; variant-level missingness is out of scope for this notebook).

The first table shows exploratory pattern labels; the second is a side-by-side numeric comparison. **`CHEK2_PATTERN`** labels are for triage; **`META_P`**, **`META_P_ADJ`**, **`PVAL_F*`**, **`BETA_F*`**, **`HET_FLAG`**, **`BEST_FREEZE`** match **6. Results** where applicable.

```{r sanity-genes, echo=FALSE}
sanity_genes <- parse_csv_param(params$genes_sanity)
if (!length(sanity_genes)) sanity_genes <- "CHEK2"
ps_chek2 <- as.numeric(params$p_support)

chek2_tbl <- merged %>%
  dplyr::filter(GENE %in% sanity_genes, TEST == "ADD") %>%
  dplyr::left_join(
    meta_add_df %>% dplyr::select(GENE, MASK, THRESH, META_P, META_P_ADJ, HET_FLAG, BEST_FREEZE),
    by = c("GENE", "MASK", "THRESH")
  ) %>%
  dplyr::mutate(
    opposite_beta = is.finite(BETA_F2) & is.finite(BETA_F3) &
      sign(BETA_F2) != 0 & sign(BETA_F3) != 0 & sign(BETA_F2) != sign(BETA_F3),
    CHEK2_PATTERN = dplyr::case_when(
      opposite_beta ~ "discordant",
      is.finite(PVAL_F3) & PVAL_F3 <= ps_chek2 &
        (!is.finite(PVAL_F2) | PVAL_F2 > ps_chek2) ~ "F3 stronger, F2 weak",
      is.finite(PVAL_F2) & PVAL_F2 <= ps_chek2 &
        (!is.finite(PVAL_F3) | PVAL_F3 > ps_chek2) ~ "F2 only",
      is.finite(PVAL_F2) & PVAL_F2 <= ps_chek2 & is.finite(PVAL_F3) & PVAL_F3 <= ps_chek2 ~ "Both support",
      TRUE ~ "unclear"
    )
  ) %>%
  dplyr::arrange(GENE, MASK, THRESH)

knitr::kable(
  chek2_tbl %>%
    dplyr::select(
      GENE, MASK, THRESH, CHEK2_PATTERN, BEST_FREEZE, META_P, META_P_ADJ,
      PVAL_F2, PVAL_F3, BETA_F2, BETA_F3, HET_FLAG
    ),
  digits = 5,
  caption = "CHEK2 ADD rows: exploratory pattern labels (not a clinical interpretation)"
)
```

```{r chek2-side-by-side, echo=FALSE}
knitr::kable(
  chek2_tbl %>%
    dplyr::select(
      GENE, MASK, THRESH, BEST_FREEZE, HET_FLAG, PVAL_F2, PVAL_F3, META_P,
      BETA_F2, BETA_F3, N_CASES_F2, N_CASES_F3
    ),
  digits = 5,
  caption = "CHEK2: side-by-side comparison (Freeze 2 vs Freeze 3)"
)
```

Interpretation (exploratory):

Differences between Freeze 2 and Freeze 3 for CHEK2 may reflect:

- differences in carrier counts  
- variant missingness or capture differences  
- power differences between freezes  

This report does not resolve these causes; it identifies candidates for variant-level follow-up.

---

# 8. Plots

Same-direction ADD rows as **6. Results**: *x*/*y* use **`PVAL_F2`** / **`PVAL_F3`** (and **`META_P`** for volcano *y* via **`LPM`** = −log10 **`META_P`**, capped).

```{r plot-data, echo=FALSE}
plot_df <- meta_add_df %>%
  mutate(
    LP2 = pmin(-log10(pmax(PVAL_F2, .Machine$double.xmin)), 50),
    LP3 = pmin(-log10(pmax(PVAL_F3, .Machine$double.xmin)), 50),
    LPM = pmin(-log10(pmax(META_P, .Machine$double.xmin)), 50)
  )
```

## P-values (F2 vs F3)

```{r plot-lp-scatter, echo=FALSE, fig.cap="ADD: −log10 p F2 vs F3 (capped 50)."}
ggplot(plot_df, aes(x = LP2, y = LP3)) +
  geom_point(alpha = 0.32, size = 2) +
  geom_abline(slope = 1, intercept = 0, linetype = 2, color = "gray50") +
  labs(
    x = expression(-log[10] ~ P[Freeze2]),
    y = expression(-log[10] ~ P[Freeze3]),
    caption = "−log10(p) capped at 50 for display."
  )
```

## Betas (F2 vs F3)

```{r plot-beta-scatter, echo=FALSE, fig.cap="ADD: BETA F2 vs F3."}
ggplot(plot_df, aes(x = BETA_F2, y = BETA_F3)) +
  geom_point(alpha = 0.32, size = 2) +
  geom_abline(slope = 1, intercept = 0, linetype = 2, color = "gray50") +
  geom_hline(yintercept = 0, color = "gray70") +
  geom_vline(xintercept = 0, color = "gray70") +
  labs(x = "BETA Freeze 2", y = "BETA Freeze 3")
```

## QQ

```{r plot-qq, echo=FALSE, fig.cap="ADD: QQ by freeze."}
qq_points <- function(p) {
  p <- p[is.finite(p) & p > 0 & p <= 1]
  n <- length(p)
  if (n < 10) return(tibble(expected = numeric(0), observed = numeric(0)))
  tibble(
    expected = -log10(ppoints(n)),
    observed = -log10(sort(p))
  )
}

qq_f2 <- qq_points(meta_add_df$PVAL_F2) %>% mutate(Freeze = "F2")
qq_f3 <- qq_points(meta_add_df$PVAL_F3) %>% mutate(Freeze = "F3")
qq_b <- bind_rows(qq_f2, qq_f3)

ggplot(qq_b, aes(x = expected, y = observed, color = Freeze)) +
  geom_point(alpha = 0.25, size = 0.8) +
  geom_abline(slope = 1, intercept = 0, color = "black", linetype = 2) +
  labs(x = expression(-log[10] ~ p[expected]), y = expression(-log[10] ~ p[observed]))
```

## Volcano

```{r plot-volcano, echo=FALSE, fig.cap="ADD volcano (−log10 META_P capped at 50)."}
nlab <- max(3L, as.integer(params$n_label_volcano))
genes_always <- parse_csv_param(params$genes_sanity)
if (!length(genes_always)) genes_always <- "CHEK2"
lab_sanity <- plot_df %>% filter(GENE %in% genes_always)
lab_ranked <- plot_df %>% arrange(META_P) %>% slice_head(n = nlab)
lab_both <- plot_df %>% filter(BOTH_SUPPORT) %>% arrange(META_P) %>% slice_head(n = 3)
volcano_labs <- bind_rows(lab_sanity, lab_ranked, lab_both) %>%
  distinct(GENE, MASK, THRESH, .keep_all = TRUE)

ggplot(plot_df, aes(x = META_BETA, y = LPM)) +
  geom_point(alpha = 0.32, size = 1.8) +
  geom_hline(yintercept = -log10(0.05), linetype = 3, color = "gray60") +
  geom_text_repel(
    data = volcano_labs,
    aes(label = GENE),
    size = 2.6,
    max.overlaps = 12,
    min.segment.length = 0
  ) +
  labs(
    x = "META_BETA",
    y = expression(-log[10] ~ P[META])
  )
```

---

# 9. Priority genes

**Columns:** same **`META_P`**, **`META_P_ADJ`**, **`HET_FLAG`** meaning as **6. Results**; **`priority`** is defined in the note after the table.

This is intentionally a short triage view (`slice_head(n = 15)`), not a full table of all genes.

```{r priority-tags, echo=FALSE}
priority_shortlist <- meta_add_df %>%
  filter(priority %in% c("High", "Medium")) %>%
  arrange(META_P) %>%
  slice_head(n = 15)
```

```{r priority-table, echo=FALSE}
show_tbl(
  priority_shortlist %>%
    select(GENE, MASK, THRESH, priority, META_P, META_P_ADJ, HET_FLAG),
  n = 25
)
```

**Priority (ADD, exploratory):** **High** — strong `META_P`, same non-zero direction, reasonable SE/beta, no `HET_FLAG`, ≥1 freeze nominal at `p_support`. **Medium** — stronger evidence in one freeze, weaker in the other. **Review** — remainder. **Discordant** — opposite-sign betas (Table D / CSV; no IVW *p*).

---

# 10. Notes

- Per-freeze analyses are primary; this report is for **cross-freeze comparison** and triage, not replication claims.  
- **IVW** here is a ranking aid, not a full joint model; **WEIGHT_RATIO** / **DOMINATED_BY_F3** flag F3-heavy influence.  
- **SKAT/SKATO:** Fisher-combined *p*-values; non-directional. Treat **ADD-SKATO** freeze gaps as exploratory (see **`only_F*_SKATO.csv`**).  
- **Gene-level only** — variant missingness and carrier-level checks belong in a separate variant follow-up workflow.

---

# 11. Follow-up (gene-level context)

`ALT_CT_*_EST` in **meta_ADD** are **expected** alt-allele counts (2×N×`A1FREQ`), not observed carrier counts.

For genes highlighted above (including **`r paste(sanity_genes, collapse = ", ")`**), use **Variant follow-up** for next steps.

---

# 12. Variant follow-up

Variant-level inspection (which variants in each mask, callability, missingness by case/control) is **not** done in this notebook. A companion **variant-level** report should take a target gene list, annotation, variant missingness, and sample freeze/ancestry — see `variant_level_followup.Rmd` (stub in this repo).

---

# 13. Output

```{r write-outputs, echo=FALSE}
add_full <- file.path(params$out_dir, "meta_ADD.csv")
add_prio <- file.path(params$out_dir, "meta_ADD_freeze_consistent_priority.csv")
add_short <- file.path(params$out_dir, "meta_ADD_priority_shortlist.csv")
add_cmp <- file.path(params$out_dir, "meta_ADD_comparison_shortlist.csv")
genes_followup_fp <- file.path(params$out_dir, "genes_for_followup.csv")
final_val_fp <- file.path(params$out_dir, "final_gene_list_for_validation.csv")
variant_targets_fp <- file.path(params$out_dir, "variant_followup_targets.csv")
add_disc <- file.path(params$out_dir, "meta_ADD_sign_discordant.csv")
skat_out <- file.path(params$out_dir, "meta_SKAT.csv")
skato_out <- file.path(params$out_dir, "meta_SKATO.csv")
skato_short_out <- file.path(params$out_dir, "meta_SKATO_shortlist.csv")

drop_internal <- c(
  "SIGN_F2", "SIGN_F3", "strong_F2", "strong_F3",
  "support_F2", "support_F3", "weak_F2", "weak_F3", "min_support_ok"
)

table_a_export <- meta_add_df %>% arrange(META_P, PVAL_F3, PVAL_F2)
table_b_export <- meta_add_df %>%
  filter(CONSISTENT_DIRECTION, reasonable_se, reasonable_beta, not_wild_diff) %>%
  arrange(META_P)
table_short_export <- meta_add_df %>%
  filter(priority %in% c("High", "Medium")) %>%
  arrange(META_P)

readr::write_csv(table_a_export, add_full)
readr::write_csv(table_b_export %>% select(-any_of(drop_internal)), add_prio)
readr::write_csv(table_short_export %>% select(-any_of(drop_internal)), add_short)
readr::write_csv(comparison_tbl, add_cmp)

final_gene_list_for_validation <- table_b_export %>%
  dplyr::filter(META_P_ADJ <= 0.05) %>%
  dplyr::arrange(META_P)
readr::write_csv(final_gene_list_for_validation, final_val_fp)

p_str_fu <- as.numeric(params$p_strong)
psup_fu <- as.numeric(params$p_support)
gf_cols <- c(
  "GENE", "MASK", "THRESH", "TEST", "BETA_F2", "BETA_F3", "SE_F2", "SE_F3",
  "PVAL_F2", "PVAL_F3", "META_P", "META_P_ADJ",
  "N_CASES_F2", "N_CASES_F3", "N_CONTROLS_F2", "N_CONTROLS_F3", "followup_reason"
)
genes_f3_f2 <- meta_add_df %>%
  dplyr::filter(
    is.finite(PVAL_F3) & PVAL_F3 <= p_str_fu &
      is.finite(PVAL_F2) & PVAL_F2 > psup_fu
  ) %>%
  dplyr::mutate(followup_reason = "F3_strong_F2_not_nominal") %>%
  dplyr::select(dplyr::any_of(gf_cols))
genes_disc <- merged_add_all %>%
  dplyr::filter(
    is.finite(BETA_F2), is.finite(BETA_F3),
    sign(BETA_F2) != 0, sign(BETA_F3) != 0,
    sign(BETA_F2) != sign(BETA_F3)
  ) %>%
  dplyr::left_join(
    meta_add_df %>% dplyr::select(GENE, MASK, THRESH, META_P, META_P_ADJ),
    by = c("GENE", "MASK", "THRESH")
  ) %>%
  dplyr::mutate(followup_reason = "discordant_beta_signs") %>%
  dplyr::select(dplyr::any_of(gf_cols))
genes_for_followup <- dplyr::bind_rows(genes_f3_f2, genes_disc) %>%
  dplyr::distinct(GENE, MASK, THRESH, TEST, followup_reason, .keep_all = TRUE)
readr::write_csv(genes_for_followup, genes_followup_fp)

vf_t3 <- merged_add_all %>%
  dplyr::filter(
    is.finite(BETA_F2), is.finite(BETA_F3),
    sign(BETA_F2) != 0, sign(BETA_F3) != 0,
    sign(BETA_F2) != sign(BETA_F3)
  ) %>%
  dplyr::left_join(
    meta_add_df %>% dplyr::select(GENE, MASK, THRESH, META_P, META_P_ADJ, PVAL_F2, PVAL_F3, priority),
    by = c("GENE", "MASK", "THRESH")
  ) %>%
  dplyr::mutate(
    handoff_tier = "Tier 3 — discordant / review",
    priority = dplyr::coalesce(priority, "Discordant")
  )
vf_t2 <- meta_add_df %>%
  dplyr::filter(
    strong_F3,
    is.finite(PVAL_F2) & PVAL_F2 > psup_fu,
    PARTIAL_DIRECTION
  ) %>%
  dplyr::mutate(handoff_tier = "Tier 2 — Freeze-3-driven")
vf_t1 <- meta_add_df %>%
  dplyr::filter(priority %in% c("High", "Medium")) %>%
  dplyr::mutate(
    handoff_tier = dplyr::if_else(priority == "High", "Tier 1 — high priority", "Tier 1 — medium priority")
  )
vf_cols <- c(
  "GENE", "MASK", "THRESH", "handoff_tier", "META_P", "META_P_ADJ",
  "PVAL_F2", "PVAL_F3", "priority", "BETA_F2", "BETA_F3"
)
variant_followup_targets <- dplyr::bind_rows(
  vf_t3 %>% dplyr::select(dplyr::any_of(vf_cols)),
  vf_t2 %>% dplyr::select(dplyr::any_of(vf_cols)),
  vf_t1 %>% dplyr::select(dplyr::any_of(vf_cols))
) %>%
  dplyr::distinct(GENE, MASK, THRESH, .keep_all = TRUE)
readr::write_csv(variant_followup_targets, variant_targets_fp)

readr::write_csv(table_d, add_disc)
readr::write_csv(meta_skat, skat_out)
readr::write_csv(meta_skato, skato_out)

skato_short <- meta_skato %>%
  arrange(FISHER_P_ADJ, FISHER_P) %>%
  slice_head(n = 20)
readr::write_csv(skato_short, skato_short_out)

message("Wrote: ", add_full)
message("Wrote: ", add_prio)
message("Wrote: ", add_short)
message("Wrote: ", add_cmp)
message("Wrote: ", final_val_fp)
message("Wrote: ", variant_targets_fp)
message("Wrote: ", genes_followup_fp)
message("Wrote: ", add_disc)
message("Wrote: ", skat_out)
message("Wrote: ", skato_out)
message("Wrote: ", skato_short_out)
```

**meta_ADD.csv** — full IVW + flags (same-direction rows only).  
**meta_ADD_comparison_shortlist.csv** — compact same-direction ADD rows, sorted **META_P** (exploratory comparison).  
**final_gene_list_for_validation.csv** — Table **B** rows with **META_P_ADJ** ≤ 0.05 (curated handoff list).  
**variant_followup_targets.csv** — union of **High/Medium** priority, **Table C** (F3-driven), and **discordant** rows, with **`handoff_tier`** labels.  
**genes_for_followup.csv** — **Diagnostic scan list** (F3 strong / F2 not nominal, or discordant signs) — **not** a validation-ready shortlist; use **`final_gene_list_for_validation.csv`** for that.  
**meta_ADD_sign_discordant.csv** — opposite-sign overlap (no IVW `META_P`).  
**meta_ADD_freeze_consistent_priority.csv** — Table B–style subset.  
**meta_ADD_priority_shortlist.csv** — **High** + **Medium** only.  
**meta_SKAT.csv** / **meta_SKATO.csv** / **meta_SKATO_shortlist.csv** — Fisher + FDR; SKATO priority High / Medium / Review.  
**only_F2.csv**, **only_F3.csv**; **only_F2_SKATO.csv**, **only_F3_SKATO.csv** (ADD-SKATO-only unmatched).  
**gene_regenie_standardized_F2_F3_long.csv** — long snapshot. Directory: `r params$out_dir`.

**Quick use:** Compare **`meta_ADD_comparison_shortlist.csv`**; full columns in **`meta_ADD.csv`**. Discordant **`meta_ADD_sign_discordant.csv`**. Diagnostic triage **`genes_for_followup.csv`**. SKAT/SKATO by **FISHER_P_ADJ**. **`THRESH`** stays character in downstream code.

---

```{r session-info, echo=FALSE}
sessionInfo()
```
