---
title: "Freeze 2 vs Freeze 3 — Gene-based ExWAS meta-analysis"
author: "Brad Wubbenhorst"
date: "`r Sys.Date()`"
knit: !expr function(inputFile, encoding) rmarkdown::render(inputFile, encoding = encoding, output_file = file.path(dirname(inputFile), paste0(tools::file_path_sans_ext(basename(inputFile)), "_", Sys.Date(), ".html")))
output:
  html_document:
    theme: paper
    toc: true
    toc_float: true
    toc_depth: 2
    code_folding: hide
params:
  out_dir: "meta"
  regenie_dir_f2: "FREEZE2/data/final"
  regenie_dir_f3: "FREEZE3/data/final"
  project_f2: "PMBB-F2-TGCT-ExWAS-17-PC-FreezeQC"
  project_f3: "PMBB-F3-TGCT-ExWAS-17-PC-FreezeQC"
  # Stem between ".chrN." and ".regenie". REGENIE appends _<phenoCol> (e.g. _STATUS) to --out, so
  # --out ...step2_gene_based_STATUS.freeze_qc_candidate → ...freeze_qc_candidate_STATUS.regenie on disk.
  step2_gene_based_mid: "step2_gene_based_STATUS.freeze_qc_candidate_STATUS"
  tests_keep: ""
  masks_keep: ""
  thresh_keep: ""
  p_strong: 0.001
  p_support: 0.05
  p_weak: 0.1
  abs_beta_max_reasonable: 5
  se_max_reasonable: 2
  abs_beta_diff_max: 1.5
  genes_sanity: "CHEK2"
  n_label_volcano: 5
---

```{=html}
<style type="text/css">
/* body */
body { font-size: 1.24rem; line-height: 1.5; }
.main-container { max-width: 52rem; }
/* title block */
h1.title { font-size: 1.65rem; margin-top: 0.5rem; }
h4.author, h4.date { font-size: 0.95rem; }
/* headings: modest vs body */
h1, h2, h3 {
  margin-top: 0.6rem;
}
h1 { font-size: 1.3rem; margin-bottom: 0.45rem; font-weight: 600; }
h2 { font-size: 1.08rem; margin-bottom: 0.3rem; font-weight: 600; }
h3 { font-size: 0.98rem; margin-bottom: 0.22rem; font-weight: 600; }
h1:first-of-type,
h2:first-of-type {
  margin-top: 0.3rem;
}
#TOC, #toc { font-size: 0.88rem; line-height: 1.35; }
#TOC .tocify-header { margin-top: 0.2rem; }
.table, table { font-size: 0.95rem; }
pre.sourceCode, pre { padding-top: 0.3rem !important; padding-bottom: 0.3rem !important; margin-top: 0.35rem; margin-bottom: 0.35rem; }
.code-folding-btn { margin-bottom: 0.25rem; }
</style>
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 10,
  fig.height = 6
)
library(tidyverse)
library(readr)
library(knitr)
library(kableExtra)
library(ggrepel)

theme_set(theme_minimal(base_size = 13))

dir.create(params$out_dir, recursive = TRUE, showWarnings = FALSE)
```

# Goal

This report evaluates **cross-freeze consistency under strong freeze/sample imbalance**: Freeze 3 carries substantially more case information than Freeze 2. **Stronger signals in Freeze 3 are therefore expected for many rare tests and are not, by themselves, evidence against biology.** Conversely, **lack of Freeze 2 support for rare signals does not imply absence of effect**—it often reflects **carrier sparsity and power imbalance**, not a clean “failed in F2” verdict.

The notebook is **not** a neutral results-ranking meta-analysis. It is a **pre-specified stratified comparison framework** for triage: where do the two freezes agree in direction and stability, where does one freeze dominate numerically, and where do estimates conflict badly enough to quarantine for review?

This report is for **exploratory comparison only**. Freeze differences are interpreted as a **mix of power imbalance and residual technical/capture structure**, not as a simple artifact-vs-truth dichotomy.

Methods in brief: for **ADD**, inverse-variance (IVW) meta is computed on all overlapping rows with finite betas and SEs in both freezes; interpretation then uses direction, heterogeneity, and Freeze-3 dominance flags. For **SKAT/SKATO**, Fisher’s method is used as the **single** exploratory combined *p*-value summary, shown alongside the per-freeze *p*-values.

Accordingly, meta-analytic summaries are interpreted in the context of freeze-specific precision and dominance, rather than as simple pooled effect estimates.

### How to use this report

- Use **Table B** (consistent-direction ADD subset) for cross-freeze consistency.  
- Use **Table C** (Freeze 3 stronger than Freeze 2) for Freeze-3-driven *p*-value patterns.  
- Use **Table D** and **CHEK2 discrepancy** for opposite-sign or pattern mismatches; use **`DOMINATED_BY_F3`** / **F3-dominated** priority for “F3 signal, limited F2 precision in the IVW” (gene-level only).  
- For **SKAT/SKATO**, read **`FISHER_P`** only together with **`PVAL_F2`** and **`PVAL_F3`**; Fisher is the combined triage metric, not a replacement for the per-freeze results.

---

# 1. Load

```{r load-files, echo=FALSE}
#' Read chr1–22 step2 gene-based .regenie for one freeze; same parsing as report_regenie.Rmd.
#' Resolves stem: REGENIE writes {--out}_STATUS.regenie for binary --phenoCol STATUS, so YAML may
#' give either ...freeze_qc_candidate or ...freeze_qc_candidate_STATUS.
gene_based_mid_candidates <- function(mid) {
  u <- unique(c(mid, paste0(mid, "_STATUS")))
  if (nzchar(mid) && endsWith(mid, "_STATUS")) {
    u <- unique(c(u, sub("_STATUS$", "", mid)))
  }
  u
}

load_regenie_gene_based <- function(project_name, regenie_dir, step2_gene_based_mid) {
  cands <- gene_based_mid_candidates(step2_gene_based_mid)
  mid_use <- NULL
  for (m in cands) {
    fp1 <- file.path(regenie_dir, paste0(project_name, ".chr1.", m, ".regenie"))
    if (file.exists(fp1)) {
      mid_use <- m
      break
    }
  }
  if (is.null(mid_use)) {
    stop(
      "No gene-based .regenie for project ", project_name, " under ", regenie_dir,
      " (tried stems: ", paste(cands, collapse = ", "), ")"
    )
  }
  message("Gene-based REGENIE stem (chrN.", mid_use, ".regenie): ", mid_use)

  lst <- list()
  for (i in 1:22) {
    fp <- file.path(regenie_dir, paste0(project_name, ".chr", i, ".", mid_use, ".regenie"))
    if (file.exists(fp)) {
      lst[[length(lst) + 1]] <- readr::read_delim(
        fp, delim = " ", col_names = TRUE, trim_ws = TRUE, comment = "#", show_col_types = FALSE
      )
    }
  }
  if (!length(lst)) {
    stop(
      "No gene-based .regenie files for project ", project_name,
      " (stem *.", mid_use, ".regenie) under ", regenie_dir
    )
  }
  tb <- bind_rows(lst)
  for (col in c("A1FREQ", "A1FREQ_CASES", "A1FREQ_CONTROLS")) {
    if (!col %in% names(tb)) tb[[col]] <- NA_real_
  }
  tb %>%
    mutate(
      across(c(CHROM, GENPOS, N, N_CASES, N_CONTROLS), as.integer),
      across(c(A1FREQ, A1FREQ_CASES, A1FREQ_CONTROLS, BETA, SE, CHISQ, LOG10P), as.numeric),
      PVAL = 10^(-LOG10P),
      GENE = sub("\\..*", "", ID),
      MASK = sub(".*\\.(M[0-9]+).*", "\\1", ID),
      THRESH_RAW = sub(".*\\.(.*)$", "\\1", ID),
      # Keep THRESH as character only (e.g. "0.01", "All variants"); do not coerce to numeric downstream.
      THRESH = as.character(ifelse(
        grepl("^\\d+\\.\\d+$", THRESH_RAW), THRESH_RAW,
        ifelse(THRESH_RAW == "all", "All variants", THRESH_RAW)
      )),
      TEST = as.character(TEST),
      OR_ADD = if_else(TEST == "ADD" & is.finite(BETA), exp(BETA), NA_real_)
    ) %>%
    transmute(
      GENE, MASK, THRESH, TEST, BETA, SE, PVAL, N_CASES, N_CONTROLS,
      OR_ADD, A1FREQ_CASES, A1FREQ_CONTROLS
    )
}

f2_long <- load_regenie_gene_based(
  params$project_f2, params$regenie_dir_f2, params$step2_gene_based_mid
)
f3_long <- load_regenie_gene_based(
  params$project_f3, params$regenie_dir_f3, params$step2_gene_based_mid
)

message(
  "REGENIE gene-based dirs — F2: ", params$regenie_dir_f2,
  " (project ", params$project_f2, ", mid ", params$step2_gene_based_mid, ") | F3: ",
  params$regenie_dir_f3, " (project ", params$project_f3, ")"
)

readr::write_csv(
  dplyr::bind_rows(
    f2_long %>% mutate(FREEZE = "F2"),
    f3_long %>% mutate(FREEZE = "F3")
  ),
  file.path(params$out_dir, "gene_regenie_standardized_F2_F3_long.csv")
)
message("Wrote: ", file.path(params$out_dir, "gene_regenie_standardized_F2_F3_long.csv"))

f2 <- f2_long %>%
  rename(
    BETA_F2 = BETA, SE_F2 = SE, PVAL_F2 = PVAL, OR_F2 = OR_ADD,
    N_CASES_F2 = N_CASES, N_CONTROLS_F2 = N_CONTROLS,
    A1FREQ_CASES_F2 = A1FREQ_CASES, A1FREQ_CONTROLS_F2 = A1FREQ_CONTROLS
  )

f3 <- f3_long %>%
  rename(
    BETA_F3 = BETA, SE_F3 = SE, PVAL_F3 = PVAL, OR_F3 = OR_ADD,
    N_CASES_F3 = N_CASES, N_CONTROLS_F3 = N_CONTROLS,
    A1FREQ_CASES_F3 = A1FREQ_CASES, A1FREQ_CONTROLS_F3 = A1FREQ_CONTROLS
  )

message("Freeze 2 rows: ", nrow(f2), " | Freeze 3 rows: ", nrow(f3))
```

```{r sample-size-summary, echo=FALSE}
ss <- dplyr::bind_rows(
  f2_long %>%
    dplyr::summarise(
      Freeze = "F2",
      gene_test_rows = dplyr::n(),
      median_N_cases = stats::median(N_CASES, na.rm = TRUE),
      median_N_controls = stats::median(N_CONTROLS, na.rm = TRUE),
      unique_N_cases = dplyr::n_distinct(N_CASES),
      unique_N_controls = dplyr::n_distinct(N_CONTROLS)
    ),
  f3_long %>%
    dplyr::summarise(
      Freeze = "F3",
      gene_test_rows = dplyr::n(),
      median_N_cases = stats::median(N_CASES, na.rm = TRUE),
      median_N_controls = stats::median(N_CONTROLS, na.rm = TRUE),
      unique_N_cases = dplyr::n_distinct(N_CASES),
      unique_N_controls = dplyr::n_distinct(N_CONTROLS)
    )
)
knitr::kable(
  ss,
  digits = 2,
  caption = "Per-freeze row counts (all gene-test rows loaded) and distribution of per-row N_CASES / N_CONTROLS (anchors the case-count imbalance before filters)."
)
```

---

# 2. Filter

Optional YAML: `tests_keep`, `masks_keep`, `thresh_keep` (REGENIE `TEST` strings, e.g. `ADD`, `ADD-SKAT`, `ADD-SKATO`).

```{r filter-tests-masks, echo=FALSE}
parse_csv_param <- function(x) {
  x <- trimws(x)
  if (is.null(x) || length(x) != 1 || x == "") return(character(0))
  strsplit(x, ",", fixed = TRUE)[[1]] %>% trimws() %>% .[. != ""]
}

tests_keep <- parse_csv_param(params$tests_keep)
masks_keep <- parse_csv_param(params$masks_keep)
thresh_keep <- parse_csv_param(params$thresh_keep)

filter_slice <- function(df) {
  out <- df
  if (length(tests_keep)) out <- out %>% filter(TEST %in% tests_keep)
  if (length(masks_keep)) out <- out %>% filter(MASK %in% masks_keep)
  if (length(thresh_keep)) out <- out %>% filter(THRESH %in% thresh_keep)
  out
}

f2_f <- filter_slice(f2)
f3_f <- filter_slice(f3)

message(
  "After filters — F2: ", nrow(f2_f), " rows; F3: ", nrow(f3_f), " rows. ",
  "TEST filter: ", if (length(tests_keep)) paste(tests_keep, collapse = ", ") else "(none)",
  " | MASK filter: ", if (length(masks_keep)) paste(masks_keep, collapse = ", ") else "(none)"
)
```

---

# 3. Join

```{r merge-check, echo=FALSE}
key <- c("GENE", "MASK", "THRESH", "TEST")

dup_f2 <- f2_f %>% count(!!!syms(key), name = "n") %>% filter(n > 1)
dup_f3 <- f3_f %>% count(!!!syms(key), name = "n") %>% filter(n > 1)

merged <- inner_join(f2_f, f3_f, by = key)

only_f2 <- anti_join(f2_f, f3_f, by = key)
only_f3 <- anti_join(f3_f, f2_f, by = key)

message(
  "Matched rows (inner join): ", nrow(merged), "\n",
  "Only in Freeze 2: ", nrow(only_f2), "\n",
  "Only in Freeze 3: ", nrow(only_f3), "\n",
  "Duplicate keys in F2 (after filter): ", nrow(dup_f2), "\n",
  "Duplicate keys in F3 (after filter): ", nrow(dup_f3)
)

if (nrow(dup_f2)) knitr::kable(head(dup_f2, 20), caption = "Unexpected duplicate keys — Freeze 2 (first 20)")
if (nrow(dup_f3)) knitr::kable(head(dup_f3, 20), caption = "Unexpected duplicate keys — Freeze 3 (first 20)")

# Freeze-specific keys (capture / join debugging; do not discard)
readr::write_csv(only_f2, file.path(params$out_dir, "only_F2.csv"))
readr::write_csv(only_f3, file.path(params$out_dir, "only_F3.csv"))
readr::write_csv(
  dplyr::filter(only_f2, TEST == "ADD-SKATO"),
  file.path(params$out_dir, "only_F2_SKATO.csv")
)
readr::write_csv(
  dplyr::filter(only_f3, TEST == "ADD-SKATO"),
  file.path(params$out_dir, "only_F3_SKATO.csv")
)
message("Wrote unmatched keys: ", file.path(params$out_dir, "only_F2.csv"), " | ", file.path(params$out_dir, "only_F3.csv"))
message("Wrote ADD-SKATO-only unmatched: ", file.path(params$out_dir, "only_F2_SKATO.csv"), " | ", file.path(params$out_dir, "only_F3_SKATO.csv"))
```

Unmatched keys in **`only_F2.csv`** / **`only_F3.csv`** can reflect **freeze-specific variant-set composition**, **differences in annotation / set construction**, or **model output present in only one freeze**—they are **not automatically a QC failure** of the join logic.

# 3b. SKAT/SKATO

Fisher’s method **combines evidence from two *p*-values**; it **does not combine effect directions** and can be **driven by one very small freeze-specific *p***. Use the SKAT/SKATO block **only for exploratory triage and ranking**, not for primary claims. In this notebook, Fisher’s method is the **only** combined summary used for SKAT/SKATO; the per-freeze *p*-values are retained for interpretation and to detect one-freeze-driven signals.

```{r meta-fisher-vc, echo=FALSE}
fisher_two_p <- function(p1, p2) {
  eps <- .Machine$double.eps
  p_floor <- 1e-300
  ok <- is.finite(p1) & is.finite(p2) & p1 > 0 & p1 <= 1 & p2 > 0 & p2 <= 1
  p1c <- pmax(pmin(p1, 1 - eps), p_floor)
  p2c <- pmax(pmin(p2, 1 - eps), p_floor)
  chi <- -2 * (log(p1c) + log(p2c))
  out <- rep(NA_real_, length(p1))
  out[ok] <- pchisq(chi[ok], df = 4, lower.tail = FALSE)
  out
}

psup <- as.numeric(params$p_support)
pwk <- as.numeric(params$p_weak)

meta_skat <- merged %>%
  dplyr::filter(TEST == "ADD-SKAT") %>%
  mutate(
    FISHER_P = fisher_two_p(PVAL_F2, PVAL_F3),
    BOTH_FREEZE_NOMINAL = PVAL_F2 <= psup & PVAL_F3 <= psup,
    BOTH_FREEZE_WEAK = PVAL_F2 <= pwk & PVAL_F3 <= pwk
  ) %>%
  mutate(FISHER_P_ADJ = p.adjust(FISHER_P, method = "BH"))

meta_skat <- meta_skat %>%
  mutate(
    priority = case_when(
      FISHER_P_ADJ <= 0.05 & BOTH_FREEZE_WEAK ~ "High",
      FISHER_P_ADJ <= 0.1 & BOTH_FREEZE_WEAK ~ "Medium",
      TRUE ~ "Review"
    )
  ) %>%
  arrange(FISHER_P_ADJ, FISHER_P)

meta_skato <- merged %>%
  dplyr::filter(TEST == "ADD-SKATO") %>%
  mutate(
    FISHER_P = fisher_two_p(PVAL_F2, PVAL_F3),
    BOTH_FREEZE_NOMINAL = PVAL_F2 <= psup & PVAL_F3 <= psup,
    BOTH_FREEZE_WEAK = PVAL_F2 <= pwk & PVAL_F3 <= pwk,
    DOMINANT_FREEZE = case_when(
      !is.finite(PVAL_F2) & !is.finite(PVAL_F3) ~ NA_character_,
      !is.finite(PVAL_F2) ~ "F3",
      !is.finite(PVAL_F3) ~ "F2",
      PVAL_F3 < PVAL_F2 ~ "F3",
      PVAL_F2 < PVAL_F3 ~ "F2",
      TRUE ~ "both"
    )
  ) %>%
  mutate(FISHER_P_ADJ = p.adjust(FISHER_P, method = "BH"))

meta_skato <- meta_skato %>%
  mutate(
    priority = case_when(
      FISHER_P_ADJ <= 0.05 & BOTH_FREEZE_WEAK ~ "High",
      FISHER_P_ADJ <= 0.1 & BOTH_FREEZE_WEAK ~ "Medium",
      TRUE ~ "Review"
    )
  ) %>%
  arrange(FISHER_P_ADJ, FISHER_P)

message("ADD-SKAT rows (merged): ", nrow(meta_skat), " | ADD-SKATO rows (merged): ", nrow(meta_skato))
```

---

# 4. ADD — IVW meta (all IVW-eligible overlap) with consistency / heterogeneity / dominance tiers

```{r meta-ivw, echo=FALSE}
# w_i = 1/SE^2; applies only to TEST == ADD (directional log-OR scale)

meta_ivw2 <- function(b1, se1, b2, se2) {
  ok <- is.finite(b1) & is.finite(b2) & is.finite(se1) & is.finite(se2) & se1 > 0 & se2 > 0
  w1 <- ifelse(ok, 1 / se1^2, NA_real_)
  w2 <- ifelse(ok, 1 / se2^2, NA_real_)
  sw <- w1 + w2
  beta_m <- ifelse(ok, (w1 * b1 + w2 * b2) / sw, NA_real_)
  se_m <- ifelse(ok, sqrt(1 / sw), NA_real_)
  z <- beta_m / se_m
  p <- 2 * pnorm(-abs(z))
  tibble(META_BETA = beta_m, META_SE = se_m, META_Z = z, META_P = p)
}

merged_add_all <- merged %>% dplyr::filter(TEST == "ADD")
merged_add_ivw <- merged_add_all %>%
  dplyr::filter(
    is.finite(BETA_F2), is.finite(BETA_F3),
    is.finite(SE_F2), is.finite(SE_F3), SE_F2 > 0, SE_F3 > 0
  )
n_discord <- merged_add_all %>%
  dplyr::filter(
    is.finite(BETA_F2), is.finite(BETA_F3),
    sign(BETA_F2) != 0, sign(BETA_F3) != 0,
    sign(BETA_F2) != sign(BETA_F3)
  ) %>%
  nrow()

message(
  "ADD rows (inner merge): ", nrow(merged_add_all),
  " | IVW-eligible (finite BETA/SE, SE>0): ", nrow(merged_add_ivw),
  " | opposite-sign overlap (Table D scope): ", n_discord
)

meta_add_df <- merged_add_ivw %>%
  bind_cols(meta_ivw2(.$BETA_F2, .$SE_F2, .$BETA_F3, .$SE_F3)) %>%
  mutate(
    META_OR = exp(META_BETA),
    META_P_ADJ = p.adjust(META_P, method = "BH")
  )

message("ADD rows entering IVW + flags (meta_add_df): ", nrow(meta_add_df))
```

---

# 5. Flags

```{r flags, echo=FALSE}
signz <- function(x) {
  x <- as.numeric(x)
  ifelse(!is.finite(x), NA_integer_, ifelse(x > 0, 1L, ifelse(x < 0, -1L, 0L)))
}

p_strong <- as.numeric(params$p_strong)
p_support <- as.numeric(params$p_support)
p_weak <- as.numeric(params$p_weak)
ab_max <- as.numeric(params$abs_beta_max_reasonable)
se_max <- as.numeric(params$se_max_reasonable)
abd_max <- as.numeric(params$abs_beta_diff_max)

meta_add_df <- meta_add_df %>%
  mutate(
    WEIGHT_F2 = ifelse(is.finite(SE_F2) & SE_F2 > 0, 1 / SE_F2^2, NA_real_),
    WEIGHT_F3 = ifelse(is.finite(SE_F3) & SE_F3 > 0, 1 / SE_F3^2, NA_real_),
    WEIGHT_RATIO = WEIGHT_F3 / (WEIGHT_F2 + WEIGHT_F3),
    DOMINATED_BY_F3 = is.finite(WEIGHT_RATIO) & WEIGHT_RATIO > 0.8,
    Q = ifelse(
      is.finite(META_BETA) & is.finite(SE_F2) & is.finite(SE_F3) & SE_F2 > 0 & SE_F3 > 0,
      (BETA_F2 - META_BETA)^2 / SE_F2^2 + (BETA_F3 - META_BETA)^2 / SE_F3^2,
      NA_real_
    ),
    Q_P = ifelse(is.finite(Q), pchisq(Q, df = 1, lower.tail = FALSE), NA_real_),
    P_HET = Q_P,
    HETEROGENEITY = is.finite(Q_P) & (Q_P < 0.05),
    HET_FLAG = is.finite(Q_P) & (Q_P < 0.05),
    SIGN_F2 = signz(BETA_F2),
    SIGN_F3 = signz(BETA_F3),
    CONSISTENT_DIRECTION = (SIGN_F2 == SIGN_F3) & (SIGN_F2 != 0) & (SIGN_F3 != 0),
    PARTIAL_DIRECTION = (SIGN_F2 == SIGN_F3) | (SIGN_F2 == 0 | SIGN_F3 == 0),
    ABS_BETA_DIFF = abs(BETA_F2 - BETA_F3),
    BEST_FREEZE = case_when(
      !is.finite(PVAL_F2) & !is.finite(PVAL_F3) ~ NA_character_,
      !is.finite(PVAL_F2) ~ "F3",
      !is.finite(PVAL_F3) ~ "F2",
      PVAL_F3 < PVAL_F2 ~ "F3",
      PVAL_F2 < PVAL_F3 ~ "F2",
      TRUE ~ "tie"
    ),
    OPPOSITE_DIRECTION = (SIGN_F2 != SIGN_F3) & (SIGN_F2 != 0) & (SIGN_F3 != 0),
    strong_F2 = is.finite(PVAL_F2) & (PVAL_F2 <= p_strong),
    strong_F3 = is.finite(PVAL_F3) & (PVAL_F3 <= p_strong),
    support_F2 = is.finite(PVAL_F2) & (PVAL_F2 <= p_support),
    support_F3 = is.finite(PVAL_F3) & (PVAL_F3 <= p_support),
    weak_F2 = is.finite(PVAL_F2) & (PVAL_F2 <= p_weak),
    weak_F3 = is.finite(PVAL_F3) & (PVAL_F3 <= p_weak),
    BOTH_WEAK_SUPPORT = weak_F2 & weak_F3,
    BOTH_SUPPORT = (strong_F2 & support_F3 & PARTIAL_DIRECTION) |
      (strong_F3 & support_F2 & PARTIAL_DIRECTION),
    ALT_CT_CASES_F2_EST = ifelse(
      is.finite(N_CASES_F2) & is.finite(A1FREQ_CASES_F2),
      2 * N_CASES_F2 * A1FREQ_CASES_F2,
      NA_real_
    ),
    ALT_CT_CONTROLS_F2_EST = ifelse(
      is.finite(N_CONTROLS_F2) & is.finite(A1FREQ_CONTROLS_F2),
      2 * N_CONTROLS_F2 * A1FREQ_CONTROLS_F2,
      NA_real_
    ),
    ALT_CT_CASES_F3_EST = ifelse(
      is.finite(N_CASES_F3) & is.finite(A1FREQ_CASES_F3),
      2 * N_CASES_F3 * A1FREQ_CASES_F3,
      NA_real_
    ),
    ALT_CT_CONTROLS_F3_EST = ifelse(
      is.finite(N_CONTROLS_F3) & is.finite(A1FREQ_CONTROLS_F3),
      2 * N_CONTROLS_F3 * A1FREQ_CONTROLS_F3,
      NA_real_
    )
  )

meta_add_df <- meta_add_df %>%
  mutate(
    reasonable_se = is.finite(SE_F2) & is.finite(SE_F3) & SE_F2 <= se_max & SE_F3 <= se_max,
    reasonable_beta = abs(BETA_F2) <= ab_max & abs(BETA_F3) <= ab_max,
    not_wild_diff = ABS_BETA_DIFF <= abd_max,
    min_support_ok = support_F2 | support_F3
  )

meta_add_df <- meta_add_df %>%
  mutate(
    priority = case_when(
      OPPOSITE_DIRECTION ~ "Discordant",
      DOMINATED_BY_F3 ~ "F3-dominated",
      CONSISTENT_DIRECTION & HET_FLAG ~ "Review",
      is.finite(META_P) & META_P <= p_strong & CONSISTENT_DIRECTION & reasonable_se &
        reasonable_beta & not_wild_diff & !HET_FLAG & min_support_ok ~ "High",
      ((strong_F2 & weak_F3) | (strong_F3 & weak_F2)) & CONSISTENT_DIRECTION ~ "Medium",
      TRUE ~ "Review"
    )
  )
```

`P_HET` = Cochran-style *Q* p-value (same as `Q_P`); `HET_FLAG` marks *P*~het~ \< 0.05 (review, do not auto-drop). Longer definitions → **Notes**.

**Stability thresholds** (`abs_beta_max_reasonable`, `se_max_reasonable`, `abs_beta_diff_max` from YAML) are **pragmatic guards** to de-prioritize rows with implausibly unstable betas/SEs or huge cross-freeze beta gaps—they are **not formal exclusion criteria** and should not be read as “curating truth.”

---

# 5b. SKAT/SKATO

**Columns in the tables below:** **`PVAL_F2`** and **`PVAL_F3`** are the REGENIE gene-based *p*-values for each freeze. **`FISHER_P`** is the **single** exploratory combined *p*-value summary used here for SKAT/SKATO; it is shown together with the per-freeze *p*-values so one-freeze-driven signals remain visible. **`FISHER_P_ADJ`** — Benjamini–Hochberg adjustment of **`FISHER_P`** within this SKAT or SKATO block. **`DOMINANT_FREEZE`** (SKATO only), **`BOTH_FREEZE_NOMINAL`**, **`BOTH_FREEZE_WEAK`**, **`priority`** — exploratory triage labels (see **Notes**): **High** requires **`FISHER_P_ADJ` ≤ 0.05** and **`BOTH_FREEZE_WEAK`** (weak support in **both** freezes).

```{r tables-vc, echo=FALSE}
vc_show <- function(df, title) {
  cat("### ", title, "\n\n", sep = "")
  if (!nrow(df)) {
    cat("_No rows after merge._\n\n")
    return(invisible(NULL))
  }
  show_tbl <- function(x, n = Inf) {
    if (is.finite(n)) x <- head(x, n)
    x %>%
      kable(digits = 4, format.args = list(scientific = FALSE)) %>%
      kable_styling(bootstrap_options = c("striped", "hover"), font_size = 13, full_width = FALSE) %>%
      scroll_box(width = "100%", height = "520px")
  }
  show_tbl(
    df %>%
      select(any_of(c(
        "GENE", "MASK", "THRESH", "PVAL_F2", "PVAL_F3", "FISHER_P", "FISHER_P_ADJ",
        "DOMINANT_FREEZE", "BOTH_FREEZE_NOMINAL", "BOTH_FREEZE_WEAK", "priority"
      )))
  )
  cat("\n")
}
vc_show(meta_skato, "SKATO")
vc_show(meta_skat, "SKAT")
```

---

# 6. Results

**ADD (inverse-variance):** **IVW** (`META_P`, `META_P_ADJ`, `META_BETA`, …) is computed on **every overlapping ADD row with finite `BETA` and `SE` in both freezes**, **without pre-filtering by direction** (`1/SE²` weights)—the usual fixed-effect two-study summary. **Interpretation** is tiered afterward: **`CONSISTENT_DIRECTION`** vs **`OPPOSITE_DIRECTION`**, **`HET_FLAG`** (*Q* heterogeneity), and **`DOMINATED_BY_F3`** (`WEIGHT_RATIO` \> 0.8; F3-driven combined estimate with limited F2 precision contribution). **Opposite-sign rows** stay in **`meta_add_df`** for a transparent pooled *p*, and are **also surfaced in Table D** because the pooled effect is **not** a directional-agreement summary.

**Columns in Tables A–C and the comparison shortlist:** **`PVAL_F2`**, **`PVAL_F3`** — REGENIE ADD *p*-values per freeze. **`META_BETA`**, **`META_SE`**, **`META_Z`**, **`META_P`** — IVW on **all IVW-eligible overlap rows**; **`META_P`** is two-sided from **`META_Z`**. **`META_P_ADJ`** — BH FDR of **`META_P`** across **`meta_add_df`** (full IVW pool). **`HET_FLAG`** — TRUE if **`Q_P`** \< 0.05 (Cochran *Q* vs **`META_BETA`**; same as **`P_HET`** in CSVs). **`CONSISTENT_DIRECTION`**, **`OPPOSITE_DIRECTION`**, **`DOMINATED_BY_F3`**, **`WEIGHT_RATIO`**. Exports include **`WEIGHT_F2`**, **`WEIGHT_F3`**, betas/SEs, **`META_OR`**.

```{r tables-abcd}
show_tbl <- function(df, n = 50) {
  if (is.finite(n)) df <- head(df, n)
  df %>%
    kable(digits = 4, format.args = list(scientific = FALSE)) %>%
    kable_styling(bootstrap_options = c("striped", "hover"), font_size = 13, full_width = FALSE) %>%
    scroll_box(width = "100%", height = "560px")
}

## A (data). All IVW-eligible overlapping ADD rows
table_a <- meta_add_df %>% arrange(META_P, PVAL_F3, PVAL_F2)

## B. Consistent (Table B filters)
table_b <- meta_add_df %>%
  filter(CONSISTENT_DIRECTION, reasonable_se, reasonable_beta, not_wild_diff) %>%
  arrange(META_P)

## C. Freeze 3 stronger (Table C filters)
table_c <- meta_add_df %>%
  filter(
    strong_F3,
    is.finite(PVAL_F2) & PVAL_F2 > p_support,
    PARTIAL_DIRECTION
  ) %>%
  arrange(PVAL_F3, META_P)

## D. Discordant ADD overlap (opposite non-zero beta signs)
table_d <- merged_add_all %>%
  dplyr::filter(
    is.finite(BETA_F2),
    is.finite(BETA_F3),
    sign(BETA_F2) != 0, sign(BETA_F3) != 0,
    sign(BETA_F2) != sign(BETA_F3)
  ) %>%
  dplyr::left_join(
    meta_add_df %>%
      dplyr::select(
        GENE, MASK, THRESH, META_P, META_P_ADJ, HET_FLAG, WEIGHT_RATIO, DOMINATED_BY_F3
      ),
    by = c("GENE", "MASK", "THRESH")
  ) %>%
  arrange(pmin(PVAL_F2, PVAL_F3, na.rm = TRUE))
```

## A. All IVW-eligible overlapping ADD rows

All overlapping ADD rows with **finite `BETA` and `SE` in both freezes** (`meta_add_df`), **IVW** applied **regardless of sign**; sorted by `META_P`. Use **`CONSISTENT_DIRECTION`**, **`HET_FLAG`**, and **`OPPOSITE_DIRECTION`** (in CSVs) for interpretation—not row drops before meta.
**Display cap: first 100 rows shown in HTML (full table is still written to CSV outputs).**

```{r table-a, echo=FALSE}
show_tbl(
  table_a %>%
    select(
      GENE, MASK, THRESH, META_P, META_P_ADJ, PVAL_F2, PVAL_F3,
      HET_FLAG, CONSISTENT_DIRECTION, OPPOSITE_DIRECTION, DOMINATED_BY_F3
    ),
  n = 100
)
```

## B. Consistent same-direction overlapping ADD rows

Subset of Table **A** restricted to same-direction rows for interpretability; meta-analysis itself is performed on all IVW-eligible rows. Same subset also applies pragmatic stability filters (`reasonable_se`, `reasonable_beta`, `not_wild_diff`; see **5. Flags**).
**Display cap: first 100 rows shown in HTML (full table is still written to CSV outputs).**

```{r table-b, echo=FALSE}
show_tbl(
  table_b %>%
    select(GENE, MASK, THRESH, META_P, META_P_ADJ, PVAL_F2, PVAL_F3, HET_FLAG),
  n = 100
)
```

## C. Freeze 3 stronger than Freeze 2

Rows with stronger evidence in Freeze 3 and weaker/non-nominal evidence in Freeze 2 (exploratory “F3-driven” view).
**Display cap: first 100 rows shown in HTML (full table is still written to CSV outputs).**

```{r table-c, echo=FALSE}
show_tbl(
  table_c %>%
    select(GENE, MASK, THRESH, META_P, META_P_ADJ, PVAL_F2, PVAL_F3, HET_FLAG),
  n = 100
)
```

## D. Discordant ADD rows

**Columns:** **`PVAL_F2`**, **`PVAL_F3`**, and (when the row is **IVW-eligible**) joined **`META_P`**, **`META_P_ADJ`**, **`HET_FLAG`**, **`WEIGHT_RATIO`**. Opposite-sign **`META_P`** is a **pooled** IVW statistic for transparency; **do not** read it as cross-freeze directional agreement.

ADD overlap rows with **opposite non-zero beta signs** between freezes (quarantined for interpretation; full overlap IVW lives in **Table A** / **`meta_ADD.csv`**).
**Display cap: first 100 rows shown in HTML (full table is still written to CSV outputs).**

```{r table-d, echo=FALSE}
show_tbl(
  table_d %>%
    select(
      GENE, MASK, THRESH, PVAL_F2, PVAL_F3, META_P, META_P_ADJ, HET_FLAG, WEIGHT_RATIO
    ),
  n = 100
)
```

## Cross-freeze comparison shortlist (ADD, IVW pool)

Sorted by `META_P` over **all IVW-eligible overlap** rows. For exploratory triage; full columns in CSV.
**Display cap: first 100 rows shown in HTML (full table is still written to CSV outputs).**

```{r comparison-table, echo=FALSE}
comparison_tbl <- meta_add_df %>%
  transmute(
    GENE,
    MASK,
    THRESH,
    P_F2 = PVAL_F2,
    P_F3 = PVAL_F3,
    P_META = META_P,
    HET_FLAG,
    META_P_ADJ
  ) %>%
  arrange(P_META, P_F3, P_F2)
```

```{r comparison-table-show, echo=FALSE}
if (nrow(comparison_tbl) > 0) {
  show_tbl(
    comparison_tbl %>%
      select(GENE, MASK, THRESH, P_META, META_P_ADJ, P_F2, P_F3, HET_FLAG),
    n = 100
  )
} else {
  cat("_No rows in comparison shortlist._\n\n")
}
```

---

# 7. CHEK2 discrepancy

CHEK2 is used here as a reference gene to understand differences between freezes (gene-level only; variant-level missingness is out of scope for this notebook).

The first table shows exploratory pattern labels; the second is a side-by-side numeric comparison. **`CHEK2_PATTERN`** labels are for triage; **`META_P`**, **`META_P_ADJ`**, **`PVAL_F*`**, **`BETA_F*`**, **`HET_FLAG`**, **`BEST_FREEZE`** match **6. Results** where applicable.

```{r sanity-genes, echo=FALSE}
sanity_genes <- parse_csv_param(params$genes_sanity)
if (!length(sanity_genes)) sanity_genes <- "CHEK2"
ps_chek2 <- as.numeric(params$p_support)

chek2_tbl <- merged %>%
  dplyr::filter(GENE %in% sanity_genes, TEST == "ADD") %>%
  dplyr::left_join(
    meta_add_df %>% dplyr::select(GENE, MASK, THRESH, META_P, META_P_ADJ, HET_FLAG, BEST_FREEZE),
    by = c("GENE", "MASK", "THRESH")
  ) %>%
  dplyr::mutate(
    opposite_beta = is.finite(BETA_F2) & is.finite(BETA_F3) &
      sign(BETA_F2) != 0 & sign(BETA_F3) != 0 & sign(BETA_F2) != sign(BETA_F3),
    CHEK2_PATTERN = dplyr::case_when(
      opposite_beta ~ "discordant",
      is.finite(PVAL_F3) & PVAL_F3 <= ps_chek2 &
        (!is.finite(PVAL_F2) | PVAL_F2 > ps_chek2) ~ "F3 stronger, F2 weak",
      is.finite(PVAL_F2) & PVAL_F2 <= ps_chek2 &
        (!is.finite(PVAL_F3) | PVAL_F3 > ps_chek2) ~ "F2 only",
      is.finite(PVAL_F2) & PVAL_F2 <= ps_chek2 & is.finite(PVAL_F3) & PVAL_F3 <= ps_chek2 ~ "Both support",
      TRUE ~ "unclear"
    )
  ) %>%
  dplyr::arrange(GENE, MASK, THRESH)

knitr::kable(
  chek2_tbl %>%
    dplyr::select(
      GENE, MASK, THRESH, CHEK2_PATTERN, BEST_FREEZE, META_P, META_P_ADJ,
      PVAL_F2, PVAL_F3, BETA_F2, BETA_F3, HET_FLAG
    ),
  digits = 5,
  caption = "CHEK2 ADD rows: exploratory pattern labels (not a clinical interpretation)"
)
```

```{r chek2-side-by-side, echo=FALSE}
knitr::kable(
  chek2_tbl %>%
    dplyr::select(
      GENE, MASK, THRESH, BEST_FREEZE, HET_FLAG, PVAL_F2, PVAL_F3, META_P,
      BETA_F2, BETA_F3, N_CASES_F2, N_CASES_F3
    ),
  digits = 5,
  caption = "CHEK2: side-by-side comparison (Freeze 2 vs Freeze 3)"
)
```

Interpretation (exploratory):

For **CHEK2**, cross-freeze differences are **most plausibly driven by power imbalance and carrier sparsity in Freeze 2**, consistent with the observed case count asymmetry between freezes. **Variant-level callability / missingness** should be assessed in a **separate** variant workflow; **this gene-level notebook does not adjudicate CHEK2 as a missingness artifact.**

This report does not resolve mechanistic causes; it highlights where freezes diverge for targeted follow-up.

---

# 8. Plots

ADD **IVW pool** rows as in **6. Results**: *x*/*y* use **`PVAL_F2`** / **`PVAL_F3`** (and **`META_P`** for volcano *y* via **`LPM`** = −log10 **`META_P`**, capped).

```{r plot-data, echo=FALSE}
plot_df <- meta_add_df %>%
  mutate(
    LP2 = pmin(-log10(pmax(PVAL_F2, .Machine$double.xmin)), 50),
    LP3 = pmin(-log10(pmax(PVAL_F3, .Machine$double.xmin)), 50),
    LPM = pmin(-log10(pmax(META_P, .Machine$double.xmin)), 50)
  )
```

```{r plot-freeze-asymmetry, echo=FALSE, fig.height=4, fig.cap="ADD: how IVW weight is split between freezes (WEIGHT_RATIO = precision_F3 / (precision_F2 + precision_F3)). Values near 1 mean the combined estimate is dominated by Freeze 3."}
ggplot(
  dplyr::filter(meta_add_df, is.finite(WEIGHT_RATIO)),
  aes(x = WEIGHT_RATIO)
) +
  geom_histogram(bins = 40, fill = "gray45", color = "white") +
  labs(x = "WEIGHT_RATIO (F3 share of IVW precision)", y = "Count")
```

```{r plot-freeze-asymmetry-bar, echo=FALSE, fig.height=3.5, fig.cap="ADD: IVW-eligible overlap (all directions) vs opposite-sign discordant overlap (Table D; META columns joined when IVW-eligible)."}
n_discord_merged <- merged_add_all %>%
  dplyr::filter(
    is.finite(BETA_F2), is.finite(BETA_F3),
    sign(BETA_F2) != 0, sign(BETA_F3) != 0,
    sign(BETA_F2) != sign(BETA_F3)
  ) %>%
  nrow()
asym_counts <- tibble::tibble(
  category = c(
    "IVW-eligible overlap (meta_add_df, all signs)",
    "Opposite-sign discordant (merged ADD, Table D)"
  ),
  n = c(nrow(meta_add_df), n_discord_merged)
)
ggplot(asym_counts, aes(x = category, y = n)) +
  geom_col(width = 0.55, fill = "gray50") +
  labs(x = NULL, y = "Rows") +
  theme(axis.text.x = element_text(angle = 12, hjust = 1))
```

## P-values (F2 vs F3)

```{r plot-lp-scatter, echo=FALSE, fig.cap="ADD: −log10 p F2 vs F3 (capped 50)."}
ggplot(plot_df, aes(x = LP2, y = LP3)) +
  geom_point(alpha = 0.32, size = 2) +
  geom_abline(slope = 1, intercept = 0, linetype = 2, color = "gray50") +
  labs(
    x = expression(-log[10] ~ P[Freeze2]),
    y = expression(-log[10] ~ P[Freeze3]),
    caption = "−log10(p) capped at 50 for display."
  )
```

## Betas (F2 vs F3)

```{r plot-beta-scatter, echo=FALSE, fig.cap="ADD: BETA F2 vs F3."}
ggplot(plot_df, aes(x = BETA_F2, y = BETA_F3)) +
  geom_point(alpha = 0.32, size = 2) +
  geom_abline(slope = 1, intercept = 0, linetype = 2, color = "gray50") +
  geom_hline(yintercept = 0, color = "gray70") +
  geom_vline(xintercept = 0, color = "gray70") +
  labs(x = "BETA Freeze 2", y = "BETA Freeze 3")
```

## QQ

```{r plot-qq, echo=FALSE, fig.cap="ADD: QQ by freeze."}
qq_points <- function(p) {
  p <- p[is.finite(p) & p > 0 & p <= 1]
  n <- length(p)
  if (n < 10) return(tibble(expected = numeric(0), observed = numeric(0)))
  tibble(
    expected = -log10(ppoints(n)),
    observed = -log10(sort(p))
  )
}

qq_f2 <- qq_points(meta_add_df$PVAL_F2) %>% mutate(Freeze = "F2")
qq_f3 <- qq_points(meta_add_df$PVAL_F3) %>% mutate(Freeze = "F3")
qq_b <- bind_rows(qq_f2, qq_f3)

ggplot(qq_b, aes(x = expected, y = observed, color = Freeze)) +
  geom_point(alpha = 0.25, size = 0.8) +
  geom_abline(slope = 1, intercept = 0, color = "black", linetype = 2) +
  labs(x = expression(-log[10] ~ p[expected]), y = expression(-log[10] ~ p[observed]))
```

## Volcano

```{r plot-volcano, echo=FALSE, fig.cap="ADD volcano (−log10 META_P capped at 50)."}
nlab <- max(3L, as.integer(params$n_label_volcano))
genes_always <- parse_csv_param(params$genes_sanity)
if (!length(genes_always)) genes_always <- "CHEK2"
lab_sanity <- plot_df %>% filter(GENE %in% genes_always)
lab_ranked <- plot_df %>% arrange(META_P) %>% slice_head(n = nlab)
lab_both <- plot_df %>% filter(BOTH_SUPPORT) %>% arrange(META_P) %>% slice_head(n = 3)
volcano_labs <- bind_rows(lab_sanity, lab_ranked, lab_both) %>%
  distinct(GENE, MASK, THRESH, .keep_all = TRUE)

ggplot(plot_df, aes(x = META_BETA, y = LPM)) +
  geom_point(alpha = 0.32, size = 1.8) +
  geom_hline(yintercept = -log10(0.05), linetype = 3, color = "gray60") +
  geom_text_repel(
    data = volcano_labs,
    aes(label = GENE),
    size = 2.6,
    max.overlaps = 12,
    min.segment.length = 0
  ) +
  labs(
    x = "META_BETA",
    y = expression(-log[10] ~ P[META])
  )
```

---

# 9. Priority genes

**Columns:** same **`META_P`**, **`META_P_ADJ`**, **`HET_FLAG`** meaning as **6. Results**; **`priority`** is defined in the note after the table.

This is intentionally a short triage view (`slice_head(n = 15)`), not a full table of all genes.

```{r priority-tags, echo=FALSE}
priority_shortlist <- meta_add_df %>%
  filter(priority %in% c("High", "Medium")) %>%
  arrange(META_P) %>%
  slice_head(n = 15)
```

```{r priority-table, echo=FALSE}
show_tbl(
  priority_shortlist %>%
    select(GENE, MASK, THRESH, priority, META_P, META_P_ADJ, HET_FLAG),
  n = 25
)
```

**Priority (ADD, exploratory):** **High** — strong `META_P`, **`CONSISTENT_DIRECTION`**, stability guards (`reasonable_se`, `reasonable_beta`, `not_wild_diff`), **`!HET_FLAG`**, ≥1 freeze nominal at `p_support`, **not** `DOMINATED_BY_F3` (IVW not almost entirely F3 precision). **Medium** — consistent direction, asymmetric per-freeze strength. **F3-dominated** — `WEIGHT_RATIO` \> 0.8: **F3-driven IVW with limited cross-freeze precision support**; **do not** treat as fully replicated. **Review** — heterogeneous consistent effects (`HET_FLAG`) or other remainder. **Discordant** — **`OPPOSITE_DIRECTION`** (Table D / CSV; joined `META_P` is pooled only, not agreement).

---

# 10. Notes

- Per-freeze analyses are primary; this report is for **cross-freeze comparison** and triage, not replication claims.  
- **`META_P` (ADD)** — **IVW** on **all IVW-eligible overlap rows** (finite `BETA`/`SE`); **`META_P_ADJ`** is BH across that full pool. **Tier interpretation** with **`CONSISTENT_DIRECTION`**, **`HET_FLAG`**, **`OPPOSITE_DIRECTION`**, and **`DOMINATED_BY_F3`**; **opposite-sign rows** appear in **Table D** because the pooled estimate is **not** a directional-agreement summary. **`HET_FLAG`** reflects statistical inconsistency between freezes; in this dataset, heterogeneity may arise from a combination of power imbalance and residual capture differences rather than pure biological disagreement.  
- **Absence of Freeze 2 support** for rare signals **often reflects limited case carriers / power**, not definitive non-replication.  
- **`DOMINATED_BY_F3`** (`WEIGHT_RATIO` \> 0.8) indicates the IVW estimate is **effectively driven by Freeze 3** due to precision imbalance; these signals should be interpreted as **Freeze-3-driven** rather than cross-freeze replication.  
- **SKAT/SKATO:** Fisher’s method is the **single** combined summary used here. It combines *p*-values only (no direction, no pooled effect size), so interpretation should always refer back to **`PVAL_F2`** and **`PVAL_F3`**. **High** SKAT/SKATO rows require at least weak support in both freezes (**`BOTH_FREEZE_WEAK`**) to avoid treating a one-freeze-only result as a cross-freeze signal. Treat **ADD-SKATO** freeze gaps as exploratory (see **`only_F*_SKATO.csv`**).  
- **Gene-level only** — variant missingness and carrier-level checks belong in a separate variant follow-up workflow.

---

# 11. Follow-up (gene-level context)

`ALT_CT_*_EST` in **meta_ADD** are **expected** alt-allele counts (2×N×`A1FREQ`), not observed carrier counts.

For genes highlighted above (including **`r paste(sanity_genes, collapse = ", ")`**), use **Variant follow-up** for next steps.

---

# 12. Variant follow-up

Variant-level inspection (which variants in each mask, callability, missingness by case/control) is **not** done in this notebook. A companion **variant-level** report should take a target gene list, annotation, variant missingness, and sample freeze/ancestry — see `variant_level_followup.Rmd` (stub in this repo).

---

# 13. Output

```{r write-outputs, echo=FALSE}
add_full <- file.path(params$out_dir, "meta_ADD.csv")
add_prio <- file.path(params$out_dir, "meta_ADD_freeze_consistent_priority.csv")
add_short <- file.path(params$out_dir, "meta_ADD_priority_shortlist.csv")
add_cmp <- file.path(params$out_dir, "meta_ADD_comparison_shortlist.csv")
genes_followup_fp <- file.path(params$out_dir, "genes_for_followup.csv")
final_val_fp <- file.path(params$out_dir, "freeze_consistent_genes_for_followup_validation.csv")
variant_targets_fp <- file.path(params$out_dir, "variant_followup_targets.csv")
add_disc <- file.path(params$out_dir, "meta_ADD_sign_discordant.csv")
skat_out <- file.path(params$out_dir, "meta_SKAT.csv")
skato_out <- file.path(params$out_dir, "meta_SKATO.csv")
skato_short_out <- file.path(params$out_dir, "meta_SKATO_shortlist.csv")

drop_internal <- c(
  "SIGN_F2", "SIGN_F3", "strong_F2", "strong_F3",
  "support_F2", "support_F3", "weak_F2", "weak_F3", "min_support_ok"
)

table_a_export <- meta_add_df %>% arrange(META_P, PVAL_F3, PVAL_F2)
table_b_export <- meta_add_df %>%
  filter(CONSISTENT_DIRECTION, reasonable_se, reasonable_beta, not_wild_diff) %>%
  arrange(META_P)
table_short_export <- meta_add_df %>%
  filter(priority %in% c("High", "Medium")) %>%
  arrange(META_P)

readr::write_csv(table_a_export, add_full)
readr::write_csv(table_b_export %>% select(-any_of(drop_internal)), add_prio)
readr::write_csv(table_short_export %>% select(-any_of(drop_internal)), add_short)
readr::write_csv(comparison_tbl, add_cmp)

freeze_consistent_genes_for_followup_validation <- table_b_export %>%
  dplyr::filter(META_P_ADJ <= 0.05) %>%
  dplyr::arrange(META_P)
readr::write_csv(freeze_consistent_genes_for_followup_validation, final_val_fp)

p_str_fu <- as.numeric(params$p_strong)
psup_fu <- as.numeric(params$p_support)
gf_cols <- c(
  "GENE", "MASK", "THRESH", "TEST", "BETA_F2", "BETA_F3", "SE_F2", "SE_F3",
  "PVAL_F2", "PVAL_F3", "META_P", "META_P_ADJ",
  "N_CASES_F2", "N_CASES_F3", "N_CONTROLS_F2", "N_CONTROLS_F3", "followup_reason"
)
genes_f3_f2 <- meta_add_df %>%
  dplyr::filter(
    is.finite(PVAL_F3) & PVAL_F3 <= p_str_fu &
      is.finite(PVAL_F2) & PVAL_F2 > psup_fu
  ) %>%
  dplyr::mutate(followup_reason = "F3_strong_F2_not_nominal") %>%
  dplyr::select(dplyr::any_of(gf_cols))
genes_disc <- merged_add_all %>%
  dplyr::filter(
    is.finite(BETA_F2), is.finite(BETA_F3),
    sign(BETA_F2) != 0, sign(BETA_F3) != 0,
    sign(BETA_F2) != sign(BETA_F3)
  ) %>%
  dplyr::left_join(
    meta_add_df %>% dplyr::select(GENE, MASK, THRESH, META_P, META_P_ADJ),
    by = c("GENE", "MASK", "THRESH")
  ) %>%
  dplyr::mutate(followup_reason = "discordant_beta_signs") %>%
  dplyr::select(dplyr::any_of(gf_cols))
genes_for_followup <- dplyr::bind_rows(genes_f3_f2, genes_disc) %>%
  dplyr::distinct(GENE, MASK, THRESH, TEST, followup_reason, .keep_all = TRUE)
readr::write_csv(genes_for_followup, genes_followup_fp)

vf_t3 <- merged_add_all %>%
  dplyr::filter(
    is.finite(BETA_F2), is.finite(BETA_F3),
    sign(BETA_F2) != 0, sign(BETA_F3) != 0,
    sign(BETA_F2) != sign(BETA_F3)
  ) %>%
  dplyr::left_join(
    meta_add_df %>% dplyr::select(GENE, MASK, THRESH, META_P, META_P_ADJ, PVAL_F2, PVAL_F3, priority),
    by = c("GENE", "MASK", "THRESH")
  ) %>%
  dplyr::mutate(
    handoff_tier = "Tier 3 — discordant / review",
    priority = dplyr::coalesce(priority, "Discordant")
  )
vf_t2 <- meta_add_df %>%
  dplyr::filter(
    strong_F3,
    is.finite(PVAL_F2) & PVAL_F2 > psup_fu,
    PARTIAL_DIRECTION
  ) %>%
  dplyr::mutate(handoff_tier = "Tier 2 — Freeze-3-driven")
vf_t1 <- meta_add_df %>%
  dplyr::filter(priority %in% c("High", "Medium")) %>%
  dplyr::mutate(
    handoff_tier = dplyr::if_else(priority == "High", "Tier 1 — high priority", "Tier 1 — medium priority")
  )
vf_cols <- c(
  "GENE", "MASK", "THRESH", "handoff_tier", "META_P", "META_P_ADJ",
  "PVAL_F2", "PVAL_F3", "priority", "BETA_F2", "BETA_F3"
)
variant_followup_targets <- dplyr::bind_rows(
  vf_t3 %>% dplyr::select(dplyr::any_of(vf_cols)),
  vf_t2 %>% dplyr::select(dplyr::any_of(vf_cols)),
  vf_t1 %>% dplyr::select(dplyr::any_of(vf_cols))
) %>%
  dplyr::distinct(GENE, MASK, THRESH, .keep_all = TRUE)
readr::write_csv(variant_followup_targets, variant_targets_fp)

readr::write_csv(table_d, add_disc)
readr::write_csv(meta_skat, skat_out)
readr::write_csv(meta_skato, skato_out)

skato_short <- meta_skato %>%
  arrange(FISHER_P_ADJ, FISHER_P) %>%
  slice_head(n = 20)
readr::write_csv(skato_short, skato_short_out)

message("Wrote: ", add_full)
message("Wrote: ", add_prio)
message("Wrote: ", add_short)
message("Wrote: ", add_cmp)
message("Wrote: ", final_val_fp)
message("Wrote: ", variant_targets_fp)
message("Wrote: ", genes_followup_fp)
message("Wrote: ", add_disc)
message("Wrote: ", skat_out)
message("Wrote: ", skato_out)
message("Wrote: ", skato_short_out)
```

**meta_ADD.csv** — full IVW + flags on **all IVW-eligible overlap** rows (finite `BETA`/`SE`).  
**meta_ADD_comparison_shortlist.csv** — compact IVW-pool ADD rows, sorted **META_P** (exploratory comparison).  
**freeze_consistent_genes_for_followup_validation.csv** — Table **B** rows with **META_P_ADJ** ≤ 0.05 (handoff list for follow-up validation; name avoids implying a single “final” causal list).  
**variant_followup_targets.csv** — union of **High/Medium** priority, **Table C** (F3-driven), and **discordant** rows, with **`handoff_tier`** labels.  
**genes_for_followup.csv** — **Diagnostic scan list** (F3 strong / F2 not nominal, or discordant signs) — **not** a validation-ready shortlist; use **`freeze_consistent_genes_for_followup_validation.csv`** for that.  
**meta_ADD_sign_discordant.csv** — opposite-sign overlap; **`META_P`** / **`META_P_ADJ`** joined when IVW-eligible (pooled IVW, not consistency).  
**meta_ADD_freeze_consistent_priority.csv** — Table B–style subset.  
**meta_ADD_priority_shortlist.csv** — **High** + **Medium** only.  
**meta_SKAT.csv** / **meta_SKATO.csv** / **meta_SKATO_shortlist.csv** — per-freeze *p*-values plus the **single** Fisher-combined summary used in this notebook; **High** = **`FISHER_P_ADJ` ≤ 0.05** and **`BOTH_FREEZE_WEAK`**.  
**only_F2.csv**, **only_F3.csv**; **only_F2_SKATO.csv**, **only_F3_SKATO.csv** (ADD-SKATO-only unmatched).  
**gene_regenie_standardized_F2_F3_long.csv** — long snapshot. Directory: `r params$out_dir`.

**Quick use:** Compare **`meta_ADD_comparison_shortlist.csv`**; full columns in **`meta_ADD.csv`**. Discordant **`meta_ADD_sign_discordant.csv`**. Diagnostic triage **`genes_for_followup.csv`**. SKAT/SKATO: **`FISHER_P`** / **`FISHER_P_ADJ`** (single Fisher summary) with **`PVAL_F2`** and **`PVAL_F3`**. **`THRESH`** stays character in downstream code.

---

```{r session-info, echo=FALSE}
sessionInfo()
```
