---
title: "ExWAS Quality Control Report"
author: "Automated QC Pipeline"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    code_folding: hide
params:
  project: "PROJECT"
  data_dir: "data"
  covariate_file: ""   # single file WITH HEADER: FID, IID, Status, Freeze (Regnie + report + pipeline)
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.width = 10, fig.height = 6)
library(tidyverse)
library(knitr)
library(yaml)

# Read config file
config <- read_yaml("config.yml")

# Project settings (params override config, for snakemake compatibility)
PROJECT <- if (!is.null(params$project)) params$project else config$project$name
DATA_DIR <- if (!is.null(params$data_dir)) params$data_dir else "data"
COVARIATE_FILE <- if (!is.null(params$covariate_file) && params$covariate_file != "") params$covariate_file else tryCatch(config$input$covariates, error = function(e) NULL)
if (is.null(COVARIATE_FILE) || length(COVARIATE_FILE) == 0) COVARIATE_FILE <- ""

# QC thresholds - always from config
MIND_THR <- as.numeric(config$qc$mind_thr)
MAF_THR <- as.numeric(config$qc$maf_thr)
GENO_THR <- as.numeric(config$qc$geno_thr)
HWE_THR <- as.numeric(config$qc$hwe_thr)
```

# Overview

This report summarizes quality control performed on whole exome sequencing data for the **`r PROJECT`** study.

## QC Pipeline Steps

1. **Sex Check**: Remove samples with ambiguous sex (chrX F-statistic)
2. **Variant Quality Filter**: Filter variants based on high-quality heterozygotes (HQHet)
3. **Variant-Level Filters**: Apply genotype missingness, MAF, and HWE filters
4. **MNP Correction**: Identify and correct multi-nucleotide variants; counts reported in QC
5. **Sample Missingness**: Remove samples with excessive missing genotypes (calculated on filtered variants)
6. **Heterozygosity**: Remove samples with outlier heterozygosity rates (calculated on filtered variants)
7. **Population Structure**: Principal component analysis on the genotype **build** (same samples as after het QC)
8. **PCA outlier filter**: Samples with **PC4 > 0.3** on that PCA are listed for removal from a **second** PCA (`build_pca_clean.*`) used for ancestry-style covariates; **genotype callset / build.psam sample count is unchanged**

---

# Load Data

One **covariate file** (with header) supplies Status and Freeze for all QC views and for the pipeline: set `config: input: covariates: path/to/covariates.txt`.

```{r load-data}
# Load sex inference results
sex_file <- file.path(DATA_DIR, "qc/reports", "sex_inference.txt")
if (!file.exists(sex_file)) stop(paste("Required file not found:", sex_file))
sex_report <- read.table(sex_file, header = TRUE, stringsAsFactors = FALSE, comment.char = "")

# Load missingness report
miss_file <- file.path(DATA_DIR, "qc/reports", "missingness_report.txt")
if (!file.exists(miss_file)) stop(paste("Required file not found:", miss_file))
miss_report <- read.csv(miss_file, stringsAsFactors = FALSE)

# Heterozygosity: build_files writes het_report.txt (1-row summary n/mean/sd) + het_exclusions.txt (failed FID IID).
# Legacy: heterozygosity_report.txt as a per-sample table.
het_sum <- file.path(DATA_DIR, "qc/reports", "het_report.txt")
het_legacy <- file.path(DATA_DIR, "qc/reports", "heterozygosity_report.txt")
het_exc <- file.path(DATA_DIR, "qc/exclusions", "het_exclusions.txt")
if (file.exists(het_sum)) {
  ht <- utils::read.table(het_sum, header = TRUE, stringsAsFactors = FALSE, comment.char = "")
  if (nrow(ht) <= 2L && all(c("n", "mean", "sd") %in% names(ht))) {
    if (file.exists(het_exc)) {
      het_report <- utils::read.table(het_exc, header = TRUE, stringsAsFactors = FALSE, comment.char = "", check.names = FALSE)
    } else {
      het_report <- data.frame(IID = character(0), stringsAsFactors = FALSE)
    }
  } else {
    het_report <- ht
  }
} else if (file.exists(het_legacy)) {
  het_report <- read.csv(het_legacy, stringsAsFactors = FALSE)
} else {
  stop("Required: qc/reports/het_report.txt (+ qc/exclusions/het_exclusions.txt) or qc/reports/heterozygosity_report.txt")
}
if (!"IID" %in% names(het_report) && ncol(het_report) >= 2) {
  names(het_report)[2] <- "IID"
}

# Load PCA results (build_files: build_pca.eigenvec; legacy: build.eigenvec)
pca_file <- file.path(DATA_DIR, "preprocess", "build_pca.eigenvec")
if (!file.exists(pca_file)) pca_file <- file.path(DATA_DIR, "preprocess", "build.eigenvec")
if (!file.exists(pca_file)) stop(paste("Required file not found: preprocess/build_pca.eigenvec or build.eigenvec"))
pca <- read.table(pca_file, header = TRUE, stringsAsFactors = FALSE, comment.char = "")

pca_exc_path <- file.path(DATA_DIR, "qc/exclusions", "pca_outliers.txt")
pca_report_path <- file.path(DATA_DIR, "qc/reports", "pca_outlier_report.txt")
has_pca_outlier_qc <- file.exists(pca_report_path)
if (has_pca_outlier_qc) {
  pca_outlier_summary <- utils::read.table(pca_report_path, header = TRUE, stringsAsFactors = FALSE, comment.char = "")
} else {
  pca_outlier_summary <- NULL
}
pca_clean_file <- file.path(DATA_DIR, "preprocess", "build_pca_clean.eigenvec")
has_pca_clean <- file.exists(pca_clean_file)
if (has_pca_clean) {
  pca_clean <- utils::read.table(pca_clean_file, header = TRUE, stringsAsFactors = FALSE, comment.char = "")
} else {
  pca_clean <- NULL
}

# Load variant type counts (prefilter)
prefilter_counts <- lapply(1:22, function(chr) {
  f <- file.path(DATA_DIR, "qc/reports", paste0("chr", chr, ".prefilter.variant_types.txt"))
  if (file.exists(f)) {
    read.table(f, header = TRUE, stringsAsFactors = FALSE, comment.char = "") %>%
      mutate(CHR = chr)
  } else {
    warning(paste("File not found:", f))
    data.frame(TYPE = character(0), COUNT = numeric(0), CHR = integer(0))
  }
}) %>% bind_rows()

# Load variant type counts (postfilter)
postfilter_counts <- lapply(1:22, function(chr) {
  f <- file.path(DATA_DIR, "qc/reports", paste0("chr", chr, ".postfilter.variant_types.txt"))
  if (file.exists(f)) {
    read.table(f, header = TRUE, stringsAsFactors = FALSE, comment.char = "") %>%
      mutate(CHR = chr)
  } else {
    warning(paste("File not found:", f))
    data.frame(TYPE = character(0), COUNT = numeric(0), CHR = integer(0))
  }
}) %>% bind_rows()

# Load variant counts after plink filtering
postplink_counts <- lapply(1:22, function(chr) {
  f <- file.path(DATA_DIR, "qc/reports", paste0("chr", chr, ".postplink.variant_count.txt"))
  if (file.exists(f)) {
    read.table(f, header = TRUE, stringsAsFactors = FALSE, comment.char = "")
  } else {
    warning(paste("File not found:", f))
    data.frame(VARIANTS = numeric(0))
  }
}) %>% bind_rows()

# Load plink filter metrics
plink_filter_metrics <- lapply(1:22, function(chr) {
  f <- file.path(DATA_DIR, "qc/reports", paste0("chr", chr, ".plink_filter_metrics.txt"))
  if (file.exists(f)) {
    read.table(f, header = TRUE, stringsAsFactors = FALSE, comment.char = "")
  } else {
    warning(paste("File not found:", f))
    data.frame(FILTER = character(0), VARIANTS_REMOVED = numeric(0))
  }
}) %>% bind_rows()

# Load variant QC metrics for distribution plots
variant_maf <- lapply(1:22, function(chr) {
  f <- file.path(DATA_DIR, "plink", paste0("chr", chr, ".afreq"))
  if (file.exists(f)) {
    read.table(f, header = TRUE, stringsAsFactors = FALSE, comment.char = "") %>%
      mutate(CHR = chr)
  } else {
    warning(paste("File not found:", f))
    data.frame(CHR = chr)
  }
}) %>% bind_rows()

variant_miss <- lapply(1:22, function(chr) {
  f <- file.path(DATA_DIR, "plink", paste0("chr", chr, ".vmiss"))
  if (file.exists(f)) {
    read.table(f, header = TRUE, stringsAsFactors = FALSE, comment.char = "") %>%
      mutate(CHR = chr)
  } else {
    warning(paste("File not found:", f))
    data.frame(CHR = chr)
  }
}) %>% bind_rows()

variant_hwe <- lapply(1:22, function(chr) {
  f <- file.path(DATA_DIR, "plink", paste0("chr", chr, ".hardy"))
  if (file.exists(f)) {
    read.table(f, header = TRUE, stringsAsFactors = FALSE, comment.char = "") %>%
      mutate(CHR = chr)
  } else {
    warning(paste("File not found:", f))
    data.frame(CHR = chr)
  }
}) %>% bind_rows()

# Load variant type counts after MNP correction (same format as prefilter/postfilter: TYPE COUNT)
postmnp_counts <- lapply(1:22, function(chr) {
  f <- file.path(DATA_DIR, "qc/reports", paste0("chr", chr, ".postmnp.variant_types.txt"))
  if (file.exists(f)) {
    read.table(f, header = TRUE, stringsAsFactors = FALSE, comment.char = "") %>%
      mutate(CHR = chr)
  } else {
    data.frame(TYPE = character(0), COUNT = numeric(0), CHR = integer(0))
  }
}) %>% bind_rows()
has_postmnp <- nrow(postmnp_counts) > 0 && "TYPE" %in% names(postmnp_counts) && "COUNT" %in% names(postmnp_counts)

# Rename sex inference Status column to Sex first
colnames(sex_report)[colnames(sex_report) == "Status"] <- "Sex"

# Status and Freeze from single covariate file (FID, IID, FREEZE, STATUS)
id_in_report <- function(d) if ("IID" %in% names(d)) trimws(as.character(d$IID)) else trimws(as.character(d[,1]))
status_lookup <- character(0)
freeze_lookup <- character(0)
# Use length nrow(.) so 0-row tables (e.g. empty missingness_report.txt) don't trigger replacement error
sex_report$Status <- rep(NA_character_, nrow(sex_report))
sex_report$Freeze <- rep(NA_character_, nrow(sex_report))
miss_report$Status <- rep(NA_character_, nrow(miss_report))
miss_report$Freeze <- rep(NA_character_, nrow(miss_report))
het_report$Status <- rep(NA_character_, nrow(het_report))
het_report$Freeze <- rep(NA_character_, nrow(het_report))
pca$Status <- rep(NA_character_, nrow(pca))
pca$Freeze <- rep(NA_character_, nrow(pca))
has_freeze <- FALSE

if (COVARIATE_FILE != "" && file.exists(COVARIATE_FILE)) {
  cov_df <- read.table(COVARIATE_FILE, header = TRUE, stringsAsFactors = FALSE, comment.char = "", check.names = FALSE)
  # Columns: 1=FID, 2=IID, 3=FREEZE, 4=STATUS
  if (nrow(cov_df) > 0 && ncol(cov_df) >= 4) {
    iids <- trimws(as.character(cov_df[[2]]))   # IID
    raw_status <- trimws(as.character(cov_df[[4]]))   # STATUS
    status_val <- ifelse(raw_status %in% c("1", "Control", "control"), "Control",
                  ifelse(raw_status %in% c("2", "Case", "case"), "Case", NA_character_))
    status_lookup <- setNames(status_val, iids)
    sex_report$Status <- as.character(status_lookup[id_in_report(sex_report)])
    miss_report$Status <- as.character(status_lookup[id_in_report(miss_report)])
    het_report$Status <- as.character(status_lookup[id_in_report(het_report)])
    pca$Status <- as.character(status_lookup[id_in_report(pca)])
    freeze_lookup <- setNames(trimws(as.character(cov_df[[3]])), iids)   # FREEZE
    sex_report$Freeze <- as.character(freeze_lookup[id_in_report(sex_report)])
    miss_report$Freeze <- as.character(freeze_lookup[id_in_report(miss_report)])
    het_report$Freeze <- as.character(freeze_lookup[id_in_report(het_report)])
    pca$Freeze <- as.character(freeze_lookup[id_in_report(pca)])
    if (!is.null(pca_clean)) {
      pca_clean$Status <- as.character(status_lookup[id_in_report(pca_clean)])
      pca_clean$Freeze <- as.character(freeze_lookup[id_in_report(pca_clean)])
    }
    has_freeze <- TRUE
  }
}

if (!is.null(pca_clean) && !"Status" %in% names(pca_clean)) {
  pca_clean$Status <- rep(NA_character_, nrow(pca_clean))
  pca_clean$Freeze <- rep(NA_character_, nrow(pca_clean))
}

if (file.exists(pca_exc_path) && file.info(pca_exc_path)$size > 0L) {
  po <- utils::read.table(pca_exc_path, header = FALSE, stringsAsFactors = FALSE, col.names = c("FID", "IID"))
  bad_iid <- trimws(as.character(po$IID))
  pca$PCA_outlier <- trimws(as.character(pca$IID)) %in% bad_iid
} else {
  pca$PCA_outlier <- FALSE
}
```
 
---

# 1. Sex Check (chrX F-statistic)

## Method

Sex is inferred from chromosome X heterozygosity using the F-statistic:

- **F > 0.8**: Male (mostly homozygous/hemizygous on X)
- **F < 0.2**: Female (heterozygous on X)
- **0.2 ≤ F ≤ 0.8**: Ambiguous/Problem (excluded)

```{r sex-check-summary}
sex_fail <- sex_report %>% filter(Sex == "AMBIGUOUS")
sex_pass <- sex_report %>% filter(Sex != "AMBIGUOUS")

total_samples <- nrow(sex_report)
failed_samples <- nrow(sex_fail)
passing_samples <- nrow(sex_pass)
```

## Results

- **Total Samples**: `r format(total_samples, big.mark = ",")`
- **Failed Sex Check**: `r format(failed_samples, big.mark = ",")` (`r round(100 * failed_samples / total_samples, 2)`%)
- **Passing Sex Check**: `r format(passing_samples, big.mark = ",")` (`r round(100 * passing_samples / total_samples, 2)`%)

```{r sex-check-by-status}
if ("Status" %in% colnames(sex_fail) && sum(!is.na(sex_report$Status)) > 0) {
  sex_fail_summary <- sex_fail %>%
    group_by(Status) %>%
    summarise(
      N_Failed = n(),
      .groups = "drop"
    )
  
  kable(sex_fail_summary, caption = "Sex Check Failures by Case/Control Status")
}
```

## F-statistic Distribution

```{r sex-f-distribution, fig.height=5}
ggplot(sex_report, aes(x = F_sex, fill = Sex)) +
  geom_histogram(bins = 50, alpha = 0.7) +
  geom_vline(xintercept = c(0.2, 0.8), linetype = "dashed", color = "red") +
  scale_fill_manual(values = c("MALE" = "steelblue", "FEMALE" = "pink", "AMBIGUOUS" = "orange")) +
  labs(
    title = "Distribution of chrX Homozygosity (Sex Inference)",
    x = "F-statistic (O(HOM)/OBS_CT)",
    y = "Count",
    fill = "Inferred Sex"
  ) +
  theme_minimal()
```

---

# 2. Variant Quality Filtering

## Method

Variants are filtered based on the presence of high-quality heterozygotes:

- **Filter**: `COUNT(GT="0/1" && DP≥10 && GQ≥20 && VAF>0.2) = 0`
- **Action**: Tag as `NoHQHet`, then exclude non-PASS variants

Variants are also filtered based on low call rate:
- **Filter**: `INFO/F_MISSING > 0.05`
- **Action**: Tag as `LowCallRate`, then exclude non-PASS variants

```{r variant-filtering}
if (nrow(prefilter_counts) > 0 && "TYPE" %in% names(prefilter_counts) && "COUNT" %in% names(prefilter_counts)) {
  prefilter_total <- sum(prefilter_counts$COUNT[prefilter_counts$TYPE == "TOTAL"], na.rm = TRUE)
} else {
  prefilter_total <- 0
}
if (nrow(postfilter_counts) > 0 && "TYPE" %in% names(postfilter_counts) && "COUNT" %in% names(postfilter_counts)) {
  postfilter_total <- sum(postfilter_counts$COUNT[postfilter_counts$TYPE == "TOTAL"], na.rm = TRUE)
} else {
  postfilter_total <- 0
}
variants_removed <- prefilter_total - postfilter_total
```

## Results

- **Variants (Pre-filter)**: `r format(prefilter_total, big.mark = ",")`
- **Variants Removed**: `r format(variants_removed, big.mark = ",")` (`r round(100 * variants_removed / prefilter_total, 2)`%)
- **Variants (Post-filter)**: `r format(postfilter_total, big.mark = ",")` (`r round(100 * postfilter_total / prefilter_total, 2)`%)

### Variant Types Before and After Filtering

```{r variant-types-table}
variant_summary <- prefilter_counts %>%
  filter(TYPE != "TOTAL") %>%
  group_by(TYPE) %>%
  summarise(Prefilter = sum(COUNT), .groups = "drop") %>%
  left_join(
    postfilter_counts %>%
      filter(TYPE != "TOTAL") %>%
      group_by(TYPE) %>%
      summarise(Postfilter = sum(COUNT), .groups = "drop"),
    by = "TYPE"
  ) %>%
  mutate(
    Removed = Prefilter - Postfilter,
    Pct_Removed = round(100 * Removed / Prefilter, 2)
  )

kable(variant_summary, 
      format.args = list(big.mark = ","),
      caption = "Variant Counts by Type")
```

```{r variant-types-plot, fig.height=5}
variant_summary_long <- variant_summary %>%
  select(TYPE, Prefilter, Postfilter) %>%
  pivot_longer(-TYPE, names_to = "Filter_Stage", values_to = "Count")

ggplot(variant_summary_long, aes(x = TYPE, y = Count, fill = Filter_Stage)) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_fill_manual(values = c("Prefilter" = "#E74C3C", "Postfilter" = "#3498DB")) +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title = "Variant Counts Before and After Quality Filtering",
    x = "Variant Type",
    y = "Count",
    fill = "Stage"
  ) +
  theme_minimal()
```

---

# 3. Variant-Level Filters (Plink)

## Method

Additional variant filters applied using Plink2 **to SNPs only** before sample QC:

- **Genotype Missingness**: `--geno `r GENO_THR`` (`r GENO_THR * 100`%)
- **Minor Allele Frequency**: `--maf `r MAF_THR`` (`r MAF_THR * 100`%)
- **Hardy-Weinberg Equilibrium**: `--hwe `r HWE_THR` midp`
- **LD Pruning**: `--indep-pairwise 200 20 0.2` (window 200 SNPs, step 20, r² threshold 0.2)

These filters are applied to SNPs passing the HQHet filter. Sample missingness and heterozygosity are then calculated on these filtered SNPs.

```{r plink-filtering}
# Calculate totals
variants_before_plink <- ifelse(length(postfilter_total) > 0 && !is.na(postfilter_total), postfilter_total, 0)
variants_after_plink <- ifelse(nrow(postplink_counts) > 0 && "VARIANTS" %in% names(postplink_counts), sum(postplink_counts$VARIANTS, na.rm = TRUE), 0)
variants_removed_plink <- variants_before_plink - variants_after_plink

# Count SNPs vs non-SNPs from postfilter variant types
if (nrow(postfilter_counts) > 0 && "TYPE" %in% names(postfilter_counts) && "COUNT" %in% names(postfilter_counts)) {
  snp_counts <- postfilter_counts %>% filter(TYPE == "SNP")
  indel_counts <- postfilter_counts %>% filter(TYPE == "INDEL")
  other_counts <- postfilter_counts %>% filter(TYPE %in% c("MNP", "OTHER"))
  
  total_snps <- sum(snp_counts$COUNT, na.rm = TRUE)
  total_indels <- sum(indel_counts$COUNT, na.rm = TRUE)
  total_other <- sum(other_counts$COUNT, na.rm = TRUE)
} else {
  total_snps <- 0
  total_indels <- 0
  total_other <- 0
}
total_non_snps <- total_indels + total_other

# Aggregate filter metrics
if (nrow(plink_filter_metrics) > 0 && "FILTER" %in% names(plink_filter_metrics) && "VARIANTS_REMOVED" %in% names(plink_filter_metrics)) {
  filter_summary <- plink_filter_metrics %>%
    group_by(FILTER) %>%
    summarise(Variants_Removed = sum(VARIANTS_REMOVED, na.rm = TRUE), .groups = "drop") %>%
    mutate(
      Pct_of_Input = ifelse(variants_before_plink > 0, round(100 * Variants_Removed / variants_before_plink, 2), 0)
    )

  ld_prune_summary <- plink_filter_metrics %>%
    filter(FILTER == "PRUNE") %>%
    mutate(CHR = factor(CHR, levels = 1:22)) %>%
    group_by(CHR) %>%
    summarise(
      Variants_Removed = sum(VARIANTS_REMOVED, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    arrange(as.numeric(as.character(CHR)))

  total_ld_pruned <- sum(ld_prune_summary$Variants_Removed, na.rm = TRUE)
  pct_ld_pruned <- ifelse(variants_before_plink > 0, round(100 * total_ld_pruned / variants_before_plink, 2), 0)
} else {
  filter_summary <- data.frame(FILTER = character(0), Variants_Removed = numeric(0), Pct_of_Input = numeric(0))
  ld_prune_summary <- data.frame(CHR = integer(0), Variants_Removed = numeric(0))
  total_ld_pruned <- 0
  pct_ld_pruned <- 0
}
```

## Results

- **Variants (Before Plink Filters)**: `r format(variants_before_plink, big.mark = ",")` (after HQHet)
  - SNPs: `r format(total_snps, big.mark = ",")` (`r round(100 * total_snps / variants_before_plink, 1)`%)
  - Indels: `r format(total_indels, big.mark = ",")` (`r round(100 * total_indels / variants_before_plink, 1)`%)
  - Other (MNPs, etc.): `r format(total_other, big.mark = ",")` (`r round(100 * total_other / variants_before_plink, 1)`%)
- **Non-SNPs Removed** (--snps-only): `r format(total_non_snps, big.mark = ",")` (`r round(100 * total_non_snps / variants_before_plink, 2)`%)
- **SNPs Removed by QC Filters**: `r format(variants_removed_plink - total_non_snps, big.mark = ",")` (`r round(100 * (variants_removed_plink - total_non_snps) / variants_before_plink, 2)`%)
- **Variants (After All Filters)**: `r format(variants_after_plink, big.mark = ",")` (SNPs only, `r round(100 * variants_after_plink / variants_before_plink, 2)`% of input)
- **LD Pruning Removed** (`prune.out`): `r format(total_ld_pruned, big.mark = ",")` (`r pct_ld_pruned`% of input SNPs)

### LD Pruning Impact

```{r ld-pruning-plot, fig.height=5}
if (nrow(ld_prune_summary) > 0) {
  ggplot(ld_prune_summary, aes(x = factor(CHR, levels = CHR), y = Variants_Removed)) +
    geom_col(fill = "#8E44AD") +
    scale_y_continuous(labels = scales::comma) +
    labs(
      title = "Variants Removed by LD Pruning",
      x = "Chromosome",
      y = "Variants Removed (prune.out)",
      caption = paste0("Total removed: ", format(total_ld_pruned, big.mark = ","), 
                       " (", pct_ld_pruned, "% of input SNPs)")
    ) +
    theme_minimal()
}
```

### Distribution Analysis

These distributions show the spread of variant-level QC metrics and where thresholds are set.

```{r maf-below-threshold, fig.height=5}
# MAF from data/plink/chr*.afreq (PLINK --freq). Variants below threshold only (what gets filtered out).
afreq_col <- if ("ALT_FREQS" %in% names(variant_maf)) "ALT_FREQS" else "ALT_FREQ"
variant_maf$MAF <- pmin(variant_maf[[afreq_col]], 1 - variant_maf[[afreq_col]])
maf_below <- variant_maf %>% filter(MAF < MAF_THR)
if (nrow(maf_below) > 0) {
  x_max <- max(MAF_THR * 1.05, max(maf_below$MAF, na.rm = TRUE) * 1.05, 0.001)
  ggplot(maf_below, aes(x = MAF)) +
    geom_histogram(bins = 50, fill = "steelblue", alpha = 0.7) +
    geom_vline(xintercept = MAF_THR, linetype = "dashed", color = "red", linewidth = 1) +
    scale_x_continuous(labels = scales::percent, limits = c(0, x_max)) +
    scale_y_continuous(labels = scales::comma) +
    labs(
      title = "MAF distribution: variants below threshold (excluded by QC)",
      subtitle = sprintf("n = %s variants excluded by MAF < %s%%", format(nrow(maf_below), big.mark = ","), MAF_THR * 100),
      x = "MAF",
      y = "Variant Count",
      caption = "Red line = MAF threshold; these variants are removed by QC"
    ) +
    theme_minimal()
} else {
  cat("No variants below the MAF threshold.\n")
}
```

```{r geno-distribution, fig.height=5}
ggplot(variant_miss, aes(x = F_MISS)) +
  geom_histogram(bins = 50, fill = "steelblue", alpha = 0.7) +
  geom_vline(xintercept = GENO_THR, linetype = "dashed", color = "red", linewidth = 1) +
  scale_x_continuous(labels = scales::percent) +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title = "Variant Genotype Missingness Distribution",
    x = "Missingness Rate",
    y = "Variant Count",
    caption = paste0("Red line = ", GENO_THR * 100, "% threshold | ", 
                     format(sum(variant_miss$F_MISS > GENO_THR), big.mark = ","), 
                     " variants (", 
                     round(100 * sum(variant_miss$F_MISS > GENO_THR) / nrow(variant_miss), 1),
                     "%) above threshold")
  ) +
  theme_minimal()
```

```{r hwe-distribution, fig.height=5}
# Transform to -log10(P) for better visualization
variant_hwe$neg_log10_P <- -log10(variant_hwe$P)
threshold_line <- -log10(HWE_THR)

ggplot(variant_hwe, aes(x = neg_log10_P)) +
  geom_histogram(bins = 50, fill = "steelblue", alpha = 0.7) +
  geom_vline(xintercept = threshold_line, linetype = "dashed", color = "red", linewidth = 1) +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title = "Hardy-Weinberg Equilibrium Distribution",
    x = "-log10(P-value)",
    y = "Variant Count",
    caption = paste0("Red line = -log10(", HWE_THR, ") = ", round(threshold_line, 1), " | ", 
                     format(sum(variant_hwe$P < HWE_THR, na.rm = TRUE), big.mark = ","), 
                     " variants (", 
                     round(100 * sum(variant_hwe$P < HWE_THR, na.rm = TRUE) / nrow(variant_hwe), 1),
                     "%) above threshold (more deviant)")
  ) +
  theme_minimal()
```

---

# 3b. MNP Correction

Multi-nucleotide variants (MNPs) are identified from VEP annotation, and genotypes are corrected. The counts below are from the **MNP-corrected** BCF (same `count_variant_types` logic as prefilter/postfilter), so you see SNP, INDEL, MNP, and OTHER in the annotation dataset after correction.

```{r postmnp-summary}
if (has_postmnp) {
  postmnp_total <- sum(postmnp_counts$COUNT[postmnp_counts$TYPE == "TOTAL"], na.rm = TRUE)
  postmnp_by_type <- postmnp_counts %>%
    filter(TYPE != "TOTAL") %>%
    group_by(TYPE) %>%
    summarise(COUNT = sum(COUNT, na.rm = TRUE), .groups = "drop")
  cat("- **Total variants (after MNP correction)**: ", format(postmnp_total, big.mark = ","), "\n", sep = "")
} else {
  postmnp_total <- 0
  postmnp_by_type <- data.frame(TYPE = character(0), COUNT = numeric(0))
  cat("Post-MNP variant type counts not found (pipeline may not have run the MNP correction step).\n")
}
```

```{r postmnp-table}
if (has_postmnp && nrow(postmnp_by_type) > 0) {
  kable(postmnp_by_type,
        format.args = list(big.mark = ","),
        caption = "Variant Counts by Type (After MNP Correction)",
        col.names = c("Type", "Count"))
}
```

```{r postmnp-plot, fig.height=5}
if (has_postmnp && nrow(postmnp_by_type) > 0) {
  ggplot(postmnp_by_type, aes(x = TYPE, y = COUNT, fill = TYPE)) +
    geom_col() +
    scale_fill_manual(values = c(SNP = "#3498DB", INDEL = "#E74C3C", MNP = "#27AE60", OTHER = "#9B59B6"), guide = "none") +
    scale_y_continuous(labels = scales::comma) +
    labs(
      title = "Variant Types After MNP Correction",
      x = "Variant Type",
      y = "Count",
      caption = paste0("Total: ", format(postmnp_total, big.mark = ","), " variants")
    ) +
    theme_minimal()
}
```

---

# 4. Sample Missingness

## Method

Sample missingness is calculated across all variants passing quality filters (HQHet + geno/maf/hwe). Samples with excessive missing genotypes are excluded.

- **Threshold**: F_MISS > `r MIND_THR` (`r MIND_THR * 100`%)

```{r missingness-summary}
total_after_sex <- passing_samples
failed_missingness <- nrow(miss_report)
passing_missingness <- total_after_sex - failed_missingness
```

## Results

- **Samples (Post Sex-Check)**: `r format(total_after_sex, big.mark = ",")`
- **Failed Missingness**: `r format(failed_missingness, big.mark = ",")` (`r round(100 * failed_missingness / total_after_sex, 2)`%)
- **Passing Missingness**: `r format(passing_missingness, big.mark = ",")` (`r round(100 * passing_missingness / total_after_sex, 2)`%)

```{r missingness-by-status}
if ("Status" %in% colnames(miss_report) && sum(!is.na(miss_report$Status)) > 0) {
  miss_fail_summary <- miss_report %>%
    group_by(Status) %>%
    summarise(
      N_Failed = n(),
      .groups = "drop"
    )
  
  kable(miss_fail_summary, caption = "Missingness Failures by Case/Control Status")
}
```

## Missingness Distribution

**WARNING**: If a large proportion of cases are failing, check the distribution below. The threshold may be too stringent or there may be a data quality issue.

```{r missingness-distribution, fig.height=5}
# Load aggregated missingness to see all samples
agg_miss <- read.table(
  file.path(DATA_DIR, "plink", "aggregated_missingness.txt"),
  header = TRUE, stringsAsFactors = FALSE, comment.char = ""
)

# Status and Freeze from covariate file (same as reports)
agg_id <- if ("IID" %in% names(agg_miss)) trimws(agg_miss$IID) else trimws(agg_miss[,1])
agg_miss$Status <- if (length(status_lookup) > 0) as.character(status_lookup[agg_id]) else "Unknown"
agg_miss$Freeze <- if (length(freeze_lookup) > 0) as.character(freeze_lookup[agg_id]) else NA_character_

ggplot(agg_miss, aes(x = F_MISS, fill = Status)) +
  geom_histogram(bins = 50, alpha = 0.7) +
  geom_vline(xintercept = MIND_THR, linetype = "dashed", color = "red", linewidth = 1) +
  scale_x_continuous(labels = scales::percent) +
  labs(
    title = "Distribution of Sample Missingness",
    x = "Missingness Rate",
    y = "Count",
    fill = "Status",
    caption = paste0("Red line = ", MIND_THR * 100, "% threshold")
  ) +
  theme_minimal()
```

```{r missingness-stats}
# Summary statistics
miss_stats <- agg_miss %>%
  group_by(Status) %>%
  summarise(
    N = n(),
    Mean = mean(F_MISS),
    Median = median(F_MISS),
    Min = min(F_MISS),
    Max = max(F_MISS),
    Above_Threshold = sum(F_MISS > MIND_THR),
    Pct_Above = round(100 * Above_Threshold / N, 2),
    .groups = "drop"
  )

kable(miss_stats, 
      digits = 4,
      caption = "Missingness Statistics by Case/Control Status")
```

```{r missingness-stats-freeze}
if (exists("has_freeze") && has_freeze && "Freeze" %in% names(agg_miss) && sum(!is.na(agg_miss$Freeze)) > 0) {
  miss_stats_freeze <- agg_miss %>%
    filter(!is.na(Freeze)) %>%
    group_by(Freeze) %>%
    summarise(
      N = n(),
      Mean = mean(F_MISS),
      Median = median(F_MISS),
      Above_Threshold = sum(F_MISS > MIND_THR),
      Pct_Above = round(100 * Above_Threshold / N, 2),
      .groups = "drop"
    )
  kable(miss_stats_freeze, digits = 4, caption = "Missingness Statistics by Freeze")
}
```

### Failing Samples Details

```{r failing-samples-distribution}
# Show distribution of failing samples (sample ID column is IID in aggregated_missingness.txt)
id_col_miss <- if ("IID" %in% names(agg_miss)) "IID" else names(agg_miss)[1]
failing_samples <- agg_miss %>%
  filter(F_MISS > MIND_THR) %>%
  arrange(desc(F_MISS)) %>%
  mutate(
    Sample_ID = .data[[id_col_miss]],
    Missingness_Pct = round(F_MISS * 100, 2)
  )

if (nrow(failing_samples) > 0) {
  cat(paste0("\n**Total Failing Samples**: ", nrow(failing_samples), "\n\n"))
  
  # Distribution plot of failing samples
  ggplot(failing_samples, aes(x = Missingness_Pct, fill = Status)) +
    geom_histogram(bins = 30, alpha = 0.7) +
    labs(
      title = paste0("Distribution of Missingness Rates for Failing Samples (n=", nrow(failing_samples), ")"),
      x = "Missingness Rate (%)",
      y = "Count",
      fill = "Status"
    ) +
    theme_minimal()
} else {
  cat("No samples exceeded the missingness threshold.\n")
}
```

```{r worst-offenders-table}
# Show worst 20 offenders
if (nrow(failing_samples) > 0) {
  worst_offenders <- failing_samples %>%
    select(Sample_ID, Status, Missingness_Pct, MISSING_CT, OBS_CT) %>%
    head(20)
  
  kable(worst_offenders, 
        format.args = list(big.mark = ","),
        caption = paste0("Top ", min(nrow(failing_samples), 20), " Worst Offenders"))
}
```

```{r case-missingness-summary}
# Specifically summarize case missingness for sharing
case_miss <- agg_miss %>%
  filter(Status == "Case") %>%
  arrange(desc(F_MISS))

if (nrow(case_miss) > 0) {
  cat("\n### Case Missingness Summary\n\n")
  cat("- **Total Cases**: ", nrow(case_miss), "\n")
  cat("- **Cases Failing (>", MIND_THR * 100, "%)**: ", sum(case_miss$F_MISS > MIND_THR), " (", 
      round(100 * sum(case_miss$F_MISS > MIND_THR) / nrow(case_miss), 2), "%)\n", sep="")
  cat("- **Mean Missingness**: ", round(mean(case_miss$F_MISS) * 100, 2), "%\n", sep="")
  cat("- **Median Missingness**: ", round(median(case_miss$F_MISS) * 100, 2), "%\n", sep="")
  cat("- **Range**: ", round(min(case_miss$F_MISS) * 100, 2), "% - ", 
      round(max(case_miss$F_MISS) * 100, 2), "%\n\n", sep="")
}
```

---

# 5. Heterozygosity

## Method

Heterozygosity rate is calculated for each sample on variants passing all quality filters. Outliers (mean ± 3 SD) are excluded as potential contamination or technical artifacts.

```{r het-summary}
total_after_miss <- passing_missingness
failed_het <- nrow(het_report)
passing_het <- total_after_miss - failed_het
```

## Results

- **Samples (Post Missingness)**: `r format(total_after_miss, big.mark = ",")`
- **Failed Heterozygosity**: `r format(failed_het, big.mark = ",")` (`r round(100 * failed_het / total_after_miss, 2)`%)
- **Passing Heterozygosity**: `r format(passing_het, big.mark = ",")` (`r round(100 * passing_het / total_after_miss, 2)`%)

```{r het-by-status}
if ("Status" %in% colnames(het_report) && sum(!is.na(het_report$Status)) > 0) {
  het_fail_summary <- het_report %>%
    group_by(Status) %>%
    summarise(
      N_Failed = n(),
      .groups = "drop"
    )
  
  kable(het_fail_summary, caption = "Heterozygosity Failures by Case/Control Status")
}
```

```{r het-by-freeze}
if (exists("has_freeze") && has_freeze && "Freeze" %in% colnames(het_report) && sum(!is.na(het_report$Freeze)) > 0) {
  het_fail_freeze <- het_report %>%
    filter(!is.na(Freeze)) %>%
    group_by(Freeze) %>%
    summarise(N_Failed = n(), .groups = "drop")
  kable(het_fail_freeze, caption = "Heterozygosity Failures by Freeze")
}
```

---

# 6. Differential Missingness (Case vs Control)

**NOTE**: This is for informational purposes only and does not affect QC filtering.

This tests whether variants have significantly different missingness rates between cases and controls, which could indicate genotyping bias or biological effects.

```{r load-diff-miss}
# Load differential missingness test results
diff_miss_files <- file.path(DATA_DIR, "plink", paste0("chr", 1:22, ".test.missing"))
diff_miss <- lapply(diff_miss_files, function(f) {
  if (file.exists(f)) {
    read.table(f, header = TRUE, stringsAsFactors = FALSE, comment.char = "")
  } else {
    NULL
  }
}) %>% bind_rows()
```

```{r diff-miss-summary}
if (!is.null(diff_miss) && nrow(diff_miss) > 0) {
  # Count significant variants at different thresholds
  sig_001 <- sum(diff_miss$P < 0.001, na.rm = TRUE)
  sig_01 <- sum(diff_miss$P < 0.01, na.rm = TRUE)
  sig_05 <- sum(diff_miss$P < 0.05, na.rm = TRUE)
  
  cat("- **Total variants tested**: ", format(nrow(diff_miss), big.mark = ","), "\n")
  cat("- **P < 0.05**: ", format(sig_05, big.mark = ","), " (", 
      round(100 * sig_05 / nrow(diff_miss), 2), "%)\n", sep="")
  cat("- **P < 0.01**: ", format(sig_01, big.mark = ","), " (", 
      round(100 * sig_01 / nrow(diff_miss), 2), "%)\n", sep="")
  cat("- **P < 0.001**: ", format(sig_001, big.mark = ","), " (", 
      round(100 * sig_001 / nrow(diff_miss), 2), "%)\n", sep="")
}
```

### Before excluding P \< 1e-5 (all variants from test-missing)

All plots in this section use the **original** case/control `--test-missing` output (no re-run). Below: density and Q-Q on the full set.

```{r diff-miss-density-before, fig.height=6}
if (!is.null(diff_miss) && nrow(diff_miss) > 0) {
  ggplot(diff_miss, aes(x = F_MISS_U, y = F_MISS_A)) +
    geom_hex(bins = 30, color = "gray65", linewidth = 0.25) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red", linewidth = 1) +
    scale_x_continuous(labels = scales::percent) +
    scale_y_continuous(labels = scales::percent) +
    scale_fill_gradient(low = "gray92", high = "darkblue", name = "Variants") +
    labs(
      title = "Variant Missingness: Cases vs Controls (before filter)",
      subtitle = sprintf("n = %s variants; fill = count per hex", format(nrow(diff_miss), big.mark = ",")),
      x = "Missingness in Controls (F_MISS_U)",
      y = "Missingness in Cases (F_MISS_A)",
      caption = "Red line = equal missingness"
    ) +
    theme_minimal()
}
```

```{r diff-miss-qq-before, fig.height=5}
if (!is.null(diff_miss) && nrow(diff_miss) > 0) {
  p_vals <- diff_miss$P[!is.na(diff_miss$P) & diff_miss$P > 0]
  n <- length(p_vals)
  if (n > 0) {
    observed <- -log10(sort(p_vals))
    expected <- -log10(ppoints(n))
    qq_df <- data.frame(expected = expected, observed = observed)
    lambda_before <- median(qchisq(1 - p_vals, df = 1), na.rm = TRUE) / qchisq(0.5, 1)
    ggplot(qq_df, aes(x = expected, y = observed)) +
      geom_point(alpha = 0.3, size = 0.8) +
      geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red", linewidth = 1) +
      labs(
        title = "Q-Q plot: Differential missingness P-values (before filter)",
        subtitle = sprintf("λ = %.4f", lambda_before),
        x = "Expected -log10(P)",
        y = "Observed -log10(P)",
        caption = "Under null, points follow the red line"
      ) +
      theme_minimal()
  }
}
```

### After excluding variants with P \< 1e-5

Same test-missing output, but density and Q-Q use only variants with P ≥ 1e-5 (the set that remains after removing differential-missingness failures).

```{r diff-miss-after-filter}
DIFF_MISS_P_THR <- 1e-5
diff_miss_cc_plot <- if (!is.null(diff_miss) && nrow(diff_miss) > 0) {
  d <- diff_miss %>% filter(!is.na(P) & P >= DIFF_MISS_P_THR)
  if (nrow(d) > 0) d else NULL
} else NULL
n_excluded_cc <- if (!is.null(diff_miss) && nrow(diff_miss) > 0) sum(diff_miss$P < DIFF_MISS_P_THR, na.rm = TRUE) else 0L
```

```{r diff-miss-density-plot, fig.height=6}
if (!is.null(diff_miss_cc_plot) && nrow(diff_miss_cc_plot) > 0) {
  ggplot(diff_miss_cc_plot, aes(x = F_MISS_U, y = F_MISS_A)) +
    geom_hex(bins = 30, color = "gray65", linewidth = 0.25) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red", linewidth = 1) +
    scale_x_continuous(labels = scales::percent) +
    scale_y_continuous(labels = scales::percent) +
    scale_fill_gradient(low = "gray92", high = "darkblue", name = "Variants") +
    labs(
      title = "Variant Missingness: Cases vs Controls (after filter)",
      subtitle = sprintf("n = %s variants (excluded P < 1e-5: %s); fill = count per hex", format(nrow(diff_miss_cc_plot), big.mark = ","), format(n_excluded_cc, big.mark = ",")),
      x = "Missingness in Controls (F_MISS_U)",
      y = "Missingness in Cases (F_MISS_A)",
      caption = "Red line = equal missingness"
    ) +
    theme_minimal()
}
```

```{r diff-miss-qq-plot, fig.height=5}
if (!is.null(diff_miss_cc_plot) && nrow(diff_miss_cc_plot) > 0) {
  p_vals <- diff_miss_cc_plot$P[!is.na(diff_miss_cc_plot$P) & diff_miss_cc_plot$P > 0]
  n <- length(p_vals)
  if (n > 0) {
    observed <- -log10(sort(p_vals))
    expected <- -log10(ppoints(n))
    qq_df <- data.frame(expected = expected, observed = observed)
    lambda_cc <- median(qchisq(1 - p_vals, df = 1), na.rm = TRUE) / qchisq(0.5, 1)
    ggplot(qq_df, aes(x = expected, y = observed)) +
      geom_point(alpha = 0.3, size = 0.8) +
      geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red", linewidth = 1) +
      labs(
        title = "Q-Q plot: Differential missingness P-values (after filter)",
        subtitle = sprintf("P ≥ 1e-5 only (λ = %.4f)", lambda_cc),
        x = "Expected -log10(P)",
        y = "Observed -log10(P)",
        caption = "Under null, points follow the red line"
      ) +
      theme_minimal()
  }
}
```

**Interpretation**: Variants should fall along the diagonal (equal missingness). Points far from the line indicate differential missingness between cases and controls, which may suggest genotyping bias. Before/after comparison shows the effect of removing variants with P \< 1e-5.

---

## Differential Missingness: Freeze 2 vs Freeze 3 (covariate only)

Freeze (2 vs 3) is a **covariate for testing and exploration only**: we test and plot differential missingness by freeze. Nothing is removed or filtered based on this; it is for interpretation only (e.g. many cases from freeze 3).

```{r load-diff-miss-freeze}
freeze_diff_miss_files <- file.path(DATA_DIR, "plink", paste0("chr", 1:22, ".test_freeze.missing"))
diff_miss_freeze <- lapply(freeze_diff_miss_files, function(f) {
  if (file.exists(f)) read.table(f, header = TRUE, stringsAsFactors = FALSE, comment.char = "") else NULL
}) %>% bind_rows()
has_freeze_diff_miss <- is.data.frame(diff_miss_freeze) && nrow(diff_miss_freeze) > 0
```

```{r diff-miss-freeze-summary}
if (has_freeze_diff_miss) {
  sig_001_f <- sum(diff_miss_freeze$P < 0.001, na.rm = TRUE)
  sig_01_f <- sum(diff_miss_freeze$P < 0.01, na.rm = TRUE)
  sig_05_f <- sum(diff_miss_freeze$P < 0.05, na.rm = TRUE)
  cat("- **Total variants tested (freeze)**: ", format(nrow(diff_miss_freeze), big.mark = ","), "\n")
  cat("- **P < 0.05**: ", format(sig_05_f, big.mark = ","), " (", round(100 * sig_05_f / nrow(diff_miss_freeze), 2), "%)\n", sep = "")
  cat("- **P < 0.01**: ", format(sig_01_f, big.mark = ","), " (", round(100 * sig_01_f / nrow(diff_miss_freeze), 2), "%)\n", sep = "")
  cat("- **P < 0.001**: ", format(sig_001_f, big.mark = ","), " (", round(100 * sig_001_f / nrow(diff_miss_freeze), 2), "%)\n", sep = "")
} else {
  cat("Freeze differential missingness was not run (set *config input.freeze_file* and run the pipeline to generate `chr*.test_freeze.missing`).\n")
}
```

```{r diff-miss-freeze-density-plot, fig.height=6}
if (has_freeze_diff_miss) {
  ggplot(diff_miss_freeze, aes(x = F_MISS_U, y = F_MISS_A)) +
    geom_hex(bins = 30, color = "gray65", linewidth = 0.25) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red", linewidth = 1) +
    scale_x_continuous(labels = scales::percent) +
    scale_y_continuous(labels = scales::percent) +
    scale_fill_gradient(low = "gray92", high = "darkblue", name = "Variants") +
    labs(
      title = "Variant Missingness: Freeze 2 vs Freeze 3",
      subtitle = sprintf("n = %s variants; fill = count per hex", format(nrow(diff_miss_freeze), big.mark = ",")),
      x = "Missingness in Freeze 2 (F_MISS_U)",
      y = "Missingness in Freeze 3 (F_MISS_A)",
      caption = "Red line = equal missingness"
    ) +
    theme_minimal()
}
```

```{r diff-miss-freeze-qq-plot, fig.height=5}
if (has_freeze_diff_miss) {
  p_vals_f <- diff_miss_freeze$P[!is.na(diff_miss_freeze$P) & diff_miss_freeze$P > 0]
  n_f <- length(p_vals_f)
  if (n_f > 0) {
    observed_f <- -log10(sort(p_vals_f))
    expected_f <- -log10(ppoints(n_f))
    qq_df_f <- data.frame(expected = expected_f, observed = observed_f)
    lambda_freeze <- median(qchisq(1 - p_vals_f, df = 1), na.rm = TRUE) / qchisq(0.5, 1)
    ggplot(qq_df_f, aes(x = expected, y = observed)) +
      geom_point(alpha = 0.3, size = 0.8) +
      geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red", linewidth = 1) +
      labs(
        title = "Q-Q plot: Differential missingness P-values (Freeze 2 vs Freeze 3)",
        subtitle = sprintf("λ = %.4f", lambda_freeze),
        x = "Expected -log10(P)",
        y = "Observed -log10(P)",
        caption = "Under null, points follow the red line; deviation at low P suggests enrichment of differential missingness"
      ) +
      theme_minimal()
  }
}
```

### Comparison: Case/Control vs Freeze

If both case/control and freeze differential missingness were run, compare how the two relate: a large proportion of cases from freeze 3 can make case/control missingness patterns overlap with freeze 2 vs 3. Variants that are differential in both may reflect freeze/batch rather than disease.

```{r diff-miss-comparison}
if (has_freeze_diff_miss && !is.null(diff_miss) && nrow(diff_miss) > 0) {
  # Merge by variant ID (CHR + SNP or ID column)
  id_col <- if ("SNP" %in% names(diff_miss)) "SNP" else if ("ID" %in% names(diff_miss)) "ID" else NULL
  if (!is.null(id_col) && id_col %in% names(diff_miss_freeze)) {
    joint <- diff_miss %>%
      select(any_of(c("CHR", "SNP", "ID", "P"))) %>%
      rename(P_case_control = P) %>%
      inner_join(
        diff_miss_freeze %>% select(any_of(c("CHR", "SNP", "ID", "P"))) %>% rename(P_freeze = P),
        by = c("CHR", id_col)
      )
    both_sig_05 <- sum(joint$P_case_control < 0.05 & joint$P_freeze < 0.05, na.rm = TRUE)
    cat("- **Variants with data in both tests**: ", format(nrow(joint), big.mark = ","), "\n")
    cat("- **Significant in both (P < 0.05)**: ", format(both_sig_05, big.mark = ","), 
        " (", round(100 * both_sig_05 / nrow(joint), 2), "% of joint set)\n", sep = "")
    cat("\nIf many variants are differential in both case/control and freeze, case/control missingness may be partly driven by freeze/batch (e.g. many cases from freeze 3).\n")
  } else {
    cat("Variant IDs could not be matched between case/control and freeze results.\n")
  }
} else if (!has_freeze_diff_miss) {
  cat("Run freeze differential missingness to see comparison with case/control.\n")
}
```

---

# 7. Final Sample Count

```{r final-summary}
final_samples <- passing_het
final_variants <- variants_after_plink

pca_fail_cases <- 0L
pca_fail_controls <- 0L
samples_after_pca_ancestry <- NA_integer_
if (exists("has_pca_outlier_qc") && has_pca_outlier_qc && !is.null(pca_outlier_summary)) {
  samples_after_pca_ancestry <- as.integer(pca_outlier_summary$n_after_pca[1])
}
if (exists("has_pca_outlier_qc") && has_pca_outlier_qc && exists("pca_exc_path") &&
    file.exists(pca_exc_path) && file.info(pca_exc_path)$size > 0L && exists("status_lookup") && length(status_lookup) > 0L) {
  po <- utils::read.table(pca_exc_path, header = FALSE, stringsAsFactors = FALSE, col.names = c("FID", "IID"))
  iid_out <- trimws(as.character(po$IID))
  st <- unname(status_lookup[iid_out])
  pca_fail_cases <- sum(st == "Case", na.rm = TRUE)
  pca_fail_controls <- sum(st == "Control", na.rm = TRUE)
}

# Count cases/controls at each stage (from Status in covariate file)
if ("Status" %in% colnames(sex_report) && sum(!is.na(sex_report$Status)) > 0) {
  start_cases <- sum(sex_report$Status == "Case", na.rm = TRUE)
  start_controls <- sum(sex_report$Status == "Control", na.rm = TRUE)
  postsex_cases <- sum(sex_pass$Status == "Case", na.rm = TRUE)
  postsex_controls <- sum(sex_pass$Status == "Control", na.rm = TRUE)
  miss_fail_cases <- sum(miss_report$Status == "Case", na.rm = TRUE)
  miss_fail_controls <- sum(miss_report$Status == "Control", na.rm = TRUE)
  postmiss_cases <- postsex_cases - miss_fail_cases
  postmiss_controls <- postsex_controls - miss_fail_controls
  het_fail_cases <- sum(het_report$Status == "Case", na.rm = TRUE)
  het_fail_controls <- sum(het_report$Status == "Control", na.rm = TRUE)
  final_cases <- postmiss_cases - het_fail_cases
  final_controls <- postmiss_controls - het_fail_controls
}
```

## Summary

- **Starting Samples**: `r format(total_samples, big.mark = ",")` (`r if(exists("start_cases")) paste0(format(start_cases, big.mark=","), " cases, ", format(start_controls, big.mark=","), " controls")`)
- **Final Samples**: `r format(final_samples, big.mark = ",")` (`r if(exists("final_cases")) paste0(format(final_cases, big.mark=","), " cases, ", format(final_controls, big.mark=","), " controls")`) (genotype **build** after het QC; unchanged by PCA outlier list)
- **Total Excluded**: `r format(total_samples - final_samples, big.mark = ",")` (`r if(exists("start_cases")) paste0(format(start_cases - final_cases, big.mark=","), " cases, ", format(start_controls - final_controls, big.mark=","), " controls")`)
`r if (exists("has_pca_outlier_qc") && has_pca_outlier_qc && !is.na(samples_after_pca_ancestry)) sprintf("- **Samples in PC-adjusted PCA** (build_pca_clean): %s (after PC4>0.3 exclusion; ancestry PCs only)", format(samples_after_pca_ancestry, big.mark = ",")) else ""`

### Samples Excluded by Stage

```{r exclusion-by-stage}
if (exists("start_cases")) {
  stg <- c("Sex Check", "Missingness", "Heterozygosity")
  cex <- c(start_cases - postsex_cases, miss_fail_cases, het_fail_cases)
  ctl <- c(start_controls - postsex_controls, miss_fail_controls, het_fail_controls)
  if (exists("has_pca_outlier_qc") && has_pca_outlier_qc) {
    stg <- c(stg, "PCA (structure)")
    cex <- c(cex, pca_fail_cases)
    ctl <- c(ctl, pca_fail_controls)
  }
  exclusion_breakdown <- data.frame(Stage = stg, Cases_Excluded = cex, Controls_Excluded = ctl) %>%
    mutate(
      Total_Excluded = Cases_Excluded + Controls_Excluded,
      Pct_Cases = ifelse(Total_Excluded > 0, round(100 * Cases_Excluded / Total_Excluded, 1), NA_real_)
    )

  kable(exclusion_breakdown,
        caption = "Sample Exclusions by QC Stage and Case/Control Status",
        col.names = c("QC Stage", "Cases Excluded", "Controls Excluded", "Total Excluded", "% Cases"))
}
```

- **Starting Variants**: `r format(prefilter_total, big.mark = ",")`
- **Final Variants**: `r format(final_variants, big.mark = ",")` (`r round(100 * final_variants / prefilter_total, 2)`% retention)
- **Total Excluded**: `r format(prefilter_total - final_variants, big.mark = ",")`

```{r exclusion-cascade}
if (exists("has_pca_outlier_qc") && has_pca_outlier_qc && !is.na(samples_after_pca_ancestry)) {
  exclusion_data <- data.frame(
    Stage = c("Starting", "Sex Check", "Missingness", "Heterozygosity", "Genotype build", "PC-adjusted PCA"),
    Samples = c(
      total_samples, passing_samples, passing_missingness, passing_het,
      passing_het, samples_after_pca_ancestry
    )
  )
} else {
  exclusion_data <- data.frame(
    Stage = c("Starting", "Sex Check", "Missingness", "Heterozygosity", "Final"),
    Samples = c(total_samples, passing_samples, passing_missingness, passing_het, final_samples)
  )
}

ggplot(exclusion_data, aes(x = factor(Stage, levels = Stage), y = Samples)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  geom_text(aes(label = format(Samples, big.mark = ",")), vjust = -0.5) +
  labs(
    title = "Sample Retention Through QC Pipeline",
    subtitle = if (exists("has_pca_outlier_qc") && has_pca_outlier_qc) "Last bar: samples remaining after PC4>0.3 (ancestry PCA only; genotype build count = previous bar)" else NULL,
    x = "QC Stage",
    y = "Number of Samples"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

---

# 8. Population Structure (PCA)

PCA is computed on autosomes **excluding chromosome 8** (chr 1–7, 9–22) to avoid known structural effects. The first PCA (`build_pca`) uses all samples in the genotype **build** (after het QC). When `identify_pca_outliers.R` runs, samples with **PC4 > 0.3** are written to `qc/exclusions/pca_outliers.txt` and PLINK recomputes PCA without them (`build_pca_clean.*`) for ancestry-style PCs. **Genotype data and `passing_samples` are not subset by that list** unless you change the pipeline.

## PCA outlier filter (report)

### How outliers are defined

- **Rule:** samples with **raw PC4 > 0.3** in **`build_pca.eigenvec`** (PLINK2 `--pca` output; **not** a z-score on PC4).
- **Implementation:** `identify_pca_outliers.R` (constants `PC_COL` / `PC_THRESHOLD` in that script; default **PC4** and **0.3**).
- **Effect:** FID/IID rows are written to **`qc/exclusions/pca_outliers.txt`** for `plink2 --remove` when computing **`build_pca_clean`** only.

```{r pca-outlier-summary-table, echo=FALSE}
if (exists("has_pca_outlier_qc") && has_pca_outlier_qc) {
  knitr::kable(pca_outlier_summary, caption = "PCA outlier step (qc/reports/pca_outlier_report.txt)")
} else {
  cat("No `qc/reports/pca_outlier_report.txt` found. Run Snakemake `identify_pca_outliers` (build_files) to generate it.\n")
}
```

```{r pca-pipeline-static-plots, echo=FALSE, out.width="100%", fig.show="hold"}
p_pca1 <- file.path(DATA_DIR, "qc/reports", "pca_outliers_pc1_pc2.png")
p_pca4 <- file.path(DATA_DIR, "qc/reports", "pca_outliers_pc1_pc2_PC4.png")
if (file.exists(p_pca1)) {
  knitr::include_graphics(p_pca1)
}
if (file.exists(p_pca4)) {
  knitr::include_graphics(p_pca4)
}
```

The **build PCA** and **build_pca_clean** subsections use the **same figure types** (variance explained for PCs 1–10, PC1 vs PC2, PC1 vs PC2 by freeze when available, PC3 vs PC4, PC3 vs PC4 by freeze, scree for all 20 PCs). **Coordinates are not comparable** across the two PCAs because eigenvectors are recomputed after sample removal.

## Principal Components (build PCA)

```{r pca-summary}
# Calculate variance explained
eigenval_file <- file.path(DATA_DIR, "preprocess", "build_pca.eigenval")
if (!file.exists(eigenval_file)) eigenval_file <- file.path(DATA_DIR, "preprocess", "build.eigenval")
if (!file.exists(eigenval_file)) stop(paste("Required file not found: preprocess/build_pca.eigenval or build.eigenval"))
eigenval <- read.table(eigenval_file, header = FALSE)
colnames(eigenval) <- "Eigenvalue"
eigenval$PC <- paste0("PC", seq_len(nrow(eigenval)))
eigenval$Variance_Explained <- 100 * eigenval$Eigenvalue / sum(eigenval$Eigenvalue)
```

### Variance Explained

```{r variance-explained, fig.height=5}
ggplot(eigenval[1:10, ], aes(x = factor(PC, levels = PC), y = Variance_Explained)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  geom_text(aes(label = sprintf("%.2f%%", Variance_Explained)), vjust = -0.5) +
  labs(
    title = "Variance Explained by Principal Components",
    x = "Principal Component",
    y = "Variance Explained (%)"
  ) +
  theme_minimal()
```

### PC1 vs PC2

```{r pca-plot-pc1-pc2, fig.height=6}
ggplot(pca, aes(x = PC1, y = PC2, color = Status)) +
  geom_point(alpha = 0.6) +
  labs(
    title = "Principal Component Analysis (PC1 vs PC2)",
    x = paste0("PC1 (", round(eigenval$Variance_Explained[1], 2), "%)"),
    y = paste0("PC2 (", round(eigenval$Variance_Explained[2], 2), "%)"),
    color = "Status"
  ) +
  theme_minimal()
```

```{r pca-plot-pc1-pc2-freeze, fig.height=6}
if (exists("has_freeze") && has_freeze && "Freeze" %in% names(pca) && sum(!is.na(pca$Freeze)) > 0) {
  ggplot(pca %>% filter(!is.na(Freeze)), aes(x = PC1, y = PC2, color = factor(Freeze))) +
    geom_point(alpha = 0.6) +
    labs(
      title = "PC1 vs PC2 by Freeze",
      x = paste0("PC1 (", round(eigenval$Variance_Explained[1], 2), "%)"),
      y = paste0("PC2 (", round(eigenval$Variance_Explained[2], 2), "%)"),
      color = "Freeze"
    ) +
    theme_minimal()
}
```

### PC3 vs PC4

```{r pca-plot-pc3-pc4, fig.height=6}
pca$pca_hit <- ifelse(pca$PCA_outlier, "PC4 > 0.3", "Kept")
ggplot(pca, aes(x = PC3, y = PC4, color = Status, shape = pca_hit)) +
  geom_point(alpha = 0.65) +
  geom_hline(yintercept = 0.3, linetype = "dashed", color = "gray40") +
  labs(
    title = "PC3 vs PC4 (dashed line: PC4 = 0.3 outlier threshold)",
    x = paste0("PC3 (", round(eigenval$Variance_Explained[3], 2), "%)"),
    y = paste0("PC4 (", round(eigenval$Variance_Explained[4], 2), "%)"),
    color = "Status",
    shape = "PCA filter"
  ) +
  theme_minimal()
```

```{r pca-plot-pc3-pc4-freeze, fig.height=6}
if (exists("has_freeze") && has_freeze && "Freeze" %in% names(pca) && sum(!is.na(pca$Freeze)) > 0) {
  ggplot(pca %>% filter(!is.na(Freeze)), aes(x = PC3, y = PC4, color = factor(Freeze))) +
    geom_point(alpha = 0.6) +
    labs(
      title = "PC3 vs PC4 by Freeze",
      x = paste0("PC3 (", round(eigenval$Variance_Explained[3], 2), "%)"),
      y = paste0("PC4 (", round(eigenval$Variance_Explained[4], 2), "%)"),
      color = "Freeze"
    ) +
    theme_minimal()
}
```

### Scree Plot (All 20 PCs)

```{r scree-plot, fig.height=5}
ggplot(eigenval, aes(x = seq_len(nrow(eigenval)), y = Eigenvalue)) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_point(color = "steelblue", size = 2) +
  labs(
    title = "Scree Plot: Eigenvalues of Principal Components",
    x = "Principal Component",
    y = "Eigenvalue"
  ) +
  theme_minimal()
```

## PCA after outlier removal (`build_pca_clean`)

Same plot layout as **build PCA** above; axes use **`build_pca_clean.eigenval`** for variance labels. PCs are **re-fit** after excluding outliers, so numeric PC values are on a **new** scale (do not compare directly to `build_pca`).

```{r pca-clean-eigenval, echo=FALSE}
eigenval_clean_file <- file.path(DATA_DIR, "preprocess", "build_pca_clean.eigenval")
has_eigenval_clean <- file.exists(eigenval_clean_file)
if (has_eigenval_clean) {
  eigenval_clean <- utils::read.table(eigenval_clean_file, header = FALSE, stringsAsFactors = FALSE)
  colnames(eigenval_clean) <- "Eigenvalue"
  eigenval_clean$PC <- paste0("PC", seq_len(nrow(eigenval_clean)))
  eigenval_clean$Variance_Explained <- 100 * eigenval_clean$Eigenvalue / sum(eigenval_clean$Eigenvalue)
} else {
  eigenval_clean <- NULL
}
```

### Variance explained (first 10 PCs)

```{r pca-clean-variance-bars, fig.height=5, eval=exists("has_eigenval_clean") && has_eigenval_clean}
ggplot(eigenval_clean[1:min(10L, nrow(eigenval_clean)), ], aes(x = factor(PC, levels = PC), y = Variance_Explained)) +
  geom_bar(stat = "identity", fill = "darkseagreen") +
  geom_text(aes(label = sprintf("%.2f%%", Variance_Explained)), vjust = -0.5) +
  labs(
    title = "Variance explained by principal components (build_pca_clean)",
    x = "Principal Component",
    y = "Variance Explained (%)"
  ) +
  theme_minimal()
```

### PC1 vs PC2

```{r pca-clean-pc1-pc2, fig.height=6, eval=exists("pca_clean") && !is.null(pca_clean) && nrow(pca_clean) > 0}
if ("Status" %in% names(pca_clean) && sum(!is.na(pca_clean$Status)) > 0) {
  ggplot(pca_clean, aes(x = PC1, y = PC2, color = Status)) +
    geom_point(alpha = 0.6) +
    labs(
      title = "PC1 vs PC2 (build_pca_clean)",
      x = if (exists("has_eigenval_clean") && has_eigenval_clean) paste0("PC1 (", round(eigenval_clean$Variance_Explained[1], 2), "%)") else "PC1",
      y = if (exists("has_eigenval_clean") && has_eigenval_clean) paste0("PC2 (", round(eigenval_clean$Variance_Explained[2], 2), "%)") else "PC2",
      color = "Status"
    ) +
    theme_minimal()
} else {
  ggplot(pca_clean, aes(x = PC1, y = PC2)) +
    geom_point(alpha = 0.6, color = "steelblue") +
    labs(title = "PC1 vs PC2 (build_pca_clean)", x = "PC1", y = "PC2") +
    theme_minimal()
}
```

```{r pca-clean-pc1-pc2-freeze, fig.height=6, eval=exists("pca_clean") && !is.null(pca_clean) && nrow(pca_clean) > 0 && exists("has_freeze") && has_freeze && "Freeze" %in% names(pca_clean) && sum(!is.na(pca_clean$Freeze)) > 0}
ggplot(pca_clean %>% dplyr::filter(!is.na(Freeze)), aes(x = PC1, y = PC2, color = factor(Freeze))) +
  geom_point(alpha = 0.6) +
  labs(
    title = "PC1 vs PC2 by Freeze (build_pca_clean)",
    x = if (exists("has_eigenval_clean") && has_eigenval_clean) paste0("PC1 (", round(eigenval_clean$Variance_Explained[1], 2), "%)") else "PC1",
    y = if (exists("has_eigenval_clean") && has_eigenval_clean) paste0("PC2 (", round(eigenval_clean$Variance_Explained[2], 2), "%)") else "PC2",
    color = "Freeze"
  ) +
  theme_minimal()
```

### PC3 vs PC4

```{r pca-clean-pc3-pc4, fig.height=6, eval=exists("pca_clean") && !is.null(pca_clean) && nrow(pca_clean) > 0 && exists("has_eigenval_clean") && has_eigenval_clean && nrow(eigenval_clean) >= 4L}
if ("Status" %in% names(pca_clean) && sum(!is.na(pca_clean$Status)) > 0) {
  ggplot(pca_clean, aes(x = PC3, y = PC4, color = Status)) +
    geom_point(alpha = 0.6) +
    labs(
      title = "PC3 vs PC4 (build_pca_clean)",
      x = paste0("PC3 (", round(eigenval_clean$Variance_Explained[3], 2), "%)"),
      y = paste0("PC4 (", round(eigenval_clean$Variance_Explained[4], 2), "%)"),
      color = "Status"
    ) +
    theme_minimal()
} else {
  ggplot(pca_clean, aes(x = PC3, y = PC4)) +
    geom_point(alpha = 0.6, color = "steelblue") +
    labs(title = "PC3 vs PC4 (build_pca_clean)", x = "PC3", y = "PC4") +
    theme_minimal()
}
```

```{r pca-clean-pc3-pc4-freeze, fig.height=6, eval=exists("pca_clean") && !is.null(pca_clean) && nrow(pca_clean) > 0 && exists("has_eigenval_clean") && has_eigenval_clean && nrow(eigenval_clean) >= 4L && exists("has_freeze") && has_freeze && "Freeze" %in% names(pca_clean) && sum(!is.na(pca_clean$Freeze)) > 0}
ggplot(pca_clean %>% dplyr::filter(!is.na(Freeze)), aes(x = PC3, y = PC4, color = factor(Freeze))) +
  geom_point(alpha = 0.6) +
  labs(
    title = "PC3 vs PC4 by Freeze (build_pca_clean)",
    x = paste0("PC3 (", round(eigenval_clean$Variance_Explained[3], 2), "%)"),
    y = paste0("PC4 (", round(eigenval_clean$Variance_Explained[4], 2), "%)"),
    color = "Freeze"
  ) +
  theme_minimal()
```

### Scree plot (all 20 PCs)

```{r pca-clean-scree, fig.height=5, eval=exists("has_eigenval_clean") && has_eigenval_clean && !is.null(eigenval_clean) && nrow(eigenval_clean) > 0}
ggplot(eigenval_clean, aes(x = seq_len(nrow(eigenval_clean)), y = Eigenvalue)) +
  geom_line(color = "darkseagreen", linewidth = 1) +
  geom_point(color = "darkseagreen", size = 2) +
  labs(
    title = "Scree plot: eigenvalues (build_pca_clean)",
    x = "Principal Component",
    y = "Eigenvalue"
  ) +
  theme_minimal()
```

---

# Session Info

```{r session-info}
sessionInfo()
```
