---
title: "ExWAS Quality Control Report"
author: "Automated QC Pipeline"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    theme: flatly
    code_folding: hide
params:
  project: "PROJECT"
  data_dir: "data"
  covariate_file: ""   # single file WITH HEADER: FID, IID, Status, Freeze (Regnie + report + pipeline)
  # Optional override; if empty, report loads anc_group.tsv next to this .Rmd or in getwd()
  anc_group_file: ""
  # Optional: override path to MNP validation summary TSV (default data/mnp/mnp.validation.metrics.tsv under data_dir)
  mnp_validation_metrics: ""
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, fig.width = 10, fig.height = 6)
library(tidyverse)
library(knitr)
library(yaml)
library(gridExtra)
# Read config file
config <- read_yaml("config.yml")

# Project settings (params override config, for snakemake compatibility)
PROJECT <- if (!is.null(params$project)) params$project else config$project$name
DATA_DIR <- if (!is.null(params$data_dir)) params$data_dir else "data"
inp <- config$input
COVARIATE_FILE <- if (!is.null(params$covariate_file) && nzchar(params$covariate_file)) {
  params$covariate_file
} else if (!is.null(inp) && !is.null(inp$covariates)) {
  as.character(inp$covariates)[1]
} else {
  ""
}
if (length(COVARIATE_FILE) == 0 || is.na(COVARIATE_FILE[1]) || !nzchar(COVARIATE_FILE[1])) COVARIATE_FILE <- ""

ANC_GROUP_FILE <- if (!is.null(params$anc_group_file) && nzchar(params$anc_group_file)) {
  params$anc_group_file
} else if (!is.null(inp) && !is.null(inp$anc_group)) {
  as.character(inp$anc_group)[1]
} else {
  ""
}
if (length(ANC_GROUP_FILE) == 0 || is.na(ANC_GROUP_FILE[1]) || !nzchar(ANC_GROUP_FILE[1])) ANC_GROUP_FILE <- ""

MNP_VALIDATION_METRICS <- if (!is.null(params$mnp_validation_metrics) && nzchar(params$mnp_validation_metrics)) {
  params$mnp_validation_metrics
} else {
  file.path(DATA_DIR, "mnp", "mnp.validation.metrics.tsv")
}

qc_cfg <- config$qc
num_qc <- function(x, default) {
  v <- suppressWarnings(as.numeric(x))
  if (length(v) < 1 || is.na(v[1])) default else v[1]
}
MIND_THR <- if (!is.null(qc_cfg) && !is.null(qc_cfg$mind_thr)) num_qc(qc_cfg$mind_thr, 0.05) else 0.05
MAF_THR <- if (!is.null(qc_cfg) && !is.null(qc_cfg$maf_thr)) num_qc(qc_cfg$maf_thr, 0.05) else 0.05
GENO_THR <- if (!is.null(qc_cfg) && !is.null(qc_cfg$geno_thr)) num_qc(qc_cfg$geno_thr, 0.01) else 0.01
HWE_THR <- if (!is.null(qc_cfg) && !is.null(qc_cfg$hwe_thr)) num_qc(qc_cfg$hwe_thr, 1e-6) else 1e-6
pca_legend_theme <- theme(
  legend.position = "right",
  legend.title = element_text(size = 13),
  legend.text = element_text(size = 12),
  legend.key.size = grid::unit(0.65, "cm")
)
pca_legend_guides <- guides(
  color = guide_legend(override.aes = list(size = 4)),
  shape = guide_legend(override.aes = list(size = 4))
)

# Side-by-side bar panels: rare types (left) vs SNP (right); no shared x-axis clutter.
plot_type_count_panels <- function(df, y_col, title, subtitle = NULL, stage_col = NULL) {
  rare_levels <- c("INDEL", "MNP", "OTHER")
  dodge <- !is.null(stage_col)
  p_rare <- df %>%
    dplyr::filter(TYPE %in% rare_levels)
  if (nrow(p_rare) == 0) {
    p_rare <- ggplot2::ggplot() +
      ggplot2::annotate("text", x = 0.5, y = 0.5, label = "No INDEL/MNP/OTHER counts") +
      ggplot2::labs(title = "INDEL / MNP / OTHER") +
      ggplot2::theme_void()
  } else if (dodge) {
    p_rare <- ggplot2::ggplot(
      p_rare,
      ggplot2::aes(
        x = factor(TYPE, levels = intersect(rare_levels, unique(TYPE))),
        y = .data[[y_col]],
        fill = .data[[stage_col]]
      )
    ) +
      ggplot2::geom_col(position = ggplot2::position_dodge(width = 0.75), width = 0.65) +
      ggplot2::scale_fill_manual(values = c("Prefilter" = "#E74C3C", "Postfilter" = "#3498DB"))
  } else {
    p_rare <- ggplot2::ggplot(
      p_rare,
      ggplot2::aes(
        x = factor(TYPE, levels = intersect(rare_levels, unique(TYPE))),
        y = .data[[y_col]],
        fill = TYPE
      )
    ) +
      ggplot2::geom_col(width = 0.65)
  }
  p_rare <- p_rare +
    ggplot2::scale_y_continuous(labels = scales::comma, expand = ggplot2::expansion(mult = c(0, 0.08))) +
    ggplot2::labs(title = "INDEL / MNP / OTHER", x = NULL, y = "Count") +
    ggplot2::theme_minimal() +
    ggplot2::theme(legend.position = "bottom")

  p_snp <- df %>% dplyr::filter(TYPE == "SNP")
  if (nrow(p_snp) == 0) {
    p_snp <- ggplot2::ggplot() +
      ggplot2::annotate("text", x = 0.5, y = 0.5, label = "No SNP counts") +
      ggplot2::labs(title = "SNP") +
      ggplot2::theme_void()
  } else if (dodge) {
    p_snp <- ggplot2::ggplot(
      p_snp,
      ggplot2::aes(x = "SNP", y = .data[[y_col]], fill = .data[[stage_col]])
    ) +
      ggplot2::geom_col(position = ggplot2::position_dodge(width = 0.75), width = 0.45) +
      ggplot2::scale_fill_manual(values = c("Prefilter" = "#E74C3C", "Postfilter" = "#3498DB"))
  } else {
    p_snp <- ggplot2::ggplot(p_snp, ggplot2::aes(x = "SNP", y = .data[[y_col]], fill = TYPE)) +
      ggplot2::geom_col(width = 0.45)
  }
  p_snp <- p_snp +
    ggplot2::scale_y_continuous(labels = scales::comma, expand = ggplot2::expansion(mult = c(0, 0.08))) +
    ggplot2::labs(title = "SNP", x = NULL, y = "Count") +
    ggplot2::theme_minimal() +
    ggplot2::theme(legend.position = "bottom")

  gridExtra::grid.arrange(
    p_rare, p_snp, ncol = 2,
    top = grid::textGrob(title, gp = grid::gpar(fontsize = 14, fontface = "bold")),
    bottom = if (!is.null(subtitle) && nzchar(subtitle)) {
      grid::textGrob(subtitle, gp = grid::gpar(fontsize = 10))
    } else {
      NULL
    }
  )
}
```

# Overview

This report summarizes quality control performed on whole exome sequencing data for the **`r PROJECT`** study.

## QC Pipeline Steps (updated: build + annotation workflow)

This report supports the **current pipeline** using `select_variants3.smk`, `build_files.smk`, and `annotation_files3.smk`:

1. **Sex check** (main workflow): ChrX F-statistic; ambiguous samples excluded.
2. **Variant QC (build)**: Geno, MAF, HWE, SNPs only per chromosome → `site-qc.var-qc` (no LD on chr files).
3. **Sample missingness (build)**: Per-chromosome missingness is computed on `site-qc.var-qc`, then aggregated to apply the MIND threshold.
   Samples passing MIND are merged across autosomes into **merge_mind**.
4. **LD pruning (build)**: Performed on **merge_mind** only (`--indep-pairwise`; chr8 and long-LD regions excluded).
   Output (`merge_mind.prune.in/.out`) is used for PCA and heterozygosity only and does not remove variants from the final **build** dataset.
5. **Heterozygosity (build)**: Computed on **merge_mind.pruned** (the LD-pruned SNP subset from **merge_mind**).
   Outlier samples are removed from **merge_mind** to create **build**, while keeping the full post-QC SNP set in **build**.
6. **PCA**: Runs on **build** samples while restricting variants to `merge_mind.prune.in` (with chr/long-LD exclusions in the PCA rule).
   Outputs include `build_pca.eigenvec`, `build_pca.eigenval`, and `build_pca.eigenvec.allele`; loading plots use positions from `build.pvar`.
7. **Annotation base (annotation_files3)**: Subset to build-passing samples (`site-qc.het_miss`) and run first no-sample VEP parse.
8. **MNP branch (mnp.smk split path)**: Detect candidate pairs, run sample-level PASS/read-check filtering, rewrite MNP genotypes, and generate MNP no-sample VEP outputs.
9. **Rejoin after MNP**: Merge adjusted MNP records back into both genotype BCF and no-sample VEP track.
10. **Rare variant freeze QC (rare_variant_qc.smk)**: Controls-only Freeze2 vs Freeze3 missingness/carrier tests; generate flagged IDs and merged exclusion list.
11. **Final annotation after exclusions**: Exclude flagged IDs from post-MNP genotype/annotation tracks, then produce final annotation pgen and `annotation.no_sample.vep.report.csv`.

Unless explicitly stated, filtering steps are cumulative, and counts reflect the dataset at that stage of the pipeline.

---

# Load Data

A single covariate file (with header) supplies Status and Freeze for all QC views and pipeline steps.

Ancestry groups for PCA use a tab-separated file with columns `FID`, `IID`, `AncGroup` (`anc_group.tsv` beside this report, or the path in `config.yml`).

```{r load-data}
# If params/config did not set ancestry path, use anc_group.tsv next to this .Rmd or in getwd()
if (!nzchar(ANC_GROUP_FILE)) {
  cand <- character()
  if (requireNamespace("knitr", quietly = TRUE)) {
    ci <- tryCatch(knitr::current_input(), error = function(e) "")
    if (is.character(ci) && length(ci) >= 1L && nzchar(ci[1])) {
      cand <- c(cand, file.path(dirname(normalizePath(ci[1], winslash = "/", mustWork = FALSE)), "anc_group.tsv"))
    }
  }
  cand <- unique(c(cand, file.path(getwd(), "anc_group.tsv")))
  cand <- cand[!is.na(cand) & nzchar(cand)]
  for (p in cand) {
    if (file.exists(p)) {
      ANC_GROUP_FILE <- p
      break
    }
  }
}

# Load sex inference results (main workflow; optional if only build/annotation ran)
sex_file <- file.path(DATA_DIR, "qc/reports", "sex_inference.txt")
if (file.exists(sex_file)) {
  sex_report <- read.table(sex_file, header = TRUE, stringsAsFactors = FALSE, comment.char = "")
} else {
  warning("sex_inference.txt not found (from main workflow). Sex check section may be empty.")
  sex_report <- data.frame(Sample_ID = character(0), O_HOM = numeric(0), E_HOM = numeric(0), OBS_CT = numeric(0), F_inbreed = numeric(0), F_sex = numeric(0), Sex = character(0), stringsAsFactors = FALSE)
}

# Load missingness report (build_files: data/qc/reports/missingness_report.txt)
miss_file <- file.path(DATA_DIR, "qc/reports", "missingness_report.txt")
if (file.exists(miss_file)) {
  miss_report <- read.csv(miss_file, stringsAsFactors = FALSE)
} else {
  warning("missingness_report.txt not found. Using empty table.")
  miss_report <- data.frame(IID = character(0), F_MISS = numeric(0), stringsAsFactors = FALSE)
}

# Load heterozygosity: prefer per-sample heterozygosity_report.txt (main); else build_files het_report.txt + het_exclusions
het_file <- file.path(DATA_DIR, "qc/reports", "heterozygosity_report.txt")
het_report_file_alt <- file.path(DATA_DIR, "qc/reports", "het_report.txt")
het_exclusions_file <- file.path(DATA_DIR, "qc/exclusions", "het_exclusions.txt")
if (file.exists(het_file)) {
  het_report <- read.csv(het_file, stringsAsFactors = FALSE)
} else if (file.exists(het_exclusions_file)) {
  het_excl <- read.table(het_exclusions_file, header = FALSE, stringsAsFactors = FALSE)
  if (ncol(het_excl) >= 2) names(het_excl) <- c("FID", "IID") else names(het_excl) <- "IID"
  het_excl$Fhat <- NA_real_
  het_report <- het_excl
  if (file.exists(het_report_file_alt)) {
    het_summary <- read.table(het_report_file_alt, header = TRUE, stringsAsFactors = FALSE)
    if (nrow(het_summary) > 0) message("Heterozygosity summary (build_files): n=", het_summary$n[1], " mean F, sd, bounds in het_report.txt")
  }
} else {
  warning("heterozygosity_report.txt and het_exclusions.txt not found. Heterozygosity section may be empty.")
  het_report <- data.frame(IID = character(0), Fhat = numeric(0), stringsAsFactors = FALSE)
}

# Load PCA results (build_files: build_pca.eigenvec / build_pca.eigenval)
pca_file <- file.path(DATA_DIR, "preprocess", "build_pca.eigenvec")
if (!file.exists(pca_file)) pca_file <- file.path(DATA_DIR, "preprocess", "build.eigenvec")
if (!file.exists(pca_file)) stop("PCA file not found: need build_pca.eigenvec or build.eigenvec in data/preprocess/")
pca <- read.table(pca_file, header = TRUE, stringsAsFactors = FALSE, comment.char = "")

# MNP read-check results (optional; for plain-English validation summary)
mnp_read_check_path <- file.path(DATA_DIR, "mnp", "mnp_read_check.results.mapq20.csv")
cases_path <- file.path(DATA_DIR, "final", "cases.txt")
controls_path <- file.path(DATA_DIR, "final", "controls.txt")

# Allele PC weights + pvar POS are read in the PCA section (build_pca.eigenvec.allele + build.pvar)

# Load variant type counts (bcftools prefilter: select_variants3; fallback plink prefilter from build_files)
prefilter_file_used <- character(22)
prefilter_counts <- lapply(1:22, function(chr) {
  f_sv <- file.path(DATA_DIR, "qc/reports", paste0("chr", chr, ".prefilter.variant_types.select_variants.txt"))
  f_bf <- file.path(DATA_DIR, "qc/reports", paste0("chr", chr, ".prefilter.variant_types.txt"))
  f <- if (file.exists(f_sv)) f_sv else f_bf
  prefilter_file_used[chr] <<- if (file.exists(f)) f else NA_character_
  if (file.exists(f)) {
    read.table(f, header = TRUE, stringsAsFactors = FALSE, comment.char = "") %>%
      mutate(CHR = chr, source_file = basename(f))
  } else {
    data.frame(TYPE = character(0), COUNT = numeric(0), CHR = integer(0), source_file = character(0))
  }
}) %>% bind_rows()
if (nrow(prefilter_counts) == 0) message("No prefilter.variant_types files found.")

# Load variant type counts (bcftools postfilter: site-qc PASS)
postfilter_file_used <- character(22)
postfilter_counts <- lapply(1:22, function(chr) {
  f <- file.path(DATA_DIR, "qc/reports", paste0("chr", chr, ".postfilter.variant_types.txt"))
  postfilter_file_used[chr] <<- if (file.exists(f)) f else NA_character_
  if (file.exists(f)) {
    read.table(f, header = TRUE, stringsAsFactors = FALSE, comment.char = "") %>%
      mutate(CHR = chr, source_file = basename(f))
  } else {
    data.frame(TYPE = character(0), COUNT = numeric(0), CHR = integer(0), source_file = character(0))
  }
}) %>% bind_rows()
if (nrow(postfilter_counts) == 0) message("No postfilter.variant_types.txt found.")

# Load variant counts after plink filtering (SNPs only; build_files)
postplink_counts <- lapply(1:22, function(chr) {
  f <- file.path(DATA_DIR, "qc/reports", paste0("chr", chr, ".postplink.variant_count.txt"))
  if (file.exists(f)) read.table(f, header = TRUE, stringsAsFactors = FALSE, comment.char = "")
  else data.frame(VARIANTS = numeric(0))
}) %>% bind_rows()

postplink_type_counts <- lapply(1:22, function(chr) {
  f <- file.path(DATA_DIR, "qc/reports", paste0("chr", chr, ".postplink.variant_types.txt"))
  if (file.exists(f)) {
    read.table(f, header = TRUE, stringsAsFactors = FALSE, comment.char = "") %>%
      mutate(CHR = chr)
  } else {
    data.frame(TYPE = character(0), COUNT = numeric(0), CHR = integer(0))
  }
}) %>% bind_rows()

# Load plink filter metrics (build_files: chr*.plink_filter_metrics.txt; has CHR, FILTER, VARIANTS_REMOVED)
plink_filter_metrics <- lapply(1:22, function(chr) {
  f <- file.path(DATA_DIR, "qc/reports", paste0("chr", chr, ".plink_filter_metrics.txt"))
  if (file.exists(f)) read.table(f, header = TRUE, stringsAsFactors = FALSE, comment.char = "")
  else data.frame(CHR = integer(0), FILTER = character(0), VARIANTS_REMOVED = numeric(0))
}) %>% bind_rows()

# Load variant QC metrics for distribution plots (build_files: chr*.site-qc.var-qc.{afreq,vmiss,hardy})
# When a file is missing, return 0-row df so we only have rows with real data (no NA placeholders).
variant_maf <- lapply(1:22, function(chr) {
  f <- file.path(DATA_DIR, "plink", paste0("chr", chr, ".site-qc.var-qc.afreq"))
  if (file.exists(f)) {
    read.table(f, header = TRUE, stringsAsFactors = FALSE, comment.char = "") %>%
      mutate(CHR = chr)
  } else data.frame()
}) %>% bind_rows()

variant_miss <- lapply(1:22, function(chr) {
  f <- file.path(DATA_DIR, "plink", paste0("chr", chr, ".site-qc.var-qc.vmiss"))
  if (file.exists(f)) {
    read.table(f, header = TRUE, stringsAsFactors = FALSE, comment.char = "") %>%
      mutate(CHR = chr)
  } else data.frame()
}) %>% bind_rows()

variant_hwe <- lapply(1:22, function(chr) {
  f <- file.path(DATA_DIR, "plink", paste0("chr", chr, ".site-qc.var-qc.hardy"))
  if (file.exists(f)) {
    d <- read.table(f, header = TRUE, stringsAsFactors = FALSE, comment.char = "")
    d$CHR <- chr
    # PLINK2 --hardy midp writes p-value in MIDP column; use as P for plotting (same as original report)
    if ("MIDP" %in% names(d)) d$P <- suppressWarnings(as.numeric(d$MIDP))
    d
  } else data.frame()
}) %>% bind_rows()

# Load variant type counts after MNP correction (same format as prefilter/postfilter).
# Expects data/qc/reports/chr{1..22}.postmnp.variant_types.txt from count_variant_types_postmnp (annotation workflow).
postmnp_paths <- file.path(DATA_DIR, "qc/reports", paste0("chr", 1:22, ".postmnp.variant_types.txt"))
n_postmnp_files_present <- sum(file.exists(postmnp_paths))
postmnp_counts <- lapply(1:22, function(chr) {
  f <- postmnp_paths[[chr]]
  if (file.exists(f)) {
    read.table(f, header = TRUE, stringsAsFactors = FALSE, comment.char = "", strip.white = TRUE) %>%
      mutate(CHR = chr)
  } else {
    data.frame(TYPE = character(0), COUNT = numeric(0), CHR = integer(0))
  }
}) %>% bind_rows()
if (nrow(postmnp_counts) > 0 && "COUNT" %in% names(postmnp_counts)) {
  postmnp_counts$COUNT <- suppressWarnings(as.numeric(gsub(",", "", trimws(as.character(postmnp_counts$COUNT)))))
}
has_postmnp <- nrow(postmnp_counts) > 0 &&
  "TYPE" %in% names(postmnp_counts) &&
  "COUNT" %in% names(postmnp_counts) &&
  sum(!is.na(postmnp_counts$COUNT) & postmnp_counts$TYPE != "", na.rm = TRUE) > 0

# MNP validation / read-check summary (wide TSV; optional)
mnp_validation_metrics <- if (file.exists(MNP_VALIDATION_METRICS)) {
  read.delim(MNP_VALIDATION_METRICS, sep = "\t", stringsAsFactors = FALSE, comment.char = "",
               strip.white = TRUE, check.names = FALSE)
} else {
  data.frame()
}
has_mnp_validation_metrics <- nrow(mnp_validation_metrics) > 0
if (has_mnp_validation_metrics && nrow(mnp_validation_metrics) == 1L) {
  mnp_validation_metrics_long <- mnp_validation_metrics %>%
    tidyr::pivot_longer(dplyr::everything(), names_to = "metric", values_to = "value") %>%
    mutate(value = suppressWarnings(as.numeric(value)))
} else if (has_mnp_validation_metrics) {
  mnp_validation_metrics_long <- mnp_validation_metrics
} else {
  mnp_validation_metrics_long <- data.frame()
}

# Plain-English MNP validation counts (from metrics TSV + optional read-check CSV)
mnp_metric_val <- function(name) {
  if (!has_mnp_validation_metrics || !"metric" %in% names(mnp_validation_metrics_long)) return(NA_real_)
  v <- mnp_validation_metrics_long$value[mnp_validation_metrics_long$metric == name]
  if (!length(v)) return(NA_real_)
  suppressWarnings(as.numeric(v[1]))
}
n_mnp_pairs_strong_cis <- mnp_metric_val("n_strong_cis_unique_pairs")
n_mnp_read_rows <- mnp_metric_val("n_read_validations_performed_rows")
mnp_case_ids <- character(0)
mnp_control_ids <- character(0)
if (file.exists(cases_path)) {
  mnp_case_ids <- trimws(readLines(cases_path, warn = FALSE))
  mnp_case_ids <- mnp_case_ids[nzchar(mnp_case_ids) & !grepl("^#", mnp_case_ids)]
}
if (file.exists(controls_path)) {
  mnp_control_ids <- trimws(readLines(controls_path, warn = FALSE))
  mnp_control_ids <- mnp_control_ids[nzchar(mnp_control_ids) & !grepl("^#", mnp_control_ids)]
}
mnp_strong_case_samples <- NA_integer_
mnp_strong_control_samples <- NA_integer_
if (file.exists(mnp_read_check_path)) {
  mnp_rc <- utils::read.csv(mnp_read_check_path, stringsAsFactors = FALSE, check.names = FALSE)
  if ("sample" %in% names(mnp_rc) && "Strong_Cis_Default" %in% names(mnp_rc)) {
    strong_samples <- unique(trimws(mnp_rc$sample[mnp_rc$Strong_Cis_Default == 1 | mnp_rc$Strong_Cis_Default == "1"]))
    mnp_strong_case_samples <- sum(strong_samples %in% mnp_case_ids)
    mnp_strong_control_samples <- sum(strong_samples %in% mnp_control_ids)
  }
}

# Rename sex inference Status column to Sex first
colnames(sex_report)[colnames(sex_report) == "Status"] <- "Sex"

# Status and Freeze from single covariate file (FID, IID, FREEZE, STATUS)
id_in_report <- function(d) if ("IID" %in% names(d)) trimws(as.character(d$IID)) else trimws(as.character(d[,1]))
status_lookup <- character(0)
freeze_lookup <- character(0)
# Use length nrow(.) so 0-row tables (e.g. empty missingness_report.txt) don't trigger replacement error
sex_report$Status <- rep(NA_character_, nrow(sex_report))
sex_report$Freeze <- rep(NA_character_, nrow(sex_report))
miss_report$Status <- rep(NA_character_, nrow(miss_report))
miss_report$Freeze <- rep(NA_character_, nrow(miss_report))
het_report$Status <- rep(NA_character_, nrow(het_report))
het_report$Freeze <- rep(NA_character_, nrow(het_report))
pca$Status <- rep(NA_character_, nrow(pca))
pca$Freeze <- rep(NA_character_, nrow(pca))
pca$AncGroup <- rep(NA_character_, nrow(pca))
pca$ANC <- rep(NA_character_, nrow(pca))
has_freeze <- FALSE
has_anc_group <- FALSE
# Filled in ancestry block so PCA plots can show *why* AncGroup is missing (not the same generic line every time)
anc_group_user_msg <- ""

# PLINK eigenvec: IID is column 2; header may be #FID / IID so names are not always "IID"
pca_sample_iid <- function() {
  if ("IID" %in% names(pca)) return(trimws(as.character(pca$IID)))
  if (ncol(pca) >= 2) return(trimws(as.character(pca[[2]])))
  trimws(as.character(pca[[1]]))
}

if (COVARIATE_FILE != "" && file.exists(COVARIATE_FILE)) {
  cov_df <- read.table(COVARIATE_FILE, header = TRUE, stringsAsFactors = FALSE, comment.char = "", check.names = FALSE)
  if (nrow(cov_df) > 0 && ncol(cov_df) >= 2) {
    cov_names_lc <- tolower(trimws(names(cov_df)))
    iid_idx <- match("iid", cov_names_lc)
    status_idx <- match("status", cov_names_lc)
    freeze_idx <- match("freeze", cov_names_lc)
    if (is.na(iid_idx)) iid_idx <- 2L
    iids <- trimws(as.character(cov_df[[iid_idx]]))

    if (!is.na(status_idx)) {
      raw_status <- trimws(as.character(cov_df[[status_idx]]))
      status_val <- ifelse(raw_status %in% c("1", "Control", "control"), "Control",
                    ifelse(raw_status %in% c("2", "Case", "case"), "Case", NA_character_))
      status_lookup <- setNames(status_val, iids)
      sex_report$Status <- as.character(status_lookup[id_in_report(sex_report)])
      miss_report$Status <- as.character(status_lookup[id_in_report(miss_report)])
      het_report$Status <- as.character(status_lookup[id_in_report(het_report)])
      pca$Status <- as.character(status_lookup[id_in_report(pca)])
    }

    if (!is.na(freeze_idx)) {
      freeze_lookup <- setNames(trimws(as.character(cov_df[[freeze_idx]])), iids)
      sex_report$Freeze <- as.character(freeze_lookup[id_in_report(sex_report)])
      miss_report$Freeze <- as.character(freeze_lookup[id_in_report(miss_report)])
      het_report$Freeze <- as.character(freeze_lookup[id_in_report(het_report)])
      pca$Freeze <- as.character(freeze_lookup[id_in_report(pca)])
      has_freeze <- TRUE
    }
  }
}

# Ancestry groups: FID, IID, AncGroup (tab-separated) → pca$AncGroup by IID (same IDs as build_pca.eigenvec)
# Path: params$anc_group_file or config input: anc_group (must be the path R can read when knitting)
anc_group_path <- ANC_GROUP_FILE
if (nzchar(anc_group_path) && file.exists(anc_group_path)) {
  anc_g <- read.delim(anc_group_path, sep = "\t", stringsAsFactors = FALSE, comment.char = "", strip.white = TRUE, check.names = FALSE)
  names(anc_g) <- trimws(names(anc_g))
  if (!"IID" %in% names(anc_g)) {
    stop("anc_group file must contain IID column: ", anc_group_path)
  }
  ag_col <- if ("AncGroup" %in% names(anc_g)) {
    "AncGroup"
  } else if ("ANC" %in% names(anc_g)) {
    "ANC"
  } else {
    NA_character_
  }
  if (is.na(ag_col)) {
    stop("anc_group file must contain AncGroup column: ", anc_group_path)
  }
  iid_key <- trimws(as.character(anc_g$IID))
  anc_lookup <- setNames(trimws(as.character(anc_g[[ag_col]])), iid_key)
  pid <- pca_sample_iid()
  pca$AncGroup <- unname(anc_lookup[pid])
  pca$ANC <- pca$AncGroup
  has_anc_group <- sum(!is.na(pca$AncGroup) & nzchar(as.character(pca$AncGroup))) > 0
  n_iid_overlap <- sum(pid %in% iid_key)
  if (!has_anc_group) {
    warning(
      "Ancestry file loaded but no non-empty AncGroup on PCA rows. IID overlap: ",
      n_iid_overlap, " of ", length(pid), ". First PCA IID: ", pid[1], "; first file IID: ", iid_key[1]
    )
    if (n_iid_overlap == 0L) {
      anc_group_user_msg <- paste0(
        "The ancestry file was read from **`", anc_group_path, "`**, but **none** of the **IID** values in **`build_pca.eigenvec`** appear in that file’s **IID** column (after trimming). ",
        "IDs must match **exactly** (e.g. same `12345` vs `sample_12345`). First PCA IID: `", pid[1], "`. First IID in ancestry file: `", iid_key[1], "`."
      )
    } else {
      anc_group_user_msg <- paste0(
        "**", n_iid_overlap, "** PCA sample(s) matched an ancestry-file **IID**, but **AncGroup** was missing or empty for every PCA row. ",
        "Check the **AncGroup** column in **`", anc_group_path, "`**."
      )
    }
  }
} else if (nzchar(anc_group_path)) {
  warning("Ancestry group file not found at path: ", anc_group_path)
  anc_group_user_msg <- paste0(
    "**`params$anc_group_file`** / **`config.yml`** `input: anc_group` points to **`", anc_group_path, "`**, but that path **does not exist** from R’s working directory when the report runs. ",
    "Use an **absolute path**, or knit from the directory where the file lives."
  )
} else {
  message("anc_group.tsv not found next to the .Rmd or in getwd(); copy it there or set params$anc_group_file / config input: anc_group")
  anc_group_user_msg <- paste0(
    "Could not find **`anc_group.tsv`** next to **`exwas_qc_report2.Rmd`** or in **`getwd()`** (", getwd(), "). ",
    "Put the file beside the Rmd, knit from the folder that contains it, or set **`params$anc_group_file`** to the full path."
  )
}
```

---

# 1. Sex Check (chrX F-statistic)

## Method

Sex is inferred from chromosome X heterozygosity using the F-statistic (PLINK-based inbreeding coefficient on chrX):

- **F > 0.8**: Male (mostly homozygous/hemizygous on X)
- **F < 0.2**: Female (heterozygous on X)
- **0.2 ≤ F ≤ 0.8**: Ambiguous/Problem (excluded)

Samples in the ambiguous range are excluded because they may represent sample swaps, aneuploidy, or poor-quality genotype data.

```{r sex-check-summary}
sex_fail <- sex_report %>% filter(Sex == "AMBIGUOUS")
sex_pass <- sex_report %>% filter(Sex != "AMBIGUOUS")

total_samples <- nrow(sex_report)
failed_samples <- nrow(sex_fail)
passing_samples <- nrow(sex_pass)
```

## Results

Percentages are computed relative to the input at each step; if an input count is zero, percentages may be undefined.

- **Total Samples**: `r format(total_samples, big.mark = ",")`
- **Failed Sex Check**: `r format(failed_samples, big.mark = ",")` (`r round(100 * failed_samples / total_samples, 2)`%)
- **Passing Sex Check**: `r format(passing_samples, big.mark = ",")` (`r round(100 * passing_samples / total_samples, 2)`%)

```{r sex-check-breakdown}
if (total_samples > 0) {
  sex_breakdown <- sex_report %>%
    mutate(Sex = ifelse(is.na(Sex) | Sex == "", "UNKNOWN", as.character(Sex))) %>%
    count(Sex, name = "N") %>%
    mutate(Percent = round(100 * N / sum(N), 2)) %>%
    arrange(factor(Sex, levels = c("MALE", "FEMALE", "AMBIGUOUS", "UNKNOWN")))

  kable(sex_breakdown,
        col.names = c("Inferred Sex", "N", "Percent (%)"),
        caption = "Inferred sex breakdown (double-check only: MALE F>0.8, FEMALE F<0.2, AMBIGUOUS 0.2-0.8). Females are not auto-removed; only AMBIGUOUS go to sexcheck_fail.txt.")
} else {
  cat(
    "**Sex inference table empty.** Expected a full per-sample file at `", sex_file, "`.\n\n",
    "Regenerate manually from existing chrX inputs (do not need full qc_pipeline):\n\n",
    "```bash\n",
    "# if chrX.het already exists:\n",
    "snakemake -s select_variants3.smk data/qc/reports/sex_inference.txt --configfile config.yml\n",
    "# or from BCF if pgen/het missing:\n",
    "snakemake -s select_variants3.smk \\\n",
    "  data/plink/chrX.pgen data/plink/chrX.het data/qc/reports/sex_inference.txt \\\n",
    "  --configfile config.yml\n",
    "```\n",
    sep = ""
  )
}
```

```{r sex-check-by-status}
if ("Status" %in% colnames(sex_fail) && sum(!is.na(sex_report$Status)) > 0) {
  sex_fail_summary <- sex_fail %>%
    group_by(Status) %>%
    summarise(
      N_Failed = n(),
      .groups = "drop"
    )

  print(kable(sex_fail_summary, caption = "Sex Check Failures by Case/Control Status"))
} else {
  cat("No case/control status was joined onto `sex_report`, so this table cannot be populated. Check the covariate file passed to the report and make sure it has `IID` and `STATUS` columns.\n")
}
```

## F-statistic Distribution

```{r sex-f-distribution, fig.height=5}
ggplot(sex_report, aes(x = F_sex, fill = Sex)) +
  geom_histogram(bins = 50, alpha = 0.7) +
  geom_vline(xintercept = c(0.2, 0.8), linetype = "dashed", color = "red") +
  scale_fill_manual(values = c("MALE" = "steelblue", "FEMALE" = "pink", "AMBIGUOUS" = "orange")) +
  labs(
    title = "Distribution of chrX Homozygosity (Sex Inference)",
    x = "F-statistic (O(HOM)/OBS_CT)",
    y = "Count",
    fill = "Inferred Sex"
  ) +
  theme_minimal()
```

---

# 2. Variant Quality Filtering

## Method

Variants are filtered based on the presence of high-quality heterozygotes:

- **Filter**: `COUNT(GT="0/1" && DP≥10 && GQ≥20 && VAF>0.2) = 0`
- **Action**: Tag as `NoHQHet`, then exclude non-PASS variants

Variants are also filtered based on low call rate:
- **Filter**: `INFO/F_MISSING > 0.05`
- **Action**: Tag as `LowCallRate`, then exclude non-PASS variants

```{r variant-filtering}
# Use typed counts only (SNP/INDEL/MNP/OTHER). Bare TOTAL can include monomorphic/ref sites
# that inflate prefilter counts and make bcftools look far more aggressive than it is.
typed_variant_types <- c("SNP", "INDEL", "MNP", "OTHER")
sum_typed <- function(df) {
  if (nrow(df) == 0 || !"TYPE" %in% names(df) || !"COUNT" %in% names(df)) return(0)
  sum(df$COUNT[df$TYPE %in% typed_variant_types], na.rm = TRUE)
}
prefilter_total <- sum_typed(prefilter_counts)
postfilter_total <- sum_typed(postfilter_counts)
variants_removed <- prefilter_total - postfilter_total
postplink_snp_total <- if (nrow(postplink_type_counts) > 0) {
  sum(postplink_type_counts$COUNT[postplink_type_counts$TYPE == "SNP"], na.rm = TRUE)
} else if (nrow(postplink_counts) > 0 && "VARIANTS" %in% names(postplink_counts)) {
  sum(postplink_counts$VARIANTS, na.rm = TRUE)
} else {
  0L
}
```

Most SNP loss happens in **Section 3** (plink geno / MAF / HWE). The bcftools step (NoHQHet + call rate) mainly drops poorly supported sites.

### Variant filter cascade

```{r variant-filter-cascade}
variant_filter_cascade <- tibble::tibble(
  Stage = c(
    "bcftools prefilter (SNP/INDEL/MNP/OTHER)",
    "bcftools postfilter (HQHet + call rate)",
    "plink SNPs after geno/MAF/HWE"
  ),
  Total_variants = c(prefilter_total, postfilter_total, postplink_snp_total),
  Removed_from_prior = c(
    NA_integer_,
    prefilter_total - postfilter_total,
    postfilter_total - postplink_snp_total
  ),
  Pct_retained = c(
    100,
    ifelse(prefilter_total > 0, round(100 * postfilter_total / prefilter_total, 2), NA_real_),
    ifelse(postfilter_total > 0, round(100 * postplink_snp_total / postfilter_total, 2), NA_real_)
  )
)
knitr::kable(
  variant_filter_cascade,
  format.args = list(big.mark = ","),
  caption = "Where variants are lost (bcftools vs plink)"
)
```

## Results

- **Variants (Pre-filter)**: `r format(prefilter_total, big.mark = ",")`
- **Variants Removed**: `r format(variants_removed, big.mark = ",")` (`r round(100 * variants_removed / prefilter_total, 2)`%)
- **Variants (Post-filter)**: `r format(postfilter_total, big.mark = ",")` (`r round(100 * postfilter_total / prefilter_total, 2)`%)

### Variant Types Before and After Filtering

```{r variant-types-table}
variant_summary <- prefilter_counts %>%
  filter(TYPE %in% c("SNP", "INDEL", "MNP", "OTHER")) %>%
  group_by(TYPE) %>%
  summarise(Prefilter = sum(COUNT), .groups = "drop") %>%
  full_join(
    postfilter_counts %>%
      filter(TYPE %in% c("SNP", "INDEL", "MNP", "OTHER")) %>%
      group_by(TYPE) %>%
      summarise(Postfilter = sum(COUNT), .groups = "drop"),
    by = "TYPE"
  ) %>%
  mutate(
    Prefilter = ifelse(is.na(Prefilter), 0L, Prefilter),
    Postfilter = ifelse(is.na(Postfilter), 0L, Postfilter),
    Removed = Prefilter - Postfilter,
    Pct_Removed = ifelse(Prefilter > 0, round(100 * Removed / Prefilter, 2), NA_real_)
  )

kable(variant_summary, 
      format.args = list(big.mark = ","),
      caption = "Variant Counts by Type (bcftools prefilter vs site-qc postfilter)")
```

```{r variant-types-plot, fig.height=5, fig.width=11}
variant_plot_df <- variant_summary %>%
  filter(TYPE != "TOTAL") %>%
  select(TYPE, Prefilter, Postfilter) %>%
  pivot_longer(c(Prefilter, Postfilter), names_to = "Stage", values_to = "Count") %>%
  mutate(Stage = factor(Stage, levels = c("Prefilter", "Postfilter")))

plot_type_count_panels(
  variant_plot_df,
  y_col = "Count",
  stage_col = "Stage",
  title = "Variant counts before and after bcftools quality filtering",
  subtitle = "Left: INDEL / MNP / OTHER. Right: SNP. (Heavy SNP filtering is in Section 3.)"
)
```

---

# 3. Variant-Level Filters (Plink)

## Method

**Per-chromosome variant QC** (applied to SNPs only, before sample QC):

- **Genotype Missingness**: `--geno `r GENO_THR`` (`r GENO_THR * 100`%)
- **Minor Allele Frequency**: `--maf `r MAF_THR`` (`r MAF_THR * 100`%)
- **Hardy-Weinberg Equilibrium**: `--hwe `r HWE_THR` midp`

These are applied per chromosome to produce `site-qc.var-qc`. Sample missingness is then calculated on these filtered SNPs. **LD pruning is not part of this step**—see **Section 5** (*LD pruning (merge_mind)*) after sample missingness.

```{r plink-filtering}
# Calculate totals
variants_before_plink <- ifelse(length(postfilter_total) > 0 && !is.na(postfilter_total), postfilter_total, 0)
variants_after_plink <- ifelse(nrow(postplink_counts) > 0 && "VARIANTS" %in% names(postplink_counts), sum(postplink_counts$VARIANTS, na.rm = TRUE), 0)
variants_removed_plink <- variants_before_plink - variants_after_plink

# Count SNPs vs non-SNPs from postfilter variant types
if (nrow(postfilter_counts) > 0 && "TYPE" %in% names(postfilter_counts) && "COUNT" %in% names(postfilter_counts)) {
  snp_counts <- postfilter_counts %>% filter(TYPE == "SNP")
  indel_counts <- postfilter_counts %>% filter(TYPE == "INDEL")
  other_counts <- postfilter_counts %>% filter(TYPE %in% c("MNP", "OTHER"))
  
  total_snps <- sum(snp_counts$COUNT, na.rm = TRUE)
  total_indels <- sum(indel_counts$COUNT, na.rm = TRUE)
  total_other <- sum(other_counts$COUNT, na.rm = TRUE)
} else {
  total_snps <- 0
  total_indels <- 0
  total_other <- 0
}
total_non_snps <- total_indels + total_other

# Aggregate per-chr Plink filter metrics (GENO, MAF, HWE, SNPs-only, etc. — not LD)
if (nrow(plink_filter_metrics) > 0 && "FILTER" %in% names(plink_filter_metrics) && "VARIANTS_REMOVED" %in% names(plink_filter_metrics)) {
  filter_summary <- plink_filter_metrics %>%
    filter(!FILTER %in% c("PRUNE", "TOTAL")) %>%
    group_by(FILTER) %>%
    summarise(Variants_Removed = sum(VARIANTS_REMOVED, na.rm = TRUE), .groups = "drop") %>%
    mutate(
      Pct_of_Input = ifelse(variants_before_plink > 0, round(100 * Variants_Removed / variants_before_plink, 2), 0)
    )
} else {
  filter_summary <- data.frame(FILTER = character(0), Variants_Removed = numeric(0), Pct_of_Input = numeric(0))
}
```

## Results

- **Variants (Before Plink Filters)**: `r format(variants_before_plink, big.mark = ",")` (after HQHet)
  - SNPs: `r format(total_snps, big.mark = ",")` (`r round(100 * total_snps / variants_before_plink, 1)`%)
  - Indels: `r format(total_indels, big.mark = ",")` (`r round(100 * total_indels / variants_before_plink, 1)`%)
  - Other (MNPs, etc.): `r format(total_other, big.mark = ",")` (`r round(100 * total_other / variants_before_plink, 1)`%)
- **Non-SNPs Removed** (--snps-only): `r format(total_non_snps, big.mark = ",")` (`r round(100 * total_non_snps / variants_before_plink, 2)`%)
- **SNPs Removed by QC Filters**: `r format(variants_removed_plink - total_non_snps, big.mark = ",")` (`r round(100 * (variants_removed_plink - total_non_snps) / variants_before_plink, 2)`%)
- **Variants (After All Filters)**: `r format(variants_after_plink, big.mark = ",")` (SNPs only, `r round(100 * variants_after_plink / variants_before_plink, 2)`% of input)

### Distribution Analysis

These distributions show the spread of variant-level QC metrics and where thresholds are set.

```{r maf-below-threshold, fig.height=5}
# MAF from data/plink/chr*.site-qc.var-qc.afreq (PLINK --freq). Variants below threshold only (what gets filtered out). Same as original.
afreq_col <- if ("ALT_FREQS" %in% names(variant_maf)) "ALT_FREQS" else "ALT_FREQ"
if (nrow(variant_maf) > 0 && afreq_col %in% names(variant_maf)) {
  variant_maf$MAF <- pmin(variant_maf[[afreq_col]], 1 - variant_maf[[afreq_col]])
  maf_below <- variant_maf %>% filter(MAF < MAF_THR, !is.na(MAF))
  if (nrow(maf_below) > 0) {
    x_max <- max(MAF_THR * 1.05, max(maf_below$MAF, na.rm = TRUE) * 1.05, 0.001)
    print(ggplot(maf_below, aes(x = MAF)) +
      geom_histogram(bins = 50, fill = "steelblue", alpha = 0.7) +
      geom_vline(xintercept = MAF_THR, linetype = "dashed", color = "red", linewidth = 1) +
      scale_x_continuous(labels = scales::percent, limits = c(0, x_max)) +
      scale_y_continuous(labels = scales::comma) +
      labs(
        title = "MAF distribution: variants below threshold (excluded by QC)",
        subtitle = sprintf("n = %s variants excluded by MAF < %s%%", format(nrow(maf_below), big.mark = ","), MAF_THR * 100),
        x = "MAF",
        y = "Variant Count",
        caption = "Red line = MAF threshold; these variants are removed by QC"
      ) +
      theme_minimal())
  } else {
    cat("No variants below the MAF threshold.\n")
  }
} else {
  cat("No MAF data: ensure data/plink/chr*.site-qc.var-qc.afreq exist (e.g. from build_files plink2_freq_missing_hardy_build).\n")
}
```

```{r geno-distribution, fig.height=5}
# Variant genotype missingness (data/plink/chr*.site-qc.var-qc.vmiss). Same as original.
if (nrow(variant_miss) > 0 && "F_MISS" %in% names(variant_miss)) {
  variant_miss_valid <- variant_miss %>% filter(!is.na(F_MISS))
  n_above <- sum(variant_miss_valid$F_MISS > GENO_THR)
  print(ggplot(variant_miss_valid, aes(x = F_MISS)) +
    geom_histogram(bins = 50, fill = "steelblue", alpha = 0.7) +
    geom_vline(xintercept = GENO_THR, linetype = "dashed", color = "red", linewidth = 1) +
    scale_x_continuous(labels = scales::percent) +
    scale_y_continuous(labels = scales::comma) +
    labs(
      title = "Variant Genotype Missingness Distribution",
      x = "Missingness Rate",
      y = "Variant Count",
      caption = paste0("Red line = ", GENO_THR * 100, "% threshold | ",
                       format(n_above, big.mark = ","), " variants (",
                       round(100 * n_above / nrow(variant_miss_valid), 1), "%) above threshold")
    ) +
    theme_minimal())
} else {
  cat("No genotype missingness data: ensure data/plink/chr*.site-qc.var-qc.vmiss exist (e.g. from build_files plink2_freq_missing_hardy_build).\n")
}
```

```{r hwe-distribution, fig.height=5}
# HWE: PLINK2 .hardy uses MIDP column (mapped to P in load). Same style as original.
if (nrow(variant_hwe) > 0 && "P" %in% names(variant_hwe)) {
  variant_hwe_valid <- variant_hwe %>% filter(!is.na(P), is.finite(P), P > 0)
  if (nrow(variant_hwe_valid) > 0) {
    variant_hwe_valid$neg_log10_P <- -log10(variant_hwe_valid$P)
    threshold_line <- -log10(HWE_THR)
    n_below_thr <- sum(variant_hwe_valid$P < HWE_THR, na.rm = TRUE)
    print(ggplot(variant_hwe_valid, aes(x = neg_log10_P)) +
      geom_histogram(bins = 50, fill = "steelblue", alpha = 0.7) +
      geom_vline(xintercept = threshold_line, linetype = "dashed", color = "red", linewidth = 1) +
      scale_y_continuous(labels = scales::comma) +
      labs(
        title = "Hardy-Weinberg Equilibrium Distribution",
        x = "-log10(P-value)",
        y = "Variant Count",
        caption = paste0("Red line = -log10(", HWE_THR, ") = ", round(threshold_line, 1), " | ",
                         format(n_below_thr, big.mark = ","), " variants (",
                         round(100 * n_below_thr / nrow(variant_hwe_valid), 1),
                         "%) above threshold (more deviant)")
      ) +
      theme_minimal())
  } else {
    cat("No valid HWE P-values in data.\n")
  }
} else {
  cat("No HWE data: ensure data/plink/chr*.site-qc.var-qc.hardy exist (e.g. from build_files plink2_freq_missing_hardy_build).\n")
}
```

---

# 3b. MNP Correction

Multi-nucleotide variants (MNPs) are identified and corrected during annotation.

This step:
- Identifies candidate variant pairs from VEP output
- Validates cis/trans configuration using read-level evidence
- Updates genotypes accordingly

Counts below reflect the post-correction dataset.

**MNP validation summary** (`mnp.validation.metrics.tsv`) is loaded when present (path: `r MNP_VALIDATION_METRICS`). **Post-MNP variant-type files** are expected at `data/qc/reports/chrN.postmnp.variant_types.txt` for `N = 1..22` (from `count_variant_types_postmnp` in the annotation workflow).

Read-check terms used in the table:
- **Total_Reads**: all reads retrieved in the query window (after filtering settings such as MAPQ).
- **Reads_with_Both**: reads supporting both variants together (same-read cis support).
- **Trans_Evidence**: reads supporting only one variant at a time (supports trans/mixed patterns).
- **Phaseable_Reads**: informative reads used for cis/trans assessment (`Reads_with_Both + Trans_Evidence`).
- **Strong_Cis_Default**: strict cis call (`Reads_with_Both >= 3`, `Phaseable_Reads >= 5`, and `Cis_Fraction_Phaseable >= 0.90`).

```{r mnp-validation-summary, results='asis'}
if (has_mnp_validation_metrics && is.finite(n_mnp_pairs_strong_cis)) {
  cat(
    sprintf(
      "**Summary:** **%s** MNP pair(s) passed strong cis read validation; **%s** sample×MNP read-check row(s) were evaluated.",
      format(as.integer(n_mnp_pairs_strong_cis), big.mark = ","),
      format(as.integer(n_mnp_read_rows), big.mark = ",")
    )
  )
  if (is.finite(mnp_strong_case_samples) && is.finite(mnp_strong_control_samples)) {
    cat(
      sprintf(
        " Strong cis evidence involved **%s** case sample(s) and **%s** control sample(s) (unique IIDs with Strong_Cis_Default = 1).",
        format(mnp_strong_case_samples, big.mark = ","),
        format(mnp_strong_control_samples, big.mark = ",")
      )
    )
  }
  cat("\n\n")
} else if (!has_mnp_validation_metrics) {
  cat("_MNP validation metrics file not found; see table below if present._\n\n")
}
```

```{r mnp-validation-metrics-table}
if (has_mnp_validation_metrics) {
  if (nrow(mnp_validation_metrics) == 1L &&
      nrow(mnp_validation_metrics_long) > 0 &&
      "metric" %in% colnames(mnp_validation_metrics_long)) {
    metric_labels <- c(
      n_mnp_candidates_identified_pairs = "Candidate MNP pairs (pre-filter)",
      n_vaf_pass_pairs = "Pairs passing VAF/sample criteria",
      n_vaf_pass_rows = "VAF-pass observations (pair-sample rows)",
      n_read_validation_targets_assigned = "Pairs selected for read validation",
      n_read_validations_performed_rows = "Read validations performed (rows)",
      n_unique_pairs_validated = "Pairs with read-level result",
      n_strong_cis_rows = "Strong cis rows",
      n_not_strong_cis_rows = "Not-strong-cis rows",
      n_strong_cis_unique_pairs = "Pairs meeting strong cis threshold",
      n_cis_only_rows = "Read outcomes: cis-only",
      n_mixed_cis_trans_rows = "Read outcomes: mixed cis/trans",
      n_trans_only_rows = "Read outcomes: trans-only",
      n_no_phaseable_rows = "Read outcomes: non-phaseable",
      n_zero_total_reads_rows = "Read outcomes: zero total reads"
    )
    metric_stage <- c(
      n_mnp_candidates_identified_pairs = "Candidate identification",
      n_vaf_pass_pairs = "VAF/sample gate",
      n_vaf_pass_rows = "VAF/sample gate",
      n_read_validation_targets_assigned = "Read-check assignment",
      n_read_validations_performed_rows = "Read-check results",
      n_unique_pairs_validated = "Read-check results",
      n_strong_cis_rows = "Read-check classification",
      n_not_strong_cis_rows = "Read-check classification",
      n_strong_cis_unique_pairs = "Read-check classification",
      n_cis_only_rows = "Read-check outcomes",
      n_mixed_cis_trans_rows = "Read-check outcomes",
      n_trans_only_rows = "Read-check outcomes",
      n_no_phaseable_rows = "Read-check outcomes",
      n_zero_total_reads_rows = "Read-check outcomes"
    )
    metric_notes <- c(
      n_mnp_candidates_identified_pairs = "Unique (id1,id2) candidate pairs from find_mnp_variants",
      n_vaf_pass_pairs = "Unique pairs with at least one PASS row",
      n_vaf_pass_rows = "Total PASS rows (pair-sample observations)",
      n_read_validation_targets_assigned = "Rows assigned for read-level validation",
      n_read_validations_performed_rows = "Rows emitted by read-check results file",
      n_unique_pairs_validated = "Unique pairs represented in read-check results",
      n_strong_cis_rows = "Rows with Strong_Cis_Default == 1",
      n_not_strong_cis_rows = "Rows with Strong_Cis_Default != 1",
      n_strong_cis_unique_pairs = "Unique strong-cis pairs",
      n_cis_only_rows = "Reads_with_Both > 0 and Trans_Evidence == 0",
      n_mixed_cis_trans_rows = "Reads_with_Both > 0 and Trans_Evidence > 0",
      n_trans_only_rows = "Reads_with_Both == 0 and Trans_Evidence > 0",
      n_no_phaseable_rows = "Phaseable_Reads == 0",
      n_zero_total_reads_rows = "Total_Reads == 0"
    )
    mnp_validation_pretty <- mnp_validation_metrics_long %>%
      mutate(
        stage = dplyr::coalesce(metric_stage[metric], "Other"),
        metric_label = dplyr::coalesce(metric_labels[metric], metric),
        note = dplyr::coalesce(metric_notes[metric], ""),
        value = suppressWarnings(as.numeric(value))
      ) %>%
      arrange(factor(stage, levels = c("Candidate identification", "VAF/sample gate", "Read-check assignment", "Read-check results", "Read-check classification", "Read-check outcomes", "Other")))
    kable(
      mnp_validation_pretty %>% select(stage, metric_label, value, note),
      format.args = list(big.mark = ","),
      caption = "MNP validation metrics (mnp.validation.metrics.tsv)",
      col.names = c("Stage", "Metric", "Value", "Definition")
    )
  } else {
    kable(
      mnp_validation_metrics,
      format.args = list(big.mark = ","),
      caption = "MNP validation metrics (mnp.validation.metrics.tsv)"
    )
  }
} else {
  cat(
    "MNP validation metrics not found at: ", MNP_VALIDATION_METRICS,
    "\n(Add `data/mnp/mnp.validation.metrics.tsv` from your MNP pipeline, or set YAML param `mnp_validation_metrics` to the file path.)\n",
    sep = ""
  )
}
```

```{r postmnp-summary}
if (has_postmnp) {
  postmnp_total <- sum(postmnp_counts$COUNT[postmnp_counts$TYPE == "TOTAL"], na.rm = TRUE)
  postmnp_by_type <- postmnp_counts %>%
    filter(TYPE != "TOTAL") %>%
    group_by(TYPE) %>%
    summarise(COUNT = sum(COUNT, na.rm = TRUE), .groups = "drop")
  cat("- **Total variants (after MNP correction)**: ", format(postmnp_total, big.mark = ","), "\n", sep = "")
} else {
  postmnp_total <- 0
  postmnp_by_type <- data.frame(TYPE = character(0), COUNT = numeric(0))
  cat("Post-MNP per-chromosome variant-type counts are not available.\n")
  cat("- **Files present**: ", n_postmnp_files_present, " of 22 (`data/qc/reports/chr*.postmnp.variant_types.txt`).\n", sep = "")
  cat("- **Typical cause**: `count_variant_types_postmnp` was not run for this project, or the report was knitted before/with a different `data_dir` than the pipeline output.\n")
  cat("- **Fix**: Ensure the annotation workflow emits those 22 files (e.g. `annotation_files` includes `postmnp_stats`, or run `generate_qc_report` so Snakemake builds them as dependencies).\n")
}
```

```{r postmnp-table}
if (has_postmnp && nrow(postmnp_by_type) > 0) {
  kable(postmnp_by_type,
        format.args = list(big.mark = ","),
        caption = "Variant Counts by Type (After MNP Correction)",
        col.names = c("Type", "Count"))
}
```

```{r postmnp-plot, fig.height=5, fig.width=11}
if (has_postmnp && nrow(postmnp_by_type) > 0) {
  plot_type_count_panels(
    postmnp_by_type %>% dplyr::rename(TYPE = TYPE, COUNT = COUNT),
    y_col = "COUNT",
    title = "Variant types after MNP correction",
    subtitle = paste0("Total variants: ", format(postmnp_total, big.mark = ","))
  )
}
```

---

# 4. Sample Missingness

## Method

Sample missingness is calculated across all variants passing quality filters (HQHet + geno/maf/hwe). Samples with excessive missing genotypes are excluded.

- **Threshold**: F_MISS > `r MIND_THR` (`r MIND_THR * 100`%)

```{r missingness-summary}
total_after_sex <- passing_samples
failed_missingness <- nrow(miss_report)
passing_missingness <- total_after_sex - failed_missingness
```

## Results

- **Samples (Post Sex-Check)**: `r format(total_after_sex, big.mark = ",")`
- **Failed Missingness**: `r format(failed_missingness, big.mark = ",")` (`r round(100 * failed_missingness / total_after_sex, 2)`%)
- **Passing Missingness**: `r format(passing_missingness, big.mark = ",")` (`r round(100 * passing_missingness / total_after_sex, 2)`%)

```{r missingness-by-status}
if ("Status" %in% colnames(miss_report) && sum(!is.na(miss_report$Status)) > 0) {
  miss_fail_summary <- miss_report %>%
    group_by(Status) %>%
    summarise(
      N_Failed = n(),
      .groups = "drop"
    )
  
  kable(miss_fail_summary, caption = "Missingness Failures by Case/Control Status")
}
```

## Missingness Distribution

**WARNING**: If a large proportion of cases are failing, check the distribution below. The threshold may be too stringent or there may be a data quality issue.

A systematic shift (e.g., cases consistently higher than controls) typically indicates batch effects or alignment/capture differences rather than random missingness.

```{r missingness-distribution, fig.height=5}
# Load aggregated missingness to see all samples
agg_miss <- read.table(
  file.path(DATA_DIR, "plink", "aggregated_missingness.txt"),
  header = TRUE, stringsAsFactors = FALSE, comment.char = ""
)

# Status and Freeze from covariate file (same as reports)
agg_id <- if ("IID" %in% names(agg_miss)) trimws(agg_miss$IID) else trimws(agg_miss[,1])
agg_miss$Status <- if (length(status_lookup) > 0) as.character(status_lookup[agg_id]) else "Unknown"
agg_miss$Freeze <- if (length(freeze_lookup) > 0) as.character(freeze_lookup[agg_id]) else NA_character_

ggplot(agg_miss, aes(x = F_MISS, fill = Status)) +
  geom_histogram(bins = 50, alpha = 0.7) +
  geom_vline(xintercept = MIND_THR, linetype = "dashed", color = "red", linewidth = 1) +
  scale_x_continuous(labels = scales::percent) +
  labs(
    title = "Distribution of Sample Missingness",
    x = "Missingness Rate",
    y = "Count",
    fill = "Status",
    caption = paste0("Red line = ", MIND_THR * 100, "% threshold")
  ) +
  theme_minimal()
```

```{r missingness-stats}
# Summary statistics
miss_stats <- agg_miss %>%
  group_by(Status) %>%
  summarise(
    N = n(),
    Mean = mean(F_MISS),
    Median = median(F_MISS),
    Min = min(F_MISS),
    Max = max(F_MISS),
    Above_Threshold = sum(F_MISS > MIND_THR),
    Pct_Above = round(100 * Above_Threshold / N, 2),
    .groups = "drop"
  )

kable(miss_stats, 
      digits = 4,
      caption = "Missingness Statistics by Case/Control Status")
```

```{r missingness-stats-freeze}
if (exists("has_freeze") && has_freeze && "Freeze" %in% names(agg_miss) && sum(!is.na(agg_miss$Freeze)) > 0) {
  miss_stats_freeze <- agg_miss %>%
    filter(!is.na(Freeze)) %>%
    group_by(Freeze) %>%
    summarise(
      N = n(),
      Mean = mean(F_MISS),
      Median = median(F_MISS),
      Above_Threshold = sum(F_MISS > MIND_THR),
      Pct_Above = round(100 * Above_Threshold / N, 2),
      .groups = "drop"
    )
  kable(miss_stats_freeze, digits = 4, caption = "Missingness Statistics by Freeze")
}
```

### Failing Samples Details

```{r failing-samples-distribution}
# Show distribution of failing samples (sample ID column is IID in aggregated_missingness.txt)
id_col_miss <- if ("IID" %in% names(agg_miss)) "IID" else names(agg_miss)[1]
failing_samples <- agg_miss %>%
  filter(F_MISS > MIND_THR) %>%
  arrange(desc(F_MISS)) %>%
  mutate(
    Sample_ID = .data[[id_col_miss]],
    Missingness_Pct = round(F_MISS * 100, 2)
  )

if (nrow(failing_samples) > 0) {
  cat(paste0("\n**Total Failing Samples**: ", nrow(failing_samples), "\n\n"))
  
  # Distribution plot of failing samples
  ggplot(failing_samples, aes(x = Missingness_Pct, fill = Status)) +
    geom_histogram(bins = 30, alpha = 0.7) +
    labs(
      title = paste0("Distribution of Missingness Rates for Failing Samples (n=", nrow(failing_samples), ")"),
      x = "Missingness Rate (%)",
      y = "Count",
      fill = "Status"
    ) +
    theme_minimal()
} else {
  cat("No samples exceeded the missingness threshold.\n")
}
```

```{r worst-offenders-table}
# Show worst 20 offenders
if (nrow(failing_samples) > 0) {
  worst_offenders <- failing_samples %>%
    select(Sample_ID, Status, Missingness_Pct, MISSING_CT, OBS_CT) %>%
    head(20)
  
  kable(worst_offenders, 
        format.args = list(big.mark = ","),
        caption = paste0("Top ", min(nrow(failing_samples), 20), " Worst Offenders"))
}
```

```{r case-missingness-summary}
# Specifically summarize case missingness for sharing
case_miss <- agg_miss %>%
  filter(Status == "Case") %>%
  arrange(desc(F_MISS))

if (nrow(case_miss) > 0) {
  cat("\n### Case Missingness Summary\n\n")
  cat("- **Total Cases**: ", nrow(case_miss), "\n")
  cat("- **Cases Failing (>", MIND_THR * 100, "%)**: ", sum(case_miss$F_MISS > MIND_THR), " (", 
      round(100 * sum(case_miss$F_MISS > MIND_THR) / nrow(case_miss), 2), "%)\n", sep="")
  cat("- **Mean Missingness**: ", round(mean(case_miss$F_MISS) * 100, 2), "%\n", sep="")
  cat("- **Median Missingness**: ", round(median(case_miss$F_MISS) * 100, 2), "%\n", sep="")
  cat("- **Range**: ", round(min(case_miss$F_MISS) * 100, 2), "% - ", 
      round(max(case_miss$F_MISS) * 100, 2), "%\n\n", sep="")
}
```

---

# 5. LD pruning (merge_mind)

## Method

In `build_files.smk`, per-chromosome SNP filters produce `site-qc.var-qc`; sample missingness (MIND) is applied; chromosomes are merged to **`merge_mind`** (mind-passing samples only). **LD pruning runs on `merge_mind`**, not on per-chr files: `--indep-pairwise 200 20 0.2`, excluding chromosome 8 and long-LD regions (e.g. HLA, 17q21.31). Outputs include `merge_mind.prune.in` and `merge_mind.prune.out`.

This step does not remove variants from the final dataset. It only defines a subset of approximately independent SNPs for downstream statistical calculations (PCA and heterozygosity).

```{r ld-pruning-merge-mind}
ld_prune_summary <- data.frame(CHR = integer(0), Variants_Removed = numeric(0))
total_ld_pruned <- 0
pct_ld_pruned <- NA_real_
n_merge_snps <- NA_integer_

snplist_file <- file.path(DATA_DIR, "plink", "merge_mind.snplist")
prune_out_file <- file.path(DATA_DIR, "plink", "merge_mind.prune.out")
pvar_file <- file.path(DATA_DIR, "plink", "merge_mind.pvar")

if (file.exists(snplist_file)) {
  n_merge_snps <- length(readLines(snplist_file))
}
if (file.exists(prune_out_file) && file.exists(pvar_file)) {
  pruned_ids <- readLines(prune_out_file)
  pruned_ids <- pruned_ids[nzchar(pruned_ids)]
  if (length(pruned_ids) > 0) {
    pvar_preview <- readLines(pvar_file, n = 200)
    hdr_idx <- grep("^#?CHROM\\b", pvar_preview)[1]
    if (!is.na(hdr_idx)) {
      hdr_line <- pvar_preview[hdr_idx]
      hdr_line <- sub("^#", "", hdr_line)
      colnames_pvar <- strsplit(hdr_line, "\t")[[1]]
      chr_idx <- which(tolower(colnames_pvar) %in% c("chrom", "chr"))[1]
      if (is.na(chr_idx)) chr_idx <- 1L
      id_idx <- which(colnames_pvar == "ID")[1]
      if (is.na(id_idx)) id_idx <- 2L
      col_cl <- rep("NULL", length(colnames_pvar))
      col_cl[chr_idx] <- "character"
      col_cl[id_idx] <- "character"
      pvar_chr_id <- read.table(
        pvar_file,
        header = FALSE,
        sep = "\t",
        skip = hdr_idx,
        col.names = colnames_pvar,
        stringsAsFactors = FALSE,
        comment.char = "",
        colClasses = col_cl
      )
      names(pvar_chr_id) <- c("CHR", "ID")
      pvar_chr_id$CHR <- suppressWarnings(as.integer(pvar_chr_id$CHR))
      pruned_chr <- pvar_chr_id %>% filter(ID %in% pruned_ids)
      ld_prune_summary <- pruned_chr %>%
        group_by(CHR) %>%
        summarise(Variants_Removed = n(), .groups = "drop") %>%
        arrange(CHR)
      total_ld_pruned <- sum(ld_prune_summary$Variants_Removed, na.rm = TRUE)
      if (!is.na(n_merge_snps) && n_merge_snps > 0) {
        pct_ld_pruned <- round(100 * total_ld_pruned / n_merge_snps, 2)
      }
    }
  }
}
```

## Results

- **SNPs in merge_mind** (input to LD pruning): `r if (!is.na(n_merge_snps) && n_merge_snps > 0) format(n_merge_snps, big.mark = ",") else "N/A (merge_mind.snplist not found)"`
- **Variants in prune.out** (pairwise redundant / thinned): `r format(total_ld_pruned, big.mark = ",")` `r if (!is.na(pct_ld_pruned)) paste0("(", pct_ld_pruned, "% of merge_mind SNPs)") else ""`

```{r ld-pruning-plot, fig.height=5}
if (nrow(ld_prune_summary) > 0) {
  cap_suffix <- if (!is.na(n_merge_snps) && n_merge_snps > 0 && !is.na(pct_ld_pruned)) {
    paste0("(", pct_ld_pruned, "% of ", format(n_merge_snps, big.mark = ","), " merge_mind SNPs). ")
  } else {
    ""
  }
  print(ggplot(ld_prune_summary, aes(x = factor(CHR, levels = sort(unique(CHR))), y = Variants_Removed)) +
    geom_col(fill = "#8E44AD") +
    scale_y_continuous(labels = scales::comma) +
    labs(
      title = "Variants in prune.out (LD pruning on merge_mind)",
      subtitle = "After MIND merge; independent SNPs (prune.in) used for PCA and het — not part of Section 3 per-chr filters.",
      x = "Chromosome",
      y = "Variants (prune.out)",
      caption = paste0(
        "Total in prune.out: ", format(total_ld_pruned, big.mark = ","), ". ",
        cap_suffix,
        "build.pgen retains all QC-pass SNPs; prune.in restricts PCA/het only."
      )
    ) +
    theme_minimal())
} else {
  cat("No LD pruning breakdown: ensure data/plink/merge_mind.prune.out and merge_mind.pvar exist (build_files plink2_ld_prune_merge_mind).\n")
}
```

---

# 6. Heterozygosity

## Method

Heterozygosity is computed on the LD-pruned SNP subset to avoid inflation due to linkage disequilibrium.

Operationally, this is **`merge_mind.pruned`** (from **`merge_mind`** + `merge_mind.prune.in`), matching `build_files.smk`. Outliers (mean ± 3 SD on F) are flagged; **`build`** is **`merge_mind`** with those samples removed, while retaining the full SNP set for association testing.

```{r het-summary}
total_after_miss <- passing_missingness
failed_het <- nrow(het_report)
passing_het <- total_after_miss - failed_het
```

## Results

- **Samples (Post Missingness)**: `r format(total_after_miss, big.mark = ",")`
- **Failed Heterozygosity**: `r format(failed_het, big.mark = ",")` (`r round(100 * failed_het / total_after_miss, 2)`%)
- **Passing Heterozygosity**: `r format(passing_het, big.mark = ",")` (`r round(100 * passing_het / total_after_miss, 2)`%)

```{r het-by-status}
if ("Status" %in% colnames(het_report) && sum(!is.na(het_report$Status)) > 0) {
  het_fail_summary <- het_report %>%
    group_by(Status) %>%
    summarise(
      N_Failed = n(),
      .groups = "drop"
    )
  
  kable(het_fail_summary, caption = "Heterozygosity Failures by Case/Control Status")
}
```

```{r het-by-freeze}
if (exists("has_freeze") && has_freeze && "Freeze" %in% colnames(het_report) && sum(!is.na(het_report$Freeze)) > 0) {
  het_fail_freeze <- het_report %>%
    filter(!is.na(Freeze)) %>%
    group_by(Freeze) %>%
    summarise(N_Failed = n(), .groups = "drop")
  kable(het_fail_freeze, caption = "Heterozygosity Failures by Freeze")
}
```

---

```{r final-summary}
final_samples <- passing_het
final_variants <- variants_after_plink

# Count cases/controls at each stage (from Status in covariate file)
if ("Status" %in% colnames(sex_report) && sum(!is.na(sex_report$Status)) > 0) {
  start_cases <- sum(sex_report$Status == "Case", na.rm = TRUE)
  start_controls <- sum(sex_report$Status == "Control", na.rm = TRUE)
  postsex_cases <- sum(sex_pass$Status == "Case", na.rm = TRUE)
  postsex_controls <- sum(sex_pass$Status == "Control", na.rm = TRUE)
  miss_fail_cases <- sum(miss_report$Status == "Case", na.rm = TRUE)
  miss_fail_controls <- sum(miss_report$Status == "Control", na.rm = TRUE)
  postmiss_cases <- postsex_cases - miss_fail_cases
  postmiss_controls <- postsex_controls - miss_fail_controls
  het_fail_cases <- sum(het_report$Status == "Case", na.rm = TRUE)
  het_fail_controls <- sum(het_report$Status == "Control", na.rm = TRUE)
  final_cases <- postmiss_cases - het_fail_cases
  final_controls <- postmiss_controls - het_fail_controls
}
```

_Final sample count is shown after the PCA sections._

---

# 7. Population Structure (PCA)

PCA uses **`build`** samples and the LD-pruned SNP list (`merge_mind.prune.in`; chr8 / long-LD regions excluded). The build genotype set itself is not LD-pruned — pruning is only for PCA (and heterozygosity).

## PCA scatterplots and variance explained

```{r pca-summary}
# Calculate variance explained (build_files: build_pca.eigenval)
eigenval_file <- file.path(DATA_DIR, "preprocess", "build_pca.eigenval")
if (!file.exists(eigenval_file)) eigenval_file <- file.path(DATA_DIR, "preprocess", "build.eigenval")
if (!file.exists(eigenval_file)) stop("Required file not found: build_pca.eigenval or build.eigenval in data/preprocess/")
eigenval <- read.table(eigenval_file, header = FALSE)
colnames(eigenval) <- "Eigenvalue"
eigenval$PC <- paste0("PC", seq_len(nrow(eigenval)))
eigenval$Variance_Explained <- 100 * eigenval$Eigenvalue / sum(eigenval$Eigenvalue)
```

### Variance explained

```{r variance-explained, fig.height=5}
ggplot(eigenval[1:10, ], aes(x = factor(PC, levels = PC), y = Variance_Explained)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  geom_text(aes(label = sprintf("%.2f%%", Variance_Explained)), vjust = -0.5) +
  labs(
    title = "Variance Explained by Principal Components",
    x = "Principal Component",
    y = "Variance Explained (%)"
  ) +
  theme_minimal()
```

### PC1 vs PC2

```{r pca-plot-pc1-pc2, fig.height=6}
ggplot(pca, aes(x = PC1, y = PC2, color = Status)) +
  geom_point(alpha = 0.55, size = 1.35, stroke = 0) +
  labs(
    title = "PC1 vs PC2 (case / control)",
    x = paste0("PC1 (", round(eigenval$Variance_Explained[1], 2), "%)"),
    y = paste0("PC2 (", round(eigenval$Variance_Explained[2], 2), "%)"),
    color = "Status"
  ) +
  theme_minimal(base_size = 13) +
  pca_legend_guides +
  pca_legend_theme +
  theme(plot.title = element_text(face = "bold"))
```

```{r pca-plot-pc1-pc2-freeze, fig.height=6}
if (exists("has_freeze") && has_freeze && "Freeze" %in% names(pca) && sum(!is.na(pca$Freeze)) > 0) {
  ggplot(pca %>% filter(!is.na(Freeze)), aes(x = PC1, y = PC2, color = factor(Freeze))) +
    geom_point(alpha = 0.55, size = 1.35, stroke = 0) +
    labs(
      title = "PC1 vs PC2 by Freeze",
      x = paste0("PC1 (", round(eigenval$Variance_Explained[1], 2), "%)"),
      y = paste0("PC2 (", round(eigenval$Variance_Explained[2], 2), "%)"),
      color = "Freeze"
    ) +
    theme_minimal(base_size = 13) +
    pca_legend_guides +
    pca_legend_theme +
    theme(plot.title = element_text(face = "bold"))
}
```

### PC1 vs PC2 by ancestry group

```{r pca-plot-pc1-pc2-ancgroup, fig.height=6.5, fig.width=9}
if (exists("has_anc_group") && has_anc_group) {
  pca_ag <- pca %>% dplyr::filter(!is.na(AncGroup), nzchar(trimws(as.character(AncGroup))))
  if (nrow(pca_ag) > 0) {
    print(
      ggplot(pca_ag, aes(x = PC1, y = PC2, color = factor(AncGroup))) +
        geom_point(alpha = 0.65, size = 1.5, stroke = 0) +
        labs(
          title = "PC1 vs PC2 (colored by AncGroup)",
          x = paste0("PC1 (", round(eigenval$Variance_Explained[1], 2), "%)"),
          y = paste0("PC2 (", round(eigenval$Variance_Explained[2], 2), "%)"),
          color = "AncGroup"
        ) +
        theme_minimal(base_size = 13) +
        pca_legend_guides +
        pca_legend_theme +
        theme(plot.title = element_text(face = "bold"))
    )
  }
} else {
  cat(
    "_**No AncGroup plot.** ",
    if (exists("anc_group_user_msg") && nzchar(anc_group_user_msg)) anc_group_user_msg else "See Load Data / console for ancestry file path and IID matching.",
    "_\n\n",
    sep = ""
  )
}
```

### PC3 vs PC4

```{r pca-plot-pc3-pc4, fig.height=6}
ggplot(pca, aes(x = PC3, y = PC4, color = Status)) +
  geom_point(alpha = 0.55, size = 1.35, stroke = 0) +
  labs(
    title = "PC3 vs PC4 (case / control)",
    x = paste0("PC3 (", round(eigenval$Variance_Explained[3], 2), "%)"),
    y = paste0("PC4 (", round(eigenval$Variance_Explained[4], 2), "%)"),
    color = "Status"
  ) +
  theme_minimal(base_size = 13) +
  pca_legend_guides +
  pca_legend_theme +
  theme(plot.title = element_text(face = "bold"))
```

```{r pca-plot-pc3-pc4-freeze, fig.height=6}
if (exists("has_freeze") && has_freeze && "Freeze" %in% names(pca) && sum(!is.na(pca$Freeze)) > 0) {
  ggplot(pca %>% filter(!is.na(Freeze)), aes(x = PC3, y = PC4, color = factor(Freeze))) +
    geom_point(alpha = 0.55, size = 1.35, stroke = 0) +
    labs(
      title = "PC3 vs PC4 by Freeze",
      x = paste0("PC3 (", round(eigenval$Variance_Explained[3], 2), "%)"),
      y = paste0("PC4 (", round(eigenval$Variance_Explained[4], 2), "%)"),
      color = "Freeze"
    ) +
    theme_minimal(base_size = 13) +
    pca_legend_guides +
    pca_legend_theme +
    theme(plot.title = element_text(face = "bold"))
}
```

### PC3 vs PC4 by ancestry group

```{r pca-plot-pc3-pc4-ancgroup, fig.height=6.5, fig.width=9}
if (exists("has_anc_group") && has_anc_group) {
  pca_ag <- pca %>% dplyr::filter(!is.na(AncGroup), nzchar(trimws(as.character(AncGroup))))
  if (nrow(pca_ag) > 0) {
    print(
      ggplot(pca_ag, aes(x = PC3, y = PC4, color = factor(AncGroup))) +
        geom_point(alpha = 0.65, size = 1.5, stroke = 0) +
        labs(
          title = "PC3 vs PC4 (colored by AncGroup)",
          x = paste0("PC3 (", round(eigenval$Variance_Explained[3], 2), "%)"),
          y = paste0("PC4 (", round(eigenval$Variance_Explained[4], 2), "%)"),
          color = "AncGroup"
        ) +
        theme_minimal(base_size = 13) +
        pca_legend_guides +
        pca_legend_theme +
        theme(plot.title = element_text(face = "bold"))
    )
  }
}
```

### Scree plot (all 20 PCs)

```{r scree-plot, fig.height=5}
ggplot(eigenval, aes(x = seq_len(nrow(eigenval)), y = Eigenvalue)) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_point(color = "steelblue", size = 2) +
  labs(
    title = "Scree Plot: Eigenvalues of Principal Components",
    x = "Principal Component",
    y = "Eigenvalue"
  ) +
  theme_minimal()
```

## Allele PC loadings (diagnostic for PCA structure)

These plots show which genomic regions drive PC structure. The dashed line marks the 99.9th percentile of absolute loading, used as a diagnostic cutoff to highlight the top 0.1% of variants for each PC.

```{r pca-loading-prep, include=FALSE}
read_plink_skip_meta <- function(path) {
  if (!file.exists(path)) return(data.frame())
  lines <- readLines(path, warn = FALSE)
  n_skip <- sum(grepl("^##", lines))
  d <- read.delim(path, skip = n_skip, sep = "\t", header = TRUE, stringsAsFactors = FALSE, check.names = FALSE, comment.char = "")
  names(d) <- sub("^#", "", names(d))
  d
}

get_loading_regions <- function(df, abs_col, thr, gap_bp = 1e6) {
  df %>%
    dplyr::filter(.data[[abs_col]] >= thr) %>%
    dplyr::arrange(CHR, POS) %>%
    dplyr::group_by(CHR) %>%
    dplyr::mutate(region_id = cumsum(c(1, diff(POS) > gap_bp))) %>%
    dplyr::group_by(CHR, region_id) %>%
    dplyr::summarise(
      start = min(POS),
      end = max(POS),
      n_snps = dplyr::n(),
      max_loading = max(.data[[abs_col]], na.rm = TRUE),
      top_pos = POS[which.max(.data[[abs_col]])],
      top_id = ID[which.max(.data[[abs_col]])],
      .groups = "drop"
    ) %>%
    dplyr::arrange(dplyr::desc(max_loading), dplyr::desc(n_snps), CHR, start)
}

pca_loading_ready <- FALSE
pca_loading_msg <- ""
allele_path <- file.path(DATA_DIR, "preprocess", "build_pca.eigenvec.allele")
pvar_path <- file.path(DATA_DIR, "preprocess", "build.pvar")

if (!file.exists(allele_path)) {
  pca_loading_msg <- paste0("**Allele loadings** missing `", allele_path, "` (e.g. **`--pca 20 allele-wts`** in **`calculate_pca`**).")
} else if (!file.exists(pvar_path)) {
  pca_loading_msg <- paste0("**Positions** missing `", pvar_path, "`; need **POS** and **ID** from **build.pvar**.")
} else {
  al <- read_plink_skip_meta(allele_path)
  pv <- read_plink_skip_meta(pvar_path)
  if (nrow(al) == 0) {
    pca_loading_msg <- "Allele loading file is empty."
  } else if (!"ID" %in% names(al) || !all(c("PC1", "PC2") %in% names(al))) {
    pca_loading_msg <- "Allele file must include columns **ID**, **PC1**, **PC2**."
  } else if (!"ID" %in% names(pv) || !"POS" %in% names(pv)) {
    pca_loading_msg <- "**build.pvar** must include **ID** and **POS**."
  } else {
    al1 <- al %>% dplyr::distinct(ID, .keep_all = TRUE)
    pos_tbl <- pv %>%
      dplyr::distinct(ID, .keep_all = TRUE) %>%
      dplyr::select(ID, POS)
    m <- dplyr::left_join(al1, pos_tbl, by = "ID")
    if (!"CHROM" %in% names(m) && "CHROM" %in% names(pv)) {
      ch <- pv %>% dplyr::distinct(ID, .keep_all = TRUE) %>% dplyr::select(ID, CHROM)
      m <- dplyr::left_join(m, ch, by = "ID")
    }
    if (!"CHROM" %in% names(m)) {
      pca_loading_msg <- "Could not find **CHROM** in allele file or **build.pvar**."
    } else {
      pv_df <- m %>%
        dplyr::transmute(
          ID = trimws(as.character(ID)),
          CHR = suppressWarnings(as.integer(gsub("^chr", "", as.character(CHROM)))),
          POS = suppressWarnings(as.numeric(POS)),
          PC1 = suppressWarnings(as.numeric(PC1)),
          PC2 = suppressWarnings(as.numeric(PC2))
        ) %>%
        dplyr::filter(!is.na(POS), is.finite(POS), is.finite(PC1), is.finite(PC2), !is.na(CHR)) %>%
        dplyr::mutate(
          abs_PC1 = abs(PC1),
          abs_PC2 = abs(PC2)
        )

      if (nrow(pv_df) == 0) {
        pca_loading_msg <- "No rows after joining allele loadings to **POS** (check **ID** matches between files)."
      } else {
        chr_order <- sort(unique(pv_df$CHR))
        gap_chr <- 8e7
        chr_bounds <- pv_df %>%
          dplyr::group_by(CHR) %>%
          dplyr::summarise(chr_max = max(POS, na.rm = TRUE), .groups = "drop") %>%
          dplyr::arrange(match(CHR, chr_order)) %>%
          dplyr::mutate(chr_start = cumsum(c(0, head(chr_max + gap_chr, -1))))

        pv2 <- pv_df %>%
          dplyr::left_join(dplyr::select(chr_bounds, CHR, chr_start), by = "CHR") %>%
          dplyr::mutate(
            xbp = chr_start + POS,
            chr_alt = as.character(as.integer(factor(CHR, levels = chr_order)) %% 2L)
          )

        chr_centers <- chr_bounds %>%
          dplyr::mutate(mid = chr_start + chr_max / 2) %>%
          dplyr::arrange(match(CHR, chr_order))

        x_scale <- scale_x_continuous(
          breaks = chr_centers$mid,
          labels = as.character(chr_centers$CHR),
          expand = expansion(mult = 0.02)
        )

        thr1 <- quantile(pv2$abs_PC1, 0.999, na.rm = TRUE)
        thr2 <- quantile(pv2$abs_PC2, 0.999, na.rm = TRUE)

        pc1_regions <- get_loading_regions(pv2, "abs_PC1", thr1)
        pc2_regions <- get_loading_regions(pv2, "abs_PC2", thr2)

        pca_loading_ready <- TRUE
      }
    }
  }
}
```

```{r pca-loading-notice, echo=FALSE, results="asis"}
if (!pca_loading_ready && nzchar(pca_loading_msg)) {
  knitr::asis_output(paste0("\n\n_", pca_loading_msg, "_\n\n"))
}
```

### Genomic drivers of PC1

```{r pca-loading-plot-pc1, echo=FALSE, fig.width=11, fig.height=4}
if (pca_loading_ready) {
  print(
    ggplot(pv2, aes(x = xbp, y = abs_PC1)) +
      geom_point(aes(color = chr_alt), size = 0.18, alpha = 0.35) +
      scale_color_manual(values = c("0" = "gray35", "1" = "gray70"), guide = "none") +
      geom_hline(yintercept = thr1, linetype = "dashed", color = "red") +
      x_scale +
      labs(
        title = "Genomic drivers of PC1",
        subtitle = paste0(
          "Dashed line = 99.9th percentile of |PC1| = ",
          signif(thr1, 5),
          " (top 0.1% of variants)"
        ),
        x = "Chromosome",
        y = "|PC1 loading|"
      ) +
      theme_minimal()
  )
}
```

Regions are defined by grouping above-cutoff variants that lie within 1 Mb of each other on the same chromosome.

### PC1 regions above loading cutoff

```{r pca-loading-table-pc1, echo=FALSE}
if (pca_loading_ready) {
  knitr::kable(
    head(pc1_regions[, c("CHR", "start", "end", "n_snps", "max_loading", "top_pos", "top_id")], 20),
    digits = 4,
    caption = "PC1 regions above loading cutoff"
  )
}
```

### Genomic drivers of PC2

```{r pca-loading-plot-pc2, echo=FALSE, fig.width=11, fig.height=4}
if (pca_loading_ready) {
  print(
    ggplot(pv2, aes(x = xbp, y = abs_PC2)) +
      geom_point(aes(color = chr_alt), size = 0.18, alpha = 0.35) +
      scale_color_manual(values = c("0" = "gray35", "1" = "gray70"), guide = "none") +
      geom_hline(yintercept = thr2, linetype = "dashed", color = "red") +
      x_scale +
      labs(
        title = "Genomic drivers of PC2",
        subtitle = paste0(
          "Dashed line = 99.9th percentile of |PC2| = ",
          signif(thr2, 5),
          " (top 0.1% of variants)"
        ),
        x = "Chromosome",
        y = "|PC2 loading|"
      ) +
      theme_minimal()
  )
}
```

### PC2 regions above loading cutoff

```{r pca-loading-table-pc2, echo=FALSE}
if (pca_loading_ready) {
  knitr::kable(
    head(pc2_regions[, c("CHR", "start", "end", "n_snps", "max_loading", "top_pos", "top_id")], 20),
    digits = 4,
    caption = "PC2 regions above loading cutoff"
  )
}
```

# 8. Final Sample Count

## Summary

- **Starting Samples**: `r format(total_samples, big.mark = ",")` (`r if(exists("start_cases")) paste0(format(start_cases, big.mark=","), " cases, ", format(start_controls, big.mark=","), " controls")`)
- **Final Samples**: `r format(final_samples, big.mark = ",")` (`r if(exists("final_cases")) paste0(format(final_cases, big.mark=","), " cases, ", format(final_controls, big.mark=","), " controls")`) (genotype build after het QC)
- **Total Excluded**: `r format(total_samples - final_samples, big.mark = ",")` (`r if(exists("start_cases")) paste0(format(start_cases - final_cases, big.mark=","), " cases, ", format(start_controls - final_controls, big.mark=","), " controls")`)

### Samples Excluded by Stage

```{r exclusion-by-stage}
if (exists("start_cases")) {
  stg <- c("Sex Check", "Missingness", "Heterozygosity")
  cex <- c(start_cases - postsex_cases, miss_fail_cases, het_fail_cases)
  ctl <- c(start_controls - postsex_controls, miss_fail_controls, het_fail_controls)
  exclusion_breakdown <- data.frame(Stage = stg, Cases_Excluded = cex, Controls_Excluded = ctl) %>%
    mutate(
      Total_Excluded = Cases_Excluded + Controls_Excluded,
      Pct_Cases = ifelse(Total_Excluded > 0, round(100 * Cases_Excluded / Total_Excluded, 1), NA_real_)
    )

  kable(exclusion_breakdown,
        caption = "Sample Exclusions by QC Stage and Case/Control Status",
        col.names = c("QC Stage", "Cases Excluded", "Controls Excluded", "Total Excluded", "% Cases"))
}
```

- **Starting Variants**: `r format(prefilter_total, big.mark = ",")`
- **Final Variants**: `r format(final_variants, big.mark = ",")` `r if (prefilter_total > 0) paste0("(", round(100 * final_variants / prefilter_total, 2), "% retention)") else "(N/A: no prefilter counts)"`
- **Total Excluded**: `r format(prefilter_total - final_variants, big.mark = ",")`

```{r exclusion-cascade}
exclusion_data <- data.frame(
  Stage = c("Starting", "Sex Check", "Missingness", "Heterozygosity", "Final"),
  Samples = c(total_samples, passing_samples, passing_missingness, passing_het, final_samples)
)

ggplot(exclusion_data, aes(x = factor(Stage, levels = Stage), y = Samples)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  geom_text(aes(label = format(Samples, big.mark = ",")), vjust = -0.5) +
  labs(
    title = "Sample Retention Through QC Pipeline",
    x = "QC Stage",
    y = "Number of Samples"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Final Dataset Definition

The final analysis dataset (**`build`**) includes:

- Samples passing sex check, missingness, and heterozygosity filters
- All SNPs passing variant QC (no LD pruning applied)

---

# 9. Session Info

```{r session-info}
sessionInfo()
```
