---
title: "ExWAS Quality Control Report"
author: "Automated QC Pipeline"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    code_folding: hide
params:
  project: "PROJECT"
  controls_file: "controls.txt"
  data_dir: "data"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.width = 10, fig.height = 6)
library(tidyverse)
library(knitr)
library(yaml)

# Read config file
config <- read_yaml("config.yml")

# Project settings (params override config, for snakemake compatibility)
PROJECT <- if (!is.null(params$project)) params$project else config$project$name
CONTROLS_FILE <- if (!is.null(params$controls_file)) params$controls_file else config$input$controls
DATA_DIR <- if (!is.null(params$data_dir)) params$data_dir else "data"

# QC thresholds - always from config
MIND_THR <- as.numeric(config$qc$mind_thr)
MAF_THR <- as.numeric(config$qc$maf_thr)
GENO_THR <- as.numeric(config$qc$geno_thr)
HWE_THR <- as.numeric(config$qc$hwe_thr)
```

# Overview

This report summarizes quality control performed on whole exome sequencing data for the **`r PROJECT`** study.

## QC Pipeline Steps

1. **Sex Check**: Remove samples with ambiguous sex (chrX F-statistic)
2. **Variant Quality Filter**: Filter variants based on high-quality heterozygotes (HQHet)
3. **Variant-Level Filters**: Apply genotype missingness, MAF, and HWE filters
4. **Sample Missingness**: Remove samples with excessive missing genotypes (calculated on filtered variants)
5. **Heterozygosity**: Remove samples with outlier heterozygosity rates (calculated on filtered variants)
6. **Population Structure**: Principal component analysis

---

# Load Data

```{r load-data}
# Load controls file and create phenotype status
if (file.exists(CONTROLS_FILE)) {
  controls <- trimws(read.table(CONTROLS_FILE, header = FALSE, stringsAsFactors = FALSE, colClasses = "character")[[1]])
} else {
  warning(paste("Controls file not found:", CONTROLS_FILE))
  controls <- character(0)
}

# Load sex inference results
sex_file <- file.path(DATA_DIR, "qc/reports", paste0(PROJECT, ".sex_inference.txt"))
if (!file.exists(sex_file)) stop(paste("Required file not found:", sex_file))
sex_report <- read.table(sex_file, header = TRUE, stringsAsFactors = FALSE, comment.char = "")

# Load missingness report
miss_file <- file.path(DATA_DIR, "qc/reports", paste0(PROJECT, ".missingness_report.txt"))
if (!file.exists(miss_file)) stop(paste("Required file not found:", miss_file))
miss_report <- read.csv(miss_file, stringsAsFactors = FALSE)

# Load heterozygosity report
het_file <- file.path(DATA_DIR, "qc/reports", paste0(PROJECT, ".heterozygosity_report.txt"))
if (!file.exists(het_file)) stop(paste("Required file not found:", het_file))
het_report <- read.csv(het_file, stringsAsFactors = FALSE)

# Load PCA results
pca_file <- file.path(DATA_DIR, "preprocess", paste0(PROJECT, ".build.eigenvec"))
if (!file.exists(pca_file)) stop(paste("Required file not found:", pca_file))
pca <- read.table(pca_file, header = TRUE, stringsAsFactors = FALSE, comment.char = "")

# Load variant type counts (prefilter)
prefilter_counts <- lapply(1:22, function(chr) {
  f <- file.path(DATA_DIR, "qc/reports", paste0(PROJECT, ".chr", chr, ".prefilter.variant_types.txt"))
  if (file.exists(f)) {
    read.table(f, header = TRUE, stringsAsFactors = FALSE, comment.char = "") %>%
      mutate(CHR = chr)
  } else {
    warning(paste("File not found:", f))
    data.frame(TYPE = character(0), COUNT = numeric(0), CHR = integer(0))
  }
}) %>% bind_rows()

# Load variant type counts (postfilter)
postfilter_counts <- lapply(1:22, function(chr) {
  f <- file.path(DATA_DIR, "qc/reports", paste0(PROJECT, ".chr", chr, ".postfilter.variant_types.txt"))
  if (file.exists(f)) {
    read.table(f, header = TRUE, stringsAsFactors = FALSE, comment.char = "") %>%
      mutate(CHR = chr)
  } else {
    warning(paste("File not found:", f))
    data.frame(TYPE = character(0), COUNT = numeric(0), CHR = integer(0))
  }
}) %>% bind_rows()

# Load variant counts after plink filtering
postplink_counts <- lapply(1:22, function(chr) {
  f <- file.path(DATA_DIR, "qc/reports", paste0(PROJECT, ".chr", chr, ".postplink.variant_count.txt"))
  if (file.exists(f)) {
    read.table(f, header = TRUE, stringsAsFactors = FALSE, comment.char = "")
  } else {
    warning(paste("File not found:", f))
    data.frame(VARIANTS = numeric(0))
  }
}) %>% bind_rows()

# Load plink filter metrics
plink_filter_metrics <- lapply(1:22, function(chr) {
  f <- file.path(DATA_DIR, "qc/reports", paste0(PROJECT, ".chr", chr, ".plink_filter_metrics.txt"))
  if (file.exists(f)) {
    read.table(f, header = TRUE, stringsAsFactors = FALSE, comment.char = "")
  } else {
    warning(paste("File not found:", f))
    data.frame(FILTER = character(0), VARIANTS_REMOVED = numeric(0))
  }
}) %>% bind_rows()

# Load variant QC metrics for distribution plots
variant_maf <- lapply(1:22, function(chr) {
  f <- file.path(DATA_DIR, "plink", paste0(PROJECT, ".chr", chr, ".afreq"))
  if (file.exists(f)) {
    read.table(f, header = TRUE, stringsAsFactors = FALSE, comment.char = "") %>%
      mutate(CHR = chr)
  } else {
    warning(paste("File not found:", f))
    data.frame(CHR = chr)
  }
}) %>% bind_rows()

variant_miss <- lapply(1:22, function(chr) {
  f <- file.path(DATA_DIR, "plink", paste0(PROJECT, ".chr", chr, ".vmiss"))
  if (file.exists(f)) {
    read.table(f, header = TRUE, stringsAsFactors = FALSE, comment.char = "") %>%
      mutate(CHR = chr)
  } else {
    warning(paste("File not found:", f))
    data.frame(CHR = chr)
  }
}) %>% bind_rows()

variant_hwe <- lapply(1:22, function(chr) {
  f <- file.path(DATA_DIR, "plink", paste0(PROJECT, ".chr", chr, ".hardy"))
  if (file.exists(f)) {
    read.table(f, header = TRUE, stringsAsFactors = FALSE, comment.char = "") %>%
      mutate(CHR = chr)
  } else {
    warning(paste("File not found:", f))
    data.frame(CHR = chr)
  }
}) %>% bind_rows()

# Rename sex inference Status column to Sex first
colnames(sex_report)[colnames(sex_report) == "Status"] <- "Sex"

# Add phenotype status based on controls
if (!is.null(controls) && length(controls) > 0) {
  sex_report$Status <- ifelse(trimws(sex_report[,1]) %in% controls, "Control", "Case")
  miss_report$Status <- ifelse(trimws(miss_report[,1]) %in% controls, "Control", "Case")
  het_report$Status <- ifelse(trimws(het_report[,1]) %in% controls, "Control", "Case")
  pca$Status <- ifelse(trimws(pca[,1]) %in% controls, "Control", "Case")
} else {
  sex_report$Status <- rep(NA_character_, nrow(sex_report))
  miss_report$Status <- rep(NA_character_, nrow(miss_report))
  het_report$Status <- rep(NA_character_, nrow(het_report))
  pca$Status <- rep(NA_character_, nrow(pca))
}
```

---

# 1. Sex Check (chrX F-statistic)

## Method

Sex is inferred from chromosome X heterozygosity using the F-statistic:

- **F > 0.8**: Male (mostly homozygous/hemizygous on X)
- **F < 0.2**: Female (heterozygous on X)
- **0.2 ≤ F ≤ 0.8**: Ambiguous/Problem (excluded)

```{r sex-check-summary}
sex_fail <- sex_report %>% filter(Sex == "AMBIGUOUS")
sex_pass <- sex_report %>% filter(Sex != "AMBIGUOUS")

total_samples <- nrow(sex_report)
failed_samples <- nrow(sex_fail)
passing_samples <- nrow(sex_pass)
```

## Results

- **Total Samples**: `r format(total_samples, big.mark = ",")`
- **Failed Sex Check**: `r format(failed_samples, big.mark = ",")` (`r round(100 * failed_samples / total_samples, 2)`%)
- **Passing Sex Check**: `r format(passing_samples, big.mark = ",")` (`r round(100 * passing_samples / total_samples, 2)`%)

```{r sex-check-by-status, eval=!is.null(controls)}
if (!is.null(controls) && "Status" %in% colnames(sex_fail)) {
  sex_fail_summary <- sex_fail %>%
    group_by(Status) %>%
    summarise(
      N_Failed = n(),
      .groups = "drop"
    )
  
  kable(sex_fail_summary, caption = "Sex Check Failures by Case/Control Status")
}
```

## F-statistic Distribution

```{r sex-f-distribution, fig.height=5}
ggplot(sex_report, aes(x = F_sex, fill = Sex)) +
  geom_histogram(bins = 50, alpha = 0.7) +
  geom_vline(xintercept = c(0.2, 0.8), linetype = "dashed", color = "red") +
  scale_fill_manual(values = c("MALE" = "steelblue", "FEMALE" = "pink", "AMBIGUOUS" = "orange")) +
  labs(
    title = "Distribution of chrX Homozygosity (Sex Inference)",
    x = "F-statistic (O(HOM)/OBS_CT)",
    y = "Count",
    fill = "Inferred Sex"
  ) +
  theme_minimal()
```

---

# 2. Variant Quality Filtering

## Method

Variants are filtered based on the presence of high-quality heterozygotes:

- **Filter**: `COUNT(GT="0/1" && DP≥10 && GQ≥20 && VAF>0.2) = 0`
- **Action**: Tag as `NoHQHet`, then exclude non-PASS variants

Variants are also filtered based on low call rate:
- **Filter**: `INFO/F_MISSING > 0.05`
- **Action**: Tag as `LowCallRate`, then exclude non-PASS variants

```{r variant-filtering}
if (nrow(prefilter_counts) > 0 && "TYPE" %in% names(prefilter_counts) && "COUNT" %in% names(prefilter_counts)) {
  prefilter_total <- sum(prefilter_counts$COUNT[prefilter_counts$TYPE == "TOTAL"], na.rm = TRUE)
} else {
  prefilter_total <- 0
}
if (nrow(postfilter_counts) > 0 && "TYPE" %in% names(postfilter_counts) && "COUNT" %in% names(postfilter_counts)) {
  postfilter_total <- sum(postfilter_counts$COUNT[postfilter_counts$TYPE == "TOTAL"], na.rm = TRUE)
} else {
  postfilter_total <- 0
}
variants_removed <- prefilter_total - postfilter_total
```

## Results

- **Variants (Pre-filter)**: `r format(prefilter_total, big.mark = ",")`
- **Variants Removed**: `r format(variants_removed, big.mark = ",")` (`r round(100 * variants_removed / prefilter_total, 2)`%)
- **Variants (Post-filter)**: `r format(postfilter_total, big.mark = ",")` (`r round(100 * postfilter_total / prefilter_total, 2)`%)

### Variant Types Before and After Filtering

```{r variant-types-table}
variant_summary <- prefilter_counts %>%
  filter(TYPE != "TOTAL") %>%
  group_by(TYPE) %>%
  summarise(Prefilter = sum(COUNT), .groups = "drop") %>%
  left_join(
    postfilter_counts %>%
      filter(TYPE != "TOTAL") %>%
      group_by(TYPE) %>%
      summarise(Postfilter = sum(COUNT), .groups = "drop"),
    by = "TYPE"
  ) %>%
  mutate(
    Removed = Prefilter - Postfilter,
    Pct_Removed = round(100 * Removed / Prefilter, 2)
  )

kable(variant_summary, 
      format.args = list(big.mark = ","),
      caption = "Variant Counts by Type")
```

```{r variant-types-plot, fig.height=5}
variant_summary_long <- variant_summary %>%
  select(TYPE, Prefilter, Postfilter) %>%
  pivot_longer(-TYPE, names_to = "Filter_Stage", values_to = "Count")

ggplot(variant_summary_long, aes(x = TYPE, y = Count, fill = Filter_Stage)) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_fill_manual(values = c("Prefilter" = "#E74C3C", "Postfilter" = "#3498DB")) +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title = "Variant Counts Before and After Quality Filtering",
    x = "Variant Type",
    y = "Count",
    fill = "Stage"
  ) +
  theme_minimal()
```

---

# 3. Variant-Level Filters (Plink)

## Method

Additional variant filters applied using Plink2 **to SNPs only** before sample QC:

- **Genotype Missingness**: `--geno `r GENO_THR`` (`r GENO_THR * 100`%)
- **Minor Allele Frequency**: `--maf `r MAF_THR`` (`r MAF_THR * 100`%)
- **Hardy-Weinberg Equilibrium**: `--hwe `r HWE_THR` midp`
- **LD Pruning**: `--indep-pairwise 200 20 0.2` (window 200 SNPs, step 20, r² threshold 0.2)

These filters are applied to SNPs passing the HQHet filter. Sample missingness and heterozygosity are then calculated on these filtered SNPs.

```{r plink-filtering}
# Calculate totals
variants_before_plink <- ifelse(length(postfilter_total) > 0 && !is.na(postfilter_total), postfilter_total, 0)
variants_after_plink <- ifelse(nrow(postplink_counts) > 0 && "VARIANTS" %in% names(postplink_counts), sum(postplink_counts$VARIANTS, na.rm = TRUE), 0)
variants_removed_plink <- variants_before_plink - variants_after_plink

# Count SNPs vs non-SNPs from postfilter variant types
if (nrow(postfilter_counts) > 0 && "TYPE" %in% names(postfilter_counts) && "COUNT" %in% names(postfilter_counts)) {
  snp_counts <- postfilter_counts %>% filter(TYPE == "SNP")
  indel_counts <- postfilter_counts %>% filter(TYPE == "INDEL")
  other_counts <- postfilter_counts %>% filter(TYPE %in% c("MNP", "OTHER"))
  
  total_snps <- sum(snp_counts$COUNT, na.rm = TRUE)
  total_indels <- sum(indel_counts$COUNT, na.rm = TRUE)
  total_other <- sum(other_counts$COUNT, na.rm = TRUE)
} else {
  total_snps <- 0
  total_indels <- 0
  total_other <- 0
}
total_non_snps <- total_indels + total_other

# Aggregate filter metrics
if (nrow(plink_filter_metrics) > 0 && "FILTER" %in% names(plink_filter_metrics) && "VARIANTS_REMOVED" %in% names(plink_filter_metrics)) {
  filter_summary <- plink_filter_metrics %>%
    group_by(FILTER) %>%
    summarise(Variants_Removed = sum(VARIANTS_REMOVED, na.rm = TRUE), .groups = "drop") %>%
    mutate(
      Pct_of_Input = ifelse(variants_before_plink > 0, round(100 * Variants_Removed / variants_before_plink, 2), 0)
    )

  ld_prune_summary <- plink_filter_metrics %>%
    filter(FILTER == "PRUNE") %>%
    mutate(CHR = factor(CHR, levels = 1:22)) %>%
    group_by(CHR) %>%
    summarise(
      Variants_Removed = sum(VARIANTS_REMOVED, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    arrange(as.numeric(as.character(CHR)))

  total_ld_pruned <- sum(ld_prune_summary$Variants_Removed, na.rm = TRUE)
  pct_ld_pruned <- ifelse(variants_before_plink > 0, round(100 * total_ld_pruned / variants_before_plink, 2), 0)
} else {
  filter_summary <- data.frame(FILTER = character(0), Variants_Removed = numeric(0), Pct_of_Input = numeric(0))
  ld_prune_summary <- data.frame(CHR = integer(0), Variants_Removed = numeric(0))
  total_ld_pruned <- 0
  pct_ld_pruned <- 0
}
```

## Results

- **Variants (Before Plink Filters)**: `r format(variants_before_plink, big.mark = ",")` (after HQHet)
  - SNPs: `r format(total_snps, big.mark = ",")` (`r round(100 * total_snps / variants_before_plink, 1)`%)
  - Indels: `r format(total_indels, big.mark = ",")` (`r round(100 * total_indels / variants_before_plink, 1)`%)
  - Other (MNPs, etc.): `r format(total_other, big.mark = ",")` (`r round(100 * total_other / variants_before_plink, 1)`%)
- **Non-SNPs Removed** (--snps-only): `r format(total_non_snps, big.mark = ",")` (`r round(100 * total_non_snps / variants_before_plink, 2)`%)
- **SNPs Removed by QC Filters**: `r format(variants_removed_plink - total_non_snps, big.mark = ",")` (`r round(100 * (variants_removed_plink - total_non_snps) / variants_before_plink, 2)`%)
- **Variants (After All Filters)**: `r format(variants_after_plink, big.mark = ",")` (SNPs only, `r round(100 * variants_after_plink / variants_before_plink, 2)`% of input)
- **LD Pruning Removed** (`prune.out`): `r format(total_ld_pruned, big.mark = ",")` (`r pct_ld_pruned`% of input SNPs)

### LD Pruning Impact

```{r ld-pruning-table}
if (nrow(ld_prune_summary) > 0) {
  ld_prune_summary <- ld_prune_summary %>%
    mutate(
      CHR = as.integer(as.character(CHR)),
      Pct_of_Input = ifelse(variants_before_plink > 0, round(100 * Variants_Removed / variants_before_plink, 2), 0)
    )
  
  kable(ld_prune_summary,
        format.args = list(big.mark = ","),
        caption = "LD Pruning Removals by Chromosome",
        col.names = c("Chromosome", "Variants Removed", "% of Input SNPs"))
} else {
  cat("No LD pruning counts were recorded in the filter metrics.\n")
}
```

```{r ld-pruning-plot, fig.height=5}
if (nrow(ld_prune_summary) > 0) {
  ggplot(ld_prune_summary, aes(x = factor(CHR, levels = CHR), y = Variants_Removed)) +
    geom_col(fill = "#8E44AD") +
    scale_y_continuous(labels = scales::comma) +
    labs(
      title = "Variants Removed by LD Pruning",
      x = "Chromosome",
      y = "Variants Removed (prune.out)",
      caption = paste0("Total removed: ", format(total_ld_pruned, big.mark = ","), 
                       " (", pct_ld_pruned, "% of input SNPs)")
    ) +
    theme_minimal()
}
```

### Distribution Analysis

These distributions show the spread of variant-level QC metrics and where thresholds are set.

```{r maf-distribution, fig.height=5}
# Calculate MAF (minimum of ALT_FREQS and 1-ALT_FREQS)
variant_maf$MAF <- pmin(variant_maf$ALT_FREQS, 1 - variant_maf$ALT_FREQS)

ggplot(variant_maf, aes(x = MAF)) +
  geom_histogram(bins = 100, fill = "steelblue", alpha = 0.7) +
  geom_vline(xintercept = MAF_THR, linetype = "dashed", color = "red", linewidth = 1) +
  scale_x_continuous(labels = scales::percent, breaks = seq(0, 0.1, 0.01), limits = c(0, 0.1)) +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title = "Minor Allele Frequency Distribution (0-10%)",
    x = "MAF",
    y = "Variant Count",
    caption = paste0("Red line = ", MAF_THR * 100, "% threshold | ", 
                     format(sum(variant_maf$MAF < MAF_THR), big.mark = ","), 
                     " variants (", 
                     round(100 * sum(variant_maf$MAF < MAF_THR) / nrow(variant_maf), 1),
                     "%) below threshold")
  ) +
  theme_minimal()
```

```{r geno-distribution, fig.height=5}
ggplot(variant_miss, aes(x = F_MISS)) +
  geom_histogram(bins = 50, fill = "steelblue", alpha = 0.7) +
  geom_vline(xintercept = GENO_THR, linetype = "dashed", color = "red", linewidth = 1) +
  scale_x_continuous(labels = scales::percent) +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title = "Variant Genotype Missingness Distribution",
    x = "Missingness Rate",
    y = "Variant Count",
    caption = paste0("Red line = ", GENO_THR * 100, "% threshold | ", 
                     format(sum(variant_miss$F_MISS > GENO_THR), big.mark = ","), 
                     " variants (", 
                     round(100 * sum(variant_miss$F_MISS > GENO_THR) / nrow(variant_miss), 1),
                     "%) above threshold")
  ) +
  theme_minimal()
```

```{r hwe-distribution, fig.height=5}
# Transform to -log10(P) for better visualization
variant_hwe$neg_log10_P <- -log10(variant_hwe$P)
threshold_line <- -log10(HWE_THR)

ggplot(variant_hwe, aes(x = neg_log10_P)) +
  geom_histogram(bins = 50, fill = "steelblue", alpha = 0.7) +
  geom_vline(xintercept = threshold_line, linetype = "dashed", color = "red", linewidth = 1) +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title = "Hardy-Weinberg Equilibrium Distribution",
    x = "-log10(P-value)",
    y = "Variant Count",
    caption = paste0("Red line = -log10(", HWE_THR, ") = ", round(threshold_line, 1), " | ", 
                     format(sum(variant_hwe$P < HWE_THR, na.rm = TRUE), big.mark = ","), 
                     " variants (", 
                     round(100 * sum(variant_hwe$P < HWE_THR, na.rm = TRUE) / nrow(variant_hwe), 1),
                     "%) above threshold (more deviant)")
  ) +
  theme_minimal()
```

---

# 4. Sample Missingness

## Method

Sample missingness is calculated across all variants passing quality filters (HQHet + geno/maf/hwe). Samples with excessive missing genotypes are excluded.

- **Threshold**: F_MISS > `r MIND_THR` (`r MIND_THR * 100`%)

```{r missingness-summary}
total_after_sex <- passing_samples
failed_missingness <- nrow(miss_report)
passing_missingness <- total_after_sex - failed_missingness
```

## Results

- **Samples (Post Sex-Check)**: `r format(total_after_sex, big.mark = ",")`
- **Failed Missingness**: `r format(failed_missingness, big.mark = ",")` (`r round(100 * failed_missingness / total_after_sex, 2)`%)
- **Passing Missingness**: `r format(passing_missingness, big.mark = ",")` (`r round(100 * passing_missingness / total_after_sex, 2)`%)

```{r missingness-by-status, eval=!is.null(controls)}
if (!is.null(controls) && "Status" %in% colnames(miss_report)) {
  miss_fail_summary <- miss_report %>%
    group_by(Status) %>%
    summarise(
      N_Failed = n(),
      .groups = "drop"
    )
  
  kable(miss_fail_summary, caption = "Missingness Failures by Case/Control Status")
}
```

## Missingness Distribution

**WARNING**: If a large proportion of cases are failing, check the distribution below. The threshold may be too stringent or there may be a data quality issue.

```{r missingness-distribution, fig.height=5}
# Load aggregated missingness to see all samples
agg_miss <- read.table(
  file.path(DATA_DIR, "plink", paste0(PROJECT, ".aggregated_missingness.txt")),
  header = TRUE, stringsAsFactors = FALSE, comment.char = ""
)

# Add status
if (!is.null(controls) && length(controls) > 0) {
  agg_miss$Status <- ifelse(trimws(agg_miss[,1]) %in% controls, "Control", "Case")
} else {
  agg_miss$Status <- "Unknown"
}

ggplot(agg_miss, aes(x = F_MISS, fill = Status)) +
  geom_histogram(bins = 50, alpha = 0.7) +
  geom_vline(xintercept = MIND_THR, linetype = "dashed", color = "red", linewidth = 1) +
  scale_x_continuous(labels = scales::percent) +
  labs(
    title = "Distribution of Sample Missingness",
    x = "Missingness Rate",
    y = "Count",
    fill = "Status",
    caption = paste0("Red line = ", MIND_THR * 100, "% threshold")
  ) +
  theme_minimal()
```

```{r missingness-stats}
# Summary statistics
miss_stats <- agg_miss %>%
  group_by(Status) %>%
  summarise(
    N = n(),
    Mean = mean(F_MISS),
    Median = median(F_MISS),
    Min = min(F_MISS),
    Max = max(F_MISS),
    Above_Threshold = sum(F_MISS > MIND_THR),
    Pct_Above = round(100 * Above_Threshold / N, 2),
    .groups = "drop"
  )

kable(miss_stats, 
      digits = 4,
      caption = "Missingness Statistics by Case/Control Status")
```

### Failing Samples Details

```{r failing-samples-distribution}
# Show distribution of failing samples
failing_samples <- agg_miss %>%
  filter(F_MISS > MIND_THR) %>%
  arrange(desc(F_MISS)) %>%
  mutate(
    Sample_ID = X.IID,
    Missingness_Pct = round(F_MISS * 100, 2)
  )

if (nrow(failing_samples) > 0) {
  cat(paste0("\n**Total Failing Samples**: ", nrow(failing_samples), "\n\n"))
  
  # Distribution plot of failing samples
  ggplot(failing_samples, aes(x = Missingness_Pct, fill = Status)) +
    geom_histogram(bins = 30, alpha = 0.7) +
    labs(
      title = paste0("Distribution of Missingness Rates for Failing Samples (n=", nrow(failing_samples), ")"),
      x = "Missingness Rate (%)",
      y = "Count",
      fill = "Status"
    ) +
    theme_minimal()
} else {
  cat("No samples exceeded the missingness threshold.\n")
}
```

```{r worst-offenders-table}
# Show worst 20 offenders
if (nrow(failing_samples) > 0) {
  worst_offenders <- failing_samples %>%
    select(Sample_ID, Status, Missingness_Pct, MISSING_CT, OBS_CT) %>%
    head(20)
  
  kable(worst_offenders, 
        format.args = list(big.mark = ","),
        caption = paste0("Top ", min(nrow(failing_samples), 20), " Worst Offenders"))
}
```

```{r case-missingness-summary}
# Specifically summarize case missingness for sharing
case_miss <- agg_miss %>%
  filter(Status == "Case") %>%
  arrange(desc(F_MISS))

if (nrow(case_miss) > 0) {
  cat("\n### Case Missingness Summary\n\n")
  cat("- **Total Cases**: ", nrow(case_miss), "\n")
  cat("- **Cases Failing (>", MIND_THR * 100, "%)**: ", sum(case_miss$F_MISS > MIND_THR), " (", 
      round(100 * sum(case_miss$F_MISS > MIND_THR) / nrow(case_miss), 2), "%)\n", sep="")
  cat("- **Mean Missingness**: ", round(mean(case_miss$F_MISS) * 100, 2), "%\n", sep="")
  cat("- **Median Missingness**: ", round(median(case_miss$F_MISS) * 100, 2), "%\n", sep="")
  cat("- **Range**: ", round(min(case_miss$F_MISS) * 100, 2), "% - ", 
      round(max(case_miss$F_MISS) * 100, 2), "%\n\n", sep="")
}
```

---

# 5. Heterozygosity

## Method

Heterozygosity rate is calculated for each sample on variants passing all quality filters. Outliers (mean ± 3 SD) are excluded as potential contamination or technical artifacts.

```{r het-summary}
total_after_miss <- passing_missingness
failed_het <- nrow(het_report)
passing_het <- total_after_miss - failed_het
```

## Results

- **Samples (Post Missingness)**: `r format(total_after_miss, big.mark = ",")`
- **Failed Heterozygosity**: `r format(failed_het, big.mark = ",")` (`r round(100 * failed_het / total_after_miss, 2)`%)
- **Passing Heterozygosity**: `r format(passing_het, big.mark = ",")` (`r round(100 * passing_het / total_after_miss, 2)`%)

```{r het-by-status, eval=!is.null(controls)}
if (!is.null(controls) && "Status" %in% colnames(het_report)) {
  het_fail_summary <- het_report %>%
    group_by(Status) %>%
    summarise(
      N_Failed = n(),
      .groups = "drop"
    )
  
  kable(het_fail_summary, caption = "Heterozygosity Failures by Case/Control Status")
}
```

---

# 6. Differential Missingness (Case vs Control)

**NOTE**: This is for informational purposes only and does not affect QC filtering.

This tests whether variants have significantly different missingness rates between cases and controls, which could indicate genotyping bias or biological effects.

```{r load-diff-miss}
# Load differential missingness test results
diff_miss_files <- paste0("data/plink/", PROJECT, ".chr", 1:22, ".test.missing")
diff_miss <- lapply(diff_miss_files, function(f) {
  if (file.exists(f)) {
    read.table(f, header = TRUE, stringsAsFactors = FALSE, comment.char = "")
  } else {
    NULL
  }
}) %>% bind_rows()
```

```{r diff-miss-summary}
if (!is.null(diff_miss) && nrow(diff_miss) > 0) {
  # Count significant variants at different thresholds
  sig_001 <- sum(diff_miss$P < 0.001, na.rm = TRUE)
  sig_01 <- sum(diff_miss$P < 0.01, na.rm = TRUE)
  sig_05 <- sum(diff_miss$P < 0.05, na.rm = TRUE)
  
  cat("- **Total variants tested**: ", format(nrow(diff_miss), big.mark = ","), "\n")
  cat("- **P < 0.05**: ", format(sig_05, big.mark = ","), " (", 
      round(100 * sig_05 / nrow(diff_miss), 2), "%)\n", sep="")
  cat("- **P < 0.01**: ", format(sig_01, big.mark = ","), " (", 
      round(100 * sig_01 / nrow(diff_miss), 2), "%)\n", sep="")
  cat("- **P < 0.001**: ", format(sig_001, big.mark = ","), " (", 
      round(100 * sig_001 / nrow(diff_miss), 2), "%)\n", sep="")
}
```

```{r diff-miss-plot, fig.height=6}
if (!is.null(diff_miss) && nrow(diff_miss) > 0) {
  # Plot missingness rates: cases vs controls
  ggplot(diff_miss, aes(x = F_MISS_U, y = F_MISS_A)) +
    geom_point(alpha = 0.1, size = 0.5) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
    scale_x_continuous(labels = scales::percent) +
    scale_y_continuous(labels = scales::percent) +
    labs(
      title = "Variant Missingness: Cases vs Controls",
      x = "Missingness in Controls (F_MISS_U)",
      y = "Missingness in Cases (F_MISS_A)",
      caption = "Red line = equal missingness | Points above line = higher missingness in cases"
    ) +
    theme_minimal()
}
```

**Interpretation**: Variants should fall along the diagonal (equal missingness). Points far from the line indicate differential missingness between cases and controls, which may suggest genotyping bias.


# 7. Final Sample Count

```{r final-summary}
final_samples <- passing_het
final_variants <- variants_after_plink

# Count cases/controls at each stage
if (!is.null(controls) && length(controls) > 0) {
  # Starting counts
  start_cases <- sum(!trimws(sex_report[,1]) %in% controls)
  start_controls <- sum(trimws(sex_report[,1]) %in% controls)
  
  # After sex check
  postsex_cases <- total_after_sex - sum(trimws(sex_report[,1]) %in% controls)
  postsex_controls <- sum(trimws(sex_report[,1]) %in% controls)
  
  # After missingness (from failing samples)
  miss_fail_cases <- sum(!trimws(miss_report[,1]) %in% controls)
  miss_fail_controls <- sum(trimws(miss_report[,1]) %in% controls)
  postmiss_cases <- postsex_cases - miss_fail_cases
  postmiss_controls <- postsex_controls - miss_fail_controls
  
  # After heterozygosity (from failing samples)
  het_fail_cases <- sum(!trimws(het_report[,1]) %in% controls)
  het_fail_controls <- sum(trimws(het_report[,1]) %in% controls)
  final_cases <- postmiss_cases - het_fail_cases
  final_controls <- postmiss_controls - het_fail_controls
}
```

## Summary

- **Starting Samples**: `r format(total_samples, big.mark = ",")` (`r if(exists("start_cases")) paste0(format(start_cases, big.mark=","), " cases, ", format(start_controls, big.mark=","), " controls")`)
- **Final Samples**: `r format(final_samples, big.mark = ",")` (`r if(exists("final_cases")) paste0(format(final_cases, big.mark=","), " cases, ", format(final_controls, big.mark=","), " controls")`) 
- **Total Excluded**: `r format(total_samples - final_samples, big.mark = ",")` (`r if(exists("start_cases")) paste0(format(start_cases - final_cases, big.mark=","), " cases, ", format(start_controls - final_controls, big.mark=","), " controls")`)

### Samples Excluded by Stage

```{r exclusion-by-stage}
if (exists("start_cases")) {
  exclusion_breakdown <- data.frame(
    Stage = c("Sex Check", "Missingness", "Heterozygosity"),
    Cases_Excluded = c(start_cases - postsex_cases, miss_fail_cases, het_fail_cases),
    Controls_Excluded = c(start_controls - postsex_controls, miss_fail_controls, het_fail_controls)
  ) %>%
    mutate(
      Total_Excluded = Cases_Excluded + Controls_Excluded,
      Pct_Cases = round(100 * Cases_Excluded / Total_Excluded, 1)
    )
  
  kable(exclusion_breakdown, 
        caption = "Sample Exclusions by QC Stage and Case/Control Status",
        col.names = c("QC Stage", "Cases Excluded", "Controls Excluded", "Total Excluded", "% Cases"))
}
```

- **Starting Variants**: `r format(prefilter_total, big.mark = ",")`
- **Final Variants**: `r format(final_variants, big.mark = ",")` (`r round(100 * final_variants / prefilter_total, 2)`% retention)
- **Total Excluded**: `r format(prefilter_total - final_variants, big.mark = ",")`

```{r exclusion-cascade}
exclusion_data <- data.frame(
  Stage = c("Starting", "Sex Check", "Missingness", "Heterozygosity", "Final"),
  Samples = c(total_samples, passing_samples, passing_missingness, passing_het, final_samples)
)

ggplot(exclusion_data, aes(x = factor(Stage, levels = Stage), y = Samples)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  geom_text(aes(label = format(Samples, big.mark = ",")), vjust = -0.5) +
  labs(
    title = "Sample Retention Through QC Pipeline",
    x = "QC Stage",
    y = "Number of Samples"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

---

# 8. Population Structure (PCA)

## Principal Components

```{r pca-summary}
# Calculate variance explained
eigenval_file <- file.path(DATA_DIR, "preprocess", paste0(PROJECT, ".build.eigenval"))
if (!file.exists(eigenval_file)) stop(paste("Required file not found:", eigenval_file))
eigenval <- read.table(eigenval_file, header = FALSE)
colnames(eigenval) <- "Eigenvalue"
eigenval$PC <- paste0("PC", seq_len(nrow(eigenval)))
eigenval$Variance_Explained <- 100 * eigenval$Eigenvalue / sum(eigenval$Eigenvalue)
```

### Variance Explained

```{r variance-explained, fig.height=5}
ggplot(eigenval[1:10, ], aes(x = factor(PC, levels = PC), y = Variance_Explained)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  geom_text(aes(label = sprintf("%.2f%%", Variance_Explained)), vjust = -0.5) +
  labs(
    title = "Variance Explained by Principal Components",
    x = "Principal Component",
    y = "Variance Explained (%)"
  ) +
  theme_minimal()
```

### PC1 vs PC2

```{r pca-plot-pc1-pc2, fig.height=6}
ggplot(pca, aes(x = PC1, y = PC2, color = Status)) +
  geom_point(alpha = 0.6) +
  labs(
    title = "Principal Component Analysis (PC1 vs PC2)",
    x = paste0("PC1 (", round(eigenval$Variance_Explained[1], 2), "%)"),
    y = paste0("PC2 (", round(eigenval$Variance_Explained[2], 2), "%)"),
    color = "Status"
  ) +
  theme_minimal()
```

### PC3 vs PC4

```{r pca-plot-pc3-pc4, fig.height=6}
ggplot(pca, aes(x = PC3, y = PC4, color = Status)) +
  geom_point(alpha = 0.6) +
  labs(
    title = "Principal Component Analysis (PC3 vs PC4)",
    x = paste0("PC3 (", round(eigenval$Variance_Explained[3], 2), "%)"),
    y = paste0("PC4 (", round(eigenval$Variance_Explained[4], 2), "%)"),
    color = "Status"
  ) +
  theme_minimal()
```

### Scree Plot (All 20 PCs)

```{r scree-plot, fig.height=5}
ggplot(eigenval, aes(x = seq_len(nrow(eigenval)), y = Eigenvalue)) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_point(color = "steelblue", size = 2) +
  labs(
    title = "Scree Plot: Eigenvalues of Principal Components",
    x = "Principal Component",
    y = "Eigenvalue"
  ) +
  theme_minimal()
```

---

# Session Info

```{r session-info}
sessionInfo()
```
