---
title: "MNP Carrier & Validation Report"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: united
    highlight: tango
params:
  stats_tsv: "data/mnp/mnp_carrier_stats.tsv"
  aaf_cutoff: 0.001
  aaf_bins: "0.0001,0.001,0.01"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, fig.width = 9, fig.height = 5)
library(tidyverse)
library(scales)
library(knitr)

aaf_bins <- as.numeric(strsplit(params$aaf_bins, ",")[[1]])
cutoff <- as.numeric(params$aaf_cutoff)
```

This report describes the candidate MNPs found in cases, how many cases and controls
*possibly* carry each (VAF-cis carriers, before CRAM read-check), where they sit relative
to the REGENIE `--aaf-bins`, and what the read-check validation showed. It justifies the
cohort-AAF cutoff below which we create + validate MNP records (full case **and** control
read-check, GT rewritten only for `Strong_Cis` carriers); everything above the cutoff is
left as separate SNVs for all samples.

```{r load}
num <- function(x) suppressWarnings(as.numeric(ifelse(x %in% c(".", "inf", "Inf", ""), NA, x)))

d <- read_tsv(params$stats_tsv, show_col_types = FALSE, col_types = cols(.default = col_character())) %>%
  mutate(
    across(c(n_case, n_control, n_other, carrier_total), as.integer),
    across(any_of(c("case_freq","control_freq","cohort_aaf","fisher_p","odds_ratio",
                    "case_cis_frac","ctrl_cis_frac")), num),
    across(any_of(c("case_checked","case_strong","ctrl_checked","ctrl_strong")), as.integer),
    target = cohort_aaf <= cutoff
  )
has_validation <- "case_cis_frac" %in% names(d)
n_mnp <- nrow(d)
```

- **Total candidate case MNPs:** `r n_mnp`
- **Cohort-AAF cutoff for MNP creation:** `r format(cutoff, scientific = TRUE)`
- **MNPs at/below cutoff (treated as MNPs):** `r sum(d$target)`
- **MNPs above cutoff (kept as separate SNVs):** `r sum(!d$target)`

## 1. Carrier-count distribution (the bimodality)

Most MNPs are carried by very few people; a long tail is carried by hundreds–thousands of
controls (common germline haplotypes). The cutoff lands in the gap.

```{r dist-control}
ggplot(d, aes(x = pmax(n_control, 0.5))) +
  geom_histogram(bins = 60, fill = "steelblue", color = "white") +
  scale_x_log10(labels = comma, breaks = c(1,2,5,10,20,50,100,500,2000,8000)) +
  labs(x = "control carriers per MNP (log scale)", y = "MNPs",
       title = "Control carriers per MNP") +
  theme_minimal()
```

## 2. Case vs control carrier frequency, with aaf-bins

Each point is an MNP. Diagonal = equal case/control carrier frequency. Dashed lines mark
the REGENIE allele-frequency bins (carrier ≈ 2×AAF). Points left of the leftmost line are
rare enough to be in the ultra-rare/rare masks.

```{r scatter}
bin_carrier <- aaf_bins * 2  # carrier freq approx
ggplot(d, aes(x = pmax(control_freq, 1e-5), y = pmax(case_freq, 1e-5), color = target)) +
  geom_point(alpha = 0.5, size = 1) +
  geom_abline(slope = 1, intercept = 0, color = "grey50") +
  geom_vline(xintercept = bin_carrier, linetype = "dashed", color = "firebrick") +
  scale_x_log10(labels = label_number(accuracy = 1e-5)) +
  scale_y_log10(labels = label_number(accuracy = 1e-5)) +
  scale_color_manual(values = c(`TRUE` = "firebrick", `FALSE` = "grey60"),
                     labels = c(`TRUE` = paste0("AAF <= ", cutoff), `FALSE` = "above cutoff"),
                     name = NULL) +
  labs(x = "control carrier frequency", y = "case carrier frequency",
       title = "Case vs control carrier frequency (dashed = aaf-bins)") +
  theme_minimal()
```

## 3. Cohort AAF and the bins

```{r aaf-bins}
bin_tbl <- tibble(bin = aaf_bins) %>%
  mutate(n_mnps = map_int(bin, ~ sum(d$cohort_aaf <= .x, na.rm = TRUE))) %>%
  add_row(bin = Inf, n_mnps = sum(d$cohort_aaf > max(aaf_bins), na.rm = TRUE))
kable(bin_tbl %>% mutate(bin = ifelse(is.infinite(bin), paste0("> ", max(aaf_bins)), paste0("<= ", bin))),
      col.names = c("cohort AAF", "MNPs"), caption = "MNPs by cohort AAF bin (cumulative for finite rows)")

ggplot(d, aes(x = pmax(cohort_aaf, 1e-6))) +
  geom_histogram(bins = 60, fill = "darkorange", color = "white") +
  geom_vline(xintercept = cutoff, color = "firebrick", linewidth = 1) +
  scale_x_log10(labels = label_number(accuracy = 1e-6)) +
  annotate("text", x = cutoff, y = Inf, label = paste0("cutoff ", cutoff),
           vjust = 2, hjust = -0.05, color = "firebrick") +
  labs(x = "cohort AAF (log scale)", y = "MNPs", title = "Cohort AAF distribution") +
  theme_minimal()
```

## 4. What the read-check validation showed

`r if (!has_validation) "No validation columns in the stats file (re-run mnp_carrier_stats.py with --read-check)." else ""`

```{r validation, eval=has_validation}
v <- d %>% filter(!is.na(case_cis_frac))
```

### Case cis-fraction vs cohort AAF

If validation confirmed the MNP, `case_cis_frac` (fraction of checked case carriers that
are `Strong_Cis`) is high. This shows whether co-occurrence reflects true cis haplotypes
across the frequency range.

```{r val-case, eval=has_validation}
ggplot(v, aes(x = pmax(cohort_aaf, 1e-6), y = case_cis_frac, size = case_checked, color = target)) +
  geom_point(alpha = 0.5) +
  geom_vline(xintercept = cutoff, color = "firebrick", linetype = "dashed") +
  scale_x_log10(labels = label_number(accuracy = 1e-6)) +
  scale_color_manual(values = c(`TRUE` = "firebrick", `FALSE` = "grey60"), name = NULL,
                     labels = c(`TRUE` = "target", `FALSE` = "above cutoff")) +
  labs(x = "cohort AAF (log scale)", y = "case cis-fraction (Strong_Cis / checked)",
       size = "cases checked", title = "Case validation vs cohort AAF") +
  theme_minimal()
```

### Case vs control cis-fraction (target MNPs)

For the MNPs we validated on both sides, do cases and controls agree on cis evidence?
Points off the diagonal are MNPs where one group is more cis than the other.

```{r val-casectrl, eval=has_validation}
vc <- v %>% filter(target, !is.na(ctrl_cis_frac))
if (nrow(vc) > 0) {
  ggplot(vc, aes(x = ctrl_cis_frac, y = case_cis_frac, size = ctrl_checked)) +
    geom_point(alpha = 0.5, color = "steelblue") +
    geom_abline(slope = 1, intercept = 0, color = "grey50") +
    labs(x = "control cis-fraction", y = "case cis-fraction",
         size = "controls checked", title = "Case vs control cis-fraction (target MNPs)") +
    theme_minimal()
} else {
  "No target MNPs with both case and control validation."
}
```

### Validation impact: carriers removed

How many "possible" carriers the read-check reclassified as not-cis (trans), i.e. would have
been false MNP carriers without validation.

```{r val-impact, eval=has_validation}
impact <- v %>% summarise(
  case_checked = sum(case_checked, na.rm = TRUE),
  case_strong  = sum(case_strong, na.rm = TRUE),
  ctrl_checked = sum(ctrl_checked, na.rm = TRUE),
  ctrl_strong  = sum(ctrl_strong, na.rm = TRUE)
) %>% mutate(
  case_removed_pct = round(100 * (1 - case_strong / pmax(case_checked, 1)), 1),
  ctrl_removed_pct = round(100 * (1 - ctrl_strong / pmax(ctrl_checked, 1)), 1)
)
kable(impact, caption = "Read-check validation totals (strong = confirmed cis carriers)")
```

## 5. Download cost vs cutoff

Target MNPs and an upper bound on control read-checks (sum of control carriers; the
**deduped** unique-CRAM count is reported by `mnp_carrier_stats.py`'s targeting table).

```{r cost}
cost <- tibble(cut = sort(unique(c(aaf_bins, cutoff)))) %>%
  mutate(
    target_mnps = map_int(cut, ~ sum(d$cohort_aaf <= .x, na.rm = TRUE)),
    control_checks_upper = map_int(cut, ~ sum(d$n_control[d$cohort_aaf <= .x], na.rm = TRUE))
  )
kable(cost, col.names = c("cohort AAF cutoff", "target MNPs", "control read-checks (upper bound)"),
      caption = "Scope and cost by cutoff")
```

## 6. Top target MNPs by case enrichment

```{r top-table}
top <- d %>% filter(target) %>%
  arrange(desc(n_case), cohort_aaf) %>%
  select(any_of(c("mnp_id","n_case","n_control","cohort_aaf","aaf_bin",
                  "case_or_ctrl_ratio","fisher_p","odds_ratio",
                  "case_cis_frac","ctrl_cis_frac"))) %>%
  head(30)
kable(top, caption = "Top 30 target MNPs by case count")
```
