---
title: "Rare Variant QC by Freeze (controls only, pairwise)"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  html_document:
    toc: true
    toc_float: true
---

This report performs **freeze-bias QC for rare variants in controls only**, comparing
**all freezes pairwise** (a variant is flagged if any freeze pair trips a threshold).

Why this is needed:
- Separate technical freeze/batch effects from true disease signal.
- If callability differs by freeze (e.g. one freeze has more missing genotypes), a variant
  can look associated downstream for technical reasons.
- Controls-only comparison avoids mixing real case enrichment into the batch check.

What this report tests per variant (across every freeze pair present):
- **Differential missingness by freeze** (Fisher's exact + absolute rate differences).
- **Differential carrier rate by freeze** among called controls (Fisher's exact + absolute differences).

The flagged list (`rare_variant_exclusions.txt`) is **built but not necessarily applied** — see
`rare_variant_qc.apply_exclusions` in config. Treat flags as a screening layer, not automatic
exclusion by p-value alone.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=FALSE,warning=FALSE,message=FALSE)
library(readr)
library(dplyr)
library(tidyr)
library(knitr)
```

```{r files}
base_dir<-"data/qc/rare_variant"
chr_files<-list.files(base_dir,pattern="^chr[0-9]+\\.rv-qc\\.tsv$",full.names=TRUE)
out_all<-file.path(base_dir,"rv-qc.freeze_pairwise.all.tsv")
out_flagged<-file.path(base_dir,"rv-qc.freeze_pairwise.flagged.tsv")
pathogenic_vus_file<-"data/preprocess/pathogenic_vus.csv"

if(length(chr_files)==0){
  stop("No per-chromosome rare variant QC files found.")
}
frames<-lapply(chr_files,function(f) read_tsv(f,show_col_types=FALSE))
qc_all<-bind_rows(frames)
qc_flagged<-qc_all %>% filter(flag==1)
write_tsv(qc_all,out_all)
write_tsv(qc_flagged,out_flagged)

freezes_seen<-if("freezes" %in% names(qc_all)) sort(unique(qc_all$freezes)) else NA

pv<-read_csv(pathogenic_vus_file,show_col_types=FALSE) %>%
  select(ID,Gene,Variant.LoF_level) %>%
  distinct() %>%
  mutate(variant_set=case_when(
    Variant.LoF_level==1~"pathogenic",
    Variant.LoF_level==2~"vus",
    TRUE~"other"
  ))
qc_all_pv<-qc_all %>% inner_join(pv,by=c("variant_id"="ID"))
qc_flagged_pv<-qc_flagged %>% inner_join(pv,by=c("variant_id"="ID"))
```

## Overview

Freezes compared: **`r if(all(is.na(freezes_seen))) "unknown" else paste(freezes_seen, collapse=", ")`**

```{r overview}
flagged_gene_n<-qc_flagged_pv %>%
  filter(!is.na(Gene),Gene!="") %>%
  distinct(Gene) %>%
  nrow()

overview<-tibble(
  metric=c(
    "Total variants tested",
    "Flagged variants",
    "Fraction flagged",
    "Unique genes with flagged variants (pathogenic_vus.csv overlap)"
  ),
  value=c(
    nrow(qc_all),
    nrow(qc_flagged),
    ifelse(nrow(qc_all)>0,sprintf("%.3f",nrow(qc_flagged)/nrow(qc_all)),"NA"),
    flagged_gene_n
  )
)
kable(overview)
```

## Exclusion Logic and Cutoffs

`rare_variant_exclusions.txt` is built from variants with `flag==1`. Rule logic (pairwise; a
variant trips if ANY freeze pair qualifies), from `rare_variant_qc.py` + `rare_variant_qc.smk`:

- Missingness absolute difference: `abs_missing_rate_diff > 0.03` (max over freeze pairs)
- High missingness in any freeze: `max_missing_rate > 0.05`
- Missingness Fisher test: `missing_fisher_p < 1e-6` (min over freeze pairs)
- Carrier Fisher test: `carrier_fisher_p < 1e-4` **and** `abs_carrier_rate_diff >= 0.01` **and** `total_called_carriers >= 3`

## Flag Reasons

```{r reasons}
if(nrow(qc_flagged)==0){
  cat("No flagged variants.")
} else {
  reason_tbl<-qc_flagged %>%
    mutate(flag_reasons=ifelse(is.na(flag_reasons)|flag_reasons=="","none",flag_reasons)) %>%
    separate_rows(flag_reasons,sep=",") %>%
    count(flag_reasons,sort=TRUE)
  kable(reason_tbl)
}
```

## Variant Type Summary

```{r variant-type-summary}
type_total<-qc_all %>% group_by(variant_type) %>% summarise(total_variants=n(),.groups='drop')
type_flagged<-qc_flagged %>% group_by(variant_type) %>% summarise(flagged_variants=n(),.groups='drop')
type_summary<-type_total %>%
  left_join(type_flagged,by="variant_type") %>%
  mutate(flagged_variants=ifelse(is.na(flagged_variants),0,flagged_variants))
kable(type_summary)
```

## Flag Reasons By Variant Type

```{r reason-by-type}
if(nrow(qc_flagged)==0){
  cat("No flagged variants.")
} else {
  reason_by_type<-qc_flagged %>%
    mutate(flag_reasons=ifelse(is.na(flag_reasons)|flag_reasons=="","none",flag_reasons)) %>%
    separate_rows(flag_reasons,sep=",") %>%
    count(variant_type,flag_reasons,sort=TRUE)
  kable(reason_by_type)
}
```

## Sparse Carrier Summary

```{r sparse-summary}
if(nrow(qc_flagged)==0){
  cat("No flagged variants.")
} else {
  sparse_tbl<-tibble(
    group=c("total_called_carriers==1","total_called_carriers==2","total_called_carriers>=3"),
    n=c(
      sum(qc_flagged$total_called_carriers==1,na.rm=TRUE),
      sum(qc_flagged$total_called_carriers==2,na.rm=TRUE),
      sum(qc_flagged$total_called_carriers>=3,na.rm=TRUE)
    )
  )
  kable(sparse_tbl)
}
```

## Top Flagged Variants

```{r top-flagged}
if(nrow(qc_flagged)==0){
  cat("No flagged variants.")
} else {
  top_flagged<-qc_flagged %>%
    arrange(missing_fisher_p,carrier_fisher_p,desc(abs_missing_rate_diff),desc(abs_carrier_rate_diff)) %>%
    select(any_of(c(
      "variant_id","variant_type","freezes",
      "abs_missing_rate_diff","max_missing_rate","missing_fisher_p","missing_worst_pair",
      "abs_carrier_rate_diff","carrier_fisher_p","carrier_worst_pair",
      "total_called_carriers","flag_reasons"))) %>%
    head(100)
  kable(top_flagged)
}
```

## Top Missingness Imbalance

```{r top-missing}
if(nrow(qc_all)==0){
  cat("No variants to display.")
} else {
  top_missing<-qc_all %>%
    arrange(desc(abs_missing_rate_diff),missing_fisher_p) %>%
    select(any_of(c(
      "variant_id","variant_type","freezes",
      "total_missing","total_called","max_missing_rate",
      "abs_missing_rate_diff","missing_fisher_p","missing_worst_pair","flag","flag_reasons"))) %>%
    head(50)
  kable(top_missing)
}
```

## Top Carrier Imbalance

```{r top-carrier}
if(nrow(qc_all)==0){
  cat("No variants to display.")
} else {
  top_carrier<-qc_all %>%
    arrange(desc(abs_carrier_rate_diff),carrier_fisher_p) %>%
    select(any_of(c(
      "variant_id","variant_type","freezes",
      "total_called","total_called_carriers",
      "abs_carrier_rate_diff","carrier_fisher_p","carrier_worst_pair","flag","flag_reasons"))) %>%
    head(50)
  kable(top_carrier)
}
```

## Pathogenic/VUS Summary

```{r pv-summary}
if(nrow(qc_all_pv)==0){
  cat("No overlaps found between QC variants and data/preprocess/pathogenic_vus.csv.")
} else {
  pv_summary<-qc_all_pv %>%
    group_by(variant_set) %>%
    summarise(total=n(),flagged=sum(flag==1),.groups='drop') %>%
    arrange(variant_set)
  kable(pv_summary)
}
```

## Flagged Pathogenic/VUS By Gene

```{r pv-gene-summary}
if(nrow(qc_flagged_pv)==0){
  cat("No flagged overlaps found in data/preprocess/pathogenic_vus.csv.")
} else {
  pv_gene_summary<-qc_flagged_pv %>%
    group_by(Gene,variant_set) %>%
    summarise(flagged=n(),.groups='drop') %>%
    arrange(desc(flagged),Gene,variant_set)
  kable(pv_gene_summary)
}
```

## CHEK2 From Pathogenic/VUS

```{r chek2-from-pv}
chek2_tbl<-qc_all_pv %>%
  filter(Gene=="CHEK2") %>%
  arrange(missing_fisher_p,carrier_fisher_p) %>%
  select(any_of(c(
    "variant_id","variant_set","variant_type","freezes",
    "abs_missing_rate_diff","max_missing_rate","missing_fisher_p","missing_worst_pair",
    "abs_carrier_rate_diff","carrier_fisher_p","carrier_worst_pair",
    "total_called_carriers","flag","flag_reasons")))
if(nrow(chek2_tbl)==0){
  cat("No CHEK2 variants from data/preprocess/pathogenic_vus.csv were found in QC output.")
} else {
  kable(chek2_tbl)
}
```
