---
title: "REGENIE Exome-Wide Association Analysis Report"
author: "Brad Wubbenhorst"
output:
  html_document:
    theme: paper
    toc: true
    toc_depth: 3
    toc_float:
      collapsed: false
      smooth_scroll: true
    code_folding: hide
date: "`r Sys.Date()`"
params:
  project_name: "TECAC_WES"
  # Stem between ".chrN." and ".regenie" for Step 2 outputs under data/final/.
  # Gene-based freeze_qc_candidate runs: --out ...step2_gene_based_STATUS.freeze_qc_candidate → ...freeze_qc_candidate_STATUS.regenie on disk.
  step2_single_variant_mid: "step2_single_variant_STATUS"
  step2_gene_based_mid: "step2_gene_based_STATUS"
---

```{=html}
<style>
/* Light section breathing room — keep TOC usable when scrolling masks */
h2 { margin-top: 2.2rem; padding-top: 0.4rem; border-top: 1px solid #e5e5e5; }
h3 { margin-top: 1.8rem; }
h4 { margin-top: 1.4rem; }
h5 { margin-top: 1.1rem; color: #444; }
.regenie-jump {
  border: 1px solid #ddd; border-left: 4px solid #2E86AB;
  background: #f7fafc; padding: 10px 14px; margin: 0.8rem 0 1.4rem 0;
  font-size: 0.95rem;
}
.regenie-jump a { margin-right: 0.85rem; white-space: nowrap; }
.regenie-block-sep {
  margin: 2rem 0 0.6rem 0; padding-top: 0.4rem;
  border-top: 2px solid #d0d7de;
}
</style>
```

## Project: `r params$project_name`

```{=html}
<div style="border:1px solid #ccc; border-left:4px solid #2E86AB; padding:12px 14px; margin:1rem 0; background:#fafafa;">
<strong>How to read this report</strong>
<ul style="margin-bottom:0;">
<li><strong>Burden-style interpretation</strong> (effect direction, OR, volcano, CHEK2, directional gene lists) uses <code>TEST == &quot;ADD&quot;</code> only.</li>
<li><strong>FDR</strong> (<code>PVAL_ADJ</code>) is Benjamini–Hochberg <em>within family</em>: ADD alone; SKAT+SKATO together. Do not mix with the other family’s FDR.</li>
<li><strong>SKAT / SKATO</strong> are non-directional (no risk/protective OR). Use the separate SKAT/SKATO export — not the ADD tables.</li>
<li>Exports: all-tests archive, ADD-only, and SKAT/SKATO-only (no BETA/OR columns).</li>
</ul>
</div>
```

<!-- 
NOTES TO SELF:
- M1 = pathogenic variants only (LoF level 1)
- M2 = VUS variants only (LoF level 2) 
- M3 = pathogenic + VUS variants combined
- M4 = synonymous negative-control variants (excluding LoF 1/2 and excluding blacklisted potential MNP-pair IDs)
- Gene×consequence exclude (preprocess): CHEK2 missense_variant dropped from all masks (literature: CHEK2 missense not disease-contributing here)
- ADD = additive burden test (BETA/SE/OR and directional interpretation apply here only)
- ADD-SKAT / ADD-SKATO = variance-component tests (non-directional; separate export; no BETA/OR interpretation)
- FDR (PVAL_ADJ): BH within ADD; separately BH within SKAT+SKATO combined
- Exome-wide significance = p < 2.5e-6 (not GWAS 5e-8)
- Lambda GC less interpretable in ExWAS due to sparse rare variants

KEY TERMS:
- BETA (Effect Size): Log odds ratio of disease association for alt allele
  * Positive BETA = alt allele increases disease risk
  * Negative BETA = alt allele is protective
- SE (Standard Error): Uncertainty around effect size estimate
  * High SE = less confidence in BETA (common for rare variants)
- PVAL/LOG10P: p-value from association test
  * LOG10P = -log10(p-value) for plotting
  * Plots cap extreme -log10(p) at 100 where needed; CSVs keep raw LOG10P
  * Smaller p-values (larger LOG10P) = stronger evidence
- PVAL_ADJ/FDR: False discovery rate-adjusted p-values (Benjamini-Hochberg)
  * Controls for multiple testing across thousands of variants/genes
  * Prevents false positives from testing many hypotheses
-->


```{r setup,echo=FALSE, message=FALSE, warning=FALSE, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width = 10, fig.height = 6)
library(dplyr)
library(readr)
library(data.table)
library(ggplot2)
library(knitr)
library(kableExtra)
library(ggrepel)
library(qqman)
library(DiagrammeR)

# Set consistent theme
theme_set(theme_minimal(base_size = 13))

# Shared HTML table look (gene col bold; consequence headers readable)
.fmt_consequence_label <- function(x) {
  # variant-list display: keep compound terms readable
  x <- as.character(x)
  x <- gsub("&", " / ", x, fixed = TRUE)
  x <- gsub("_", " ", x, fixed = TRUE)
  x
}
.fmt_consequence_header <- function(x) {
  # matrix column headers: single terms, spaces not underscores
  x <- as.character(x)
  x <- gsub("_", " ", x, fixed = TRUE)
  x
}

.style_report_kable <- function(
  k,
  font_size = 12,
  full_width = TRUE,
  gene_col = 1L,
  scroll_h = NULL,
  header_wrap = FALSE
) {
  out <- kableExtra::kable_styling(
    k,
    full_width = full_width,
    bootstrap_options = c("striped", "hover", "condensed"),
    font_size = font_size
  )
  if (!is.null(gene_col) && length(gene_col) && gene_col[1] > 0) {
    out <- kableExtra::column_spec(out, gene_col[1], bold = TRUE)
  }
  if (isTRUE(header_wrap)) {
    out <- kableExtra::row_spec(
      out, 0,
      align = "c",
      extra_css = paste(
        "vertical-align: middle;",
        "white-space: normal;",
        "line-height: 1.25;",
        "min-height: 4.2em;",
        "max-width: 8.5em;",
        "word-break: break-word;",
        "hyphens: auto;",
        "padding-top: 10px;",
        "padding-bottom: 10px;"
      )
    )
  }
  if (!is.null(scroll_h)) {
    out <- kableExtra::scroll_box(out, width = "100%", height = scroll_h)
  }
  out
}
```

```{r load-data,echo=FALSE,message=FALSE,warning=FALSE}
# Validate project name parameter
if(is.null(params$project_name) || params$project_name == "") {
  stop("ERROR: project_name parameter is required but was not provided or is empty. ",
       "Please specify params=list(project_name='YOUR_PROJECT_NAME') when rendering.")
}

# Load single variant data from all chromosomes
single_variant_list <- list()

for(i in 1:22) {
  file_path <- paste0("data/final/", params$project_name, ".chr", i, ".", params$step2_single_variant_mid, ".regenie")
  if(file.exists(file_path)) {
    single_variant_list[[i]] <- read_delim(file_path, delim = " ", col_names = TRUE, trim_ws = TRUE)
  }
}

# Combine all single variant data
single_variant <- bind_rows(single_variant_list) %>%
   mutate(across(c(CHROM, GENPOS, N, N_CASES, N_CONTROLS), as.integer),
         across(c(A1FREQ, A1FREQ_CASES, A1FREQ_CONTROLS, BETA, SE, CHISQ, LOG10P), as.numeric)) %>%
  mutate(CHROM = as.factor(CHROM)) %>% 
  arrange(CHROM, GENPOS) %>% 
  mutate(PVAL = 10^(-LOG10P),
         PVAL_ADJ = p.adjust(PVAL, method = "BH"),
         LOG10P_ADJ = -log10(PVAL_ADJ)) %>%
  mutate(POS_INDEX = row_number(),
         OR = exp(BETA)) %>%
  # Parse REF/ALT from trailing ID fields (…_POS_REF_ALT)
  mutate(
    REF = vapply(ID, function(x) {
      p <- strsplit(as.character(x), "_", fixed = TRUE)[[1L]]
      if (length(p) >= 4L) p[[length(p) - 1L]] else NA_character_
    }, character(1L)),
    ALT = vapply(ID, function(x) {
      p <- strsplit(as.character(x), "_", fixed = TRUE)[[1L]]
      if (length(p) >= 4L) p[[length(p)]] else NA_character_
    }, character(1L))
  )

# Thin pathogenic_vus / synonymous (small extracts only — do not load full VEP or full mask stats here)
pathogenic_vus_file <- paste0("data/final/", params$project_name, ".pathogenic_vus.csv")
if (file.exists(pathogenic_vus_file)) {
  pathogenic_annotations <- read_csv(pathogenic_vus_file, show_col_types = FALSE)
} else {
  warning("Pathogenic/VUS annotation file not found.")
  pathogenic_annotations <- NULL
}

synonymous_file <- paste0("data/final/", params$project_name, ".synonymous.csv")
if (file.exists(synonymous_file)) {
  synonymous_annotations <- read_csv(synonymous_file, show_col_types = FALSE)
} else {
  synonymous_annotations <- NULL
}

# Full mask stats may be ~700k rows (all mask sites). Never load whole file — subset by ID later.
.mask_stats_candidates <- c(
  file.path("data/regenie", "mask_variant_stats.tsv"),
  paste0("data/final/", params$project_name, ".mask_variant_stats.tsv")
)
.mask_stats_path <- .mask_stats_candidates[file.exists(.mask_stats_candidates)][1]
if (is.na(.mask_stats_path)) .mask_stats_path <- ""
if (nzchar(.mask_stats_path)) {
  message("Mask variant stats available (subset-on-demand): ", .mask_stats_path)
}

# Pull only needed ID rows from mask_variant_stats (data.table; no shell/awk)
.mask_stats_dt <- NULL
.id_aliases <- function(ids) {
  ids <- as.character(ids)
  out <- ids
  out <- c(out, ifelse(startsWith(ids, "chr"), substring(ids, 4L), paste0("chr", ids)))
  unique(out[nzchar(out) & !is.na(out)])
}
.stats_for_ids <- function(ids) {
  ids <- unique(as.character(ids))
  ids <- ids[nzchar(ids) & !is.na(ids)]
  if (!length(ids)) return(tibble())
  want <- .id_aliases(ids)

  if (nzchar(.mask_stats_path) && file.exists(.mask_stats_path)) {
    got <- tryCatch({
      if (is.null(.mask_stats_dt)) {
        .mask_stats_dt <<- data.table::fread(
          .mask_stats_path,
          sep = "\t",
          showProgress = FALSE,
          nThread = 1L
        )
        if (!"ID" %in% names(.mask_stats_dt)) {
          stop("mask_variant_stats.tsv missing ID column")
        }
        message("Loaded mask_variant_stats into memory: ", nrow(.mask_stats_dt), " rows")
      }
      sub <- .mask_stats_dt[ID %in% want]
      if (nrow(sub) == 0L) {
        tibble()
      } else {
        # remap stats ID back to the query ID form used by the caller
        alias_to_query <- new.env(parent = emptyenv())
        for (q in ids) {
          for (a in .id_aliases(q)) assign(a, q, envir = alias_to_query)
        }
        sub <- tibble::as_tibble(sub)
        sub$ID <- vapply(as.character(sub$ID), function(a) {
          if (exists(a, envir = alias_to_query, inherits = FALSE)) get(a, envir = alias_to_query) else a
        }, character(1))
        distinct(sub, ID, .keep_all = TRUE)
      }
    }, error = function(e) {
      message("mask_stats subset failed: ", conditionMessage(e))
      NULL
    })
    if (!is.null(got) && nrow(got) > 0) return(got)
  }

  # fallback: thin pathogenic/synonymous extracts
  bits <- list()
  if (!is.null(pathogenic_annotations) && nrow(pathogenic_annotations) > 0) {
    bits[[length(bits) + 1L]] <- pathogenic_annotations %>% filter(ID %in% want)
  }
  if (!is.null(synonymous_annotations) && nrow(synonymous_annotations) > 0) {
    bits[[length(bits) + 1L]] <- synonymous_annotations %>% filter(ID %in% want)
  }
  if (!length(bits)) return(tibble(ID = character()))
  bind_rows(bits) %>% distinct(ID, .keep_all = TRUE)
}

.join_gene_carriers <- function(df) {
  if (is.null(report_gene_carriers) || nrow(report_gene_carriers) == 0L) {
    return(df %>% mutate(Cases = NA_integer_, Controls = NA_integer_))
  }
  carr <- report_gene_carriers %>%
    select(GENE, MASK, THRESH, Cases = n_case_carriers, Controls = n_control_carriers)
  df %>% left_join(carr, by = c("GENE", "MASK", "THRESH"))
}

.fill_variant_carriers_from_afcc <- function(df) {
  # Rare-variant approx: alt allele count ≈ carrier count when mostly heterozygous
  if (!all(c("A1FREQ_CASES", "A1FREQ_CONTROLS", "N_CASES", "N_CONTROLS") %in% names(df))) {
    return(df)
  }
  if (!"n_case_carrier" %in% names(df)) df$n_case_carrier <- NA_real_
  if (!"n_control_carrier" %in% names(df)) df$n_control_carrier <- NA_real_
  df %>%
    mutate(
      n_case_carrier = if_else(
        is.na(n_case_carrier) & is.finite(A1FREQ_CASES) & is.finite(N_CASES),
        as.integer(round(A1FREQ_CASES * 2 * N_CASES)),
        as.integer(n_case_carrier)
      ),
      n_control_carrier = if_else(
        is.na(n_control_carrier) & is.finite(A1FREQ_CONTROLS) & is.finite(N_CONTROLS),
        as.integer(round(A1FREQ_CONTROLS * 2 * N_CONTROLS)),
        as.integer(n_control_carrier)
      )
    )
}

# Prebuilt slim tables from build_regenie_report_tables.py — required
.report_prefix <- paste0("data/final/", params$project_name, ".report")
.pre_top_fp <- paste0(.report_prefix, ".top_genes_ADD.tsv")
.pre_skat_fp <- paste0(.report_prefix, ".top_genes_SKAT.tsv")
.pre_contrib_fp <- paste0(.report_prefix, ".variant_contrib.tsv")
.pre_cons_fp <- paste0(.report_prefix, ".consequence_matrix.tsv")
.pre_carr_fp <- paste0(.report_prefix, ".gene_carriers.tsv")
.pre_chek2_carr_fp <- paste0(.report_prefix, ".chek2_carriers.tsv")
.required_report_files <- c(
  .pre_top_fp, .pre_skat_fp, .pre_contrib_fp, .pre_cons_fp,
  .pre_carr_fp, .pre_chek2_carr_fp
)
.missing_report <- .required_report_files[!file.exists(.required_report_files)]
if (length(.missing_report)) {
  stop(
    "Missing required prebuilt report tables. Run build_regenie_report_tables first.\n",
    paste(" -", .missing_report, collapse = "\n")
  )
}
report_top_genes <- read_tsv(.pre_top_fp, show_col_types = FALSE)
report_top_skat <- read_tsv(.pre_skat_fp, show_col_types = FALSE)
report_contrib <- read_tsv(.pre_contrib_fp, show_col_types = FALSE)
report_consequence <- read_tsv(.pre_cons_fp, show_col_types = FALSE)
report_gene_carriers <- read_tsv(.pre_carr_fp, show_col_types = FALSE)
report_chek2_carriers <- read_tsv(.pre_chek2_carr_fp, show_col_types = FALSE)
message("Loaded prebuilt report tables: ", .report_prefix, ".*")

# Snplist not needed when prebuilt contrib/consequence tables are present
mask_snplist <- tibble(
  mask_id = character(), ID = character(),
  GENE = character(), MASK = character(), THRESH = character()
)
message("Skipping full snplist load (using prebuilt report tables)")

# OR 95% CI from BETA ± 1.96*SE
.add_or_ci <- function(df) {
  df %>%
    mutate(
      OR_L95 = if_else(is.finite(BETA) & is.finite(SE), exp(BETA - 1.96 * SE), NA_real_),
      OR_U95 = if_else(is.finite(BETA) & is.finite(SE), exp(BETA + 1.96 * SE), NA_real_)
    )
}
single_variant <- .add_or_ci(single_variant)

# Load gene-based data from all chromosomes
gene_based_list <- list()

for(i in 1:22) {
  file_path <- paste0("data/final/", params$project_name, ".chr", i, ".", params$step2_gene_based_mid, ".regenie")
  if(file.exists(file_path)) {
    # Skip comment lines (e.g. "# Regenie step 2") so header is CHROM GENPOS ... BETA ...; wrong skip misaligns BETA.
    gene_based_list[[i]] <- read_delim(file_path, delim = " ", col_names = TRUE, trim_ws = TRUE, comment = "#")
  }
}

# Combine all gene-based rows (all TEST types) — FDR applied within family below
gene_based_all <- bind_rows(gene_based_list) %>%
  mutate(across(c(CHROM, GENPOS, N, N_CASES, N_CONTROLS), as.integer),
         across(c(A1FREQ, A1FREQ_CASES, A1FREQ_CONTROLS, BETA, SE, CHISQ, LOG10P), as.numeric)) %>%
  mutate(CHROM = as.factor(CHROM),
         PVAL = 10^(-LOG10P)) %>%
  # OR only meaningful for additive burden (ADD)
  mutate(OR = if_else(TEST == "ADD" & is.finite(BETA), exp(BETA), NA_real_))

if(any(is.na(gene_based_all$LOG10P))) {
  warning(sprintf("Warning: %d gene-based tests have NA LOG10P values", sum(is.na(gene_based_all$LOG10P))))
}
if(any(is.infinite(gene_based_all$LOG10P))) {
  warning(sprintf("Warning: %d gene-based tests have infinite LOG10P values (p=0)", sum(is.infinite(gene_based_all$LOG10P))))
}
add_beta <- gene_based_all$BETA[gene_based_all$TEST == "ADD"]
if(length(add_beta) && any(is.finite(add_beta) & abs(add_beta) > 20)) {
  warning("Some ADD burden BETA values are very large (|BETA| > 20). Check Regenie column alignment (CHROM GENPOS ... BETA SE LOG10P).")
}

# REGENIE gene-based ID: GENE.MASK.<aaf_suffix> — suffix can contain dots (0.01, 0.001, …).
.regenie_split_gene_mask_thresh <- function(ids) {
  ids <- as.character(ids)
  n <- length(ids)
  g <- rep(NA_character_, n)
  m <- rep(NA_character_, n)
  tr <- rep(NA_character_, n)
  for (i in seq_len(n)) {
    p <- strsplit(ids[i], ".", fixed = TRUE)[[1L]]
    if (length(p) < 3L) next
    g[i] <- p[[1L]]
    m[i] <- p[[2L]]
    tr[i] <- paste(p[seq.int(3L, length(p))], collapse = ".")
  }
  list(GENE = g, MASK = m, THRESH_RAW = tr)
}
gene_id_parts <- .regenie_split_gene_mask_thresh(gene_based_all$ID)
gene_based_all <- gene_based_all %>%
  mutate(
    GENE = gene_id_parts$GENE,
    MASK = gene_id_parts$MASK,
    THRESH_RAW = gene_id_parts$THRESH_RAW,
    THRESH = as.character(ifelse(THRESH_RAW == "all", "All variants", THRESH_RAW)),
    MAF_Filter = THRESH
  )

# BH within family: ADD alone; SKAT+SKATO together; any other TEST each alone
.bh_within <- function(df) {
  if (nrow(df) == 0L) return(df)
  df %>%
    mutate(
      PVAL_ADJ = {
        ok <- is.finite(PVAL) & PVAL > 0 & PVAL <= 1
        out <- rep(NA_real_, n())
        if (any(ok)) out[ok] <- p.adjust(PVAL[ok], method = "BH")
        out
      },
      LOG10P_ADJ = -log10(PVAL_ADJ)
    )
}

gene_based_add <- gene_based_all %>% filter(TEST == "ADD") %>% .bh_within() %>% .add_or_ci()
gene_based_skat <- gene_based_all %>%
  filter(TEST %in% c("ADD-SKAT", "ADD-SKATO")) %>%
  .bh_within()
gene_based_other <- gene_based_all %>%
  filter(!TEST %in% c("ADD", "ADD-SKAT", "ADD-SKATO")) %>%
  group_by(TEST) %>%
  group_modify(~ .bh_within(.x)) %>%
  ungroup()

# Archive table carries family-specific FDR (not one mega-BH across all TESTs)
gene_based_all <- bind_rows(gene_based_add, gene_based_skat, gene_based_other)
```

## Summary and quality

### Analysis summary

```{r summary-stats,echo=FALSE,message=FALSE,warning=FALSE}
# Load phenotype file for actual case/control numbers
pheno_file <- paste0("data/final/", params$project_name, ".regenie.pheno.txt")
pheno <- read_delim(pheno_file, delim = " ", col_names = TRUE, trim_ws = TRUE)

# Build proper summary table with scalar values
summary_df <- data.frame(
  Metric = c("Total Samples", "Cases", "Controls",
             "Variants Tested (Single-Variant)",
             "Unique genes tested (all TEST rows)",
             "Unique genes tested (ADD only)"),
  Count = c(
    length(unique(pheno$IID)),
    sum(pheno$STATUS == 1),
    sum(pheno$STATUS == 0),
    nrow(single_variant),
    length(unique(gene_based_all$GENE)),
    length(unique(gene_based_add$GENE))
  )
)

kable(summary_df, format = "html", col.names = c("Metric", "Count")) %>%
  .style_report_kable(font_size = 13, full_width = FALSE, gene_col = 1L)
```

### Quality metrics

```{r quality-metrics,echo=FALSE,message=FALSE,warning=FALSE}
# Calculate lambda GC for both analyses (finite p in (0,1])
single_lambda <- {
  pv <- single_variant$PVAL[is.finite(single_variant$PVAL) & single_variant$PVAL > 0 & single_variant$PVAL < 1]
  if (length(pv) < 10L) NA_real_ else median(qchisq(1 - pv, 1), na.rm = TRUE) / qchisq(0.5, 1)
}
gene_lambda <- {
  pv <- gene_based_add$PVAL[gene_based_add$TEST == "ADD" & is.finite(gene_based_add$PVAL) & gene_based_add$PVAL > 0 & gene_based_add$PVAL < 1]
  if (length(pv) < 10L) NA_real_ else median(qchisq(1 - pv, 1), na.rm = TRUE) / qchisq(0.5, 1)
}

# Simple quality metrics (burden ADD only for gene-based rows)
sig_add_or <- gene_based_add$OR[gene_based_add$PVAL_ADJ < 0.05 & is.finite(gene_based_add$OR)]
max_or_sig <- if (length(sig_add_or)) round(max(sig_add_or, na.rm = TRUE), 2) else NA_real_

quality_df <- data.frame(
    Metric = c(
        "Significant Single-Variants (p < 2.5e-6)",
        "Significant Burden Tests (ADD, FDR < 0.05)",
        "Lambda GC (Single-Variant)",
        "Lambda GC (gene-based ADD only)",
        "Max OR (FDR-significant ADD only)"
    ),
    Value = c(
        format(sum(single_variant$LOG10P > 5.60206, na.rm = TRUE), big.mark = ","),
        format(sum(gene_based_add$PVAL_ADJ < 0.05, na.rm = TRUE), big.mark = ","),
        round(single_lambda, 3),
        round(gene_lambda, 3),
        ifelse(is.finite(max_or_sig), format(max_or_sig, nsmall = 2), "—")
    )
)

kable(quality_df, format = "html",
      col.names = c("Quality Metric", "Value")) %>%
    .style_report_kable(font_size = 13, full_width = FALSE, gene_col = 1L)
```

## Single-variant results

Step 2 tests all variants in the annotation pgen; gene-based M1–M4 use only `regenie.annotation.txt` / `regenie.set.txt`.

### Significant variants

```{r significant-single,echo=FALSE,message=FALSE,warning=FALSE,results='asis'}
sig_log10p_thresh <- 5.60206  # ~2.5e-6 exome-wide

# HTML kable output must use results='asis' or knitr wraps it as verbatim (shows as `## <div...` text).
fmt_small_p <- function(x) {
  x <- as.numeric(x)
  ifelse(!is.finite(x), "",
    ifelse(x < 1e-4, formatC(x, format = "e", digits = 2), format(round(x, 5), trim = TRUE))
  )
}

significant_variants <- single_variant %>%
  filter(!is.na(LOG10P), LOG10P > sig_log10p_thresh) %>%
  arrange(desc(LOG10P))

.annot_join <- function(ids_df) {
  # subset stats/annot for just these IDs (never load full mask table)
  stats <- .stats_for_ids(ids_df$ID)
  if (nrow(stats) == 0L) return(ids_df)
  ids_df %>% left_join(stats, by = "ID", suffix = c("", ".ann"))
}

if (nrow(significant_variants) == 0L) {
  cat("None above exome-wide threshold (LOG10P > ", sig_log10p_thresh, ").\n", sep = "")
  significant_variants_export <- significant_variants
} else {
  sig_joined <- .annot_join(significant_variants)
  sig_joined <- .fill_variant_carriers_from_afcc(sig_joined)
  if (!"Gene" %in% names(sig_joined)) sig_joined$Gene <- NA_character_
  if (!"HGVSc" %in% names(sig_joined)) sig_joined$HGVSc <- NA_character_
  if (!"HGVSp" %in% names(sig_joined)) sig_joined$HGVSp <- NA_character_
  if (!"ClinVar.SIG" %in% names(sig_joined)) sig_joined[["ClinVar.SIG"]] <- NA_character_
  if (!"AutoGVP" %in% names(sig_joined)) sig_joined$AutoGVP <- NA_character_
  if (!"n_case_carrier" %in% names(sig_joined)) sig_joined$n_case_carrier <- NA_integer_
  if (!"n_control_carrier" %in% names(sig_joined)) sig_joined$n_control_carrier <- NA_integer_

  sig_display <- sig_joined %>%
    mutate(
      Variant = paste0("chr", CHROM, ":", GENPOS, " ", REF, ">", ALT),
      Gene_lab = ifelse(is.na(Gene) | Gene == "", "—", as.character(Gene))
    ) %>%
    select(
      Variant, ID, Gene_lab, HGVSc, HGVSp, `ClinVar.SIG`, AutoGVP,
      n_case_carrier, n_control_carrier,
      A1FREQ, OR, OR_L95, OR_U95, LOG10P, PVAL, PVAL_ADJ
    ) %>%
    arrange(PVAL)

  significant_variants_export <- sig_display

  sig_view <- sig_display %>%
    mutate(PVAL = fmt_small_p(PVAL), PVAL_ADJ = fmt_small_p(PVAL_ADJ))

  knitr::kable(
    sig_view,
    format = "html",
    digits = 4,
    col.names = c(
      "Variant", "ID", "Gene", "HGVSc", "HGVSp", "ClinVar.SIG", "AutoGVP",
      "n case", "n ctrl",
      "Alt freq", "OR", "OR L95", "OR U95", "−log10(p)", "P", "FDR P"
    ),
    caption = paste0(
      "Exome-wide hits (n = ", nrow(sig_display),
      "). n case/ctrl = carriers from mask_variant_stats when available; ",
      "else ≈ alt allele count from REGENIE --af-cc (A1FREQ × 2 × N)."
    )
  ) %>%
    .style_report_kable(font_size = 12, full_width = TRUE, gene_col = 3L, scroll_h = "420px") %>%
    print()
}
```

### Manhattan plot

```{r manhattan-plot,echo=FALSE,message=FALSE,warning=FALSE, fig.cap="Single-variant Manhattan (y capped for display)."}
chrom_colors <- rep(c("#2E86AB", "#FF6B35"), length.out = length(unique(single_variant$CHROM)))
hi <- max(single_variant$LOG10P, na.rm = TRUE)
if (!is.finite(hi)) hi <- 10
ylim_u <- min(80, hi + 2)
n_exome_wide <- sum(single_variant$LOG10P > 5.60206, na.rm = TRUE)

manhattan_plot <- ggplot(single_variant, aes(x = POS_INDEX, y = LOG10P, color = CHROM)) +
    geom_point(alpha = 0.55, size = 1.35) +
    scale_color_manual(values = chrom_colors) +
    labs(x = "Genomic position", y = expression(-log[10](italic(p)))) +
    geom_hline(yintercept = 5.602, color = "red", linetype = "dashed", linewidth = 0.6) +
    coord_cartesian(ylim = c(0, ylim_u)) +
    labs(subtitle = paste0("Points above red line: exome-wide (n = ", n_exome_wide, ")")) +
    theme(legend.position = "none")

print(manhattan_plot)
```

### QQ plot

```{r qq-plot,echo=FALSE,message=FALSE,warning=FALSE, fig.cap="Single-variant QQ with genomic inflation (λ) in the subtitle."}
qq_n <- single_variant %>% filter(is.finite(LOG10P))
qq_data <- tibble(
  observed = sort(qq_n$LOG10P, decreasing = TRUE),
  expected = sort(-log10(ppoints(nrow(qq_n))), decreasing = TRUE)
)

lam_txt <- if (exists("single_lambda") && is.finite(single_lambda)) {
  paste0("λ = ", round(single_lambda, 3))
} else {
  "λ = NA (insufficient finite P-values)"
}

qq_plot <- ggplot(qq_data, aes(x = expected, y = observed)) +
    geom_point(alpha = 0.55, color = "#2E86AB") +
    geom_abline(intercept = 0, slope = 1, color = "red", linetype = "dashed") +
    labs(
      x = "Expected -log10(p)",
      y = "Observed -log10(p)",
      subtitle = lam_txt
    )

print(qq_plot)
```

Deviation from the diagonal at the tail reflects excess signal. **λ** is the median χ² (1 df) inflation vs null (same idea as in the QC report); it is also listed under **Quality metrics**.

## Gene-based results

```{=html}
<div class="regenie-jump">
<strong>Jump to mask results:</strong>
<a href="#m1-pathogenic">M1</a>
<a href="#m2-vus">M2</a>
<a href="#m3-pathogenic--vus">M3</a>
<a href="#m4-synonymous-negative-control">M4</a>
&nbsp;|&nbsp;
<a href="#variants-in-burden-masks-for-top-genes-add">Mask variants</a>
<a href="#chek2-add">CHEK2</a>
<a href="#volcano-add">Volcano</a>
</div>
```

Exports: **`r paste0(params$project_name, "_gene_based_results_alltests_", format(Sys.Date(), "%Y%m%d"), ".csv")`** (archive), **`r paste0(params$project_name, "_gene_based_results_ADDonly_", format(Sys.Date(), "%Y%m%d"), ".csv")`** (directional burden), **`r paste0(params$project_name, "_gene_based_results_SKAT_SKATO_", format(Sys.Date(), "%Y%m%d"), ".csv")`** (non-directional; no BETA/OR), and **`r paste0(params$project_name, "_gene_based_meta_input_", format(Sys.Date(), "%Y%m%d"), ".csv")`**. FDR is within family (ADD vs SKAT+SKATO).

### Burden (ADD) — summary by mask

```{r mask-summary,echo=FALSE,message=FALSE,warning=FALSE}
mask_summary <- gene_based_add %>%
  group_by(MASK) %>%
  summarize(
    Tests = n(),
    Significant_Tests = sum(PVAL_ADJ < 0.05, na.rm = TRUE),
    Most_Significant_Gene = GENE[which.max(LOG10P)],
    Max_Log10P = max(pmin(LOG10P, 100), na.rm = TRUE),
    Mean_Beta = mean(BETA, na.rm = TRUE)
  ) %>%
  mutate(Significance_Rate = Significant_Tests / Tests * 100)

kable(mask_summary, format = "html", digits = 3,
      col.names = c("Mask", "Tests (gene×mask×threshold)", "FDR significant tests", "Top Gene", "Max −log10(P) (display cap 100)", "Mean Beta", "Significance Rate (%)")) %>%
  .style_report_kable(font_size = 13, full_width = FALSE, gene_col = 1L)
```

**Mask note:** `M4` is a negative-control mask built from synonymous variants only.
It explicitly excludes `M1/M2` variants and a blacklist of potential MNP-pair IDs that did not pass full validation, to reduce false-negative MNP contamination in controls.

**Gene×consequence excludes:** sites matching rules in `mask_gene_consequence_exclude.txt` are dropped from **all** masks at REGENIE preprocess (annotation/set). Current rule: **CHEK2 + missense_variant** (including compound VEP consequences that contain missense). Rationale: CHEK2 missense variants are not treated as disease-contributing in this analysis despite some AutoGVP likely-pathogenic calls. Dropped IDs are written to `data/regenie/mask_gene_consequence_excluded.csv`.

### Top genes by clinical category (burden ADD only)

**Sorted by:** gene ADD **−log10(P)** (highest first) within each mask × AAF bin.

**HTML:** top **10** genes for **0.01**, **0.001**, and **singleton**.

**CSV:** top **20** genes for **0.01, 0.001, 0.0001, singleton, All variants** (All variants is CSV-only, not shown in HTML).

| Bin | Meaning |
|-----|---------|
| **0.01** | AAF ≤ 1% |
| **0.001** | AAF ≤ 0.1% |
| **0.0001** | AAF ≤ 0.01% (CSV) |
| **singleton** | MAC=1 sites burdened together — **not** single-variant GWAS |
| **All variants** | no AAF ceiling (CSV only) |

OR 95% CI = `exp(BETA ± 1.96×SE)`. Cases/Controls = unique samples carrying ≥1 mask variant in the gene (from prebuilt `*.report.gene_carriers.tsv`).

Prefer prebuilt tables from `build_regenie_report_tables.py` (`data/final/{project}.report.*.tsv`) when present — report only displays.

```{r topgenes-add-helper, include=FALSE}
HTML_THRESH <- c("0.01", "0.001", "singleton")
CSV_THRESH <- c("0.01", "0.001", "0.0001", "singleton", "All variants")

.qq_gene_add <- function(mask, threshs = HTML_THRESH) {
  # One QQ per AAF bin for this mask (ADD only), faceted — λ GC labeled on each panel
  dat <- gene_based_add %>%
    filter(MASK == mask, THRESH %in% threshs, is.finite(PVAL), PVAL > 0, PVAL <= 1) %>%
    mutate(THRESH = factor(THRESH, levels = threshs))
  if (nrow(dat) == 0L) {
    cat("_No ADD p-values for QQ: ", mask, "._\n\n", sep = "")
    return(invisible(NULL))
  }
  lam <- dat %>%
    group_by(THRESH) %>%
    summarise(
      n_tests = n(),
      lambda = {
        pv <- PVAL
        if (length(pv) < 10L) {
          NA_real_
        } else {
          median(qchisq(1 - pv, 1), na.rm = TRUE) / qchisq(0.5, 1)
        }
      },
      .groups = "drop"
    ) %>%
    mutate(
      lam_lab = ifelse(is.finite(lambda), sprintf("%.3f", lambda), "NA"),
      facet_lab = sprintf("%s  |  λ = %s", as.character(THRESH), lam_lab)
    )
  panel_levels <- lam$facet_lab[match(threshs, as.character(lam$THRESH))]
  panel_levels <- panel_levels[!is.na(panel_levels)]
  qq <- dat %>%
    left_join(lam %>% select(THRESH, facet_lab), by = "THRESH") %>%
    mutate(facet_lab = factor(facet_lab, levels = panel_levels)) %>%
    group_by(facet_lab) %>%
    arrange(PVAL, .by_group = TRUE) %>%
    mutate(
      expected = ppoints(n()),
      observed = PVAL,
      neglog10_exp = -log10(expected),
      neglog10_obs = -log10(observed)
    ) %>%
    ungroup()
  lim <- max(qq$neglog10_exp, qq$neglog10_obs, na.rm = TRUE)
  lim <- max(lim, 1)
  cat(sprintf(
    "\n**ADD burden QQ: %s** (one panel per AAF bin; λ = genomic inflation from median χ²)\n\n",
    mask
  ))
  print(
    ggplot(qq, aes(x = neglog10_exp, y = neglog10_obs)) +
      geom_abline(slope = 1, intercept = 0, color = "grey50", linetype = "dashed") +
      geom_point(size = 1.35, alpha = 0.65, color = "#2E86AB") +
      facet_wrap(~facet_lab, nrow = 1) +
      coord_equal(xlim = c(0, lim), ylim = c(0, lim)) +
      labs(
        x = expression(Expected ~ -log[10](p)),
        y = expression(Observed ~ -log[10](p)),
        title = paste0("Gene-based ADD QQ — ", mask)
      ) +
      theme(
        legend.position = "none",
        strip.text = element_text(size = 12, face = "bold"),
        plot.title = element_text(size = 14, face = "bold"),
        axis.title = element_text(size = 12),
        axis.text = element_text(size = 11)
      )
  )
  invisible(NULL)
}

regenie_topgenes_add_table <- function(mask, thresh, n = 10L) {
  if (!is.null(report_top_genes) && nrow(report_top_genes) > 0) {
    tab <- report_top_genes %>%
      filter(MASK == mask, THRESH == thresh) %>%
      arrange(rank) %>%
      slice_head(n = n) %>%
      select(GENE, MASK, THRESH, BETA, SE, OR, OR_L95, OR_U95, LOG10P, PVAL, PVAL_ADJ)
  } else {
    tab <- gene_based_add %>%
      filter(MASK == mask, THRESH == thresh) %>%
      slice_max(LOG10P, n = n, with_ties = FALSE) %>%
      select(GENE, MASK, THRESH, BETA, SE, OR, OR_L95, OR_U95, LOG10P, PVAL, PVAL_ADJ) %>%
      arrange(desc(LOG10P))
  }
  if (nrow(tab) == 0L) return(invisible(NULL))
  carr <- report_gene_carriers %>%
    filter(MASK == mask, THRESH == thresh) %>%
    select(GENE, Cases = n_case_carriers, Controls = n_control_carriers)
  tab <- tab %>%
    left_join(carr, by = "GENE") %>%
    transmute(
      Gene = GENE,
      Cases, Controls,
      Beta = BETA, SE, OR, `OR L95` = OR_L95, `OR U95` = OR_U95,
      `-log10(P)` = LOG10P, P = PVAL, `FDR P` = PVAL_ADJ
    )
  cat(sprintf("\n**Top %d genes by ADD −log10(P): %s — %s**\n\n", n, mask, thresh))
  knitr::kable(tab, format = "html", digits = 4) %>%
    .style_report_kable(font_size = 13, full_width = FALSE, gene_col = 1L)
}
```

#### M1 (Pathogenic)

##### M1 — 0.01

```{r topgenes-m1-af01, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
regenie_topgenes_add_table("M1", "0.01")
```

##### M1 — 0.001

```{r topgenes-m1-af001, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
regenie_topgenes_add_table("M1", "0.001")
```

##### M1 — singleton

```{r topgenes-m1-singleton, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
regenie_topgenes_add_table("M1", "singleton")
```

#### M2 (VUS)

##### M2 — 0.01

```{r topgenes-m2-af01, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
regenie_topgenes_add_table("M2", "0.01")
```

##### M2 — 0.001

```{r topgenes-m2-af001, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
regenie_topgenes_add_table("M2", "0.001")
```

##### M2 — singleton

```{r topgenes-m2-singleton, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
regenie_topgenes_add_table("M2", "singleton")
```

#### M3 (Pathogenic + VUS)

##### M3 — 0.01

```{r topgenes-m3-af01, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
regenie_topgenes_add_table("M3", "0.01")
```

##### M3 — 0.001

```{r topgenes-m3-af001, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
regenie_topgenes_add_table("M3", "0.001")
```

##### M3 — singleton

```{r topgenes-m3-singleton, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
regenie_topgenes_add_table("M3", "singleton")
```

#### M4 (Synonymous negative-control)

##### M4 — 0.01

```{r topgenes-m4-af01, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
regenie_topgenes_add_table("M4", "0.01")
```

##### M4 — 0.001

```{r topgenes-m4-af001, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
regenie_topgenes_add_table("M4", "0.001")
```

##### M4 — singleton

```{r topgenes-m4-singleton, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
regenie_topgenes_add_table("M4", "singleton")
```

### Variants in burden masks for top genes (ADD)

Snplist-exact membership from `data/regenie/chr*.step2_gene_based_masks.snplist`. HTML: top 10 × 0.01/0.001/singleton. CSV: top 20 × 0.01/0.001/0.0001/singleton/All variants.

```{r variant-contrib-helper, include=FALSE}
top_genes_for <- function(mask, thresh, n = 10L) {
  if (!is.null(report_top_genes) && nrow(report_top_genes) > 0) {
    return(
      report_top_genes %>%
        filter(MASK == mask, THRESH == thresh) %>%
        arrange(rank) %>%
        slice_head(n = n) %>%
        transmute(
          GENE,
          gene_LOG10P = LOG10P,
          gene_PVAL_ADJ = PVAL_ADJ,
          gene_BETA = BETA,
          gene_OR = OR,
          gene_OR_L95 = OR_L95,
          gene_OR_U95 = OR_U95,
          gene_A1FREQ = A1FREQ,
          gene_A1FREQ_CASES = A1FREQ_CASES,
          gene_A1FREQ_CONTROLS = A1FREQ_CONTROLS
        )
    )
  }
  gene_based_add %>%
    filter(MASK == mask, THRESH == thresh) %>%
    slice_max(LOG10P, n = n, with_ties = FALSE) %>%
    arrange(desc(LOG10P)) %>%
    transmute(
      GENE,
      gene_LOG10P = LOG10P,
      gene_PVAL_ADJ = PVAL_ADJ,
      gene_BETA = BETA,
      gene_OR = OR,
      gene_OR_L95 = OR_L95,
      gene_OR_U95 = OR_U95,
      gene_A1FREQ = A1FREQ,
      gene_A1FREQ_CASES = A1FREQ_CASES,
      gene_A1FREQ_CONTROLS = A1FREQ_CONTROLS
    )
}

.enrich_ids <- function(df) {
  stats <- .stats_for_ids(df$ID)
  if (nrow(stats) == 0L) return(df)
  keep <- intersect(names(stats), c(
    "ID", "Gene", "Variant.LoF_level", "Variant.Class", "Variant.Consequence",
    "HGVSc", "HGVSp", "ClinVar.SIG", "AutoGVP",
    "n_case_carrier", "n_control_carrier", "AF", "AC"
  ))
  df %>% left_join(stats %>% select(all_of(keep)), by = "ID")
}

contrib_table <- function(mask, thresh, n = 10L) {
  if (!is.null(report_contrib) && nrow(report_contrib) > 0) {
    tg <- top_genes_for(mask, thresh, n)
    if (nrow(tg) == 0L) return(tibble())
    return(
      report_contrib %>%
        filter(MASK == mask, THRESH == thresh, GENE %in% tg$GENE) %>%
        arrange(desc(gene_LOG10P), GENE, ID)
    )
  }
  tg <- top_genes_for(mask, thresh, n)
  if (nrow(tg) == 0L || nrow(mask_snplist) == 0L) return(tibble())
  mem <- mask_snplist %>%
    filter(MASK == mask, THRESH == thresh, GENE %in% tg$GENE) %>%
    select(GENE, ID, MASK, THRESH)
  if (nrow(mem) == 0L) return(tibble())
  mem %>%
    left_join(tg, by = "GENE") %>%
    .enrich_ids() %>%
    arrange(desc(gene_LOG10P), GENE, ID)
}

contrib_consequence_matrix <- function(mask, thresh, tab) {
  # Split VEP compound consequences on &; each atom gets +1 (e.g. missense&splice → both)
  if (nrow(tab) == 0L || !"Variant.Consequence" %in% names(tab)) return(tibble())
  rows <- tab %>%
    filter(!is.na(Variant.Consequence), Variant.Consequence != "", Variant.Consequence != ".") %>%
    select(GENE, cons = Variant.Consequence)
  if (nrow(rows) == 0L) return(tibble())
  rows %>%
    mutate(atom = strsplit(as.character(cons), "&", fixed = TRUE)) %>%
    tidyr::unnest_longer(atom) %>%
    mutate(atom = trimws(atom)) %>%
    filter(nzchar(atom)) %>%
    count(GENE, atom, name = "n") %>%
    tidyr::pivot_wider(names_from = atom, values_from = n, values_fill = 0)
}

render_contrib <- function(mask, thresh, n = 10L) {
  cat(sprintf(
    '<div class="regenie-block-sep"><p style="margin:0 0 0.4rem 0; font-weight:600;">%s — %s</p></div>\n',
    mask, thresh
  ))
  tab <- contrib_table(mask, thresh, n = n)
  if (nrow(tab) == 0L) {
    cat("_No snplist variants for ", mask, " — ", thresh, "._\n\n", sep = "")
    return(invisible(NULL))
  }
  if (!"HGVSc" %in% names(tab)) tab$HGVSc <- NA_character_
  if (!"HGVSp" %in% names(tab)) tab$HGVSp <- NA_character_
  if (!"Variant.Consequence" %in% names(tab)) tab$Variant.Consequence <- NA_character_
  if (!"ClinVar.SIG" %in% names(tab)) tab[["ClinVar.SIG"]] <- NA_character_
  if (!"AutoGVP" %in% names(tab)) tab$AutoGVP <- NA_character_
  if (!"n_case_carrier" %in% names(tab)) tab$n_case_carrier <- NA_integer_
  if (!"n_control_carrier" %in% names(tab)) tab$n_control_carrier <- NA_integer_

  show <- tab %>%
    transmute(
      Gene = GENE, ID, HGVSc, HGVSp,
      Consequence = .fmt_consequence_label(Variant.Consequence),
      `ClinVar.SIG`, AutoGVP,
      n_case = n_case_carrier, n_ctrl = n_control_carrier,
      `Gene -log10(P)` = round(gene_LOG10P, 3),
      `Gene FDR` = signif(gene_PVAL_ADJ, 3),
      `Gene OR` = gene_OR
    )
  cat(sprintf(
    "\n**Mask variants (snplist): %s — %s** (top %d genes by ADD −log10(P))\n\n",
    mask, thresh, n
  ))
  print(
    knitr::kable(show, format = "html", digits = 4) %>%
      .style_report_kable(font_size = 12, full_width = TRUE, gene_col = 1L, scroll_h = "360px")
  )

  cat('<div style="height:1.6rem;"></div>\n')
  mx <- contrib_consequence_matrix(mask, thresh, tab)
  if (nrow(mx) > 0) {
    cons_cols <- setdiff(names(mx), "GENE")
    if (length(cons_cols)) {
      names(mx)[names(mx) %in% cons_cols] <- .fmt_consequence_header(cons_cols)
    }
    names(mx)[names(mx) == "GENE"] <- "Gene"
    cat(sprintf("\n**Consequence counts: %s — %s**\n\n", mask, thresh))
    cat("Compound VEP terms split on `&`; each term counted.\n\n")
    print(
      knitr::kable(mx, format = "html") %>%
        .style_report_kable(
          font_size = 12, full_width = TRUE, gene_col = 1L,
          header_wrap = TRUE, scroll_h = "320px"
        )
    )
  }
}
```

#### M1 mask variants (HTML: 0.01 / 0.001 / singleton)

```{r qq-m1-add, echo=FALSE, message=FALSE, warning=FALSE, fig.width=12.5, fig.height=4.8, results='asis'}
.qq_gene_add("M1")
```

```{r variant-contrib-m1, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
for (th in HTML_THRESH) render_contrib("M1", th, n = 10L)
```

#### M2 mask variants (HTML: 0.01 / 0.001 / singleton)

```{r qq-m2-add, echo=FALSE, message=FALSE, warning=FALSE, fig.width=12.5, fig.height=4.8, results='asis'}
.qq_gene_add("M2")
```

```{r variant-contrib-m2, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
for (th in HTML_THRESH) render_contrib("M2", th, n = 10L)
```

#### M3 mask variants (HTML: 0.01 / 0.001 / singleton)

```{r qq-m3-add, echo=FALSE, message=FALSE, warning=FALSE, fig.width=12.5, fig.height=4.8, results='asis'}
.qq_gene_add("M3")
```

```{r variant-contrib-m3, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
for (th in HTML_THRESH) render_contrib("M3", th, n = 10L)
```

#### M4 mask variants (HTML: 0.01 / 0.001 / singleton)

```{r qq-m4-add, echo=FALSE, message=FALSE, warning=FALSE, fig.width=12.5, fig.height=4.8, results='asis'}
.qq_gene_add("M4")
```

```{r variant-contrib-m4, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
for (th in HTML_THRESH) render_contrib("M4", th, n = 10L)
```

### Suggestive ADD burden signals

_Exploratory table: strong raw p but not FDR-significant after BH among ADD tests only. Cases/Controls = unique snplist carriers (same as top-gene tables)._

```{r suggestive-genes,echo=FALSE,message=FALSE,warning=FALSE}
top_by_raw_p <- gene_based_add %>%
  group_by(MASK) %>%
  slice_max(LOG10P, n = 10, with_ties = FALSE) %>%
  ungroup()

suggestive_genes <- top_by_raw_p %>%
  filter(LOG10P > 3, PVAL_ADJ >= 0.05) %>%
  arrange(PVAL) %>%
  slice_head(n = 15) %>%
  select(GENE, MASK, THRESH, BETA, LOG10P, PVAL, PVAL_ADJ, OR) %>%
  .join_gene_carriers() %>%
  select(GENE, MASK, THRESH, Cases, Controls, BETA, LOG10P, PVAL, PVAL_ADJ, OR)

if(nrow(suggestive_genes) > 0) {
    kable(suggestive_genes,
          format = "html", digits = 4,
          caption = "Suggestive ADD burden signals (exploratory)",
          col.names = c("Gene", "Mask", "MAF Filter", "Cases", "Controls",
                        "Beta", "Raw -log10(P)", "Raw P-value", "FDR P-value", "OR")) %>%
        .style_report_kable(font_size = 12, full_width = FALSE, gene_col = 1L) %>%
        column_spec(c(6, 10), color = "orange")
} else {
    cat("No suggestive associations found")
}
```

### Volcano (ADD) {#volcano-add}

```{r volcano-plot,echo=FALSE,message=FALSE,warning=FALSE, fig.cap="Burden ADD by mask (−log10 p capped at 100 in plot)."}
gene_based_plot <- gene_based_add %>%
  mutate(
    Significant = PVAL_ADJ < 0.05,
    LOG10P_CAP = pmin(LOG10P, 100)
  )

top_hits <- gene_based_plot %>%
  filter(Significant) %>%
  group_by(MASK) %>%
  slice_max(LOG10P_CAP, n = 3, with_ties = FALSE) %>%
  ungroup()

gene_based_plot <- gene_based_plot %>%
  mutate(Label = ifelse(GENE %in% top_hits$GENE, GENE, NA))

fdr_cutoff <- suppressWarnings(min(gene_based_add$LOG10P[gene_based_add$PVAL_ADJ < 0.05], na.rm = TRUE))
if (!is.finite(fdr_cutoff)) fdr_cutoff <- NA_real_
fdr_cutoff_plot <- if (is.finite(fdr_cutoff)) pmin(fdr_cutoff, 100) else NA_real_

volcano_plot <- ggplot(gene_based_plot, aes(x = as.numeric(BETA), y = LOG10P_CAP, color = Significant)) +
  geom_point(alpha = 0.55, size = 1.35)
if (is.finite(fdr_cutoff_plot)) {
  volcano_plot <- volcano_plot +
    geom_hline(yintercept = fdr_cutoff_plot, linetype = "dashed", color = "blue", alpha = 0.65)
}
volcano_plot <- volcano_plot +
  geom_vline(xintercept = 0, linetype = "dotted", color = "grey40") +
  geom_text_repel(aes(label = Label), size = 2.6, max.overlaps = 12,
                  box.padding = 0.3, point.padding = 0.1) +
  scale_color_manual(values = c("gray60", "red"),
                     labels = c("Not Significant", "FDR < 0.05")) +
  scale_x_continuous(limits = c(NA, NA), oob = scales::oob_keep) +
  labs(
       x = "Beta (log-OR)",
       y = "-log10(p), capped at 100",
       color = NULL) +
  facet_wrap(~MASK, scales = "free_y") +
  theme(legend.position = "bottom",
        strip.text = element_text(size = 11, face = "bold"))

print(volcano_plot)
```

### CHEK2 (ADD)

Gene-level ADD burden for **M1 (pathogenic) only**, then M1 snplist variants with HGVSc/p and case/control **carrier** counts. M2/M3/M4 CHEK2 results are omitted here for now.

**Note:** CHEK2 **missense** sites are excluded from all masks upstream (`mask_gene_consequence_exclude.txt`), so they should not appear in this snplist table.

```{r chek2-analysis, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
.load_snplist_for_gene <- function(gene) {
  files <- Sys.glob("data/regenie/chr*.step2_gene_based_masks.snplist")
  if (!length(files)) return(tibble())
  prefix <- paste0(gene, ".")
  rows <- list()
  for (fp in files) {
    lines <- readLines(fp, warn = FALSE)
    for (ln in lines) {
      ln <- trimws(ln)
      if (!nzchar(ln) || startsWith(ln, "#") || !startsWith(ln, prefix)) next
      parts <- strsplit(ln, "\t", fixed = TRUE)[[1]]
      if (length(parts) < 2L) next
      mid <- parts[[1]]
      vids <- strsplit(parts[[2]], ",", fixed = TRUE)[[1]]
      vids <- vids[nzchar(vids)]
      if (!length(vids)) next
      sp <- strsplit(mid, ".", fixed = TRUE)[[1]]
      if (length(sp) < 3L) next
      thr_raw <- paste(sp[seq.int(3L, length(sp))], collapse = ".")
      rows[[length(rows) + 1L]] <- tibble(
        GENE = sp[[1]], MASK = sp[[2]],
        THRESH = ifelse(thr_raw == "all", "All variants", thr_raw),
        ID = vids
      )
    }
  }
  if (!length(rows)) return(tibble())
  bind_rows(rows)
}

chek2_results <- gene_based_add %>%
  filter(GENE == "CHEK2", MASK == "M1") %>%
  select(GENE, MASK, THRESH, A1FREQ, BETA, OR, OR_L95, OR_U95, SE, LOG10P, PVAL, PVAL_ADJ) %>%
  arrange(MASK, THRESH)

if (nrow(chek2_results) > 0) {
  print(
    knitr::kable(
      chek2_results, format = "html", digits = 4,
      col.names = c(
        "Gene", "Mask", "Threshold", "Alt freq",
        "Beta", "OR", "OR L95", "OR U95", "SE", "-log10(P)", "P", "FDR P"
      ),
      caption = "CHEK2 ADD burden (M1 pathogenic only)"
    ) %>%
      .style_report_kable(font_size = 12, full_width = FALSE, gene_col = 1L)
  )
} else {
  cat("_CHEK2 M1 not found in ADD burden results._\n")
}

cat('<div style="height:1.2rem;"></div>\n')

chek2_mem <- .load_snplist_for_gene("CHEK2")
chek2_html_bins <- c("0.01", "0.001")
if (nrow(chek2_mem) > 0) {
  chek2_mem <- chek2_mem %>% filter(MASK == "M1", THRESH %in% chek2_html_bins)
}
chek2_var <- tibble()
if (nrow(chek2_mem) > 0) {
  gene_join <- chek2_results %>%
    select(GENE, MASK, THRESH, gene_LOG10P = LOG10P, gene_PVAL_ADJ = PVAL_ADJ, gene_OR = OR)
  chek2_var <- chek2_mem %>%
    left_join(gene_join, by = c("GENE", "MASK", "THRESH")) %>%
    .enrich_ids() %>%
    arrange(MASK, THRESH, ID)
}

if (nrow(chek2_var) == 0L) {
  cat("_No CHEK2 M1 snplist variants found for HTML bins 0.01 / 0.001._\n")
} else {
  if (!"HGVSc" %in% names(chek2_var)) chek2_var$HGVSc <- NA_character_
  if (!"HGVSp" %in% names(chek2_var)) chek2_var$HGVSp <- NA_character_
  if (!"Variant.Consequence" %in% names(chek2_var)) chek2_var$Variant.Consequence <- NA_character_
  if (!"ClinVar.SIG" %in% names(chek2_var)) chek2_var[["ClinVar.SIG"]] <- NA_character_
  if (!"AutoGVP" %in% names(chek2_var)) chek2_var$AutoGVP <- NA_character_
  if (!"n_case_carrier" %in% names(chek2_var)) chek2_var$n_case_carrier <- NA_integer_
  if (!"n_control_carrier" %in% names(chek2_var)) chek2_var$n_control_carrier <- NA_integer_

  chek2_show <- chek2_var %>%
    transmute(
      Mask = MASK, Bin = THRESH, ID, HGVSc, HGVSp,
      Consequence = .fmt_consequence_label(Variant.Consequence),
      `ClinVar.SIG`, AutoGVP,
      n_case = n_case_carrier, n_ctrl = n_control_carrier,
      `Gene -log10(P)` = round(gene_LOG10P, 3),
      `Gene FDR` = signif(gene_PVAL_ADJ, 3),
      `Gene OR` = gene_OR
    )
  print(
    knitr::kable(
      chek2_show, format = "html", digits = 4,
      caption = "CHEK2 M1 mask variants (snplist): HGVSc/p and case/control carriers"
    ) %>%
      .style_report_kable(font_size = 12, full_width = TRUE, gene_col = NULL, scroll_h = "420px")
  )
}

cat('<div style="height:1.4rem;"></div>\n')
```

<!-- CHEK2 PCA (parked — set eval=TRUE and uncomment markdown to restore)
### CHEK2 carriers in PCA space

Samples carrying any **CHEK2 M1 (pathogenic)** mask variant (AAF bins 0.01 and 0.001 only). PCs / status / AncGroup from hardcoded `sample_meta.csv`.
-->

```{r chek2-pca, eval=FALSE, echo=FALSE, message=FALSE, warning=FALSE, fig.width=10, fig.height=4.5, results='asis'}
.sample_meta_fp <- "sample_meta.csv"
.pc_cols <- c(
  "PC1_build", "PC2_build", "PC3_build", "PC4_build",
  "PC1_pmbb", "PC2_pmbb", "PC3_pmbb", "PC4_pmbb"
)
.meta_need <- c("IID", "STATUS", "AncGroup", .pc_cols)
if (!file.exists(.sample_meta_fp)) stop("Missing required file: ", .sample_meta_fp)
if (nrow(report_chek2_carriers) == 0L) {
  stop("CHEK2 carrier table is empty: ", .pre_chek2_carr_fp)
}

sample_meta <- read_csv(.sample_meta_fp, show_col_types = FALSE)
.meta_miss <- setdiff(.meta_need, names(sample_meta))
if (length(.meta_miss)) {
  stop("sample_meta.csv missing columns: ", paste(.meta_miss, collapse = ", "))
}

carr_iids <- report_chek2_carriers %>%
  mutate(IID = as.character(IID)) %>%
  distinct(IID)

meta <- sample_meta %>%
  mutate(
    IID = as.character(IID),
    STATUS = as.integer(STATUS),
    Status = ifelse(STATUS == 1L, "Case", "Control"),
    AncGroup = as.character(AncGroup),
    across(all_of(.pc_cols), as.numeric)
  ) %>%
  select(all_of(.meta_need), Status)

plot_df <- carr_iids %>% left_join(meta, by = "IID")

missing_meta <- plot_df$IID[is.na(plot_df$STATUS)]
missing_pcs <- plot_df$IID[
  !is.na(plot_df$STATUS) &
    (is.na(plot_df$PC1_build) | is.na(plot_df$PC1_pmbb))
]
if (length(missing_meta)) {
  stop(
    length(missing_meta), " CHEK2 carrier IID(s) not in sample_meta.csv. Examples: ",
    paste(head(missing_meta, 10), collapse = ", ")
  )
}
if (length(missing_pcs)) {
  stop(
    length(missing_pcs), " CHEK2 carrier IID(s) are in sample_meta.csv but PC1_build and/or PC1_pmbb is NA. ",
    "Those samples were likely missing from a PC source when sample_meta was built. Examples: ",
    paste(head(missing_pcs, 10), collapse = ", ")
  )
}
if (!any(nzchar(trimws(plot_df$AncGroup[!is.na(plot_df$AncGroup)])))) {
  stop("CHEK2 carriers have no AncGroup labels in sample_meta.csv")
}

cat(sprintf(
  "CHEK2 carriers in `sample_meta.csv`: **%d** (%d cases, %d controls).\n\n",
  nrow(plot_df),
  sum(plot_df$STATUS == 1L, na.rm = TRUE),
  sum(plot_df$STATUS == 0L, na.rm = TRUE)
))
cat(sprintf(
  "Unique PC values among carriers — build: PC1=%d PC2=%d; pmbb: PC1=%d PC2=%d.\n\n",
  dplyr::n_distinct(plot_df$PC1_build), dplyr::n_distinct(plot_df$PC2_build),
  dplyr::n_distinct(plot_df$PC1_pmbb), dplyr::n_distinct(plot_df$PC2_pmbb)
))
if (dplyr::n_distinct(plot_df$PC1_pmbb) < max(5L, floor(nrow(plot_df) / 5))) {
  cat(paste0(
    "_Note: `PC*_pmbb` look heavily rounded/quantized (few unique PC1 values), ",
    "which produces vertical stacks. Re-export PMBB PCs with more decimal places if available._\n\n"
  ))
}

.plot_chek2_pcs <- function(df, x, y, color_col, title, xlab, ylab) {
  ggplot(df, aes(x = .data[[x]], y = .data[[y]], color = .data[[color_col]])) +
    geom_point(size = 2.4, alpha = 0.85) +
    labs(title = title, x = xlab, y = ylab, color = NULL) +
    theme(legend.position = "bottom")
}

# Build PCs
print(.plot_chek2_pcs(plot_df, "PC1_build", "PC2_build", "Status",
                      "CHEK2: PC1_build vs PC2_build (case / control)", "PC1_build", "PC2_build") +
        scale_color_manual(values = c(Case = "#c0392b", Control = "#2980b9")))
print(.plot_chek2_pcs(plot_df, "PC3_build", "PC4_build", "Status",
                      "CHEK2: PC3_build vs PC4_build (case / control)", "PC3_build", "PC4_build") +
        scale_color_manual(values = c(Case = "#c0392b", Control = "#2980b9")))
print(.plot_chek2_pcs(plot_df, "PC1_build", "PC2_build", "AncGroup",
                      "CHEK2: PC1_build vs PC2_build (AncGroup)", "PC1_build", "PC2_build"))
print(.plot_chek2_pcs(plot_df, "PC3_build", "PC4_build", "AncGroup",
                      "CHEK2: PC3_build vs PC4_build (AncGroup)", "PC3_build", "PC4_build"))

# PMBB PCs
print(.plot_chek2_pcs(plot_df, "PC1_pmbb", "PC2_pmbb", "Status",
                      "CHEK2: PC1_pmbb vs PC2_pmbb (case / control)", "PC1_pmbb", "PC2_pmbb") +
        scale_color_manual(values = c(Case = "#c0392b", Control = "#2980b9")))
print(.plot_chek2_pcs(plot_df, "PC3_pmbb", "PC4_pmbb", "Status",
                      "CHEK2: PC3_pmbb vs PC4_pmbb (case / control)", "PC3_pmbb", "PC4_pmbb") +
        scale_color_manual(values = c(Case = "#c0392b", Control = "#2980b9")))
print(.plot_chek2_pcs(plot_df, "PC1_pmbb", "PC2_pmbb", "AncGroup",
                      "CHEK2: PC1_pmbb vs PC2_pmbb (AncGroup)", "PC1_pmbb", "PC2_pmbb"))
print(.plot_chek2_pcs(plot_df, "PC3_pmbb", "PC4_pmbb", "AncGroup",
                      "CHEK2: PC3_pmbb vs PC4_pmbb (AncGroup)", "PC3_pmbb", "PC4_pmbb"))
```

### SKAT and SKATO — non-directional supporting tests (no OR)

These tests ask whether rare variants in a gene associate with disease **without** assuming a shared effect direction. Mixed risk/protective effects can still produce a small p-value. **There is no signed BETA/OR to interpret as risk vs protective** — use ADD burden for direction.

FDR (`PVAL_ADJ`) here is Benjamini–Hochberg among SKAT+SKATO rows only.

**Cases/Controls** = unique samples carrying ≥1 variant in that gene’s mask snplist (same definition as ADD top-gene tables). REGENIE’s own `N_CASES`/`N_CONTROLS` are cohort sizes in the model (nearly constant) and are not shown.

```{r gene-based-skat-skato-top, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
.fmt_small_p <- function(x) {
  x <- as.numeric(x)
  ifelse(!is.finite(x), "",
    ifelse(x < 1e-4, formatC(x, format = "e", digits = 2), format(round(x, 5), trim = TRUE))
  )
}

if (!is.null(report_top_skat) && nrow(report_top_skat) > 0) {
  skat_skato <- report_top_skat %>%
    arrange(PVAL_ADJ, desc(LOG10P)) %>%
    slice_head(n = 30L) %>%
    select(any_of(c("GENE", "TEST", "MASK", "THRESH", "PVAL", "PVAL_ADJ", "LOG10P")))
} else {
  skat_skato <- gene_based_skat %>%
    arrange(PVAL_ADJ, PVAL) %>%
    slice_head(n = 30L) %>%
    select(GENE, TEST, MASK, THRESH, PVAL, PVAL_ADJ, LOG10P)
}

if (nrow(skat_skato) == 0L) {
  cat("_No SKAT/SKATO rows found (`TEST` in `ADD-SKAT`, `ADD-SKATO`). Confirm gene-based Step 2 outputs include those tests._\n")
} else {
  skat_view <- skat_skato %>%
    .join_gene_carriers() %>%
    mutate(
      PVAL = .fmt_small_p(PVAL),
      PVAL_ADJ = .fmt_small_p(PVAL_ADJ)
    ) %>%
    select(GENE, TEST, MASK, THRESH, Cases, Controls, PVAL, PVAL_ADJ, LOG10P)

  knitr::kable(
    skat_view,
    format = "html",
    digits = 4,
    col.names = c(
      "Gene", "Test", "Mask", "MAF filter", "Cases", "Controls",
      "P-value", "FDR P-value", "−log10(P)"
    ),
    caption = "Top SKAT/SKATO signals (FDR within SKAT+SKATO only). Cases/Controls = unique mask carriers. No OR."
  ) %>%
    .style_report_kable(font_size = 12, full_width = TRUE, gene_col = 1L, scroll_h = "480px") %>%
    kableExtra::column_spec(c(7, 8, 9), extra_css = "white-space: nowrap;") %>%
    print()
}
```

## Directional burden (ADD)

### Risk-increasing genes (Beta > 0, FDR < 0.05)

```{r risk-genes,echo=FALSE,message=FALSE,warning=FALSE}
risk_genes <- gene_based_add %>%
  filter(PVAL_ADJ < 0.05, BETA > 0) %>%
  select(GENE, MASK, THRESH, BETA, PVAL_ADJ) %>%
  arrange(PVAL_ADJ) %>%
  .join_gene_carriers() %>%
  select(GENE, MASK, THRESH, Cases, Controls, BETA, PVAL_ADJ)

if(nrow(risk_genes) > 0) {
    kable(risk_genes,
          format = "html", digits = 4,
          col.names = c("Gene", "Mask", "MAF Filter", "Cases", "Controls", "Beta", "FDR P-value")) %>%
      .style_report_kable(font_size = 12, full_width = FALSE, gene_col = 1L) %>%
      column_spec(c(6, 7), color = "red")
} else {
    cat("No risk-increasing gene associations found")
}
```

### Protective genes (Beta < 0, FDR < 0.05)

```{r protective-genes,echo=FALSE,message=FALSE,warning=FALSE}
protective_genes <- gene_based_add %>%
  filter(PVAL_ADJ < 0.05, BETA < 0) %>%
  select(GENE, MASK, THRESH, BETA, PVAL_ADJ) %>%
  arrange(PVAL_ADJ) %>%
  .join_gene_carriers() %>%
  select(GENE, MASK, THRESH, Cases, Controls, BETA, PVAL_ADJ)

if(nrow(protective_genes) > 0) {
    kable(protective_genes,
          format = "html", digits = 4,
          col.names = c("Gene", "Mask", "MAF Filter", "Cases", "Controls", "Beta", "FDR P-value")) %>%
      .style_report_kable(font_size = 12, full_width = FALSE, gene_col = 1L) %>%
      column_spec(c(6, 7), color = "steelblue")
} else {
    cat("No protective gene associations found")
}
```

## Output files

```{r save-results,echo=FALSE,message=FALSE,warning=FALSE}
date_stamp <- format(Sys.Date(), "%Y%m%d")
single_file <- paste0(params$project_name, "_single_variant_results_", date_stamp, ".csv")
write_csv(single_variant, single_file)
cat("**All single variant results saved to:** `", single_file, "`\n\n")

gene_file_all <- paste0(params$project_name, "_gene_based_results_alltests_", date_stamp, ".csv")
gene_file_add <- paste0(params$project_name, "_gene_based_results_ADDonly_", date_stamp, ".csv")
gene_file_skat <- paste0(params$project_name, "_gene_based_results_SKAT_SKATO_", date_stamp, ".csv")
gene_meta_fp <- paste0(params$project_name, "_gene_based_meta_input_", date_stamp, ".csv")
write_csv(gene_based_all, gene_file_all)
write_csv(gene_based_add, gene_file_add)

# SKAT/SKATO export: p/FDR only — omit BETA/SE/OR so direction cannot be misread
skat_export_cols <- c(
  "CHROM", "GENPOS", "ID", "GENE", "TEST", "MASK", "THRESH",
  "LOG10P", "PVAL", "PVAL_ADJ", "LOG10P_ADJ",
  "N", "N_CASES", "N_CONTROLS",
  "A1FREQ", "A1FREQ_CASES", "A1FREQ_CONTROLS"
)
skat_export <- gene_based_skat %>% select(any_of(skat_export_cols))
write_csv(skat_export, gene_file_skat)

meta_cols <- c("GENE", "MASK", "THRESH", "TEST", "BETA", "SE", "PVAL", "N_CASES", "N_CONTROLS", "A1FREQ_CASES", "A1FREQ_CONTROLS")
gene_meta_input <- gene_based_all %>% select(any_of(meta_cols))
write_csv(gene_meta_input, gene_meta_fp)

cat("**Gene-based REGENIE tables:**\n\n")
cat("- **All tests (archive):** `", gene_file_all, "`\n")
cat("- **ADD only (burden / direction):** `", gene_file_add, "`\n")
cat("- **SKAT/SKATO only (non-directional; no BETA/OR):** `", gene_file_skat, "`\n")
cat("- **Meta input:** `", gene_meta_fp, "`\n\n")

write_csv(
  dplyr::filter(gene_based_all, GENE == "CHEK2", MASK == "M1"),
  paste0(params$project_name, "_CHEK2_M1_alltests_", date_stamp, ".csv")
)
write_csv(
  dplyr::filter(gene_based_add, GENE == "CHEK2", MASK == "M1"),
  paste0(params$project_name, "_CHEK2_M1_ADDonly_", date_stamp, ".csv")
)
cat("**CHEK2 M1 quick extracts:** `", params$project_name, "_CHEK2_M1_alltests_", date_stamp, ".csv` and `", params$project_name, "_CHEK2_M1_ADDonly_", date_stamp, ".csv`\n\n")

top_gene_signals <- gene_based_add %>%
  filter(PVAL_ADJ < 0.05) %>%
  arrange(PVAL, GENE, MASK, THRESH)
top_signals_fp <- paste0(params$project_name, "_top_gene_signals_ADDonly_", date_stamp, ".csv")
write_csv(top_gene_signals, top_signals_fp)
cat("**Top ADD gene signals (FDR < 0.05):** `", top_signals_fp, "`\n\n")

if(nrow(significant_variants) > 0) {
    sig_single_file <- paste0(params$project_name, "_significant_single_variants_", date_stamp, ".csv")
    write_csv(significant_variants_export, sig_single_file)
    cat("**Exome-wide singles:** `", sig_single_file, "`\n\n", sep = "")
} else {
    cat("**No variants reached exome-wide significance (p < 2.5e-6)**\n\n")
}

if(exists("suggestive_genes") && nrow(suggestive_genes) > 0) {
    sig_gene_file <- paste0(params$project_name, "_suggestive_gene_associations_ADDonly_", date_stamp, ".csv")
    write_csv(suggestive_genes, sig_gene_file)
    cat("**Suggestive ADD-only (exploratory) saved to:** `", sig_gene_file, "`\n\n")
} else {
    cat("**No suggestive ADD-only gene associations found**\n\n")
}

if (!is.null(report_top_genes) && nrow(report_top_genes) > 0) {
  topgene_csv <- report_top_genes %>%
    filter(THRESH %in% CSV_THRESH) %>%
    arrange(MASK, THRESH, rank)
} else {
  topgene_csv_list <- list()
  for (m in c("M1", "M2", "M3", "M4")) {
    for (th in CSV_THRESH) {
      tg20 <- top_genes_for(m, th, n = 20L)
      if (nrow(tg20) > 0) {
        tg20$MASK <- m
        tg20$THRESH <- th
        topgene_csv_list[[paste(m, th, "genes")]] <- tg20
      }
    }
  }
  topgene_csv <- if (length(topgene_csv_list)) bind_rows(topgene_csv_list) %>% relocate(MASK, THRESH) else tibble()
}
if (nrow(topgene_csv) > 0) {
  topgene_fp <- paste0(params$project_name, "_top_genes_ADDonly_", date_stamp, ".csv")
  write_csv(topgene_csv, topgene_fp)
  cat("**Top 20 genes per mask×bin (ADD):** `", topgene_fp, "`\n\n", sep = "")
}

if (!is.null(report_contrib) && nrow(report_contrib) > 0) {
  contrib_all <- report_contrib %>% filter(THRESH %in% CSV_THRESH)
} else {
  contrib_list <- list()
  for (m in c("M1", "M2", "M3", "M4")) {
    for (th in CSV_THRESH) {
      t <- contrib_table(m, th, n = 20L)
      if (nrow(t) > 0) {
        t$MASK <- m
        t$THRESH <- th
        contrib_list[[paste(m, th)]] <- t
      }
    }
  }
  contrib_all <- if (length(contrib_list)) bind_rows(contrib_list) %>% relocate(MASK, THRESH) else tibble()
}
if (nrow(contrib_all) > 0) {
  contrib_fp <- paste0(params$project_name, "_variant_contributions_topgenes_", date_stamp, ".csv")
  write_csv(contrib_all, contrib_fp)
  cat("**Mask variants for top 20 genes (snplist×annot):** `", contrib_fp, "`\n\n", sep = "")
}

cat("**Analysis completed on:**", format(Sys.time(), "%Y-%m-%d %H:%M:%S"))
```

## Appendix: `--minMAC` — single-variant vs gene-based

This run uses **`--minMAC 1`** in both Step 2 single-variant and Step 2 gene-based (**REGENIE’s default is 5**; earlier comparison runs used **10**). The flag does **not** mean the same thing in both analyses.

### Single-variant Step 2

- `--minMAC` filters **each tested site**.
- With `--minMAC 10`, sites with MAC &lt; 10 are **not tested** and do not appear in the single-variant `.regenie` files.
- With `--minMAC 1`, almost every polymorphic site is tested (only monomorphic / MAC 0 drop out).
- Effect on the report: many more ultra-rare tests → noisier QQ / higher genomic inflation (λ). That is expected; it does not by itself mean the null model is broken.
- SPA / Firth approximations are also less well calibrated for MAC ≲ 5–10, so some extra inflation under `--minMAC 1` is expected.

### Gene-based Step 2 (burden / SKAT / SKATO)

- Individual rare sites are still allowed into masks (AAF bins / singleton bin decide **which** sites enter which mask).
- `--minMAC` then filters the **collapsed mask genotype**, not each constituent variant.
- REGENIE’s log wording is along the lines of: masks with MAC below the threshold are ignored.
- A gene can still be tested even if every site in the mask is a singleton or doubleton, as long as the finished mask (the number of samples carrying at least one qualifying variant) has MAC ≥ `--minMAC`.

**Practical example of the confusion:** people count “4 case carriers + 1 control carrier” across several ultra-rare sites in a gene and ask why that gene appears when they are thinking “minMAC 10 per variant.” Answer: gene-based minMAC is on the **mask**, and those carriers can accumulate across different rare sites. Single-variant minMAC 10 would have dropped each of those sites from the per-site GWAS, but they can still contribute to the gene mask.

### Related knobs (not the same as minMAC)

| Knob | Role |
|------|------|
| `--aaf-bins` | Upper AAF cutoffs for building separate masks (e.g. 0.01, 0.001, 0.0001). |
| **singleton** mask | Separate burden of MAC=1 sites (REGENIE always builds this with aaf-bins). **Not** the same as single-variant GWAS. |
| `--build-mask max` | How dosages collapse across sites in a gene (max ALT across sites per sample). |
| `--vc-tests skat,skato` | Extra non-directional tests on the same masks; no OR / direction. |
| `--vc-MACthr` | **Different** from `--minMAC`. Default **10**. See below. |

### `--vc-MACthr` (SKAT-family only — we have not changed this)

REGENIE default is **`--vc-MACthr 10`**. Our runs leave that default alone.

It applies **only** when `--vc-tests` runs SKAT / SKATO / ACAT-type tests. It does **not** change the ADD burden test (`TEST == ADD`), which already collapses every site in the mask via `--build-mask`.

**What it does (REGENIE docs):**  
Within each gene mask for a VC test, ultra-rare sites with **MAC ≤ `--vc-MACthr`** are **collapsed into one internal burden component**. That single collapsed genotype is what enters the SKAT/ACAT machinery **instead of** treating each ultra-rare site as its own term. Sites with MAC above the threshold stay as individual contributors to the VC test.

**Default 10 means:** under the default of 10, nearly all sites in a typical rare-variant mask are collapsed into one internal component before SKAT is run. SKAT therefore sees mostly that one aggregate rare-burden term plus any less-rare sites that remain.

**Lowering `--vc-MACthr` (e.g. toward 1):**  
- Fewer sites qualify as “ultra-rare” for collapsing.  
- More individual rare variants enter SKAT as separate terms.  
- SKAT becomes more “per-site variance” and less dependent on a single collapsed ultra-rare component.  
- Can change which genes look strong on SKAT/SKATO **without** changing ADD results.  
- Same calibration caveat as low-MAC singles: very rare per-site terms are noisy.

**Raising `--vc-MACthr`:**  
- More collapsing into the single collapsed ultra-rare component.  
- SKAT leans even more on one aggregate rare-burden direction within the VC model.

**If the goal is to retain ultra-rare signal in the gene-based results:**

| Knob | What it filters / collapses | Affects ADD? | Affects SKAT/SKATO? |
|------|------------------------------|--------------|---------------------|
| `--minMAC` (gene-based) | Finished **mask** MAC (test or drop the whole mask) | Yes | Yes (whether that mask is tested at all) |
| `--minMAC` (single-variant) | Each **site** | N/A | N/A |
| `--vc-MACthr` | Inside a VC test: which **sites** are collapsed into an ultra-rare component vs kept separate | **No** | **Yes** |

- `--minMAC` controls whether a low-MAC mask is tested at all (affects both ADD and SKAT).
- `--vc-MACthr` only changes how SKAT treats the rare sites inside a mask that already passed the minMAC filter. It does not alter ADD results and does not affect single-variant inflation.
- We left `--vc-MACthr` at its default of 10. SKAT results therefore reflect the software’s standard collapsing behavior.

### What “Cases / Controls” mean in this report

- **Gene tables** (top ADD, suggestive, SKAT, risk/protective): unique samples carrying ≥1 variant in that gene×mask×bin snplist (from `*.report.gene_carriers.tsv`).
- **Variant tables** (mask contrib, CHEK2 sites): per-site carriers from `mask_variant_stats.tsv` when available.
- **Single-variant hits:** prefer mask-stats carriers; if missing (site not in the LoF/syn mask-stats set), approximate alt allele count from REGENIE `--af-cc` (`A1FREQ × 2 × N`), which for rare hets is ≈ carrier count.
- REGENIE’s own `N_CASES` / `N_CONTROLS` columns are **cohort sizes in the model** (nearly constant) — not “how many people carry the variant,” so we do not use those as the collaborator-facing carrier columns.

## Appendix: REGENIE pipeline flowchart

```{r regenie_flowchart, echo=FALSE, fig.height=7}
grViz("
digraph regenie_pipeline {

  graph [rankdir = TB, fontsize = 12, fontname = Helvetica]
  node [fontname = Helvetica, fontsize = 11, shape = box, style = filled, color = DarkBlue]

  input_vcf [label = 'Multi-chromosome VCF\\nfiles', fillcolor = Khaki, fontcolor = Black, shape = folder]
  samples [label = 'Sample list', fillcolor = Khaki, fontcolor = Black, shape = note]
  pheno [label = 'Phenotypes', fillcolor = Khaki, fontcolor = Black, shape = note]
  covar [label = 'Covariates', fillcolor = Khaki, fontcolor = Black, shape = note]

  preprocess [label = 'VCF Preprocessing\\n(normalize, filter, QC)', fillcolor = LightSkyBlue]
  pgen_convert [label = 'Convert to PLINK2\\nPGEN format', fillcolor = LightSkyBlue]

  step1_chr1 [label = 'Step 1\\nChr 1\\n(covariate adjustment)', fillcolor = LavenderBlush, shape = parallelogram]
  step1_chr_ellipsis [label = '...', fillcolor = LavenderBlush, shape = parallelogram]
  step1_chr22 [label = 'Step 1\\nChr 22\\n(covariate adjustment)', fillcolor = LavenderBlush, shape = parallelogram]

  merge_step1 [label = 'Merge Step 1\\nresults', fillcolor = MistyRose, shape = diamond]

  step2_single [label = 'Step 2 Single-Variant\\nAssociation Testing\\nChr 1-22', fillcolor = HoneyDew, shape = parallelogram]
  step2_gene [label = 'Step 2 Gene-Based\\nAssociation Testing\\nChr 1-22', fillcolor = HoneyDew, shape = parallelogram]

  vep_annotate [label = 'VEP Annotation\\n(pathogenic/VUS variants)', fillcolor = LemonChiffon]
  report_gen [label = 'Generate HTML Report\\n(statistics, plots, tables)', fillcolor = AliceBlue, shape = note]

  input_vcf -> preprocess
  samples -> preprocess
  preprocess -> pgen_convert
  preprocess -> vep_annotate

  pgen_convert -> step1_chr1
  pgen_convert -> step1_chr_ellipsis
  pgen_convert -> step1_chr22

  step1_chr1 -> merge_step1 [style = dotted]
  step1_chr_ellipsis -> merge_step1 [style = dotted]
  step1_chr22 -> merge_step1 [style = dotted]

  merge_step1 -> step2_single
  merge_step1 -> step2_gene
  vep_annotate -> step2_gene

  step2_single -> report_gen
  step2_gene -> report_gen

  {rank = same; step1_chr1; step1_chr_ellipsis; step1_chr22}
  {rank = same; step2_single; step2_gene}

  subgraph cluster_parallel {
    label = 'Parallel Processing\\n(Chromosome-level)'
    fontsize = 14
    style = filled
    fillcolor = WhiteSmoke
    step1_chr1
    step1_chr_ellipsis
    step1_chr22
    step2_single
    step2_gene
  }
}
")
```