---
title: "REGENIE Exome-Wide Association Analysis Report"
author: "Brad Wubbenhorst"
output: 
  html_document:
    theme: paper
    toc: true
    toc_float: true
    code_folding: hide
date: "`r Sys.Date()`"
params:
  project_name: "TECAC_WES"
---

**Note:** This report is configured for Exome-Wide Association Studies (ExWAS) using Regenie burden and single-variant results.

<!-- 
NOTES TO SELF:
- M1 = pathogenic variants only (LoF level 1)
- M2 = VUS variants only (LoF level 2) 
- M3 = pathogenic + VUS variants combined
- ADD = additive burden test (use for BETA interpretation)
- SKAT/SKATO = other burden tests (not used in this report)
- Exome-wide significance = p < 2.5e-6 (not GWAS 5e-8)
- Lambda GC less interpretable in ExWAS due to sparse rare variants

KEY TERMS:
- BETA (Effect Size): Log odds ratio of disease association for alt allele
  * Positive BETA = alt allele increases disease risk
  * Negative BETA = alt allele is protective
- SE (Standard Error): Uncertainty around effect size estimate
  * High SE = less confidence in BETA (common for rare variants)
- PVAL/LOG10P: p-value from association test
  * LOG10P = -log10(p-value) for plotting
  * Smaller p-values (larger LOG10P) = stronger evidence
- PVAL_ADJ/FDR: False discovery rate-adjusted p-values (Benjamini-Hochberg)
  * Controls for multiple testing across thousands of variants/genes
  * Prevents false positives from testing many hypotheses
-->


```{r setup,echo=FALSE, message=FALSE, warning=FALSE, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width = 10, fig.height = 6)
library(dplyr)
library(readr)
library(data.table)
library(ggplot2)
library(knitr)
library(kableExtra)
library(ggrepel)
library(qqman)

# Set consistent theme
theme_set(theme_minimal(base_size = 12))
```

```{r load-data,echo=FALSE,message=FALSE,warning=FALSE}
# Validate project name parameter
if(is.null(params$project_name) || params$project_name == "") {
  stop("ERROR: project_name parameter is required but was not provided or is empty. ",
       "Please specify params=list(project_name='YOUR_PROJECT_NAME') when rendering.")
}

# Load single variant data from all chromosomes
single_variant_list <- list()

for(i in 1:22) {
  file_path <- paste0("data/work/regenie/", params$project_name, ".chr", i, ".step2_single_variant_STATUS.regenie")
  if(file.exists(file_path)) {
    single_variant_list[[i]] <- read_delim(file_path, delim = " ", col_names = TRUE, trim_ws = TRUE)
  }
}

# Combine all single variant data
single_variant <- bind_rows(single_variant_list) %>%
   mutate(across(c(CHROM, GENPOS, N, N_CASES, N_CONTROLS), as.integer),
         across(c(A1FREQ, A1FREQ_CASES, A1FREQ_CONTROLS, BETA, SE, CHISQ, LOG10P), as.numeric)) %>%
  mutate(CHROM = as.factor(CHROM)) %>% 
  arrange(CHROM, GENPOS) %>% 
  mutate(PVAL = 10^(-LOG10P),
         PVAL_ADJ = p.adjust(PVAL, method = "BH"),
         LOG10P_ADJ = -log10(PVAL_ADJ)) %>%
  mutate(POS_INDEX = row_number(),
         OR = exp(BETA)) %>%
  # Parse ID into REF and ALT for better variant presentation
  mutate(ID_PARTS = strsplit(ID, "_")) %>%
  rowwise() %>%
  mutate(REF = ID_PARTS[[2]], ALT = ID_PARTS[[3]]) %>%
  ungroup() %>%
  select(-ID_PARTS)

# Load pre-filtered pathogenic/VUS annotations
pathogenic_vus_file <- "PMBB-TGCT-ExWAS.pathogenic_vus.csv"
if(file.exists(pathogenic_vus_file)) {
  pathogenic_annotations <- read_csv(pathogenic_vus_file, show_col_types = FALSE) %>%
    select(ID, Gene, Variant.LoF_level, HGVSc, HGVSp)
} else {
  warning("Pathogenic/VUS annotation file not found. Table will show basic variant info only.")
  pathogenic_annotations <- NULL
}

# Load gene-based data from all chromosomes
gene_based_list <- list()

for(i in 1:22) {
  file_path <- paste0("data/work/regenie/", params$project_name, ".chr", i, ".step2_gene_based_STATUS.regenie")
  if(file.exists(file_path)) {
    gene_based_list[[i]] <- read_delim(file_path, delim = " ", col_names = TRUE, trim_ws = TRUE, skip = 1)
  }
}

# Combine all gene-based data
gene_based <- bind_rows(gene_based_list) %>%
  mutate(across(c(CHROM, GENPOS, N, N_CASES, N_CONTROLS), as.integer),
         across(c(A1FREQ, A1FREQ_CASES, A1FREQ_CONTROLS, BETA, SE, CHISQ, LOG10P), as.numeric)) %>%
  mutate(CHROM = as.factor(CHROM),
         PVAL = 10^(-LOG10P),
         PVAL_ADJ = p.adjust(PVAL, method = "BH"),
         LOG10P_ADJ = -log10(PVAL_ADJ),
         OR = exp(BETA)) %>%
  # Extract gene and mask information from ID column
  mutate(
    GENE = sub("\\..*", "", ID),
    MASK = sub(".*\\.(M[0-9]+).*", "\\1", ID),
    THRESH_RAW = sub(".*\\.(.*)$", "\\1", ID),  # Extract everything after last dot
    THRESH = ifelse(grepl("^\\d+\\.\\d+$", THRESH_RAW), as.numeric(THRESH_RAW),
                   ifelse(THRESH_RAW == "all", "All variants", THRESH_RAW))
  ) %>%
  # Filter to ADD test only for BETA-based analysis
  filter(TEST == "ADD")
```

## Analysis Summary

```{r summary-stats,echo=FALSE,message=FALSE,warning=FALSE}
# Load phenotype file for actual case/control numbers
pheno_file <- paste0(params$project_name, ".regenie.pheno.txt")
pheno <- read_delim(pheno_file, delim = " ", col_names = TRUE, trim_ws = TRUE)

# Build proper summary table with scalar values
summary_df <- data.frame(
  Metric = c("Total Samples", "Cases", "Controls",
             "Variants Tested (Single-Variant)",
             "Genes Tested (Burden Test)"),
  Count = c(
    length(unique(pheno$IID)),
    sum(pheno$STATUS == 1),
    sum(pheno$STATUS == 0),
    nrow(single_variant),
    length(unique(gene_based$GENE))  # important: gene count, not row count
  )
)

kable(summary_df, format = "html", col.names = c("Metric", "Count")) %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover")) %>%
  column_spec(1, bold = TRUE)
```

## Quality Metrics

```{r quality-metrics,echo=FALSE,message=FALSE,warning=FALSE}
# Calculate lambda GC for both analyses
single_lambda <- median(qchisq(1 - single_variant$PVAL, 1), na.rm = TRUE) / qchisq(0.5, 1)
gene_lambda <- median(qchisq(1 - gene_based$PVAL, 1), na.rm = TRUE) / qchisq(0.5, 1)

# Simple quality metrics
quality_df <- data.frame(
    Metric = c(
        "Significant Single-Variants (p < 2.5e-6)",
        "Significant Gene-Based (FDR < 0.05)",
        "Lambda GC (Single-Variant)",
        "Lambda GC (Gene-Based)",
        "Max Effect Size (Gene-Based OR)"
    ),
    Value = c(
        format(sum(single_variant$LOG10P > 5.60206, na.rm = TRUE), big.mark = ","),
        format(sum(gene_based$PVAL_ADJ < 0.05, na.rm = TRUE), big.mark = ","),
        round(single_lambda, 3),
        round(gene_lambda, 3),
        round(max(exp(gene_based$BETA[gene_based$PVAL_ADJ < 0.05]), na.rm = TRUE), 2)
    )
)

kable(quality_df, format = "html",
      col.names = c("Quality Metric", "Value")) %>%
    kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover")) %>%
    column_spec(1, bold = TRUE)
```

## Single-Variant Exome Association Results

### Significant Variants (p < 2.5e-6)

```{r significant-single,echo=FALSE,message=FALSE,warning=FALSE}
# Filter significant variants and join with pathogenic/VUS annotations
significant_variants <- single_variant %>%
    filter(!is.na(LOG10P), LOG10P > 5.60206) %>%  # Exome-wide threshold ~2.5e-6
    arrange(desc(LOG10P))

# Join with pathogenic/VUS annotations if available
if(!is.null(pathogenic_annotations) && nrow(pathogenic_annotations) > 0) {
  significant_pathogenic <- significant_variants %>%
    inner_join(pathogenic_annotations, by = "ID") %>%
    select(CHROM, GENPOS, Gene, Variant.LoF_level, HGVSc, HGVSp, A1FREQ, OR, PVAL, PVAL_ADJ) %>%
    arrange(desc(PVAL))  # Sort by significance

  if(nrow(significant_pathogenic) > 0) {
    # Display all significant pathogenic/VUS variants (should be small number)
    display_variants <- significant_pathogenic %>%
      mutate(Position = paste0(CHROM, ":", GENPOS),
             LoF = case_when(Variant.LoF_level == 1 ~ "Pathogenic",
                           Variant.LoF_level == 2 ~ "VUS",
                           TRUE ~ as.character(Variant.LoF_level))) %>%
      select(Position, Gene, LoF, HGVSc, HGVSp, A1FREQ, OR, PVAL, PVAL_ADJ)

    cat(paste0("**Showing ", nrow(significant_pathogenic), " significant pathogenic/VUS variants**\n\n"))

    kable(display_variants,
          format = "html", digits = 4,
          col.names = c("Chr:Pos", "Gene", "LoF Level", "HGVSc", "HGVSp", "Alt Freq", "OR", "P-value", "FDR P-value")) %>%
        kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover"),
                      font_size = 11) %>%
        column_spec(1, bold = TRUE, width = "80px") %>%
        column_spec(8:9, color = "red") %>%
        scroll_box(width = "100%", height = "400px")
  } else {
    cat("**No significant variants are classified as pathogenic or VUS**")
  }
} else {
  # Fallback: show basic variant info if no annotations available
  display_variants <- significant_variants %>%
    slice_head(n = min(20, nrow(significant_variants))) %>%
    select(CHROM, GENPOS, REF, ALT, A1FREQ, OR, PVAL, PVAL_ADJ) %>%
    mutate(Variant = paste0("chr", CHROM, ":", GENPOS, " ", REF, ">", ALT)) %>%
    select(Variant, A1FREQ, OR, PVAL, PVAL_ADJ)

  kable(display_variants,
        format = "html", digits = 4,
        col.names = c("Variant (chr:pos ref>alt)", "Alt Freq", "OR", "P-value", "FDR P-value")) %>%
      kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover"),
                    font_size = 12) %>%
      column_spec(1, bold = TRUE, width = "250px") %>%
      column_spec(4:5, color = "red") %>%
      scroll_box(width = "100%", height = "400px")

  if(nrow(significant_variants) > 20) {
    cat(paste0("\n**Showing top 20 of ", nrow(significant_variants), " significant variants (annotations not available)**"))
  }
}
```

### Manhattan Plot

```{r manhattan-plot,echo=FALSE,message=FALSE,warning=FALSE}
# Manhattan Plot with improved styling
chrom_colors <- rep(c("#2E86AB", "#FF6B35"), length.out = length(unique(single_variant$CHROM)))

manhattan_plot <- ggplot(single_variant, aes(x = POS_INDEX, y = LOG10P, color = CHROM)) +
    geom_point(alpha = 0.7, size = 0.8) +
    scale_color_manual(values = chrom_colors) +
    labs(title = "Manhattan Plot: Single-Variant Exome Association",
         subtitle = paste("Total variants:", nrow(single_variant), "| Exome-wide significance threshold: p < 2.5e-6"),
         x = "Genomic Position", y = "-log10(p-value)") +
    geom_hline(yintercept = 5.602, color = "red", linetype = "dashed", size = 0.8) +
    theme(legend.position = "none",
          plot.title = element_text(size = 14, face = "bold"),
          plot.subtitle = element_text(size = 10, color = "gray50"))

print(manhattan_plot)
```

### QQ Plot

```{r qq-plot,echo=FALSE,message=FALSE,warning=FALSE}
# QQ Plot for single variant p-values
qq_data <- single_variant %>%
    filter(!is.na(LOG10P)) %>%
    arrange(desc(LOG10P)) %>%
    mutate(
        observed = sort(LOG10P, decreasing = TRUE),
        expected = sort(-log10(ppoints(length(LOG10P))), decreasing = TRUE)
    )

# Calculate lambda GC correctly
chisq <- qchisq(1 - single_variant$PVAL, 1)
lambda_gc <- median(chisq, na.rm = TRUE) / qchisq(0.5, 1)
cat("Lambda GC formula: median(qchisq(1 - P, 1)) / qchisq(0.5, 1) =", round(lambda_gc, 3), "\n")

qq_plot <- ggplot(qq_data, aes(x = expected, y = observed)) +
    geom_point(alpha = 0.6, color = "#2E86AB") +
    geom_abline(intercept = 0, slope = 1, color = "red", linetype = "dashed") +
    labs(title = "Q-Q Plot: Single-Variant P-values",
         subtitle = paste("Lambda GC:", round(lambda_gc, 3), "| Note: Inflation metrics are less interpretable in ExWAS due to sparse, low-frequency variant testing"),
         x = "Expected -log10(p-value)", 
         y = "Observed -log10(p-value)") +
    theme(plot.title = element_text(size = 14, face = "bold"),
          plot.subtitle = element_text(size = 10, color = "gray50"))

print(qq_plot)
```

## Gene-Based Burden Testing Results

### Summary by Mask

```{r mask-summary,echo=FALSE,message=FALSE,warning=FALSE}
# Summary statistics per mask
mask_summary <- gene_based %>%
  group_by(MASK) %>%
  summarize(
    Genes_Tested = n(),
    Significant_Genes = sum(PVAL_ADJ < 0.05, na.rm = TRUE),
    Most_Significant_Gene = GENE[which.max(LOG10P)],
    Max_Log10P = max(LOG10P, na.rm = TRUE),
    Mean_Beta = mean(BETA, na.rm = TRUE)
  ) %>%
  mutate(Significance_Rate = Significant_Genes / Genes_Tested * 100)

kable(mask_summary, format = "html", digits = 3,
      col.names = c("Mask", "Genes Tested", "FDR Significant", "Top Gene", "Max -log10(P)", "Mean Beta", "Significance Rate (%)")) %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover")) %>%
  column_spec(1, bold = TRUE)
```

### Top Genes by Clinical Category

#### M1 (Pathogenic variants only)

```{r top-genes-m1,echo=FALSE,message=FALSE,warning=FALSE}
# M1 (Pathogenic) - Top 10 genes
m1_top <- gene_based %>%
  filter(MASK == "M1") %>%
  slice_max(LOG10P, n = 10) %>%
  select(GENE, THRESH, BETA, SE, LOG10P, PVAL, PVAL_ADJ, OR) %>%
  arrange(desc(LOG10P))

if(nrow(m1_top) > 0) {
  kable(m1_top, format = "html", digits = 4,
        col.names = c("Gene", "MAF Filter", "Beta", "SE", "-log10(P)", "P-value", "FDR P-value", "OR")) %>%
    kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover")) %>%
    column_spec(1, bold = TRUE) %>%
    column_spec(c(3, 5, 6), color = "red")
} else {
  cat("No results")
}
```

#### M2 (VUS variants only)

```{r top-genes-m2,echo=FALSE,message=FALSE,warning=FALSE}
# M2 (VUS) - Top 10 genes
m2_top <- gene_based %>%
  filter(MASK == "M2") %>%
  slice_max(LOG10P, n = 10) %>%
  select(GENE, THRESH, BETA, SE, LOG10P, PVAL, PVAL_ADJ, OR) %>%
  arrange(desc(LOG10P))

if(nrow(m2_top) > 0) {
  kable(m2_top, format = "html", digits = 4,
        col.names = c("Gene", "MAF Filter", "Beta", "SE", "-log10(P)", "P-value", "FDR P-value", "OR")) %>%
    kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover")) %>%
    column_spec(1, bold = TRUE) %>%
    column_spec(c(3, 5, 6), color = "red")
} else {
  cat("No results")
}
```

#### M3 (Pathogenic + VUS combined)

```{r top-genes-m3,echo=FALSE,message=FALSE,warning=FALSE}
# M3 (Pathogenic + VUS) - Top 10 genes
m3_top <- gene_based %>%
  filter(MASK == "M3") %>%
  slice_max(LOG10P, n = 10) %>%
  select(GENE, THRESH, BETA, SE, LOG10P, PVAL, PVAL_ADJ, OR) %>%
  arrange(desc(LOG10P))

if(nrow(m3_top) > 0) {
  kable(m3_top, format = "html", digits = 4,
        col.names = c("Gene", "MAF Filter", "Beta", "SE", "-log10(P)", "P-value", "FDR P-value", "OR")) %>%
    kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover")) %>%
    column_spec(1, bold = TRUE) %>%
    column_spec(c(3, 5, 6), color = "red")
} else {
  cat("No results")
}
```

### Suggestive Gene-Based Associations (Raw p < 0.001, FDR ≥ 0.05)

```{r suggestive-genes,echo=FALSE,message=FALSE,warning=FALSE}
# Show genes with strong raw associations but don't meet FDR significance
# These are "near misses" - promising candidates that need more evidence

# Get genes that are in top 10 by raw p-value per mask but don't meet FDR threshold
top_by_raw_p <- gene_based %>%
  group_by(MASK) %>%
  slice_max(LOG10P, n = 10) %>%
  ungroup()

# Filter to suggestive associations: strong raw p (< 0.001) but not FDR significant
suggestive_genes <- top_by_raw_p %>%
  filter(LOG10P > 3, PVAL_ADJ >= 0.05) %>%  # Raw p < 0.001, FDR >= 0.05
  arrange(PVAL) %>%  # Sort by raw p-value strength
  slice_head(n = 15) %>%  # Show top 15 suggestive associations
  select(GENE, MASK, THRESH, BETA, LOG10P, PVAL, PVAL_ADJ, OR)

if(nrow(suggestive_genes) > 0) {
    cat("**Genes with strong evidence (raw p < 0.001) but not FDR significant after multiple testing correction**\n\n")
    kable(suggestive_genes,
          format = "html", digits = 4,
          col.names = c("Gene", "Mask", "MAF Filter", "Beta", "Raw -log10(P)", "Raw P-value", "FDR P-value", "OR")) %>%
        kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover"),
                      font_size = 11) %>%
        column_spec(1, bold = TRUE) %>%
        column_spec(c(4, 8), color = "orange")
} else {
    cat("No suggestive associations found")
}
```

### Volcano Plot by Mask

```{r volcano-plot,echo=FALSE,message=FALSE,warning=FALSE}
# Volcano Plot with FDR significance, faceted by mask
gene_based_plot <- gene_based %>%
  mutate(Significant = PVAL_ADJ < 0.05)

# Get top FDR significant genes for labeling (top 5 per mask)
top_hits <- gene_based_plot %>% 
  filter(Significant) %>%
  group_by(MASK) %>%
  slice_max(LOG10P, n = 5) %>%
  ungroup()

gene_based_plot <- gene_based_plot %>%
  mutate(Label = ifelse(GENE %in% top_hits$GENE, GENE, NA))

# Calculate FDR cutoff for the line
fdr_cutoff <- min(gene_based$LOG10P[gene_based$PVAL_ADJ < 0.05], na.rm = TRUE)

volcano_plot <- ggplot(gene_based_plot, aes(x = BETA, y = LOG10P, color = Significant)) +
  geom_point(alpha = 0.6, size = 1.2) +
  geom_hline(yintercept = fdr_cutoff, linetype = "dashed", color = "blue", alpha = 0.7) +
  geom_vline(xintercept = 0, linetype = "dotted", color = "grey40") +
  geom_text_repel(aes(label = Label), size = 2.5, max.overlaps = 30,
                  box.padding = 0.3, point.padding = 0.1) +
  scale_color_manual(values = c("gray60", "red"), 
                     labels = c("Not Significant", "FDR < 0.05")) +
  labs(title = "Volcano Plot: Gene-Based Burden Testing by Mask",
       subtitle = paste("FDR significant genes:", sum(gene_based_plot$Significant)),
       x = "Effect Size (Beta)",
       y = "-log10(p-value)",
       color = "Significance") +
  facet_wrap(~MASK, scales = "free_y") +
  theme(plot.title = element_text(size = 14, face = "bold"),
        plot.subtitle = element_text(size = 10, color = "gray50"),
        legend.position = "bottom",
        strip.text = element_text(size = 12, face = "bold"))

print(volcano_plot)
```

### CHEK2 Gene Analysis

```{r chek2-analysis,echo=FALSE,message=FALSE,warning=FALSE}
# Extract CHEK2 results specifically
chek2_results <- gene_based %>%
  filter(GENE == "CHEK2") %>%
  select(GENE, MASK, THRESH, CHROM, GENPOS, ALLELE1, A1FREQ, BETA, OR, SE, LOG10P, PVAL, PVAL_ADJ, LOG10P_ADJ) %>%
  arrange(MASK, THRESH)

if(nrow(chek2_results) > 0) {
  cat("**CHEK2 Gene Results (Chr22):**\n\n")
  
  kable(chek2_results, 
        format = "html", digits = 4,
        col.names = c("Gene", "Mask", "Threshold", "Chr", "Pos", "Alt Allele", "Alt Freq", 
                     "Beta", "OR", "SE", "-log10(P)", "P-value", "FDR P-value", "-log10(FDR P)")) %>%
    kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover")) %>%
    column_spec(c(8, 10, 11), color = ifelse(chek2_results$PVAL_ADJ < 0.05, "red", "black"))
  
  # Summary of CHEK2 results
  cat("\n**CHEK2 Summary:**\n")
  cat("- **Total tests performed:**", nrow(chek2_results), "\n")
  cat("- **Best p-value:**", sprintf("%.2e", min(chek2_results$PVAL, na.rm = TRUE)), "\n")
  cat("- **Best FDR p-value:**", sprintf("%.2e", min(chek2_results$PVAL_ADJ, na.rm = TRUE)), "\n")
  cat("- **Most significant mask:**", chek2_results$MASK[which.min(chek2_results$PVAL)], "\n")
  cat("- **Most significant threshold:**", chek2_results$THRESH[which.min(chek2_results$PVAL)], "\n")
  
} else {
  cat("**CHEK2 gene not found in results**")
}
```

## Risk-Increasing Genes (Beta > 0)

```{r risk-genes,echo=FALSE,message=FALSE,warning=FALSE}
# Risk-increasing genes (positive associations)
risk_genes <- gene_based %>%
  filter(PVAL_ADJ < 0.05, BETA > 0) %>%
  select(GENE, MASK, THRESH, BETA, PVAL_ADJ) %>%
  arrange(PVAL_ADJ)

if(nrow(risk_genes) > 0) {
    cat("**Found", nrow(risk_genes), "risk-increasing gene associations**\n\n")
    kable(risk_genes,
          format = "html", digits = 4,
          col.names = c("Gene", "Mask", "MAF Filter", "Beta", "FDR P-value")) %>%
      kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover"),
                    font_size = 11) %>%
      column_spec(1, bold = TRUE) %>%
      column_spec(c(4, 5), color = "red")
} else {
    cat("No risk-increasing gene associations found")
}

# Summary count
risk_count <- gene_based %>%
  filter(PVAL_ADJ < 0.05) %>%
  summarise(Risk_Increasing = sum(BETA > 0, na.rm = TRUE),
            Protective = sum(BETA < 0, na.rm = TRUE))

if(nrow(risk_count) > 0 && risk_count$Risk_Increasing > 0) {
  cat("\n\n**Summary:**", risk_count$Risk_Increasing, "risk-increasing vs",
      risk_count$Protective, "protective associations")
}
```

## Output Files

```{r save-results,echo=FALSE,message=FALSE,warning=FALSE}
# Save ALL results to CSV files
date_stamp <- format(Sys.Date(), "%Y%m%d")
single_file <- paste0(params$project_name, "_single_variant_results_", date_stamp, ".csv")
gene_file <- paste0(params$project_name, "_gene_based_results_", date_stamp, ".csv")

write_csv(single_variant, single_file)
cat("**All single variant results saved to:** `", single_file, "`\n\n")

write_csv(gene_based, gene_file)
cat("**All gene-based results saved to:** `", gene_file, "`\n\n")

# Also save significant results if any exist
if(nrow(significant_variants) > 0) {
    sig_single_file <- paste0(params$project_name, "_significant_single_variants_", date_stamp, ".csv")
    write_csv(significant_variants, sig_single_file)
    cat("**Significant single variants saved to:** `", sig_single_file, "`\n\n")
} else {
    cat("**No variants reached exome-wide significance (p < 2.5e-6)**\n\n")
}

if(exists("suggestive_genes") && nrow(suggestive_genes) > 0) {
    sig_gene_file <- paste0(params$project_name, "_suggestive_gene_associations_", date_stamp, ".csv")
    write_csv(suggestive_genes, sig_gene_file)
    cat("**Suggestive gene-based associations saved to:** `", sig_gene_file, "`\n\n")
} else {
    cat("**No suggestive gene associations found**\n\n")
}

cat("**Analysis completed on:**", format(Sys.time(), "%Y-%m-%d %H:%M:%S"))
```