---
title: "REGENIE Exome-Wide Association Analysis Report"
author: "Brad Wubbenhorst"
output:
  html_document:
    theme: paper
    toc: true
    toc_depth: 2
    toc_float: true
    code_folding: hide
date: "`r Sys.Date()`"
params:
  project_name: "TECAC_WES"
---

```{=html}
<div style="border:1px solid #ccc; border-left:4px solid #2E86AB; padding:12px 14px; margin:1rem 0; background:#fafafa;">
<strong>How to read this report</strong>
<ul style="margin-bottom:0;">
<li><strong>Burden-style interpretation</strong> (effect direction, OR, volcano, CHEK2, directional gene lists) uses <code>TEST == &quot;ADD&quot;</code> only.</li>
<li>An <strong>all-tests</strong> export includes <code>ADD</code>, <code>ADD-SKAT</code>, <code>ADD-SKATO</code>, … for archive and downstream meta-analysis.</li>
<li><strong>Gene-based FDR</strong> (<code>PVAL_ADJ</code>) is Benjamini–Hochberg across <em>all</em> gene-based rows (every TEST × gene × mask × threshold), not within each TEST.</li>
</ul>
</div>
```

<!-- 
NOTES TO SELF:
- M1 = pathogenic variants only (LoF level 1)
- M2 = VUS variants only (LoF level 2) 
- M3 = pathogenic + VUS variants combined
- M4 = synonymous negative-control variants (excluding LoF 1/2 and excluding blacklisted potential MNP-pair IDs)
- ADD = additive burden test (BETA/SE/OR and directional interpretation apply here only)
- ADD-SKAT / ADD-SKATO = variance-component tests (`TEST` in REGENIE; p-values preserved in full export; not interpreted via BETA/OR in this report)
- Exome-wide significance = p < 2.5e-6 (not GWAS 5e-8)
- Lambda GC less interpretable in ExWAS due to sparse rare variants

KEY TERMS:
- BETA (Effect Size): Log odds ratio of disease association for alt allele
  * Positive BETA = alt allele increases disease risk
  * Negative BETA = alt allele is protective
- SE (Standard Error): Uncertainty around effect size estimate
  * High SE = less confidence in BETA (common for rare variants)
- PVAL/LOG10P: p-value from association test
  * LOG10P = -log10(p-value) for plotting
  * Plots cap extreme -log10(p) at 100 where needed; CSVs keep raw LOG10P
  * Smaller p-values (larger LOG10P) = stronger evidence
- PVAL_ADJ/FDR: False discovery rate-adjusted p-values (Benjamini-Hochberg)
  * Controls for multiple testing across thousands of variants/genes
  * Prevents false positives from testing many hypotheses
-->


```{r setup,echo=FALSE, message=FALSE, warning=FALSE, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width = 10, fig.height = 6)
library(dplyr)
library(readr)
library(data.table)
library(ggplot2)
library(knitr)
library(kableExtra)
library(ggrepel)
library(qqman)
library(DiagrammeR)

# Set consistent theme
theme_set(theme_minimal(base_size = 13))
```

```{r load-data,echo=FALSE,message=FALSE,warning=FALSE}
# Validate project name parameter
if(is.null(params$project_name) || params$project_name == "") {
  stop("ERROR: project_name parameter is required but was not provided or is empty. ",
       "Please specify params=list(project_name='YOUR_PROJECT_NAME') when rendering.")
}

# Load single variant data from all chromosomes
single_variant_list <- list()

for(i in 1:22) {
  file_path <- paste0("data/final/", params$project_name, ".chr", i, ".step2_single_variant_STATUS.regenie")
  if(file.exists(file_path)) {
    single_variant_list[[i]] <- read_delim(file_path, delim = " ", col_names = TRUE, trim_ws = TRUE)
  }
}

# Combine all single variant data
single_variant <- bind_rows(single_variant_list) %>%
   mutate(across(c(CHROM, GENPOS, N, N_CASES, N_CONTROLS), as.integer),
         across(c(A1FREQ, A1FREQ_CASES, A1FREQ_CONTROLS, BETA, SE, CHISQ, LOG10P), as.numeric)) %>%
  mutate(CHROM = as.factor(CHROM)) %>% 
  arrange(CHROM, GENPOS) %>% 
  mutate(PVAL = 10^(-LOG10P),
         PVAL_ADJ = p.adjust(PVAL, method = "BH"),
         LOG10P_ADJ = -log10(PVAL_ADJ)) %>%
  mutate(POS_INDEX = row_number(),
         OR = exp(BETA)) %>%
  # Parse REF/ALT from trailing ID fields (…_POS_REF_ALT)
  mutate(
    REF = vapply(ID, function(x) {
      p <- strsplit(as.character(x), "_", fixed = TRUE)[[1L]]
      if (length(p) >= 4L) p[[length(p) - 1L]] else NA_character_
    }, character(1L)),
    ALT = vapply(ID, function(x) {
      p <- strsplit(as.character(x), "_", fixed = TRUE)[[1L]]
      if (length(p) >= 4L) p[[length(p)]] else NA_character_
    }, character(1L))
  )

# Load pre-filtered pathogenic/VUS annotations
pathogenic_vus_file <- paste0("data/final/", params$project_name, ".pathogenic_vus.csv")
if(file.exists(pathogenic_vus_file)) {
  pathogenic_annotations <- read_csv(pathogenic_vus_file, show_col_types = FALSE) %>%
    select(ID, Gene, Variant.LoF_level, HGVSc, HGVSp)
} else {
  warning("Pathogenic/VUS annotation file not found. Table will show basic variant info only.")
  pathogenic_annotations <- NULL
}

# Load gene-based data from all chromosomes
gene_based_list <- list()

for(i in 1:22) {
  file_path <- paste0("data/final/", params$project_name, ".chr", i, ".step2_gene_based_STATUS.regenie")
  if(file.exists(file_path)) {
    # Skip comment lines (e.g. "# Regenie step 2") so header is CHROM GENPOS ... BETA ...; wrong skip misaligns BETA.
    gene_based_list[[i]] <- read_delim(file_path, delim = " ", col_names = TRUE, trim_ws = TRUE, comment = "#")
  }
}

# Combine all gene-based rows (all TEST types: ADD, ADD-SKAT, ADD-SKATO, …) — do not filter here
gene_based_all <- bind_rows(gene_based_list) %>%
  mutate(across(c(CHROM, GENPOS, N, N_CASES, N_CONTROLS), as.integer),
         across(c(A1FREQ, A1FREQ_CASES, A1FREQ_CONTROLS, BETA, SE, CHISQ, LOG10P), as.numeric)) %>%
  mutate(CHROM = as.factor(CHROM),
         PVAL = 10^(-LOG10P)) %>%
  mutate(
    PVAL_ADJ = p.adjust(PVAL, method = "BH"),
    LOG10P_ADJ = -log10(PVAL_ADJ)
  ) %>%
  # OR only meaningful for additive burden (ADD); leave NA for ADD-SKAT / ADD-SKATO etc.
  mutate(OR = if_else(TEST == "ADD" & is.finite(BETA), exp(BETA), NA_real_))

# Data quality check (all tests)
if(any(is.na(gene_based_all$LOG10P))) {
  warning(sprintf("Warning: %d gene-based tests have NA LOG10P values", sum(is.na(gene_based_all$LOG10P))))
}
if(any(is.infinite(gene_based_all$LOG10P))) {
  warning(sprintf("Warning: %d gene-based tests have infinite LOG10P values (p=0)", sum(is.infinite(gene_based_all$LOG10P))))
}
add_beta <- gene_based_all$BETA[gene_based_all$TEST == "ADD"]
if(length(add_beta) && any(is.finite(add_beta) & abs(add_beta) > 20)) {
  warning("Some ADD burden BETA values are very large (|BETA| > 20). Check Regenie column alignment (CHROM GENPOS ... BETA SE LOG10P).")
}

# REGENIE gene-based ID: GENE.MASK.<aaf_suffix> — suffix can contain dots (0.01, 0.001, …).
# Do NOT use sub(".*\\.(.*)$", ID): for WDR31.M1.0.01 that wrongly yields "01" instead of "0.01".
.regenie_split_gene_mask_thresh <- function(ids) {
  ids <- as.character(ids)
  n <- length(ids)
  g <- rep(NA_character_, n)
  m <- rep(NA_character_, n)
  tr <- rep(NA_character_, n)
  for (i in seq_len(n)) {
    p <- strsplit(ids[i], ".", fixed = TRUE)[[1L]]
    if (length(p) < 3L) next
    g[i] <- p[[1L]]
    m[i] <- p[[2L]]
    tr[i] <- paste(p[seq.int(3L, length(p))], collapse = ".")
  }
  list(GENE = g, MASK = m, THRESH_RAW = tr)
}
gene_id_parts <- .regenie_split_gene_mask_thresh(gene_based_all$ID)
gene_based_all <- gene_based_all %>%
  mutate(
    GENE = gene_id_parts$GENE,
    MASK = gene_id_parts$MASK,
    THRESH_RAW = gene_id_parts$THRESH_RAW,
    THRESH = as.character(ifelse(THRESH_RAW == "all", "All variants", THRESH_RAW)),
    MAF_Filter = THRESH
  )

# Burden (ADD) only: used for BETA/OR, directional language, volcano, mask summaries (mean beta), CHEK2 burden
gene_based_add <- gene_based_all %>% filter(TEST == "ADD")
```

## Summary and quality

### Analysis summary

```{r summary-stats,echo=FALSE,message=FALSE,warning=FALSE}
# Load phenotype file for actual case/control numbers
pheno_file <- paste0("data/final/", params$project_name, ".regenie.pheno.txt")
pheno <- read_delim(pheno_file, delim = " ", col_names = TRUE, trim_ws = TRUE)

# Build proper summary table with scalar values
summary_df <- data.frame(
  Metric = c("Total Samples", "Cases", "Controls",
             "Variants Tested (Single-Variant)",
             "Unique genes tested (all TEST rows)",
             "Unique genes tested (ADD only)"),
  Count = c(
    length(unique(pheno$IID)),
    sum(pheno$STATUS == 1),
    sum(pheno$STATUS == 0),
    nrow(single_variant),
    length(unique(gene_based_all$GENE)),
    length(unique(gene_based_add$GENE))
  )
)

kable(summary_df, format = "html", col.names = c("Metric", "Count")) %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover"), font_size = 13) %>%
  column_spec(1, bold = TRUE)
```

### Quality metrics

```{r quality-metrics,echo=FALSE,message=FALSE,warning=FALSE}
# Calculate lambda GC for both analyses (finite p in (0,1])
single_lambda <- {
  pv <- single_variant$PVAL[is.finite(single_variant$PVAL) & single_variant$PVAL > 0 & single_variant$PVAL < 1]
  if (length(pv) < 10L) NA_real_ else median(qchisq(1 - pv, 1), na.rm = TRUE) / qchisq(0.5, 1)
}
gene_lambda <- {
  pv <- gene_based_add$PVAL[gene_based_add$TEST == "ADD" & is.finite(gene_based_add$PVAL) & gene_based_add$PVAL > 0 & gene_based_add$PVAL < 1]
  if (length(pv) < 10L) NA_real_ else median(qchisq(1 - pv, 1), na.rm = TRUE) / qchisq(0.5, 1)
}

# Simple quality metrics (burden ADD only for gene-based rows)
sig_add_or <- gene_based_add$OR[gene_based_add$PVAL_ADJ < 0.05 & is.finite(gene_based_add$OR)]
max_or_sig <- if (length(sig_add_or)) round(max(sig_add_or, na.rm = TRUE), 2) else NA_real_

quality_df <- data.frame(
    Metric = c(
        "Significant Single-Variants (p < 2.5e-6)",
        "Significant Burden Tests (ADD, FDR < 0.05)",
        "Lambda GC (Single-Variant)",
        "Lambda GC (gene-based ADD only)",
        "Max OR (FDR-significant ADD only)"
    ),
    Value = c(
        format(sum(single_variant$LOG10P > 5.60206, na.rm = TRUE), big.mark = ","),
        format(sum(gene_based_add$PVAL_ADJ < 0.05, na.rm = TRUE), big.mark = ","),
        round(single_lambda, 3),
        round(gene_lambda, 3),
        ifelse(is.finite(max_or_sig), format(max_or_sig, nsmall = 2), "—")
    )
)

kable(quality_df, format = "html",
      col.names = c("Quality Metric", "Value")) %>%
    kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover"), font_size = 13) %>%
    column_spec(1, bold = TRUE)
```

## Single-variant results

Step 2 tests all variants in the annotation pgen; gene-based M1–M4 use only `regenie.annotation.txt` / `regenie.set.txt`.

### Significant variants

```{r significant-single,echo=FALSE,message=FALSE,warning=FALSE,results='asis'}
sig_log10p_thresh <- 5.60206  # ~2.5e-6 exome-wide

# HTML kable output must use results='asis' or knitr wraps it as verbatim (shows as `## <div...` text).
fmt_small_p <- function(x) {
  x <- as.numeric(x)
  ifelse(!is.finite(x), "",
    ifelse(x < 1e-4, formatC(x, format = "e", digits = 2), format(round(x, 5), trim = TRUE))
  )
}

significant_variants <- single_variant %>%
  filter(!is.na(LOG10P), LOG10P > sig_log10p_thresh) %>%
  arrange(desc(LOG10P))

if (nrow(significant_variants) == 0L) {
  cat("None above exome-wide threshold (LOG10P > ", sig_log10p_thresh, ").\n", sep = "")
} else {
  if (!is.null(pathogenic_annotations) && nrow(pathogenic_annotations) > 0L) {
    sig_display <- significant_variants %>%
      dplyr::left_join(pathogenic_annotations, by = "ID") %>%
      dplyr::mutate(
        Variant = paste0("chr", CHROM, ":", GENPOS, " ", REF, ">", ALT),
        burden_extract = dplyr::case_when(
          suppressWarnings(as.numeric(Variant.LoF_level)) == 1 ~ "Pathogenic (in M1 extract)",
          suppressWarnings(as.numeric(Variant.LoF_level)) == 2 ~ "VUS (in M2 extract)",
          TRUE ~ "Not in pathogenic/VUS extract"
        ),
        Gene_lab = ifelse(is.na(Gene) | Gene == "", "—", as.character(Gene))
      ) %>%
      dplyr::select(Variant, ID, Gene_lab, burden_extract, HGVSc, HGVSp, A1FREQ, OR, LOG10P, PVAL, PVAL_ADJ) %>%
      dplyr::arrange(PVAL)

    sig_cap <- paste0("Exome-wide hits (n = ", nrow(sig_display), "); join: pathogenic_vus.csv (M1/M2 extract).")
    sig_view <- sig_display %>%
      dplyr::mutate(PVAL = fmt_small_p(PVAL), PVAL_ADJ = fmt_small_p(PVAL_ADJ))

    knitr::kable(
      sig_view,
      format = "html",
      digits = 4,
      col.names = c(
        "Variant", "ID", "Gene (extract)", "Pathogenic/VUS extract",
        "HGVSc", "HGVSp", "Alt freq", "OR", "−log10(p)", "P-value", "FDR P-value"
      ),
      caption = sig_cap
    ) %>%
      kableExtra::kable_styling(
        full_width = TRUE,
        bootstrap_options = c("striped", "hover", "condensed"),
        font_size = 11
      ) %>%
      kableExtra::column_spec(1, bold = TRUE, extra_css = "word-break: break-all; font-size: 0.92em; max-width: 9rem;") %>%
      kableExtra::column_spec(2, extra_css = "word-break: break-all; font-size: 0.92em; max-width: 11rem;") %>%
      kableExtra::column_spec(3:6, extra_css = "font-size: 0.92em;") %>%
      kableExtra::column_spec(9:11, color = "red") %>%
      kableExtra::scroll_box(width = "100%", height = "420px") %>%
      print()
  } else {
    sig_display <- significant_variants %>%
      dplyr::mutate(Variant = paste0("chr", CHROM, ":", GENPOS, " ", REF, ">", ALT)) %>%
      dplyr::select(Variant, ID, A1FREQ, OR, LOG10P, PVAL, PVAL_ADJ) %>%
      dplyr::arrange(PVAL)

    sig_view <- sig_display %>%
      dplyr::mutate(PVAL = fmt_small_p(PVAL), PVAL_ADJ = fmt_small_p(PVAL_ADJ))

    knitr::kable(
      sig_view,
      format = "html",
      digits = 4,
      caption = paste0("Exome-wide hits (n = ", nrow(sig_display), "); pathogenic_vus.csv not available.")
    ) %>%
      kableExtra::kable_styling(
        full_width = TRUE,
        bootstrap_options = c("striped", "hover", "condensed"),
        font_size = 11
      ) %>%
      kableExtra::column_spec(1, bold = TRUE, extra_css = "word-break: break-all; font-size: 0.92em; max-width: 9rem;") %>%
      kableExtra::column_spec(2, extra_css = "word-break: break-all; font-size: 0.92em; max-width: 11rem;") %>%
      kableExtra::scroll_box(width = "100%", height = "360px") %>%
      print()
  }

  significant_variants_export <- sig_display
}
```

### Manhattan plot

```{r manhattan-plot,echo=FALSE,message=FALSE,warning=FALSE, fig.cap="Single-variant Manhattan (y capped for display)."}
chrom_colors <- rep(c("#2E86AB", "#FF6B35"), length.out = length(unique(single_variant$CHROM)))
hi <- max(single_variant$LOG10P, na.rm = TRUE)
if (!is.finite(hi)) hi <- 10
ylim_u <- min(80, hi + 2)
n_exome_wide <- sum(single_variant$LOG10P > 5.60206, na.rm = TRUE)

manhattan_plot <- ggplot(single_variant, aes(x = POS_INDEX, y = LOG10P, color = CHROM)) +
    geom_point(alpha = 0.55, size = 1.35) +
    scale_color_manual(values = chrom_colors) +
    labs(x = "Genomic position", y = expression(-log[10](italic(p)))) +
    geom_hline(yintercept = 5.602, color = "red", linetype = "dashed", linewidth = 0.6) +
    coord_cartesian(ylim = c(0, ylim_u)) +
    labs(subtitle = paste0("Points above red line: exome-wide (n = ", n_exome_wide, ")")) +
    theme(legend.position = "none")

print(manhattan_plot)
```

### QQ plot

```{r qq-plot,echo=FALSE,message=FALSE,warning=FALSE, fig.cap="Single-variant QQ with genomic inflation (λ) in the subtitle."}
qq_data <- single_variant %>%
    filter(!is.na(LOG10P)) %>%
    arrange(desc(LOG10P)) %>%
    mutate(
        observed = sort(LOG10P, decreasing = TRUE),
        expected = sort(-log10(ppoints(length(LOG10P))), decreasing = TRUE)
    )

lam_txt <- if (is.finite(single_lambda)) paste0("λ = ", round(single_lambda, 3)) else "λ = NA (insufficient finite P-values)"

qq_plot <- ggplot(qq_data, aes(x = expected, y = observed)) +
    geom_point(alpha = 0.55, color = "#2E86AB") +
    geom_abline(intercept = 0, slope = 1, color = "red", linetype = "dashed") +
    labs(
      x = "Expected -log10(p)",
      y = "Observed -log10(p)",
      subtitle = lam_txt
    )

print(qq_plot)
```

Deviation from the diagonal at the tail reflects excess signal. **λ** is the median χ² (1 df) inflation vs null (same idea as in the QC report); it is also listed under **Quality metrics**.

## Gene-based results

Exports use explicit names: **`r paste0(params$project_name, "_gene_based_results_alltests_", format(Sys.Date(), "%Y%m%d"), ".csv")`** (all `TEST` values), **`r paste0(params$project_name, "_gene_based_results_ADDonly_", format(Sys.Date(), "%Y%m%d"), ".csv")`** (ADD only), and **`r paste0(params$project_name, "_gene_based_meta_input_", format(Sys.Date(), "%Y%m%d"), ".csv")`** (columns used for cross-freeze meta). Use **all-tests** for archive/SKAT/SKATO; **meta_input** matches downstream meta pipelines.

### Burden (ADD) — summary by mask

```{r mask-summary,echo=FALSE,message=FALSE,warning=FALSE}
mask_summary <- gene_based_add %>%
  group_by(MASK) %>%
  summarize(
    Tests = n(),
    Significant_Tests = sum(PVAL_ADJ < 0.05, na.rm = TRUE),
    Most_Significant_Gene = GENE[which.max(LOG10P)],
    Max_Log10P = max(pmin(LOG10P, 100), na.rm = TRUE),
    Mean_Beta = mean(BETA, na.rm = TRUE)
  ) %>%
  mutate(Significance_Rate = Significant_Tests / Tests * 100)

kable(mask_summary, format = "html", digits = 3,
      col.names = c("Mask", "Tests (gene×mask×threshold)", "FDR significant tests", "Top Gene", "Max −log10(P) (display cap 100)", "Mean Beta", "Significance Rate (%)")) %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover"), font_size = 13) %>%
  column_spec(1, bold = TRUE)
```

**Mask note:** `M4` is a negative-control mask built from synonymous variants only.
It explicitly excludes `M1/M2` variants and a blacklist of potential MNP-pair IDs that did not pass full validation, to reduce false-negative MNP contamination in controls.

### Top genes by clinical category (burden ADD only)

Gene-based REGENIE is run with **`--aaf-bins 0.01,0.001,0.0001`**, so each gene is tested under several **allele-frequency ceilings** in addition to **singleton** and **all variants** (REGENIE’s `all` bin is shown here as **“All variants”**). Stricter MAF bins keep rarer variation; broader bins add more variants per gene (often lower power per variant, different correlation structure).

```{r topgenes-add-helper, include=FALSE}
# Used by one chunk per MAF bin below (real Markdown headings; kable as last expr per chunk).
regenie_topgenes_add_table <- function(mask, thresh) {
  mask_chr <- mask
  thresh_chr <- thresh
  tab <- gene_based_add %>%
    dplyr::filter(MASK == mask_chr, THRESH == thresh_chr) %>%
    dplyr::slice_max(LOG10P, n = 10L, with_ties = FALSE) %>%
    dplyr::select(GENE, BETA, SE, LOG10P, PVAL, PVAL_ADJ, OR) %>%
    dplyr::arrange(dplyr::desc(LOG10P))
  if (nrow(tab) == 0L) return(invisible(NULL))
  knitr::kable(
    tab,
    format = "html",
    digits = 4,
    col.names = c("Gene", "Beta", "SE", "-log10(P)", "P-value", "FDR P-value", "OR"),
    caption = paste0("Top genes (ADD): ", mask, " — ", thresh)
  ) %>%
    kableExtra::kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover"), font_size = 13) %>%
    kableExtra::column_spec(1, bold = TRUE) %>%
    kableExtra::column_spec(c(2, 4, 5), color = "red")
}
```

#### M1 (Pathogenic variants only)

##### M1 — singleton

```{r topgenes-m1-singleton, echo=FALSE, message=FALSE, warning=FALSE}
regenie_topgenes_add_table("M1", "singleton")
```

##### M1 — All variants

```{r topgenes-m1-all-variants, echo=FALSE, message=FALSE, warning=FALSE}
regenie_topgenes_add_table("M1", "All variants")
```

##### M1 — 0.01

```{r topgenes-m1-af01, echo=FALSE, message=FALSE, warning=FALSE}
regenie_topgenes_add_table("M1", "0.01")
```

##### M1 — 0.001

```{r topgenes-m1-af001, echo=FALSE, message=FALSE, warning=FALSE}
regenie_topgenes_add_table("M1", "0.001")
```

##### M1 — 0.0001

```{r topgenes-m1-af0001, echo=FALSE, message=FALSE, warning=FALSE}
regenie_topgenes_add_table("M1", "0.0001")
```

#### M2 (VUS variants only)

##### M2 — singleton

```{r topgenes-m2-singleton, echo=FALSE, message=FALSE, warning=FALSE}
regenie_topgenes_add_table("M2", "singleton")
```

##### M2 — All variants

```{r topgenes-m2-all-variants, echo=FALSE, message=FALSE, warning=FALSE}
regenie_topgenes_add_table("M2", "All variants")
```

##### M2 — 0.01

```{r topgenes-m2-af01, echo=FALSE, message=FALSE, warning=FALSE}
regenie_topgenes_add_table("M2", "0.01")
```

##### M2 — 0.001

```{r topgenes-m2-af001, echo=FALSE, message=FALSE, warning=FALSE}
regenie_topgenes_add_table("M2", "0.001")
```

##### M2 — 0.0001

```{r topgenes-m2-af0001, echo=FALSE, message=FALSE, warning=FALSE}
regenie_topgenes_add_table("M2", "0.0001")
```

#### M3 (Pathogenic + VUS combined)

##### M3 — singleton

```{r topgenes-m3-singleton, echo=FALSE, message=FALSE, warning=FALSE}
regenie_topgenes_add_table("M3", "singleton")
```

##### M3 — All variants

```{r topgenes-m3-all-variants, echo=FALSE, message=FALSE, warning=FALSE}
regenie_topgenes_add_table("M3", "All variants")
```

##### M3 — 0.01

```{r topgenes-m3-af01, echo=FALSE, message=FALSE, warning=FALSE}
regenie_topgenes_add_table("M3", "0.01")
```

##### M3 — 0.001

```{r topgenes-m3-af001, echo=FALSE, message=FALSE, warning=FALSE}
regenie_topgenes_add_table("M3", "0.001")
```

##### M3 — 0.0001

```{r topgenes-m3-af0001, echo=FALSE, message=FALSE, warning=FALSE}
regenie_topgenes_add_table("M3", "0.0001")
```

#### M4 (Synonymous negative-control)

```{r top-genes-m4,echo=FALSE,message=FALSE,warning=FALSE}
# M4 (synonymous negative-control) - Top 10 genes
m4_top <- gene_based_add %>%
  filter(MASK == "M4") %>%
  slice_max(LOG10P, n = 10, with_ties = FALSE) %>%
  select(GENE, THRESH, BETA, SE, LOG10P, PVAL, PVAL_ADJ, OR) %>%
  arrange(desc(LOG10P))

if(nrow(m4_top) > 0) {
  kable(m4_top, format = "html", digits = 4,
        col.names = c("Gene", "MAF Filter", "Beta", "SE", "-log10(P)", "P-value", "FDR P-value", "OR")) %>%
    kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover"), font_size = 13) %>%
    column_spec(1, bold = TRUE) %>%
    column_spec(c(3, 5, 6), color = "red")
} else {
  invisible(NULL)
}
```

### Suggestive ADD burden signals

_Exploratory table: strong raw p but not FDR-significant after genome-wide BH across all gene-based tests._

```{r suggestive-genes,echo=FALSE,message=FALSE,warning=FALSE}
top_by_raw_p <- gene_based_add %>%
  group_by(MASK) %>%
  slice_max(LOG10P, n = 10, with_ties = FALSE) %>%
  ungroup()

suggestive_genes <- top_by_raw_p %>%
  filter(LOG10P > 3, PVAL_ADJ >= 0.05) %>%
  arrange(PVAL) %>%
  slice_head(n = 15) %>%
  select(GENE, MASK, THRESH, BETA, LOG10P, PVAL, PVAL_ADJ, OR)

if(nrow(suggestive_genes) > 0) {
    kable(suggestive_genes,
          format = "html", digits = 4,
          caption = "Suggestive ADD burden signals (exploratory)",
          col.names = c("Gene", "Mask", "MAF Filter", "Beta", "Raw -log10(P)", "Raw P-value", "FDR P-value", "OR")) %>%
        kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover"),
                      font_size = 13) %>%
        column_spec(1, bold = TRUE) %>%
        column_spec(c(4, 8), color = "orange")
} else {
    cat("No suggestive associations found")
}
```

### Volcano (ADD)

```{r volcano-plot,echo=FALSE,message=FALSE,warning=FALSE, fig.cap="Burden ADD by mask (−log10 p capped at 100 in plot)."}
gene_based_plot <- gene_based_add %>%
  mutate(
    Significant = PVAL_ADJ < 0.05,
    LOG10P_CAP = pmin(LOG10P, 100)
  )

top_hits <- gene_based_plot %>%
  filter(Significant) %>%
  group_by(MASK) %>%
  slice_max(LOG10P_CAP, n = 3, with_ties = FALSE) %>%
  ungroup()

gene_based_plot <- gene_based_plot %>%
  mutate(Label = ifelse(GENE %in% top_hits$GENE, GENE, NA))

fdr_cutoff <- suppressWarnings(min(gene_based_add$LOG10P[gene_based_add$PVAL_ADJ < 0.05], na.rm = TRUE))
if (!is.finite(fdr_cutoff)) fdr_cutoff <- NA_real_
fdr_cutoff_plot <- if (is.finite(fdr_cutoff)) pmin(fdr_cutoff, 100) else NA_real_

volcano_plot <- ggplot(gene_based_plot, aes(x = as.numeric(BETA), y = LOG10P_CAP, color = Significant)) +
  geom_point(alpha = 0.55, size = 1.35)
if (is.finite(fdr_cutoff_plot)) {
  volcano_plot <- volcano_plot +
    geom_hline(yintercept = fdr_cutoff_plot, linetype = "dashed", color = "blue", alpha = 0.65)
}
volcano_plot <- volcano_plot +
  geom_vline(xintercept = 0, linetype = "dotted", color = "grey40") +
  geom_text_repel(aes(label = Label), size = 2.6, max.overlaps = 12,
                  box.padding = 0.3, point.padding = 0.1) +
  scale_color_manual(values = c("gray60", "red"),
                     labels = c("Not Significant", "FDR < 0.05")) +
  scale_x_continuous(limits = c(NA, NA), oob = scales::oob_keep) +
  labs(
       x = "Beta (log-OR)",
       y = "-log10(p), capped at 100",
       color = NULL) +
  facet_wrap(~MASK, scales = "free_y") +
  theme(legend.position = "bottom",
        strip.text = element_text(size = 11, face = "bold"))

print(volcano_plot)
```

### CHEK2 (ADD only)

_Exploratory gene-level summary only; it does not address variant-level missingness, capture, or per-sample carrier data._

```{r chek2-analysis,echo=FALSE,message=FALSE,warning=FALSE}
chek2_results <- gene_based_add %>%
  filter(GENE == "CHEK2") %>%
  select(GENE, MASK, THRESH, CHROM, GENPOS, ALLELE1, A1FREQ, BETA, OR, SE, LOG10P, PVAL, PVAL_ADJ, LOG10P_ADJ) %>%
  arrange(MASK, THRESH)

if(nrow(chek2_results) > 0) {
  kable(chek2_results,
        format = "html", digits = 4,
        col.names = c("Gene", "Mask", "Threshold", "Chr", "Pos", "Alt Allele", "Alt Freq",
                     "Beta", "OR", "SE", "-log10(P)", "P-value", "FDR P-value", "-log10(FDR P)")) %>%
    kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover"), font_size = 13) %>%
    column_spec(c(8, 10, 11), color = ifelse(!is.na(chek2_results$PVAL_ADJ) & chek2_results$PVAL_ADJ < 0.05, "red", "black"))
} else {
  cat("_CHEK2 not found in ADD burden results._")
}
```

## Directional burden (ADD)

### Risk-increasing genes (Beta > 0, FDR < 0.05)

```{r risk-genes,echo=FALSE,message=FALSE,warning=FALSE}
risk_genes <- gene_based_add %>%
  filter(PVAL_ADJ < 0.05, BETA > 0) %>%
  select(GENE, MASK, THRESH, BETA, PVAL_ADJ) %>%
  arrange(PVAL_ADJ)

if(nrow(risk_genes) > 0) {
    kable(risk_genes,
          format = "html", digits = 4,
          col.names = c("Gene", "Mask", "MAF Filter", "Beta", "FDR P-value")) %>%
      kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover"),
                    font_size = 13) %>%
      column_spec(1, bold = TRUE) %>%
      column_spec(c(4, 5), color = "red")
} else {
    cat("No risk-increasing gene associations found")
}
```

### Protective genes (Beta < 0, FDR < 0.05)

```{r protective-genes,echo=FALSE,message=FALSE,warning=FALSE}
protective_genes <- gene_based_add %>%
  filter(PVAL_ADJ < 0.05, BETA < 0) %>%
  select(GENE, MASK, THRESH, BETA, PVAL_ADJ) %>%
  arrange(PVAL_ADJ)

if(nrow(protective_genes) > 0) {
    kable(protective_genes,
          format = "html", digits = 4,
          col.names = c("Gene", "Mask", "MAF Filter", "Beta", "FDR P-value")) %>%
      kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover"),
                    font_size = 13) %>%
      column_spec(1, bold = TRUE) %>%
      column_spec(c(4, 5), color = "steelblue")
} else {
    cat("No protective gene associations found")
}
```

## Output files

```{r save-results,echo=FALSE,message=FALSE,warning=FALSE}
date_stamp <- format(Sys.Date(), "%Y%m%d")
single_file <- paste0(params$project_name, "_single_variant_results_", date_stamp, ".csv")
write_csv(single_variant, single_file)
cat("**All single variant results saved to:** `", single_file, "`\n\n")

gene_file_all <- paste0(params$project_name, "_gene_based_results_alltests_", date_stamp, ".csv")
gene_file_add <- paste0(params$project_name, "_gene_based_results_ADDonly_", date_stamp, ".csv")
gene_meta_fp <- paste0(params$project_name, "_gene_based_meta_input_", date_stamp, ".csv")
write_csv(gene_based_all, gene_file_all)
write_csv(gene_based_add, gene_file_add)

meta_cols <- c("GENE", "MASK", "THRESH", "TEST", "BETA", "SE", "PVAL", "N_CASES", "N_CONTROLS", "A1FREQ_CASES", "A1FREQ_CONTROLS")
gene_meta_input <- gene_based_all %>% select(any_of(meta_cols))
write_csv(gene_meta_input, gene_meta_fp)

cat("**Gene-based REGENIE tables:**\n\n")
cat("- **All tests:** `", gene_file_all, "`\n")
cat("- **ADD only (burden):** `", gene_file_add, "`\n")
cat("- **Meta input (columns for downstream meta):** `", gene_meta_fp, "`\n\n")

write_csv(
  dplyr::filter(gene_based_all, GENE == "CHEK2"),
  paste0(params$project_name, "_CHEK2_alltests_", date_stamp, ".csv")
)
write_csv(
  dplyr::filter(gene_based_add, GENE == "CHEK2"),
  paste0(params$project_name, "_CHEK2_ADDonly_", date_stamp, ".csv")
)
cat("**CHEK2 quick extracts:** `", params$project_name, "_CHEK2_alltests_", date_stamp, ".csv` and `", params$project_name, "_CHEK2_ADDonly_", date_stamp, ".csv`\n\n")

top_gene_signals <- gene_based_add %>%
  filter(PVAL_ADJ < 0.05) %>%
  arrange(PVAL, GENE, MASK, THRESH)
top_signals_fp <- paste0(params$project_name, "_top_gene_signals_ADDonly_", date_stamp, ".csv")
write_csv(top_gene_signals, top_signals_fp)
cat("**Top ADD gene signals (FDR < 0.05):** `", top_signals_fp, "`\n\n")

if(nrow(significant_variants) > 0) {
    sig_single_file <- paste0(params$project_name, "_significant_single_variants_", date_stamp, ".csv")
    write_csv(significant_variants_export, sig_single_file)
    cat("**Exome-wide singles:** `", sig_single_file, "`\n\n", sep = "")
} else {
    cat("**No variants reached exome-wide significance (p < 2.5e-6)**\n\n")
}

if(exists("suggestive_genes") && nrow(suggestive_genes) > 0) {
    sig_gene_file <- paste0(params$project_name, "_suggestive_gene_associations_ADDonly_", date_stamp, ".csv")
    write_csv(suggestive_genes, sig_gene_file)
    cat("**Suggestive ADD-only (exploratory) saved to:** `", sig_gene_file, "`\n\n")
} else {
    cat("**No suggestive ADD-only gene associations found**\n\n")
}

cat("**Analysis completed on:**", format(Sys.time(), "%Y-%m-%d %H:%M:%S"))
```

## Appendix: REGENIE pipeline flowchart

```{r regenie_flowchart, echo=FALSE, fig.height=7}
grViz("
digraph regenie_pipeline {

  graph [rankdir = TB, fontsize = 12, fontname = Helvetica]
  node [fontname = Helvetica, fontsize = 11, shape = box, style = filled, color = DarkBlue]

  input_vcf [label = 'Multi-chromosome VCF\\nfiles', fillcolor = Khaki, fontcolor = Black, shape = folder]
  samples [label = 'Sample list', fillcolor = Khaki, fontcolor = Black, shape = note]
  pheno [label = 'Phenotypes', fillcolor = Khaki, fontcolor = Black, shape = note]
  covar [label = 'Covariates', fillcolor = Khaki, fontcolor = Black, shape = note]

  preprocess [label = 'VCF Preprocessing\\n(normalize, filter, QC)', fillcolor = LightSkyBlue]
  pgen_convert [label = 'Convert to PLINK2\\nPGEN format', fillcolor = LightSkyBlue]

  step1_chr1 [label = 'Step 1\\nChr 1\\n(covariate adjustment)', fillcolor = LavenderBlush, shape = parallelogram]
  step1_chr_ellipsis [label = '...', fillcolor = LavenderBlush, shape = parallelogram]
  step1_chr22 [label = 'Step 1\\nChr 22\\n(covariate adjustment)', fillcolor = LavenderBlush, shape = parallelogram]

  merge_step1 [label = 'Merge Step 1\\nresults', fillcolor = MistyRose, shape = diamond]

  step2_single [label = 'Step 2 Single-Variant\\nAssociation Testing\\nChr 1-22', fillcolor = HoneyDew, shape = parallelogram]
  step2_gene [label = 'Step 2 Gene-Based\\nAssociation Testing\\nChr 1-22', fillcolor = HoneyDew, shape = parallelogram]

  vep_annotate [label = 'VEP Annotation\\n(pathogenic/VUS variants)', fillcolor = LemonChiffon]
  report_gen [label = 'Generate HTML Report\\n(statistics, plots, tables)', fillcolor = AliceBlue, shape = note]

  input_vcf -> preprocess
  samples -> preprocess
  preprocess -> pgen_convert
  preprocess -> vep_annotate

  pgen_convert -> step1_chr1
  pgen_convert -> step1_chr_ellipsis
  pgen_convert -> step1_chr22

  step1_chr1 -> merge_step1 [style = dotted]
  step1_chr_ellipsis -> merge_step1 [style = dotted]
  step1_chr22 -> merge_step1 [style = dotted]

  merge_step1 -> step2_single
  merge_step1 -> step2_gene
  vep_annotate -> step2_gene

  step2_single -> report_gen
  step2_gene -> report_gen

  {rank = same; step1_chr1; step1_chr_ellipsis; step1_chr22}
  {rank = same; step2_single; step2_gene}

  subgraph cluster_parallel {
    label = 'Parallel Processing\\n(Chromosome-level)'
    fontsize = 14
    style = filled
    fillcolor = WhiteSmoke
    step1_chr1
    step1_chr_ellipsis
    step1_chr22
    step2_single
    step2_gene
  }
}
")
```