---
title: "Variant-level follow-up"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```

```{r libs}
library(readr)
library(dplyr)
library(tibble)
library(knitr)
```

```{r config}
# Edit here only. No YAML params, no render(params=...). Snakemake GENES should match genes_to_report.
data_dir <- "data/variant"
genes_to_report <- c("CHEK2")
genes_source_file <- "" # genes-from-file chunk: set path when eval=TRUE
```

```{r genes-from-file, eval=FALSE}
# eval=TRUE: set genes_source_file in chunk config; table needs column Gene.
if (!nzchar(genes_source_file)) {
  stop("genes-from-file: set genes_source_file in chunk config")
}
if (!file.exists(genes_source_file)) {
  stop("genes-from-file: file not found: ", genes_source_file)
}
pv_all <- read_csv(genes_source_file, show_col_types = FALSE)
if (!"Gene" %in% names(pv_all)) {
  stop("genes-from-file: input must contain a column named Gene")
}
genes_to_report <- sort(unique(as.character(pv_all$Gene[!is.na(pv_all$Gene) & nzchar(as.character(pv_all$Gene))])))
if (!length(genes_to_report)) {
  stop("genes-from-file: no non-empty Gene values in file")
}
```

Brief: **Genes** and **`data_dir`** are set in chunk **`config`** only. Optional: **`genes-from-file`** with **`eval=TRUE`**. Subtables: **Overall**, **By STATUS**, **By FREEZE**, **By STATUS × FREEZE**.

---

```{r gene-loop, results='asis'}
style_kable <- function(x, caption = NULL) {
  k <- knitr::kable(x, caption = caption, row.names = FALSE)
  if (requireNamespace("kableExtra", quietly = TRUE)) {
    k <- kableExtra::kable_styling(
      k,
      bootstrap_options = c("striped", "hover", "condensed"),
      full_width = FALSE,
      position = "left"
    )
  }
  k
}

read_gene_summary <- function(gene, variant_set) {
  subdir <- data_dir
  fname <- if (identical(variant_set, "pathogenic")) {
    sprintf("variants-pathogenic-summary.%s.tsv", gene)
  } else {
    sprintf("variants-vus-summary.%s.tsv", gene)
  }
  path <- file.path(subdir, fname)
  if (!file.exists(path)) {
    return(tibble())
  }
  read_tsv(path, show_col_types = FALSE)
}

split_by_scope <- function(df) {
  if (!nrow(df)) {
    return(list(
      overall = tibble(),
      status = tibble(),
      freeze = tibble(),
      status_freeze = tibble()
    ))
  }
  list(
    overall = df %>% filter(scope == "overall"),
    status = df %>% filter(scope == "status"),
    freeze = df %>% filter(scope == "freeze"),
    status_freeze = df %>% filter(scope == "status_freeze")
  )
}

cols_summary <- c(
  "variant_id", "variant_type", "status", "freeze",
  "n_total", "n_called", "n_missing", "n_ref", "n_het", "n_hom", "call_rate"
)

for (g in genes_to_report) {
  cat("\n\n")
  cat("## ", g, "\n\n", sep = "")

  for (vs_lab in c("pathogenic", "VUS")) {
    vs_key <- if (vs_lab == "pathogenic") "pathogenic" else "vus"
    sum_df <- read_gene_summary(g, vs_key)
    sec_title <- if (vs_lab == "pathogenic") "Pathogenic" else "VUS"
    cat("### ", sec_title, "\n\n", sep = "")
    if (!nrow(sum_df)) {
      miss_path <- file.path(
        data_dir,
        paste0("variants-", vs_key, "-summary.", g, ".tsv")
      )
      cat("_Missing summary file for ", vs_lab, " (expected: `", miss_path, "`)._ \n\n", sep = "")
      next
    }
    if (all(c("variant_id", "variant_type") %in% names(sum_df))) {
      cat("#### Variants in this file\n\n")
      id_tbl <- sum_df %>%
        distinct(variant_id, variant_type) %>%
        arrange(variant_id)
      print(style_kable(
        id_tbl,
        caption = paste0(g, " — ", vs_lab, " — variant IDs in summary TSV")
      ))
      cat("\n\n")
    }
    parts <- split_by_scope(sum_df)
    cat("#### Overall\n\n")
    if (nrow(parts$overall)) {
      print(style_kable(
        parts$overall %>% select(any_of(cols_summary)),
        caption = paste0(g, " — ", vs_lab, " — overall")
      ))
    } else {
      cat("_No overall rows._\n\n")
    }
    cat("\n\n#### By STATUS\n\n")
    if (nrow(parts$status)) {
      print(style_kable(
        parts$status %>% select(any_of(cols_summary)),
        caption = paste0(g, " — ", vs_lab, " — by STATUS")
      ))
    } else {
      cat("_No status-stratified rows._\n\n")
    }
    cat("\n\n#### By FREEZE\n\n")
    if (nrow(parts$freeze)) {
      print(style_kable(
        parts$freeze %>% select(any_of(cols_summary)),
        caption = paste0(g, " — ", vs_lab, " — by FREEZE")
      ))
    } else {
      cat("_No freeze-stratified rows._\n\n")
    }
    cat("\n\n#### By STATUS × FREEZE\n\n")
    if (nrow(parts$status_freeze)) {
      print(style_kable(
        parts$status_freeze %>% select(any_of(cols_summary)),
        caption = paste0(g, " — ", vs_lab, " — by STATUS and FREEZE")
      ))
    } else {
      cat("_No STATUS×FREEZE rows._\n\n")
    }
    cat("\n\n")
  }

  cat(
    "<!-- Plots / tests (e.g. carrier rates by STATUS, FREEZE, interaction): add chunks below for ",
    g,
    " -->\n\n",
    sep = ""
  )
}
```

---

# Notes

- **Gene list:** `genes_to_report` in chunk `config` (keep in sync with Snakemake `GENES`). Or `genes-from-file` with `eval=TRUE` and `genes_source_file` set in `config`.
- **Comparisons:** pathogenic vs VUS are separate summary files; within each, strata are **control/case** (from STATUS), **FREEZE**, and both. Formal tests can use `n_ref` / `n_het` / `n_hom` / `n_missing`.
- **Inputs:** `data_dir` in chunk `config` — `variants-pathogenic-summary.{GENE}.tsv`, `variants-vus-summary.{GENE}.tsv`.
