---
title: "Variant QC Report"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: hide
    theme: flatly
params:
  project_name: "PROJECT"
  date: !r format(Sys.Date(), "%Y%m%d")
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE, fig.width=10, fig.height=6)
library(tidyverse)
library(knitr)
library(kableExtra)

project <- params$project_name
```

# Quality Control Summary

**Project:** `r project`  
**Date:** `r params$date`

## Overview

This report documents the quality control steps applied to exome sequencing data prior to association analysis. QC was performed on individual chromosome files (chr1-22, chrX) with sequential filtering steps.

---

# Sex Check (Chromosome X)

Sex discordance can indicate sample mix-ups or contamination. For this male-only study, only samples with confirmed male genetic sex are retained.

```{r sex_check}
sexcheck_file <- paste0("data/qc/temp/", project, ".chrX.sexcheck")
sex_exclusions_file <- paste0("data/qc/exclusions/", project, ".sexcheck_fail.txt")
sex_info_file <- paste0("data/qc/", project, ".sex_info.txt")

if(file.exists(sexcheck_file)) {
  sexcheck <- read_table(sexcheck_file, col_types=cols())
  
  sexcheck <- sexcheck %>%
    mutate(
      Genetic_Sex=case_when(
        PEDSEX==1 ~ "Male",
        PEDSEX==2 ~ "Female",
        TRUE ~ "Unknown"
      ),
      Status=case_when(
        STATUS=="OK" & PEDSEX==1 ~ "Pass",
        STATUS=="PROBLEM" ~ "Problem",
        PEDSEX==2 ~ "Female",
        TRUE ~ "Fail"
      )
    )
  
  sex_summary <- sexcheck %>%
    count(Status) %>%
    mutate(Percentage=paste0(round(n/sum(n)*100,2),"%"))
  
  kable(sex_summary, format="html", caption="Sex Check Results (Male-Only Study)") %>%
    kable_styling(bootstrap_options=c("striped","hover","condensed"), full_width=FALSE)
  
  failed_count <- sum(sexcheck$Status=="Fail")
  problem_count <- sum(sexcheck$Status=="Problem")
  female_count <- sum(sexcheck$Status=="Female")
  
  cat("\n\n**Sex Check Summary:**\n")
  cat("- Sex check problems:", problem_count, "\n")
  cat("- Females detected:", female_count, "\n")
  cat("- Other failures:", failed_count, "\n")
  cat("- Total excluded:", problem_count + female_count + failed_count, "\n\n")
  
  if(female_count > 0) {
    cat("**WARNING: Females detected in male-only study. Review exclusion list.**\n\n")
  }
  
  passing_samples <- sexcheck %>%
    filter(Status=="Pass") %>%
    select(`#FID`, IID) %>%
    mutate(SEX="M")
  
  write.table(passing_samples, sex_info_file, quote=FALSE, row.names=FALSE, col.names=TRUE, sep="\t")
  cat("**Sex info file created:**", sex_info_file, "\n")
  cat("**Passing samples:**", nrow(passing_samples), "\n\n")
  
} else {
  cat("Sex check file not found.\n")
}
```

---

# Variant-Level QC (Per Chromosome)

After applying sex check exclusions and bcftools filters (NoHQHet, PASS only), variants were filtered based on:

- **Missingness:** GENO > 5%
- **Minor Allele Frequency:** MAF < 1%
- **Hardy-Weinberg Equilibrium:** HWE p < 1e-6

```{r variant_qc}
# Read variant filtering logs
chr_list <- 1:22
variant_stats <- data.frame()

for(chr in chr_list) {
  log_file <- paste0("data/qc/temp/", project, ".chr", chr, ".variants_filtered.log")
  
  if(file.exists(log_file)) {
    log_lines <- readLines(log_file)
    
    variants_before <- log_lines[grep("variants loaded", log_lines)]
    variants_after <- log_lines[grep("variants remaining", log_lines)]
    
    if(length(variants_before) > 0 && length(variants_after) > 0) {
      before <- as.numeric(str_extract(variants_before, "[0-9]+"))
      after <- as.numeric(str_extract(variants_after, "[0-9]+"))
      removed <- before - after
      
      variant_stats <- rbind(variant_stats, data.frame(
        Chromosome=paste0("chr",chr),
        Variants_Before=before,
        Variants_After=after,
        Variants_Removed=removed,
        Percent_Removed=round(removed/before*100,2)
      ))
    }
  }
}

if(nrow(variant_stats) > 0) {
  kable(variant_stats, format="html", caption="Variant Filtering Results by Chromosome") %>%
    kable_styling(bootstrap_options=c("striped","hover","condensed"))
  
  total_before <- sum(variant_stats$Variants_Before)
  total_after <- sum(variant_stats$Variants_After)
  total_removed <- sum(variant_stats$Variants_Removed)
  
  cat("\n\n**Total variants across chr1-22:**\n\n")
  cat("- Before filtering:", format(total_before, big.mark=","), "\n")
  cat("- After filtering:", format(total_after, big.mark=","), "\n")
  cat("- Removed:", format(total_removed, big.mark=","), "(", round(total_removed/total_before*100,2), "%)\n\n")
} else {
  cat("No variant filtering logs found.\n")
}
```

---

# Sample-Level QC

## Sample Missingness (Chr1-22)

Samples with high missingness rates indicate poor sequencing quality or technical failures.

```{r sample_missingness}
miss_report_file <- paste0("data/qc/reports/", project, ".missingness_report.txt")
miss_exclusions_file <- paste0("data/qc/exclusions/", project, ".high_missingness.txt")

if(file.exists(miss_report_file)) {
  cat(readLines(miss_report_file), sep="\n")
  cat("\n\n")
}

if(file.exists(miss_exclusions_file)) {
  miss_exclusions <- read_table(miss_exclusions_file, col_names=c("FID","IID"), col_types=cols())
  n_excluded <- nrow(miss_exclusions)
  cat("**Samples excluded due to high missingness (MIND > 2%):**", n_excluded, "\n\n")
} else {
  cat("No samples excluded due to high missingness.\n\n")
}
```

## Heterozygosity (Chr1-22)

Extreme heterozygosity rates can indicate contamination or inbreeding.

```{r heterozygosity}
het_exclusions_file <- paste0("data/qc/exclusions/", project, ".het_outliers.txt")

if(file.exists(het_exclusions_file)) {
  het_exclusions <- read_table(het_exclusions_file, col_names=c("FID","IID"), col_types=cols())
  n_excluded <- nrow(het_exclusions)
  cat("**Samples excluded due to heterozygosity outliers (±3 SD):**", n_excluded, "\n\n")
} else {
  cat("No samples excluded due to heterozygosity outliers.\n\n")
}

het_agg_file <- paste0("data/qc/temp/", project, ".aggregated_het.txt")
if(file.exists(het_agg_file)) {
  het_data <- read_table(het_agg_file, col_types=cols())
  
  if(nrow(het_data) > 0 && "Fhat" %in% colnames(het_data)) {
    ggplot(het_data, aes(x=Fhat)) +
      geom_histogram(bins=50, fill="steelblue", color="black", alpha=0.7) +
      geom_vline(xintercept=mean(het_data$Fhat, na.rm=TRUE), color="red", linetype="dashed", size=1) +
      geom_vline(xintercept=mean(het_data$Fhat, na.rm=TRUE) + 3*sd(het_data$Fhat, na.rm=TRUE), color="orange", linetype="dotted") +
      geom_vline(xintercept=mean(het_data$Fhat, na.rm=TRUE) - 3*sd(het_data$Fhat, na.rm=TRUE), color="orange", linetype="dotted") +
      labs(title="Distribution of Sample Heterozygosity (Fhat)", x="Fhat", y="Count") +
      theme_minimal()
  }
}
```

---

# Principal Component Analysis

PCA on merged autosomal chromosomes (chr1-22) after all sample exclusions.

## PCA Plots

```{r pca}
pca_file <- paste0("data/plink/", project, ".pca.eigenvec")
site_file <- "sites.txt"
pca_outliers_file <- paste0("data/qc/exclusions/", project, ".pca_outliers.txt")

if(file.exists(pca_file)) {
  pca_data <- read_table(pca_file, col_types=cols())
  
  pca_data <- pca_data %>%
    mutate(
      PC_dist=sqrt(PC1^2+PC2^2+PC3^2+PC4^2),
      mean_dist=mean(PC_dist, na.rm=TRUE),
      sd_dist=sd(PC_dist, na.rm=TRUE),
      z_score=(PC_dist-mean_dist)/sd_dist,
      Outlier=abs(z_score)>6
    )
  
  if(file.exists(site_file)) {
    site_data <- read_table(site_file, col_types=cols())
    pca_data <- pca_data %>%
      left_join(site_data, by=c("#FID"="FID","IID"="IID"))
    
    if("SITE" %in% colnames(pca_data)) {
      ggplot(pca_data, aes(x=PC1, y=PC2, color=SITE, shape=Outlier)) +
        geom_point(alpha=0.6, size=2) +
        scale_shape_manual(values=c("FALSE"=16,"TRUE"=4)) +
        labs(title="Principal Component Analysis (PC1 vs PC2)", x="PC1", y="PC2") +
        theme_minimal() +
        theme(legend.position="right")
      
      ggplot(pca_data, aes(x=PC3, y=PC4, color=SITE, shape=Outlier)) +
        geom_point(alpha=0.6, size=2) +
        scale_shape_manual(values=c("FALSE"=16,"TRUE"=4)) +
        labs(title="Principal Component Analysis (PC3 vs PC4)", x="PC3", y="PC4") +
        theme_minimal() +
        theme(legend.position="right")
      
      site_summary <- pca_data %>%
        count(SITE) %>%
        arrange(desc(n))
      
      kable(site_summary, format="html", caption="Sample Counts by Site") %>%
        kable_styling(bootstrap_options=c("striped","hover","condensed"), full_width=FALSE)
      
    } else {
      cat("SITE column not found in sites.txt file.\n")
    }
  } else {
    cat("sites.txt file not found. Plotting without site information.\n\n")
    
    ggplot(pca_data, aes(x=PC1, y=PC2, color=Outlier)) +
      geom_point(alpha=0.6, size=2) +
      scale_color_manual(values=c("FALSE"="steelblue","TRUE"="red")) +
      labs(title="Principal Component Analysis (PC1 vs PC2)", subtitle="Outliers identified as |z| > 6 SD from centroid", x="PC1", y="PC2") +
      theme_minimal()
    
    ggplot(pca_data, aes(x=PC3, y=PC4, color=Outlier)) +
      geom_point(alpha=0.6, size=2) +
      scale_color_manual(values=c("FALSE"="steelblue","TRUE"="red")) +
      labs(title="Principal Component Analysis (PC3 vs PC4)", subtitle="Outliers identified as |z| > 6 SD from centroid", x="PC3", y="PC4") +
      theme_minimal()
  }
  
  outliers <- pca_data %>% filter(Outlier)
  n_outliers <- nrow(outliers)
  
  cat("\n\n**PCA Outlier Detection:**\n")
  cat("- Distance threshold: |z-score| > 6 SD from centroid (PC1-4)\n")
  cat("- Mean distance:", round(mean(pca_data$PC_dist, na.rm=TRUE), 3), "\n")
  cat("- SD distance:", round(sd(pca_data$PC_dist, na.rm=TRUE), 3), "\n")
  cat("- Outliers detected:", n_outliers, "\n\n")
  
  if(n_outliers>0) {
    outliers %>%
      select(`#FID`, IID) %>%
      write.table(pca_outliers_file, quote=FALSE, row.names=FALSE, col.names=FALSE, sep=" ")
    
    cat("**PCA outlier exclusion list created:**", pca_outliers_file, "\n\n")
    
    outliers_summary <- outliers %>%
      select(`#FID`, IID, PC1, PC2, PC3, PC4, PC_dist, z_score) %>%
      arrange(desc(abs(z_score)))
    
    kable(outliers_summary, format="html", caption="PCA Outliers (|z| > 6)", digits=3) %>%
      kable_styling(bootstrap_options=c("striped","hover","condensed"))
  } else {
    cat("No PCA outliers detected.\n\n")
  }
  
} else {
  cat("PCA file not found.\n")
}
```

---

# Final Sample Count

```{r final_count}
final_bcf <- paste0("data/qc/", project, ".chr1.qc.bcf")

if(file.exists(sexcheck_file)) {
  initial_samples <- nrow(read_table(sexcheck_file, col_types=cols()))
  
  sex_fail <- 0
  if(file.exists(sex_exclusions_file)) {
    sex_fail <- nrow(read_table(sex_exclusions_file, col_names=FALSE, col_types=cols()))
  }
  
  miss_fail <- 0
  if(file.exists(miss_exclusions_file)) {
    miss_fail <- nrow(read_table(miss_exclusions_file, col_names=FALSE, col_types=cols()))
  }
  
  het_fail <- 0
  if(file.exists(het_exclusions_file)) {
    het_fail <- nrow(read_table(het_exclusions_file, col_names=FALSE, col_types=cols()))
  }
  
  pca_fail <- 0
  if(file.exists(pca_outliers_file)) {
    pca_fail <- nrow(read_table(pca_outliers_file, col_names=FALSE, col_types=cols()))
  }
  
  final_samples <- initial_samples - sex_fail - miss_fail - het_fail - pca_fail
  
  summary_df <- data.frame(
    Step=c("Initial Samples", "Sex Check Failures", "High Missingness", "Heterozygosity Outliers", "PCA Outliers", "Final Sample Count"),
    Count=c(initial_samples, -sex_fail, -miss_fail, -het_fail, -pca_fail, final_samples)
  )
  
  kable(summary_df, format="html", caption="Sample QC Summary") %>%
    kable_styling(bootstrap_options=c("striped","hover","condensed"), full_width=FALSE)
}
```

---

# Session Info

```{r session_info}
sessionInfo()
```

